### 2022-03-12 pilot-publication/manuscript.Rmd
Start drafting pilot publication based on protocol

---
## compile with:
## Rscript -e "rmarkdown::render('manuscript-clean.Rmd', output_format = 'all')"
title: A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes  
author: Trauma life support training Effectiveness Research Network (TERN) collaborators
bibliography: bibliography.bib
csl: nature.csl
output:
    bookdown::word_document2:
      toc: false
      number_sections: false
      reference_docx: "reference.docx"
    md_document: default
---

## Trial registration
This pilot study was registered with Clinical Trials Registry - India (reg. no XX) and ClinicalTrials.gov (reg. no XX).

## Corresponding author
Martin Gerdin Wärnberg (MGW)

Department of Global Public Health  
Karolinska Institutet  
171 77 Stockholm  
Sweden

martin.gerdin@ki.se  
+46 708 53 95 98 (mobile)

## Authors

## Contributors

Trauma life support Effectiveness Research Network (TERN) collaborators:

##### Martin Gerdin Wärnberg (MGW)
Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

Function Perioperative Medicine and Intensive Care, Karolinska University Hospital, Solna, Sweden.

##### Johanna Berg (JB)
Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

Emergency Medicine, Department of Internal and Emergency Medicine, Skåne University Hospital, Malmö, Sweden.

##### Prashant Bhandarkar (PB)
WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

##### Anirban Chatterjee (AC)
Department of Orthopaedic Sciences, Medica Superspecialty Hospital, Kolkata, India.

##### Shamita Chatterjee (SC)
Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.

##### Chintamani (C)
Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.

##### Li Felländer-Tsai (LFT)
Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.

Department of Reconstructive Orthopedics, Karolinska University Hospital, Stockholm, Sweden.

##### Anita Gadgil (AG)
WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

##### Geeta Ghag (GG)
Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.

##### Marie Hasselberg (MH)
Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

##### Catherine Juillard (CJ)
Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.

##### Monty Khajanchi (MK)
WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.

##### Deepa Kizhakke Veetil (DKV)
Department of Surgery, Manipal Hospital, Dwarka, New Delhi.

##### Vineet Kumar (VK)
Department of Surgery, Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai, India.

##### Debabrata Kundu (DK)
Department of Surgery, Medical College Kolkata, Kolkata, India.

##### Anurag Mishra (AM)
Department of Surgery, Maulana Azad Medical College, New Delhi, India.

##### Priti Patil (PP)
WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

##### Nobhojit Roy (NR)
WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.

Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

##### Amit Roy (AR)
Department of Surgery, Sir Nil Ratan Sircar Medical College & Hospital, Kolkata, India.

##### Siddarth David (SD)
Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

Doctors For You, Mumbai, India.

##### Rajdeep Singh (RS)
Department of Surgery, Maulana Azad Medical College, New Delhi, India.

##### Harris Solomon (HS)
Department of Cultural Anthropology and the Duke Global Health Institute, Duke University, Durham, USA.

##### Kapil Dev Soni (KDS)
Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.

##### Lovisa Strömmer (LS)
Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.

Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.

##### Megha Tandon (MT)
Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.

## Keywords
Trauma management, Accident and emergency medicine, Education and training.

## Word Count
4494

## Role of study sponsor and funders
The funding sources had no role in the design of this study and will not
have any role during its execution, analyses, interpretation of the
data, or decision to submit results.

\newpage

# Abstract

**Introduction** Trauma accounts for nearly 10% of the global burden of disease. Several trauma life support programs aim to improve trauma outcomes. There is no evidence from controlled trials to show the effect of these programs on patient outcomes. We describe the protocol of a pilot study that aims to assess the feasibility of conducting a cluster randomised controlled trial comparing Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care. **Methods and analysis** We will pilot a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We will recruit tertiary hospitals and include trauma patients and residents managing these patients. Two hospitals will be randomised to ATLS, two to PTC, and two to standard care. The primary outcome will be all cause mortality at 30 days from the time of arrival to the emergency department. Our secondary outcomes will include patient, provider, and process measures. All outcomes except time to event outcomes will be measured both as final values as well as change from baseline. We will compare outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care using absolute and relative differences along with associated confidence intervals. We will conduct subgroup analyses across the clinical subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly. In parallel to the pilot study we will conduct community consultations to inform the planning of the full-scale trial. **Ethics and dissemination** We will apply for ethics approvals to the local Institutional Review Board (IRB) in each hospital. The protocol will be published to Clinical Trials Registry - India and ClinicalTrials.gov. The results will be published and the anonymized data and code for analysis will be released publicly.

\newpage

# Article Summary

Strengths and limitations of this study:

- Cluster randomized controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated project officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.

\newpage

# Introduction

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response, causes 4.5 millions deaths every year
[@Naghavi2017]. Almost 10% of the global burden of disease is due
to trauma and trauma is the top contributor to the burden of disease in
children and adults aged 10 to 49 years [@Vos2020].

Trauma care is time sensitive and early management of life or limb
threatening conditions is crucial. Several trauma life support training
programs have been developed to improve the early management of patients
as they arrive at hospital by providing a structured framework to
assessment and treatment [@Mohammad2013; @Jayaraman2014;
@Kadhum2020].

The proprietary Advanced Trauma Life Support (ATLS) is the most
established trauma life support training program and more than one
million doctors in over 80 countries have been trained in the program
[@acsAtls2018]. Uptake in low- and middle income countries (LMIC)
has been slow, potentially due to high costs [@Kadhum2020].

The free Primary Trauma Care (PTC) program is the most widely spread
alternative program. The goal of PTC is to improve trauma care in LMIC
[@ptcfPtc2018]. Like ATLS, doctors in over 80 countries have been
trained in PTC, and the program has been endorsed by the World Health
Organization (WHO), among other international organizations
including several professional societies [@ptcfPtc2018].

Despite the widespread use of these training programs there are no
controlled trials showing that they impact patient outcomes
[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. But there is level 1
evidence that these programs improve provider skills and practices
[@Ali1995; @Ali1996], and observational data suggesting that
they also improve patient outcomes [@Qin2019].

We will perform a pilot study that aims to assess the feasibility of
conducting a cluster randomised controlled trial comparing ATLS and
PTC with standard care. Recent methodological guidelines indicate that
the design of efficient cluster randomised controlled trials requires
data on probable or target effect sizes, proportion of participants
with the outcome (if binary), and the intracluster correlation
coefficient [@Hemming2017]. The objectives of this pilot study will be
to:

- Estimate probable effect sizes on patient outcomes associated with
  ATLS and PTC compared with standard care, estimate the proportion of
  participants with the outcome (if binary), and estimate the
  intracluster correlation coefficient, as a basis for future sample
  size calculations.

- Assess the feasibility of recruiting participants and collecting
  data on primary and secondary outcomes, such as mortality,
  in-hospital complications, length of stay, and quality of life.

- Assess how the effect sizes and directions of these effects of ATLS
  and PTC may differ across clinically important subgroups.

# Methods

## Trial Design
This study will pilot a pragmatic three-armed parallel, cluster
randomised, controlled trial. There will be two intervention arms,
ATLS and PTC training, and one control arm, standard care. We will
collect data for four months in all three arms, first during a one
month observation phase and then during a three month intervention
phase (or continued observation in the control arm). This design will
allow us to assess outcomes both as final values and as change from
baseline. Our study is a pilot study because its objectives involves
estimating quantities, such as the probable effect sizes, proportion
of participants with the outcome (if binary), and the intracluster
correlation coefficient, needed for the sample size calculations of a
full-scale trial [@Hemming2017]. The full-scale trial will be planned
regardless of the effect-sizes identified in this pilot
study. This pilot study will also establish how many
participants that can be enrolled, as well as likely drop out rates,
and the feasibility of collecting primary and secondary outcomes.

## Study Setting
We conducted this pilot in seven Indian tertiary hospitals, where
neither ATLS, PTC, nor any other trauma life support training program
is routinely taught. India is the world's second most populous country
and has 20% of the world's trauma deaths. The trauma system is still
developing, with limited prehospital care, and the in hospital trauma
mortality as well as the proportion of preventable deaths remain
high. Lack of standard trauma training for healthcare providers,
limited hospital resources, inadequate processes of care, overcrowding
emergency departments - are some of the factors that contribute to the
high mortality and morbidity. During recent years efforts have been
made to improve hospital trauma care, through capacity building for
trained trauma care providers, augmenting facilities, and developing
care protocols within the hospitals. We will conduct this trial in
India because training providers in a trauma life support program is
not yet the standard.

The seven hospitals were King Edward Memorial Seth G. S. Medical
College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
College and General Hospital, Mumbai; HBT Medical College And Dr. R N
Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
Medical Education & Research, Chandigarh.

## Eligibility Criteria for Participants and Clusters
There will be two groups of participants: patients and resident
doctors.

### Patient Participants
Adults (15 years or older) who present to the emergency department at
participating hospitals with a history of trauma. History of trauma is
here defined as having any of the external causes of morbidity and
mortality listed in block V01-Y36, chapter XX of the International
Classification of Disease version 10 (ICD-10) codebook as reason for
presenting. We will explore intervention effects across the following
clinical subgroups: men, women, blunt multisystem trauma, penetrating
trauma, shock, severe traumatic brain injury, and elderly, as defined by
Hornor et al [@Hornor2018]. The consent form for patients are available
as Supplemental Material 1.

### Resident Doctor Participants
Resident doctors doing their speciality training in surgery (in India
it is very rare to have emergency medicine managing trauma patients in
the emergency department, why we decide to focus on surgical units in
this pilot), who manage trauma patients in the emergency department,
and who are expected to remain in the participating hospitals for at
least one year. To facilitate administration each surgical department
is divided into units, which manages the out patient department,
emergency department, operating rooms etc on different days each
week. One or two, out of typically six, units'
residents will be selected from each hospital. One unit consists of at
least three faculty and three to twelve residents.

To be eligible, units should have a maximum of 25% of the doctors
trained in either ATLS, PTC, or similar training programs before the
start of the pilot (hospitals that have so far agreed to participate
have no or single current residents trained in any program). Those
residents who have received training in the last five years will be
considered as trained. The figure of 25% was decided through consensus
in the research team, to balance feasibility and contamination of
results. We will select the units by conducting a prior survey to
ascertain this criteria. Consent will be sought from the residents in
each of the intervention groups before they undergo the ATLS or PTC
training. The consent form for residents are available as
Supplemental Material 2. We will not ask
for consent from residents at the units in the control hospitals as
their practice will not be affected by this pilot and we will not
collect any personal identifiable data on them. This is in line with
ethical regulations in the study setting.

### Clusters
Indian tertiary care hospitals that admit 400-800 adult patients with
trauma each year. We randomise on the cluster (hospital) level to avoid
contamination between intervention and control arms. To be eligible for
inclusion hospitals have to provide the following services round the
clock: operation theatres, X-ray, CT, and ultrasound facilities, and
blood bank. In addition the baseline admission rate should be more than
35 adult patients with major trauma per month.

## Interventions
In each intervention arm one or two units', out of typically
six, residents per hospital providing
emergency care to trauma patients will be trained in either ATLS or
PTC. For the purpose of this pilot study, we will target to train a
minimum of 75% of residents in each unit. If residents drop out or
change units after training but before data collection is completed we
will conduct additional training if needed to meet the 75% criterion.
We will not train the units' faculty, as they are typically not
involved in the initial management of trauma patients.

The ATLS training will be conducted in the nearest ATLS certified
training centre in India according to the standard ATLS curriculum
[@acsAtls2018]. The PTC training will be arranged in hospitals
randomized to the PTC arm, according to the standard PTC curriculum
[@ptcfPtc2018]. These courses will be conducted over a period of
2.5 to 3 days. The residents certified "pass" will be considered as
trained in respective courses.

The control group provides standard care with no intervention.

### Modifications
Both ATLS and PTC are standard training programs with fixed curricula
[@acsAtls2018; @ptcfPtc2018]. We will not modify the
delivery or content of these programs during this pilot.

### Adherence
The intervention is the training in either ATLS or PTC. Participants are
required to adhere to, i.e. participate in, the training, to be eligible
for passing. We will not consider adherence to training contents during
care delivery as adherence to the trial intervention, but rather as a
provider level outcome.

### Concomitant Care

#### Baseline Training
The care provided by all participating hospitals at baseline is based on
the training curriculum formulated by The National Medical Council of
India for post graduation in General Surgery
[@mciMsSurgery2019]. Regarding trauma, these guidelines state
that the student should:

a.  Have knowledge about response to trauma; burns: causes, prevention
    and management; wounds of scalp and its management; recognition,
    diagnosis and monitoring of patients with head injury, Glasgow coma
    scale.
b.  Be able to provide and coordinate emergency resuscitative measures
    in acute surgical situations including trauma.
c.  Choose, perform and interpret appropriate imaging in trauma -
    ultrasound Focused Abdominal Sonography in Trauma (FAST).
d.  Undergo advanced trauma and cardiac life support course (certified)
    before appearing in final examination.
e.  Undergo clinical posting in emergency and trauma.
f.  Present or discuss cases of blunt abdominal trauma.

Although training in an advanced trauma life support course is part of
the curriculum it is optional and not doing this training does not
result in failure to obtain post graduation completion.

#### Standard of Care
At most medical colleges in India trauma patients present to the
emergency department where they are assessed by a doctor and referred
to the surgical bay for further management. In the surgical bay a
second or third year general surgery resident sees all the major
trauma and provide the initial care, including
initiating treatment and investigations.  This resident informs the
consultant on call who is generally an Assistant Professor. Most
procedures like intercostal drainage, open wound suturing, intubation
etc. would be done in the surgical resuscitation area, by the surgical
resident.

Compared to other settings where a trauma team approach is adopted,
nurses and other healthcare professionals are involved to a limited
extent during the initial management. Their roles include assisting
during intubation and other bedside procedures, charting the vitals
(not recording) and giving injections. They also accompany the
resident during transfers of serious
patients.

After completing the assessment and starting initial resuscitation, the
resident decides to send the patient for imaging (X-rays/FAST/CT-scan)
or to the operation room in consultation with or after assessment by the
on-call consultant. A portable X-ray and an ultrasonography machine to
conduct FAST may or may not be available in the surgical bay. The
patients who are operated, managed conservatively, not intubated, or
with minor trauma will be sent to the surgical ward. Those who need
increased monitoring or mechanical ventilation remain in the surgical
bay or in the intensive care unit (ICU) depending on the availability of
ICU beds. The further treatment continues in the respective ward or ICU
and patients are finally discharged from the ward.

## Outcomes
Our pilot study include both participant and feasibility
outcomes. Prior to deciding on these participant outcomes we searched
the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's
database but were unable to identify appropriate core outcome sets for
our populations of participants.

The primary participant outcome will be all cause mortality within 30
days from the time of arrival to the emergency department. The
primary outcome and most secondary outcome will be assessed and
compared both as final values and as change from
baseline. All outcomes that pertain to the
individual participant level are detailed in Supplemental Material
3. We decided to include a large number of outcomes,
including some more exploratory, so that we can test their feasibility
and relevance. We may remove secondary participant outcomes during the
course of the pilot study, if they prove to be too difficult to
collect. If we remove outcomes we will document the reasons for doing
so.

We will also assess the following feasibility outcomes, which pertain
both to overall study population as well as to the individual cluster
level:

- Recruitment rate. For both patients and residents this will equal
  the proportion of participants enrolled, out of the total number of
  eligible participants, over the course of the pilot study.
- Lost to follow up rate. This will apply only to patients and equal
  the proportion of patients that do not complete 30 day follow up,
  out of all enrolled patients, over the course of the pilot study.
- Pass rate. This will apply only to residents in the intervention
  arms and equal the proportion of residents that pass the training
  programme, out of the total number of trained residents, over the
  course of the pilot study.
- Missing data rate. This will apply to each outcome and variable and
  equal the proportion of missing data, over the course of the pilot
  study.
- Differences in distributions of observed and extracted data. This
  will apply to each outcome and variable and will compare the
  distributions of data collected by observations versus extracted
  from hospital records. For quantitative variables this will be the
  difference in means, standard deviations, medians, interquartile
  ranges, and ranges. For qualitative variables this will be the
  differences in absolute counts and percentages, across categories.

## Participant Timeline

### Patients
Patients will be screened for eligibility as they arrive at the
emergency department. Eligible patients will be approached in the
emergency department to consent to follow up, if they are
conscious. If they are unconscious a patient representative will be
approached to consent to follow up. Once the patient is conscious we
will approach the patient to affirm the patient representative's
consent. We will follow up patients at discharge, at 24 hours after
arrival at the emergency department, and at 30 days after arrival at
the emergency department.

### Residents
Surgical units will be screened for eligibility once hospitals confirm
their participation. All residents in eligible units will be approached
to consent to training if their hospital is randomised to either of the
intervention arms. Training will be conducted as soon as possible after
the study starts. Resident participants will be followed up 30 days
after training, if they are in the intervention arms, or 30 days after
the study started, if they are in the control arm.

## Sample size
Given budget and time constraints, including the rotation of surgical
units in Indian hospitals (which often happen on a six months basis)
the feasible data collection period is four months. Each of the
surgical units see 2-4 trauma patients per week. If we select a
minimum of one unit per hospital then each hospital will enrol 8-16
patients per month and 32-64 patients during the four months of this
pilot. With a 20% attrition rate we expect each hospital to enrol
26-51 patients, coming to a total sample size of between 156 and 306
patients for this pilot study.

## Recruitment
To ensure adequate recruitment we only approach hospitals with trauma
volumes high enough to allow us to reach the sample size goals detailed
above. Patients will be enrolled by a dedicated project officer as they
arrive at the emergency department. The recruitment period will be three
months. Recruitment will be monitored weekly through online conferences.
No financial or non-financial incentives will be provided to trial
investigators or participants for enrolment.

## Allocation

### Sequence generation
We used simple randomisation to allocate sites to trial arms. We
prepared seven sealed envelopes from which one representative from
each pilot site selected one in a common meeting on March 8, 2022
[^dev1]. The content of the envelopes dictated what trial arm (ATLS,
PTC, or standard care) each hospital was allocated to. There were two
hospitals in each of the intervention arms and three hospitals in the
control arm.

[^dev1]: Initially the plan was to enroll six hospitals (clusters) but
    we ended up enrolling seven.

### Concealment Mechanism
We did not conceal the sequence.

### Implementation
The random allocation sequence was generated by MGW. Clusters were
enrolled by the project's core team, who also enrolled
clusters. Patient participants will be included if they present during
the project officers shift. Resident participants are enrolled if they
are in the units selected for training. We will use simple random
sampling to select units if there are more than two eligible units in
a hospital. For patient participants consent for follow up is sought
after randomisation from patients or patient relatives as
appropriate. For resident participants consent is sought before
randomisation. If residents in a unit decline to participate, so that
the target of training 75% of residents in a given unit cannot be met,
another unit will be selected for participation.

## Blinding
It will not be possible to blind investigators or participants to
interventions. We will not blind the data analysts during this pilot,
but we plan to blind the data analysts during the full-scale
trial.

## Data Collection
Data collection will start one month before the training is delivered,
to establish a baseline. A variability of three months of the date when
data collection is started between hospitals will be accepted. Each
participating hospital will have a dedicated project officer to collect
data. The project officers will have a masters in a health science field
and should have experience in data collection.

Because participating residents are assigned designated days for trauma
care for a period of 6 months, data will be collected during those
particular days and shifts when these trained doctors are in the
emergency department. The project officers will collect data both by
observing the care delivered and by interviewing the participants, and
by extracting data from hospital records.

Data collection will continue for a minimum of three months after
training. The research officers will collect data of all patients, who
present with trauma in the surgical bay during their duty hours. Those
patients who are admitted will be followed up for complications and
other in-hospital outcome measures, for example length of stay. Patients
who are not admitted will be followed up telephonically for mortality
outcomes and quality of life outcomes. The follow up period will be 30
days. The project officers will make at most three attempts to reach a
participant or participant representative telephonically, after which
the data will be recorded as missing.

The project officer will administer the study information and informed
consent (consent will only be sought for data collection including
follow up) to the patient, or the patient's representative as
appropriate, once the patient is stabilised. They will continue to
collect data once they have received the consent.

Details of data of those patients/relatives not willing to give consent
will be removed from the analysis. The number of patients who opt out
from data collection will be collected, as well as limited data on their
age and sex. Patients will be followed up in the ward regularly for the
various outcome variables. They will also be followed up telephonically
after they have been discharged.

## Variables
The project officers will collect data on demographics, time of injury
to arrival at the participating hospital, time to recording vital
signs, vital signs, and times to and management details including
imaging and surgery. Details of any injury sustained will be collected
and coded using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD
10 coders will undergo the WHO online ICD 10 training module and for
AIS they will be accredited. Based on AIS we will calculate the Injury
Severity Score (ISS) and the New ISS (NISS). Supplemental Material 4
contains a full variable list, with definitions.

## Patient and public involvement
In this study, we will conduct community consultations to collect
inputs from patients, their caregivers, patient groups, and resident
doctors to be used in the selection of outcome measures and
implementation of the full-scale trial; following the Guidance for
Reporting Involvement of Patients and the Public (GRIIPS 2)
[@Staniszewska2017].

During the pilot study, interviews will be conducted with
post-discharge trauma patients and their caregivers to identify
outcomes most relevant to them. These patients will be identified
through the medical registers of the participating hospitals,
contacted through telephone, and after receiving their consent be
interviewed as per their convenience. Their consent form is available
as Supplemental Material 5.  Additionally, members from non-government
organizations working with trauma patients and the hospital Social
Service Section will also be contacted for their views on contextual
patient-centred outcomes for trauma patients. Their consent form is
available as Supplemental Material 6. For feasibility, these
interviews will be held in each of the cities where the participating
centres are located. The commonest patient-centred outcomes reported
across all the locations will be incorporated into the evaluation of
the effects of the different training programs and standardized care
on patient outcomes.

Similarly, the inputs of resident doctor participants at each
participating centre will be collected during the pilot study. A
discussion and periodic surveys will be conducted to document any
challenges or suggestions they may have in the scheduling or
implementation of the training programs. These inputs will be
incorporated in the final study.

A summary of the findings of the study as well as their inputs will be
shared with those who participated in the interviews and surveys. A
meeting will be held with the patient participants, at each city, where
the changes in the measured patient-centered outcomes would be presented
to them. Another meeting will be held with the resident doctors at each
hospital to present the confidence of the residents after being trained.
Any suggestions and reflections from the participants during the
meetings will be used as inputs for planning the final study.

## Data management
We will supply an online data collection tool, accessible only over a
virtual private network (VPN), for each participating hospital to upload
pseudonymised data to secure servers. Data validation techniques like
restricted values or values of a specific range will be used to avoid
ambiguous data entries and ensure the validity of the data. Ambiguous
responses, data errors, if any, will be resolved after discussion with
the core team during weekly meetings. An instruction manual or codebook
for data variables will be prepared to ensure consistency in data entry.
This manual will be referred to during the project data collection and
variable descriptions are visible for each variable in the online data
collection tool. Pseudonymised data will be stored at the centralised
server. The data will be accessible by the project's principal
investigator or by delegation of the project principal investigator
only.

## Data monitoring
Weekly meetings with the core team and project officers will take
place and for this meeting a data status report will be automatically
generated highlighting missing data and number of patients awaiting
followup. Cluster specific interim analysis will take place after two
months. The results of this will be presented to the core team, this
team will decide if the pilot should be terminated. Although we will
not have formal termination criteria because of the short duration of
the study, reasons not to continue could include that the collection
of key variables, such as mortality outcomes, is unfeasible or that
patients are not consenting to be included in the data
collection. A data monitoring committee
will not be used in the pilot study due to its limited scope.

## Statistical Methods
We will analyse all pilot data using descriptive
statistics. Quantitative variables will be summarised as mean +/-
standard deviation, median, interquartile range and range. Qualitative
variables will be presented as absolute numbers and
percentages. Feasibility outcomes will be summarised both on the
overall sample level as well as on the individual cluster level. We
will use an empty generalised linear mixed model<!-- see [@Sommet2017]
--> to estimate the intracluster correlation coefficient.

We will compare participant outcomes in three combinations of trial
arms: ATLS versus PTC, ATLS versus standard care, and PTC versus
standard care. In each combination we will compare both differences in
final values and differences in change from baseline. For example, for
the primary participant outcome of all cause mortality within 30 days
from the time of arrival to the emergency department, comparing ATLS
versus PTC, we will compare both the difference in mortality between
the ATLS and PTC arms as well as the difference in the change from
baseline in mortality between the ATLS and PTC arms.

For the intervention arms the change from baseline will be calculated as
the difference between the one month period of data collection before
the training was undertaken and the three month period after the
training. For the control arm the data collection period will be four
months and the difference from baseline will be calculated as the
difference between the first one month and the following three months.

Within each combination of trial arms we will conduct subgroup
analyses of men, women, blunt multisystem trauma, penetrating trauma,
shock, severe traumatic brain injury, and elderly. Table S7.1 in
Supplemental Material 7 shows which outcomes will be assessed in which
subgroups, decided through consensus in the research team. We will
further compare the results of all subgroups with the results in the
whole cohort, and compare the results in the female subgroup with the
male subgroup, and the results in the blunt multisystem trauma
subgroup with the penetrating trauma subgroup. We are aware that the
numbers in some of these subgroups are likely to be small, but we
include them to help guide the formulation of the statistical analysis
plan for the full-scale trial.

We will calculate both absolute and relative differences for each
comparison, along with 75, 85, and 95% confidence intervals. We will use
an empirical bootstrap procedure with 1000 draws to estimate these
confidence intervals. We will not perform any formal hypothesis tests
during the analysis of this pilot's data [@Lee2014]. We will also
compare the data collected through observations and interviews with the
data collected from hospital records, to assess the feasibility of
collecting data from hospital records in the full-scale trial.

## Ethics and Dissemination
We were granted research ethics approval from each participating
hospital. The approval numbers were IEC(II)/OUT/134/2022 for King
Edward Memorial Seth G. S. Medical College and K.E.M. Hospital,
Mumbai; XXX for Lokmanya Tilak Municipal Medical College and General
Hospital, Mumbai; XXX for HBT Medical College And Dr. R N Cooper
Municipal General Hospital, Mumbai; CREC/2022/FEB/1(ii) for Medica
Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata;
NRSmC/IEC/93/2021 for Sir Nil Ratan Sircar Medical College & Hospital,
Kolkata; XXX for Postgraduate Institute of Medical Education & Research,
Chandigarh.

The protocol was published in BMJ Open [ADD CITATION] and was
registered with Clinical Trials Registry - India (reg. no XXX) and
ClinicalTrials.gov (reg. no XXX). Amendments to the protocol after
publication were determined by the core research group and updated on
Clinical Trials Registry - India and ClinicalTrials.gov. Substantial
amendments, such as modifications to the eligibility criteria or
outcomes will also be resubmitted to the journal. Declaration of
interest will be submitted from all participating researchers both in
the core team and at each site.

The final anonymized dataset and code for analysis are released
publicly. Authorship follows the International Committee of
Medical Journal Editors (ICMJE) guidelines.

# Results

# Discussion

# Contributorship statement
MGW conceived of the study. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the study. DKV, KDS, MK, and MGW drafted the first version of the protocol. AG, HS, and SD drafted the first version of the patient and public involvement activities. JB and PP drafted the first versions of the data management sections and wrote the data management plan. PB and PP drafted the first versions of the statistical analysis section. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB, PP, RS, SC, SD, and VK contributed to the refinement of the protocol. AR, AC, C, DK, GG, MK, MT, and VK are representatives of prospective participating hospitals.

# Competing Interests
Several authors are Advanced Trauma Life Support
instructors.

# Funding
Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).

# Data Sharing Statement
The final anonymized dataset and code for analysis will be released publicly.

# References

### 2022-03-12 pilot-publication/manuscript.Rmd
Correct meeting date
Add meeting notes reference

@@
 prepared seven sealed envelopes from which one representative from
-each pilot site selected one in a common meeting on March 8, 2022
-[^dev1]. The content of the envelopes dictated what trial arm (ATLS,
-PTC, or standard care) each hospital was allocated to. There were two
-hospitals in each of the intervention arms and three hospitals in the
-control arm.
+each pilot site selected one in a common meeting on March 7, 2022
+[^dev1]<!-- Meeting notes ../meetings/20220307/ -->. The content of
+the envelopes dictated what trial arm (ATLS, PTC, or standard care)
+each hospital was allocated to. There were two hospitals in each of
+the intervention arms and three hospitals in the control arm.
 

### 2022-06-09 pilot-publication/manuscript.Rmd
Add IEC registration number for Cooper
Add ethical approval number for LTMGH to manuscrtipt
Add protocol publication citation

@@
 Edward Memorial Seth G. S. Medical College and K.E.M. Hospital,
-Mumbai; XXX for Lokmanya Tilak Municipal Medical College and General
-Hospital, Mumbai; XXX for HBT Medical College And Dr. R N Cooper
-Municipal General Hospital, Mumbai; CREC/2022/FEB/1(ii) for Medica
-Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata;
-NRSmC/IEC/93/2021 for Sir Nil Ratan Sircar Medical College & Hospital,
-Kolkata; XXX for Postgraduate Institute of Medical Education & Research,
-Chandigarh.
+Mumbai; IEC/214/22 for Lokmanya Tilak Municipal Medical College and
+General Hospital, Mumbai; HBTMC/266/SURGERY for HBT Medical College
+And Dr. R N Cooper Municipal General Hospital, Mumbai;
+CREC/2022/FEB/1(ii) for Medica Superspecialty Hospital, Kolkata;
+Medical College Kolkata, Kolkata; NRSmC/IEC/93/2021 for Sir Nil Ratan
+Sircar Medical College & Hospital, Kolkata; XXX for Postgraduate
+Institute of Medical Education & Research, Chandigarh.
 
-The protocol was published in BMJ Open [ADD CITATION] and was
+The protocol was published in BMJ Open [@GerdinWärnberg2022] and was
 registered with Clinical Trials Registry - India (reg. no XXX) and

### 2022-09-06 pilot-publication/manuscript.Rmd
Add ethics from PGI Chandigarh

@@
 Medical College Kolkata, Kolkata; NRSmC/IEC/93/2021 for Sir Nil Ratan
-Sircar Medical College & Hospital, Kolkata; XXX for Postgraduate
-Institute of Medical Education & Research, Chandigarh.
+Sircar Medical College & Hospital, Kolkata; and PGI/IEC/2022/000/003
+for Postgraduate Institute of Medical Education & Research,
+Chandigarh.
 

### 2022-09-06 pilot-publication/manuscript.Rmd
Add ClinicalTrials registration to manuscript and add section on protocol deviations

@@
 ## Trial registration
-This pilot study was registered with Clinical Trials Registry - India (reg. no XX) and ClinicalTrials.gov (reg. no XX).
+This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).
 
@@
 # Competing Interests
-Several authors are Advanced Trauma Life Support
-instructors.
+Several authors are Advanced Trauma Life Support or Primary Trauma Care instructors
 
@@
 
+# Protocol Deviations
+
+## Trial Registration
+We intended to also register our trial with Clinical Trials Registry - India but because of time constrains we had to initiate the study before registration was finalised, and Clinical Trials Registry - India only allow prospective registration.
+
+## Number of Participating Centres
+We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
+
+## Resident Participants
+Emergency medicine in addition to surgery.
+
 # References

### 2022-11-07 pilot-publication/manuscript.Rmd
Start drafting pilot publication results

@@
 
-For the intervention arms the change from baseline will be calculated as
-the difference between the one month period of data collection before
-the training was undertaken and the three month period after the
-training. For the control arm the data collection period will be four
-months and the difference from baseline will be calculated as the
+For the intervention arms the change from baseline will be calculated
+as the difference between the one month period of data collection
+before the training was undertaken and the three month period after
+the training. For the control arm the data collection period will be
+four months and the difference from baseline will be calculated as the
 difference between the first one month and the following three months.
@@
 We will calculate both absolute and relative differences for each
-comparison, along with 75, 85, and 95% confidence intervals. We will use
-an empirical bootstrap procedure with 1000 draws to estimate these
+comparison, along with 75, 85, and 95% confidence intervals. We will
+use an empirical bootstrap procedure with 1000 draws to estimate these
 confidence intervals. We will not perform any formal hypothesis tests
 during the analysis of this pilot's data [@Lee2014]. We will also
-compare the data collected through observations and interviews with the
-data collected from hospital records, to assess the feasibility of
+compare the data collected through observations and interviews with
+the data collected from hospital records, to assess the feasibility of
 collecting data from hospital records in the full-scale trial.
@@
 The protocol was published in BMJ Open [@GerdinWärnberg2022] and was
-registered with Clinical Trials Registry - India (reg. no XXX) and
-ClinicalTrials.gov (reg. no XXX). Amendments to the protocol after
-publication were determined by the core research group and updated on
-Clinical Trials Registry - India and ClinicalTrials.gov. Substantial
-amendments, such as modifications to the eligibility criteria or
-outcomes will also be resubmitted to the journal. Declaration of
-interest will be submitted from all participating researchers both in
-the core team and at each site.
+registered with ClinicalTrials.gov (reg. no NCT05417243). Amendments
+to the protocol after publication were determined by the core research
+group and updated on ClinicalTrials.gov.
 
@@
 # Results
+Between X and X, we enrolled X trauma patients from X participating
+centres. The ATLS arm enrolled X patients, the PTC arm X patients, and
+the standard care arm X patients. The study flowchart is shown in
+Figure X and sample characteristics across trial arms are shown in
+Table X. Overall, the number of females were X (X%), the median age
+was X (IQR X), and the median ISS was X (IQR X). A total of X (X%)
+patients had the primary outcome of mortality at 30 days after arrival
+to the emergency department.
+
+## Patient Participant Outcomes
+In the post training period, a total of X (X%) patients in the ATLS
+arm had the primary outcome, compared to X (X%) patients in the PTC
+arm, and X (X%) patients in the standard care arm. The differences
+between the arms were X (95% CI X - X) between the ATLS and PTC arms,
+X (95% CI X - X) between the ATLS and standard care arms, and X (95%
+CI X - X) between the PTC and standard care arms. The change from
+baseline for the primary outcome was X in the ATLS arm, X in the PTC
+arm, and X in the standard care arm. The differences between changes
+from baseline were X (95% CI X - X) between the ATLS and PTC arms, X
+(95% CI X - X) between the ATLS and standard care arms, and X (95% CI
+X - X) between the PTC and standard care arms. Table X shows
+differences in the primary outcome between trial arms with 75, 85, and
+95% CIs. Differences in secondary outcomes between trial arms and
+subgroups are available as Supplementary material X.
+
+## Resident Participant Outcomes
+A total of X residents were trained during this pilot. Overall their
+confidence in managing trauma patients was X (IQR X - X) on
+
+## Feasibility Outcomes
 
@@
 # Contributorship statement
-MGW conceived of the study.
-AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the study.
-DKV, KDS, MK, and MGW drafted the first version of the protocol.
-AG, HS, and SD drafted the first version of the patient and public involvement activities.
-JB and PP drafted the first versions of the data management sections and wrote the data management plan.
-PB and PP drafted the first versions of the statistical analysis section.
-AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB, PP, RS, SC, SD, and VK contributed to the refinement of the protocol.
-AR, AC, C, DK, GG, MK, MT, and VK are representatives of prospective participating hospitals.
+MGW conceived of the study. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH,
+MGW, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
+study. DKV, KDS, MK, and MGW drafted the first version of the
+protocol. AG, HS, and SD drafted the first version of the patient and
+public involvement activities. JB and PP drafted the first versions of
+the data management sections and wrote the data management plan. PB
+and PP drafted the first versions of the statistical analysis
+section. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB,
+PP, RS, SC, SD, and VK contributed to the refinement of the
+protocol. AR, AC, C, DK, GG, MK, MT, and VK are representatives of
+prospective participating hospitals.
 

### 2022-11-07 pilot-publication/manuscript.Rmd
Update manuscript

@@
 A total of X residents were trained during this pilot. Overall their
-confidence in managing trauma patients was X (IQR X - X) on
+confidence in managing trauma patients was X (IQR X - X) on a 10 point
+Likert scale with 10 being most confident. In the post training
+period, the confidence in the ATLS arm was X (IQR X - X), compared to
+X (IQR X - X) in the PTC arm, and X (IQR X - X) in the standard care
+arm.
 
 ## Feasibility Outcomes
+Out of X potentially eligible patients, X (X%) patients were enrolled
+in the study. Out of X potentially eligible residents, X (X%) were
+enrolled in the study. The loss to follow up rate was X, with X out of
+X patients not completing 30 day follow up. Among residents the pass
+rate, after two attempts, was X (X%). The missing data rate ranged
+from X to X, with details per variable and outcome in Table X. The
+differences in distributions between observed and extracted data, for
+variables that were collected through observation or interview, are
+shown in Table X.
+
+## Community Consultations
+The results of interviews with patients, caregivers and social workers
+will be published separately. Residents felt that...
 
@@
 # Competing Interests
-Several authors are Advanced Trauma Life Support or Primary Trauma Care instructors
+Several authors are Advanced Trauma Life Support or Primary Trauma
+Care instructors.
 
 # Funding
-Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).
+Doctors for You through grants awarded to Karolinska Institutet by the
+Swedish Research Council (grant number 2020-03779) and the Laerdal
+Foundation (grant number 2021-0048).
 
@@
 ## Trial Registration
-We intended to also register our trial with Clinical Trials Registry - India but because of time constrains we had to initiate the study before registration was finalised, and Clinical Trials Registry - India only allow prospective registration.
+We intended to also register our trial with Clinical Trials Registry -
+India but because of time constrains we had to initiate the study
+before registration was finalised, and Clinical Trials Registry -
+India only allow prospective registration.
 
 ## Number of Participating Centres
-We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
+We ended up recruiting seven centres instead of six and therefore
+assigned two centres each to the intervention arms and three centres
+to the control arm.
 

### 2022-11-09 pilot-publication/manuscript.Rmd
Start defining basic results

@@
 # Results
-Between X and X, we enrolled X trauma patients from X participating
-centres. The ATLS arm enrolled X patients, the PTC arm X patients, and
-the standard care arm X patients. The study flowchart is shown in
-Figure X and sample characteristics across trial arms are shown in
+
+**Note that participating centres are randomly assigned to the
+different arms every time this manuscript is compiled**
+
+Between `r start.date` and `r end.date`, we enrolled `r n.patients`
+trauma patients from `r n.centres` participating centres. The ATLS arm
+enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the
+standard care arm `r n.control` patients. The study flowchart is shown
+in Figure X and sample characteristics across trial arms are shown in
 Table X. Overall, the number of females were X (X%), the median age

### 2022-11-09 pilot-publication/main.R

@@ -1,5 +1,7 @@
 ## Load packages
 library(dotenv)
+library(dplyr)
+library(lubridate)
 
 ## Import data
 data <- readr::read_csv(Sys.getenv("DATA_DIR"))
@@ -8,6 +10,26 @@ data <- readr::read_csv(Sys.getenv("DATA_DIR"))
 codebook.arguments <- lapply(c("URL", "UID", "USERNAME", "PASSWORD"),
                              function(x) Sys.getenv(paste0("KOBO_", x)))
 codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
+
+## Define basic results
+arrival.dates <- data %>% pull("incident/date_of_arrival") %>% sort()
+format_date <- function(date) paste0(month(date[1], label = TRUE, abbr = FALSE), " ", year(date[1]))
+start.date <- format_date(arrival.dates[1])
+end.date <- format_date(rev(arrival.dates)[1])
+n.patients <- nrow(data)
+centre.ids <- data %>% pull("id/reg_hospital_id") %>% unique() 
+n.centres <-  centre.ids %>% length()
+shuffled.centres <- centre.ids %>% sample(n.centres)
+atls.centres <- shuffled.centres[1:2]
+ptc.centres <- shuffled.centres[3:4]
+control.centres <- shuffled.centres[5:n.centres]
+centre.data <- data %>% split(data$`id/reg_hospital_id`)
+atls.data <- do.call(rbind, centre.data[as.character(atls.centres)])
+ptc.data <- do.call(rbind, centre.data[as.character(ptc.centres)])
+control.data <- do.call(rbind, centre.data[as.character(control.centres)])
+n.atls <- nrow(atls.data)
+n.ptc <- nrow(ptc.data)
+n.control <- nrow(control.data)
 
 ## Label variables
 

### 2022-11-21 pilot-publication/manuscript.Rmd
Add Jin review

@@
 controlled trials showing that they impact patient outcomes
-[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. But there is level 1
-evidence that these programs improve provider skills and practices
-[@Ali1995; @Ali1996], and observational data suggesting that
-they also improve patient outcomes [@Qin2019].
+[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. A systematic review
+report that hospital trauma training is associated with a reduction of
+patient mortality with 29% (95% CI 4-47%) based on seven studies that
+were either observational or uncontrolled trials [@Jin2021].
 

### 2022-11-22 pilot-publication/manuscript.Rmd
Start writing discussion

--- 2022-11-09 pilot-publication/manuscript.Rmd
@@
 controlled trials showing that they impact patient outcomes
-[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. But there is level 1
-evidence that these programs improve provider skills and practices
-[@Ali1995; @Ali1996], and observational data suggesting that
-they also improve patient outcomes [@Qin2019].
+[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. A systematic review
+report that hospital trauma life support training is associated with a
+reduction in patient mortality, based on studies that were either
+observational or quasi-experimental without a control group
+[@Jin2021].<!-- Note that some of the studies that Jin include state
+that there was no effect of ATLS -->
 
@@
 # Discussion
+We show that conducting a cluster randomized controlled trial
+comparing ATLS with PTC and standard care is feasible in this setting
+provided that there are dedicated project officers to enrol
+participants, collect data, and follow up participants. Missing data
+rates were low for the primary outcome and many secondary outcomes, as
+well as for key variables. Some variables had very high missing data
+rates and may not be feasible to include in a full-scale trial.
+
+<!-- Should we include dependency at discharge, for example using the
+Modified Oxford Handicap Scale used in @Roberts2013?-->
+
+We found that the ATLS arm had lower 30-day mortality compared to the
+PTC and standard care arms.
+
+We found that the PTC arm had lower 30-day mortality compared to the
+ATLS and standard care arms.
+
+We found that the standard care arm had lower 30-day mortality
+compared to the ATLS and PTC arms.
+
+We found that 30-day mortality was similar in the ATLS and PTC arms
+but lower than the standard care arm.
+
+We found that 30-day mortality was similar in the ATLS and standard
+care arms but lower than the PTC arm.
+
+We found that 30-day mortality was similar in the PTC and standard
+care arms but lower than the ATLS arm.
+
+Previous studies on the effect of in-hospital trauma life support
+training on patient outcomes are observational or quasi-experimental
+without a control group, with heterogeneous results
+[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
+@Wang2010]. Studies from Trinidad and Tobago and, El Salvador, Rwanda,
+and Cambodia found no significant effect on patient mortality after
+implementing in-hospital trauma life support training
+[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
+whereas one study from China on 820 patients found a significant
+reduction in mortality, from 20 to 15%, after implementing ATLS
+[@Wang2010].
+
+Several randomised controlled trials show that ATLS is associated with
+improved provider skills.
+
+The primary patient participant outcome, all-cause 30-day mortality,
+was missing in X% of patients. This may appear high, especially
+compared to for example the CRASH-2 and REACT-2 trials, which report
+missing primary outcome in less than 0.01% of patients [@Roberts2013;
+@Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2
+used in-hospital mortality as their primary outcome measure, whereas
+we attempted to follow up patients after discharge. The missing data
+rate for in-hospital mortality was only X%, comparable to previous
+trials.
+
+Compare missing data rates.
+
+Non-inferiority vs superiority.
 
@@
 
+## Periodic suverys to residents
+We did not distribute periodic surveys to the participating residents
+but discussed challenges and suggestions that they had regarding the
+scheduling or implementation of the training programs.
+
 # References

### 2022-11-23 pilot-publication/manuscript.Rmd
Complete preliminary discussion

@@
 generated highlighting missing data and number of patients awaiting
-followup. Cluster specific interim analysis will take place after two
-months. The results of this will be presented to the core team, this
-team will decide if the pilot should be terminated. Although we will
-not have formal termination criteria because of the short duration of
-the study, reasons not to continue could include that the collection
-of key variables, such as mortality outcomes, is unfeasible or that
-patients are not consenting to be included in the data
-collection. A data monitoring committee
-will not be used in the pilot study due to its limited scope.
+followup. We conducted an interim analysis on October 12, 2022, and
+decided to complete the study as participants were consenting to be
+included in the study and key variables including mortality outcomes
+could be collected. No outcomes were dropped. We did not use a data
+monitoring committee in the pilot study due to its limited scope.
 
@@
 
-Previous studies on the effect of in-hospital trauma life support
-training on patient outcomes are observational or quasi-experimental
-without a control group, with heterogeneous results
-[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
-@Wang2010]. Studies from Trinidad and Tobago and, El Salvador, Rwanda,
-and Cambodia found no significant effect on patient mortality after
-implementing in-hospital trauma life support training
-[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
-whereas one study from China on 820 patients found a significant
-reduction in mortality, from 20 to 15%, after implementing ATLS
-[@Wang2010].
-
-Several randomised controlled trials show that ATLS is associated with
-improved provider skills.
-
 The primary patient participant outcome, all-cause 30-day mortality,
@@
 
-Compare missing data rates.
+<!-- Consider discussing non-inferiority vs superiority. -->
 
-Non-inferiority vs superiority.
+During the course of this pilot we deviated from the protocol in
+multiple ways, and provide a detailed list as Supplementary
+material. Some key limitations of this pilot and therefore lessons to
+be learned and factored into the design of the full-scale trial
+include the lower than expected enrolment rates of some centres,
+centre specific management routines, and difficulties in collecting
+data on complications and cause of death. We minimised the impact of
+the lower than expected enrolment rates by including a seventh centre,
+but on-site observations of patient volumes are likely to be needed
+for the full-scale trial. We decided to be pragmatic in selecting
+which residents to train and how to structure the data collection
+depending on how and by whom patients were initially managed, but this
+flexibility will need to be built into the full-scale trial
+protocol. Finally, we found that data on complications and cause of
+death were hard to identify and therefore the full-scale trial will
+need to include longer training of project officers.
+
+Previous studies on the effect of in-hospital trauma life support
+training on patient outcomes are observational or quasi-experimental
+without a control group, with heterogeneous results
+[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
+@Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
+Cambodia found no significant effect on patient mortality after
+implementing in-hospital trauma life support training
+[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
+whereas one study from China that included 820 patients found a
+significant reduction in mortality, from 20 to 15%, after implementing
+ATLS [@Wang2010].
+
+Several controlled trials, including at least two randomized
+controlled trials [@Ali1996; @Ali1999], show that ATLS is associated
+with improved provider skills [@Mohammad2013]. Observational evidence
+indicates that PTC also leads to improved provider skills
+[@Kadhum2020]. The missing link is then how, and if, these improved
+skills translate into improved patient outcomes. As trauma care
+providers we assume, and probably rightly so, that we deliver better
+care if we train. The question is then how we should train, especially
+considering the costs associated with some of the programs
+offered.
+
+We conclude that a full-scale cluster randomised trial should be
+feasible after incorporating the lessons of this pilot, and that this
+full-scale trial should, regardless of its outcome, influence how we
+train trauma care providers in the future.
 
@@
 
+## Data collection from records
+We decided to record data from records only for a subset of patients
+to reduce the project officers' workload.
+
 # References

### 2022-11-23 pilot-publication/manuscript.Rmd
Draft abstract

@@
 
-**Introduction** Trauma accounts for nearly 10% of the global burden of disease.
-Several trauma life support programs aim to improve trauma outcomes.
-There is no evidence from controlled trials to show the effect of these programs on patient outcomes.
-We describe the protocol of a pilot study that aims to assess the feasibility of conducting a cluster randomised controlled trial comparing Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care.
-**Methods and analysis** We will pilot a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught.
-We will recruit tertiary hospitals and include trauma patients and residents managing these patients.
-Two hospitals will be randomised to ATLS, two to PTC, and two to standard care.
-The primary outcome will be all cause mortality at 30 days from the time of arrival to the emergency department.
-Our secondary outcomes will include patient, provider, and process measures.
-All outcomes except time to event outcomes will be measured both as final values as well as change from baseline.
-We will compare outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care using absolute and relative differences along with associated confidence intervals.
-We will conduct subgroup analyses across the clinical subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
-In parallel to the pilot study we will conduct community consultations to inform the planning of the full-scale trial.
-**Ethics and dissemination** We will apply for ethics approvals to the local Institutional Review Board (IRB) in each hospital.
-The protocol will be published to Clinical Trials Registry - India and ClinicalTrials.gov.
-The results will be published and the anonymized data and code for analysis will be released publicly.
+**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes.
+We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught.
+We recruited tertiary hospitals and included trauma patients and residents managing these patients.
+Two hospitals were randomised to ATLS, two to PTC, and three to standard care.
+The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
+We conducted community consultations in parallel with the pilot trial.
+**Ethics and dissemination** We obtained ethical approval from all participating centres.
+**Results** We included X patients and X residents.
+The patient recruitment rate was X and the resident recruitment rate was X.
+The loss to follow up rate was X.
+The all-cause 30 day mortality was X in the ATLS arm, X in the PTC arm, and X in the standard care arm.
+**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
 

### 2022-11-23 pilot-publication/manuscript.Rmd
Work on code

@@
 ## Word Count
-4494
+<!-- 4494 -->
 
@@
 \newpage
+
+<!-- Source main code to make global objects available in the text -->
+```{r source, echo = FALSE, include = FALSE}
+source("main.R")
+```
 
@@
 **Ethics and dissemination** We obtained ethical approval from all participating centres.
-**Results** We included X patients and X residents.
+**Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres.
 The patient recruitment rate was X and the resident recruitment rate was X.

### 2022-11-23 pilot-publication/main.R

@@ -17,6 +17,11 @@ format_date <- function(date) paste0(month(date[1], label = TRUE, abbr = FALSE),
 start.date <- format_date(arrival.dates[1])
 end.date <- format_date(rev(arrival.dates)[1])
 n.patients <- nrow(data)
+n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre
+n.atls.residents.passed.first.attempt <- 3 + 2 # The number of residents who had passed ATLS after the first attempt, per centre
+n.atls.residents.passed.first.attempt <- 4 + 2 # The number of residents who had passed ATLS after the second attempt, per centre
+n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
+n.residents <- n.atls.residents + n.ptc.residents
 centre.ids <- data %>% pull("id/reg_hospital_id") %>% unique() 
 n.centres <-  centre.ids %>% length()
 shuffled.centres <- centre.ids %>% sample(n.centres)
@@ -30,6 +35,17 @@ control.data <- do.call(rbind, centre.data[as.character(control.centres)])
 n.atls <- nrow(atls.data)
 n.ptc <- nrow(ptc.data)
 n.control <- nrow(control.data)
+## List of potentially eligible patient participants per site, replace
+## IDs with actual IDs once known and replace figures with actual
+## figures once known
+n.potentially.eligible = list(n.potentially.eligible.centre.1 = 999,
+                              n.potentially.eligible.centre.2 = 999,
+                              n.potentially.eligible.centre.3 = 999,
+                              n.potentially.eligible.centre.4 = 999,
+                              n.potentially.eligible.centre.5 = 999,
+                              n.potentially.eligible.centre.6 = 999,
+                              n.potentially.eligible.centre.7 = 999)
+recruitment.rate.patients <- 
 
 ## Label variables
 

### 2022-11-25 pilot-publication/manuscript.Rmd
Work on results

@@
 ## compile with:
-## Rscript -e "rmarkdown::render('manuscript-clean.Rmd', output_format = 'all')"
+## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
 title: A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes  
@@
 enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the
-standard care arm `r n.control` patients. The study flowchart is shown
-in Figure X and sample characteristics across trial arms are shown in
-Table X. Overall, the number of females were X (X%), the median age
-was X (IQR X), and the median ISS was X (IQR X). A total of X (X%)
-patients had the primary outcome of mortality at 30 days after arrival
-to the emergency department.
+standard care arm `r n.control` patients. We trained a total of 
+`r n.residents` residents, `r n.atls.residents` in ATLS, and 
+`r n.ptc.residents` in PTC.
+
+The study flowchart is shown in Figure X and patient sample
+characteristics across trial arms are shown in Table X. Overall, the
+number of females were X (X%), the median age was X (IQR X), and the
+median ISS was X (IQR X). A total of X (X%) patients had the primary
+outcome of mortality at 30 days after arrival to the emergency
+department.
 

### 2022-11-25 pilot-publication/functions/create_descriptive_table.R

@@ -1,0 +1,26 @@
+#' Create a Table of Descriptive Statistics
+#'
+#' Function to create a table of descriptive statistics, such as mean
+#' and median and proportions, given a dataset.
+#' @param table.data A data.frame. The data to use when creating the
+#'     table. No default.
+#' @param strata Character or NULL. The variable to use as the stratifying variable. If NULL the table is not stratified. Defaults to NULL.
+#' @param include.overall Logical. If TRUE a column called "Overall" is included in the table, which includes descriptive statistics for the whole sample. Defaults to TRUE.
+create_descriptive_table <- function(table.data, strata = NULL, include.overall = TRUE) {
+    ## Check arguments
+    assertthat::assert_that(is.data.frame(table.data))
+    assertthat::assert_that(is.character(strata) | is.null(strata))
+    assertthat::assert_that(is.logical(include.overall))
+
+    ## Create table
+    if (!is.null(strata)) {
+        table.data$strata <- table.data[, strata]
+        table.data[, strata] <- NULL
+    }
+    if (include.overall) {
+        overall.data <- table.data
+        overall.data$strata <- "Overall"
+        table.data <- cbind(overall.data, table.data)
+    }
+        
+}

### 2022-11-25 pilot-publication/main.R

@@ -45,8 +45,10 @@ n.potentially.eligible = list(n.potentially.eligible.centre.1 = 999,
                               n.potentially.eligible.centre.5 = 999,
                               n.potentially.eligible.centre.6 = 999,
                               n.potentially.eligible.centre.7 = 999)
-recruitment.rate.patients <- 
+recruitment.rate.patients <- round(n.patients/do.call(sum, n.potentially.eligible) * 100)
+
+## Create table of sample characteristics
+table.data <- data
+table.sample.characteristics <- create_descriptive_table(data = table.data)
 
 ## Label variables
-
-

### 2022-11-27 pilot-publication/manuscript.Rmd
Update manuscript.Rmd
Work on table of patient sample characteristics

@@
 output:
-    bookdown::word_document2:
+    bookdown::pdf_document2:
       toc: false
       number_sections: false
-      reference_docx: "reference.docx"
-    md_document: default
+##    md_document: default
+##    bookdown::word_document2:
+##      toc: false
+##      number_sections: false
+##      reference_docx: "reference.docx"
 ---
@@
 The study flowchart is shown in Figure X and patient sample
-characteristics across trial arms are shown in Table X. Overall, the
+characteristics across trial arms are shown in Table 
+\@ref(tab:sample-characteristics). Overall, the
 number of females were X (X%), the median age was X (IQR X), and the
@@
 department.
+
+```{r sample-characteristics, echo = FALSE, results = "asis"}
+create_descriptive_table(table.data,
+                         strata = "patinfo/pt_gender",
+                         caption = "Patient sample characteristics")
+```
 
@@
 without a control group, with heterogeneous results
-[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
+[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008;
 @Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
@@
 implementing in-hospital trauma life support training
-[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
+[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008],
 whereas one study from China that included 820 patients found a

### 2022-11-27 pilot-publication/functions/create_descriptive_table.R

@@ -6,21 +6,28 @@
 #'     table. No default.
 #' @param strata Character or NULL. The variable to use as the stratifying variable. If NULL the table is not stratified. Defaults to NULL.
 #' @param include.overall Logical. If TRUE a column called "Overall" is included in the table, which includes descriptive statistics for the whole sample. Defaults to TRUE.
-create_descriptive_table <- function(table.data, strata = NULL, include.overall = TRUE) {
+#' @param caption Character or NULL. The table caption. Defaults to NULL, as in no caption.
+create_descriptive_table <- function(table.data, strata = NULL, include.overall = TRUE, caption = NULL) {
     ## Check arguments
     assertthat::assert_that(is.data.frame(table.data))
     assertthat::assert_that(is.character(strata) | is.null(strata))
     assertthat::assert_that(is.logical(include.overall))
+    assertthat::assert_that(is.character(caption) | is.null(caption))
 
     ## Create table
+    table.data <- as.data.frame(table.data)
+    strata.data <- rep("Overall", nrow(table.data))
     if (!is.null(strata)) {
-        table.data$strata <- table.data[, strata]
+        strata.data <- table.data[, strata]
         table.data[, strata] <- NULL
     }
-    if (include.overall) {
-        overall.data <- table.data
-        overall.data$strata <- "Overall"
-        table.data <- cbind(overall.data, table.data)
-    }
-        
+    strata.list <- base::split(table.data, as.factor(strata.data))
+    if (!is.null(strata) & include.overall)
+        strata.list <- c(strata.list, list("Overall" = table.data))
+    labels <- list(variables = setNames(nm = names(table.data)))
+    descriptive.table <- table1::t1kable((table1::table1(strata.list,
+                                                         labels = labels,
+                                                         caption = caption,
+                                                         render.missing=table1::render.missing.default)))
+    return(descriptive.table)
 }

### 2022-11-27 pilot-publication/functions/prepare_data.R

@@ -1,0 +1,15 @@
+#' Prepare data
+#'
+#' This functions prepares the data by changing column types, merging
+#' categories, labelling columns and so on.
+#' @param data A data.frame. The data to be prepared. No default.
+#' @param codebook A list. A list describing the columns of data or
+#'     NULL. Defaults to NULL.
+prepare_data <- function(data, codebook = NULL) {
+    ## Check arguments
+    assertthat::assert_that(is.data.frame(data))
+    assertthat::assert_that(is.list(codebook) | is.null(codebook))
+    ## Prepare data
+    prepared.data <- data
+    return (prepared.data)
+}

### 2022-11-27 pilot-publication/main.R

@@ -3,6 +3,9 @@ library(dotenv)
 library(dplyr)
 library(lubridate)
 
+## Source functions
+noacsr::source_all_functions()
+
 ## Import data
 data <- readr::read_csv(Sys.getenv("DATA_DIR"))
 
@@ -10,6 +13,9 @@ data <- readr::read_csv(Sys.getenv("DATA_DIR"))
 codebook.arguments <- lapply(c("URL", "UID", "USERNAME", "PASSWORD"),
                              function(x) Sys.getenv(paste0("KOBO_", x)))
 codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
+
+## Prepare data
+data <- prepare_data(data)
 
 ## Define basic results
 arrival.dates <- data %>% pull("incident/date_of_arrival") %>% sort()
@@ -48,7 +54,10 @@ n.potentially.eligible = list(n.potentially.eligible.centre.1 = 999,
 recruitment.rate.patients <- round(n.patients/do.call(sum, n.potentially.eligible) * 100)
 
 ## Create table of sample characteristics
-table.data <- data
-table.sample.characteristics <- create_descriptive_table(data = table.data)
+table.variables <- c("patinfo/pt_age", "patinfo/pt_gender",
+                     "incident/dominating_injury_type",
+                     "patvitals/ed_rr", "patvitals/ed_sat",
+                     "patvitals/ed_hr", "patvitals/ed_sbp")
+table.data <- data[, table.variables]
 
 ## Label variables

### 2022-11-29 pilot-publication/manuscript.Rmd
Start working on patient participants outcomes

@@
 characteristics across trial arms are shown in Table
-\@ref(tab:sample-characteristics). Overall, the
-number of females were X (X%), the median age was X (IQR X), and the
-median ISS was X (IQR X). A total of X (X%) patients had the primary
-outcome of mortality at 30 days after arrival to the emergency
-department.
+\@ref(tab:sample-characteristics). Overall, the number of females were
+`r n.females` (`r p.females`%), the median age was `r median.age`
+years (IQR `r iqr.age`), and the median ISS was X (IQR X). A total of
+`r n.m30d` (`r p.m30d`%) patients had the primary outcome of mortality
+at 30 days after arrival to the emergency department.
 
@@
 ## Patient Participant Outcomes
-In the post training period, a total of X (X%) patients in the ATLS
-arm had the primary outcome, compared to X (X%) patients in the PTC
-arm, and X (X%) patients in the standard care arm. The differences
-between the arms were X (95% CI X - X) between the ATLS and PTC arms,
-X (95% CI X - X) between the ATLS and standard care arms, and X (95%
-CI X - X) between the PTC and standard care arms. The change from
-baseline for the primary outcome was X in the ATLS arm, X in the PTC
-arm, and X in the standard care arm. The differences between changes
-from baseline were X (95% CI X - X) between the ATLS and PTC arms, X
-(95% CI X - X) between the ATLS and standard care arms, and X (95% CI
-X - X) between the PTC and standard care arms. Table X shows
-differences in the primary outcome between trial arms with 75, 85, and
-95% CIs. Differences in secondary outcomes between trial arms and
-subgroups are available as Supplementary material X.
+In the post training period, a total of `r n.m30d.atls` 
+(`r p.m30d.atls`%) patients in the ATLS arm had the primary outcome,
+compared to `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, 
+and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard
+care arm. The absolute risk difference between the ATLS and PTC arm 
+was X (95% CI X - X), between the ATLS and standard care arms X 
+(95% CI X - X), and between the PTC and standard care arms X 
+(95% CI X - X).
+
+The change from baseline for the primary outcome was X in the ATLS
+arm, X in the PTC arm, and X in the standard care arm. The differences
+between changes from baseline were X (95% CI X - X) between the ATLS
+and PTC arms, X (95% CI X - X) between the ATLS and standard care
+arms, and X (95% CI X - X) between the PTC and standard care
+arms. Table X shows differences in the primary outcome between trial
+arms with 75, 85, and 95% CIs.
+
+Differences in secondary outcomes between trial arms and subgroups are
+available as Supplementary material X.
 
@@
 
+<!-- Consider discussing waiver of informed consent for initial 
+data collection -->
+
 Previous studies on the effect of in-hospital trauma life support

### 2022-11-29 pilot-publication/functions/outcomes.R

@@ -1,0 +1,34 @@
+#' Return vector of all binary outcomes
+binary_outcomes <- function() {
+    c("outcomes__alive_24h",
+      "outcomes__discharge_alive",
+      "outcomes__selfambulatory_at_discharge",
+      "outcomes__alive_after_30_days",
+      "outcomes__return_to_work",
+      "complications__pulmonary_complication",
+      "complications__complication_septic_shock",
+      "complications__complication_renal_failure",
+      "complications__complication_coagulopathy",      
+      "complications__need_for_reexploration_or_resurgery",
+      "complications__failure_of_conservative_management")
+}
+
+categorical_outcomes <- function() {
+    c("outcomes__eq5dm",
+      "outcomes__eq5dsc",
+      "outcomes__eq5dua",
+      "outcomes__eq5dpd", 
+      "outcomes__eq5dad",
+      "outcomes__eq5dm30d",
+      "outcomes__eq5dsc30d", 
+      "outcomes__eq5dua30d",
+      "outcomes__eq5dpd30d",
+      "outcomes__eq5dad30d")
+}
+
+quantitative_outcomes <- function() {
+    c("complications__number_of_hospitalizations_for_this_injury",
+      "outcomes__eq5dhs",
+      "outcomes__eq5dhs30d",
+      "outcomes__cost_of_treatment")
+}

### 2022-11-29 pilot-publication/functions/prepare_data.R

@@ -13,9 +13,11 @@ prepare_data <- function(data, codebook = NULL) {
     variable.names <- codebook$survey$name
     codebook$survey$type <- setNames(codebook$survey$type, nm = variable.names)
     codebook$survey$label <- setNames(codebook$survey$label, nm = variable.names)
+    ## Rename variables
+    names(data) <- gsub("/", "__", names(data), fixed = TRUE)
     ## Prepare data
     prepared.data <- data
-    prepared.data$`patinfo/pt_age` <- as.numeric(prepared.data$`patinfo/pt_age`)
+    prepared.data$`patinfo/pt_age` <- as.numeric(prepared.data$patinfo__pt_age)
     ## Replace with missing
     prepared.data <- prepared.data %>%
         naniar::replace_with_na_if(.predicate = is.character,
@@ -34,7 +36,7 @@ prepare_data <- function(data, codebook = NULL) {
 
 label_variable <- function(variable.data, name, codebook) {
     relabelled.data <- variable.data
-    name.components <- strsplit(name, "/", fixed = TRUE)
+    name.components <- strsplit(name, "__", fixed = TRUE)
     variable.name <- name.components[[1]][2]
     variable.label <- codebook$survey$label[variable.name]
     type <- codebook$survey$type[variable.name]

### 2022-11-29 pilot-publication/main.R

@@ -18,7 +18,7 @@ codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
 data <- prepare_data(data, codebook)
 
 ## Define basic results
-arrival.dates <- data %>% pull("incident/date_of_arrival") %>% sort()
+arrival.dates <- data %>% pull(incident__date_of_arrival) %>% sort()
 format_date <- function(date) paste0(month(date[1], label = TRUE, abbr = FALSE), " ", year(date[1]))
 start.date <- format_date(arrival.dates[1])
 end.date <- format_date(rev(arrival.dates)[1])
@@ -28,19 +28,41 @@ n.atls.residents.passed.first.attempt <- 3 + 2 # The number of residents who had
 n.atls.residents.passed.first.attempt <- 4 + 2 # The number of residents who had passed ATLS after the second attempt, per centre
 n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
 n.residents <- n.atls.residents + n.ptc.residents
-centre.ids <- data %>% pull("id/reg_hospital_id") %>% unique() 
+centre.ids <- data %>% pull(id__reg_hospital_id) %>% unique() 
 n.centres <-  centre.ids %>% length()
 shuffled.centres <- centre.ids %>% sample(n.centres)
 atls.centres <- shuffled.centres[1:2]
 ptc.centres <- shuffled.centres[3:4]
 control.centres <- shuffled.centres[5:n.centres]
-centre.data <- data %>% split(data$`id/reg_hospital_id`)
+centre.data <- data %>% split(data$id__reg_hospital_id)
 atls.data <- do.call(rbind, centre.data[as.character(atls.centres)])
 ptc.data <- do.call(rbind, centre.data[as.character(ptc.centres)])
 control.data <- do.call(rbind, centre.data[as.character(control.centres)])
+tlsp.data.list <- list(atls = atls.data, ptc = ptc.data, control = control.data)
 n.atls <- nrow(atls.data)
 n.ptc <- nrow(ptc.data)
 n.control <- nrow(control.data)
+n.females <- with(data, sum(patinfo__pt_gender == "Female"))
+p.females <- round(n.females/nrow(data) * 100)
+median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
+iqr.age <- quantile(data$patinfo__pt_age, probs = c(0.25, 0.75), na.rm = TRUE) %>% paste0(collapse = "-")
+
[…]

### 2022-12-02 pilot-publication/manuscript.Rmd
Finalise early but incomplete draft

@@
 
+<!--
 ## Corresponding author
@@
 
+-->
+
 ## Keywords
@@
 
+<!--
 ## Word Count
-<!-- 4494 -->
+4494
+ -->
 
 ## Role of study sponsor and funders
-The funding sources had no role in the design of this study and will not
-have any role during its execution, analyses, interpretation of the
-data, or decision to submit results.
+The funding sources had no role in the design of this study nor during
+its execution, analyses, interpretation of the data, or decision to
+submit the results.
 
-\newpage
+<!-- \newpage -->
 
@@
 **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres.
-The patient recruitment rate was X and the resident recruitment rate was X.
-The loss to follow up rate was X.
-The all-cause 30 day mortality was X in the ATLS arm, X in the PTC arm, and X in the standard care arm.
+The patient recruitment rate was X and the resident recruitment rate was `r recruitment.rate.residents`%.
+The lost to follow up rate was `r rate.lost.to.follow.up`%.
+`r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm.
 **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
@@
 
+<!--
 # Article Summary
@@
 - Participating centers' heterogeneity may affect the study estimates and bias the results.
-
+-->
 \newpage
@@
 with the outcome (if binary), and the intracluster correlation
-coefficient [@Hemming2017]. The objectives of this pilot study will be
+coefficient [@Hemming2017]. The objectives of this pilot study were
 to:
@@
 ## Trial Design
-This study will pilot a pragmatic three-armed parallel, cluster
-randomised, controlled trial. There will be two intervention arms,
-ATLS and PTC training, and one control arm, standard care. We will
-collect data for four months in all three arms, first during a one
-month observation phase and then during a three month intervention
-phase (or continued observation in the control arm). This design will
-allow us to assess outcomes both as final values and as change from
-baseline. Our study is a pilot study because its objectives involves
-estimating quantities, such as the probable effect sizes, proportion
-of participants with the outcome (if binary), and the intracluster
+We piloted a three-armed cluster randomised controlled trial. There
+were two intervention arms, ATLS and PTC training, and one control
+arm, standard care. We collected data for four months in all three
+arms, first during a one month observation phase and then during a
+three month intervention phase (or continued observation in the
+control arm). This design allowed us to assess outcomes both as
+final values and as change from baseline.
+
+Our study is a pilot study because its objectives involves estimating
+quantities, such as the probable effect sizes, proportion of
+participants with the outcome (if binary), and the intracluster
 correlation coefficient, needed for the sample size calculations of a
-full-scale trial [@Hemming2017]. The full-scale trial will be planned
-regardless of the effect-sizes identified in this pilot
-study. This pilot study will also establish how many
-participants that can be enrolled, as well as likely drop out rates,
-and the feasibility of collecting primary and secondary outcomes.
+full-scale trial [@Hemming2017]. This pilot study also establish how
+many participants that can be enrolled, as well as likely drop out
+rates, and the feasibility of collecting primary and secondary
+outcomes.
 
@@
 neither ATLS, PTC, nor any other trauma life support training program
-is routinely taught. India is the world's second most populous country
-and has 20% of the world's trauma deaths. The trauma system is still
-developing, with limited prehospital care, and the in hospital trauma
-mortality as well as the proportion of preventable deaths remain
-high. Lack of standard trauma training for healthcare providers,
-limited hospital resources, inadequate processes of care, overcrowding
-emergency departments - are some of the factors that contribute to the
-high mortality and morbidity. During recent years efforts have been
-made to improve hospital trauma care, through capacity building for
-trained trauma care providers, augmenting facilities, and developing
-care protocols within the hospitals. We will conduct this trial in
-India because training providers in a trauma life support program is
-not yet the standard.
+is routinely taught. The seven hospitals were King Edward Memorial
+Seth G. S. Medical College and K.E.M. Hospital, Mumbai; Lokmanya Tilak
+Municipal Medical College and General Hospital, Mumbai; HBT Medical
+College And Dr. R N Cooper Municipal General Hospital, Mumbai; Medica
+Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata;
+Sir Nil Ratan Sircar Medical College & Hospital, Kolkata; Postgraduate
+Institute of Medical Education & Research, Chandigarh.
 
-The seven hospitals were King Edward Memorial Seth G. S. Medical
-College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
-College and General Hospital, Mumbai; HBT Medical College And Dr. R N
-Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
-Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
-Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
-Medical Education & Research, Chandigarh.
+<!--
+The trauma system is still developing, with
+limited prehospital care, and the in hospital trauma mortality as well
+as the proportion of preventable deaths remain high. Lack of standard
+trauma training for healthcare providers, limited hospital resources,
+inadequate processes of care, overcrowding emergency departments - are
+some of the factors that contribute to the high mortality and
+morbidity. During recent years efforts have been made to improve
+hospital trauma care, through capacity building for trained trauma
+care providers, augmenting facilities, and developing care protocols
+within the hospitals. 
+-->
 
 ## Eligibility Criteria for Participants and Clusters
-There will be two groups of participants: patients and resident
-doctors.
+There were two groups of participants: patients and resident doctors.
 
@@
 Classification of Disease version 10 (ICD-10) codebook as reason for
-presenting. We will explore intervention effects across the following
+presenting. We explored intervention effects across the following
 clinical subgroups: men, women, blunt multisystem trauma, penetrating
-trauma, shock, severe traumatic brain injury, and elderly, as defined by
-Hornor et al [@Hornor2018]. The consent form for patients are available
-as Supplemental Material 1.
+trauma, shock, severe traumatic brain injury, and elderly, as defined
+by Hornor et al [@Hornor2018]. The consent form for patients are
+available as Supplemental Material 1.
 
 ### Resident Doctor Participants
-Resident doctors doing their speciality training in surgery (in India
-it is very rare to have emergency medicine managing trauma patients in
-the emergency department, why we decide to focus on surgical units in
-this pilot), who manage trauma patients in the emergency department,
-and who are expected to remain in the participating hospitals for at
-least one year. To facilitate administration each surgical department
-is divided into units, which manages the out patient department,
-emergency department, operating rooms etc on different days each
-week. One or two, out of typically six, units'
-residents will be selected from each hospital. One unit consists of at
+Resident doctors doing their speciality training in surgery or
+emergency medicine managing trauma patients in the emergency
+department and who were expected to remain in the participating
+hospitals for at least one year. To facilitate administration each
+surgical department is divided into units, which manages the out
+patient department, emergency department, operating rooms etc on
+different days each week. One or two, out of typically six, units'
+residents were selected from each hospital. One unit consists of at
 least three faculty and three to twelve residents.
 
-To be eligible, units should have a maximum of 25% of the doctors
+To be eligible, units had to have a maximum of 25% of the doctors
 trained in either ATLS, PTC, or similar training programs before the
-start of the pilot (hospitals that have so far agreed to participate
-have no or single current residents trained in any program). Those
-residents who have received training in the last five years will be
-considered as trained. The figure of 25% was decided through consensus
-in the research team, to balance feasibility and contamination of
-results. We will select the units by conducting a prior survey to
-ascertain this criteria. Consent will be sought from the residents in
-each of the intervention groups before they undergo the ATLS or PTC
-training. The consent form for residents are available as
-Supplemental Material 2. We will not ask
-for consent from residents at the units in the control hospitals as
-their practice will not be affected by this pilot and we will not
-collect any personal identifiable data on them. This is in line with
-ethical regulations in the study setting.
+start of the pilot. Those residents who had received training in the
+last five years were considered as trained. The figure of 25% was
+decided through consensus in the research team, to balance feasibility
+and contamination of results.
+
+Consent was sought from the residents in each of the intervention
+groups before they underwent the ATLS or PTC training. The consent
+form for residents is available as Supplemental Material 2. We did
+not ask for consent from residents at the units in the control
+hospitals as their practice was not affected by this pilot and we
+did not collect any personal identifiable data on them.
 
@@
 Indian tertiary care hospitals that admit 400-800 adult patients with
-trauma each year. We randomise on the cluster (hospital) level to avoid
-contamination between intervention and control arms. To be eligible for
-inclusion hospitals have to provide the following services round the
-clock: operation theatres, X-ray, CT, and ultrasound facilities, and
-blood bank. In addition the baseline admission rate should be more than
-35 adult patients with major trauma per month.
+trauma each year. We randomised on the cluster (hospital) level to
+avoid contamination between intervention and control arms. To be
+eligible for inclusion hospitals had to provide the following services
+round the clock: operation theatres, X-ray, CT, and ultrasound
+facilities, and blood bank. In addition the baseline admission rate
+had to be more than 35 adult patients with major trauma per month.
 
 ## Interventions
-In each intervention arm one or two units', out of typically
-six, residents per hospital providing
-emergency care to trauma patients will be trained in either ATLS or
-PTC. For the purpose of this pilot study, we will target to train a
-minimum of 75% of residents in each unit. If residents drop out or
-change units after training but before data collection is completed we
-will conduct additional training if needed to meet the 75% criterion.
-We will not train the units' faculty, as they are typically not
-involved in the initial management of trauma patients.
+In each intervention arm one or two units', out of typically six,
+residents per hospital providing emergency care to trauma patients
+were trained in either ATLS or PTC. For the purpose of this pilot
+study, we will target to train a minimum of 75% of residents in each
+unit. If residents dropped out or changed units after training but
+before data collection was completed we planned to conduct additional
+training if needed to meet the 75% criterion, but this was not
+required. We did not train the units' faculty, as they are typically
+not involved in the initial management of trauma patients.
 
-The ATLS training will be conducted in the nearest ATLS certified
-training centre in India according to the standard ATLS curriculum
-[@acsAtls2018]. The PTC training will be arranged in hospitals
-randomized to the PTC arm, according to the standard PTC curriculum
-[@ptcfPtc2018]. These courses will be conducted over a period of
-2.5 to 3 days. The residents certified "pass" will be considered as
-trained in respective courses.
+The ATLS training was conducted in an ATLS certified training centre
+in Mumbai, according to the standard ATLS curriculum
+[@acsAtls2018]. The PTC training was conducted in New Delhi, according
+to the standard PTC curriculum [@ptcfPtc2018]. These courses were
+conducted over a period of 2.5 to 3 days. The residents certified
+"pass" were considered as trained in respective courses.
 
@@
 Both ATLS and PTC are standard training programs with fixed curricula
-[@acsAtls2018; @ptcfPtc2018]. We will not modify the
-delivery or content of these programs during this pilot.
+[@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
+content of these programs during this pilot.
 
 ### Adherence
-The intervention is the training in either ATLS or PTC. Participants are
-required to adhere to, i.e. participate in, the training, to be eligible
-for passing. We will not consider adherence to training contents during
-care delivery as adherence to the trial intervention, but rather as a
-provider level outcome.
+The intervention was the training in either ATLS or PTC and resident
+participants were required to adhere to, i.e. participate in, the
+training, to be eligible for passing. We did not consider adherence to
+training contents during care delivery as adherence to the trial
+intervention.
 
@@
 #### Baseline Training
-The care provided by all participating hospitals at baseline is based on
+
+**This section needs to be updated with the emergency medicine curriculum**
+
+The care provided by all participating hospitals at baseline was based on
 the training curriculum formulated by The National Medical Council of
@@
 #### Standard of Care
-At most medical colleges in India trauma patients present to the
+
+**This section needs to be revised to include information from sites
+where patients were not managed by surgical residents**
+
+At most tertiary hospitals in India trauma patients present to the
 emergency department where they are assessed by a doctor and referred
@@
 second or third year general surgery resident sees all the major
-trauma and provide the initial care, including
-initiating treatment and investigations.  This resident informs the
-consultant on call who is generally an Assistant Professor. Most
-procedures like intercostal drainage, open wound suturing, intubation
-etc. would be done in the surgical resuscitation area, by the surgical
-resident.
+trauma and provide the initial care, including initiating treatment
+and investigations.  This resident informs the consultant on call who
+is generally an Assistant Professor. Most procedures like intercostal
+drainage, open wound suturing, intubation etc. would be done in the
+surgical resuscitation area, by the surgical resident.
 
@@
 (not recording) and giving injections. They also accompany the
-resident during transfers of serious
-patients.
+resident during transfers of serious patients.
 
@@
 ## Outcomes
-Our pilot study include both participant and feasibility
+Our pilot study included both participant and feasibility
 outcomes. Prior to deciding on these participant outcomes we searched
@@
 
-The primary participant outcome will be all cause mortality within 30
-days from the time of arrival to the emergency department. The
-primary outcome and most secondary outcome will be assessed and
-compared both as final values and as change from
-baseline. All outcomes that pertain to the
-individual participant level are detailed in Supplemental Material
+The primary participant outcome was all cause mortality within 30 days
+from the time of arrival to the emergency department. The primary
+outcome and most secondary outcome were assessed and compared both as
+final values and as change from baseline. All outcomes that pertain to
+the individual participant level are detailed in Supplemental Material
 3. We decided to include a large number of outcomes,
 including some more exploratory, so that we can test their feasibility
-and relevance. We may remove secondary participant outcomes during the
-course of the pilot study, if they prove to be too difficult to
-collect. If we remove outcomes we will document the reasons for doing
-so.
+and relevance.
 
-We will also assess the following feasibility outcomes, which pertain
+We also assessed the following feasibility outcomes, which pertain
 both to overall study population as well as to the individual cluster
@@
 
-- Recruitment rate. For both patients and residents this will equal
+- Recruitment rate. For both patients and residents this was equal to
   the proportion of participants enrolled, out of the total number of
   eligible participants, over the course of the pilot study.
-- Lost to follow up rate. This will apply only to patients and equal
-  the proportion of patients that do not complete 30 day follow up,
+- Lost to follow up rate. This applied only to patients and was equal
+  to the proportion of patients that do not complete 30 day follow up,
   out of all enrolled patients, over the course of the pilot study.
-- Pass rate. This will apply only to residents in the intervention
+- Pass rate. This applied only to residents in the intervention
   arms and equal the proportion of residents that pass the training
@@
   course of the pilot study.
-- Missing data rate. This will apply to each outcome and variable and
-  equal the proportion of missing data, over the course of the pilot
-  study.
+- Missing data rate. This applied to each outcome and variable and was
+  equal to the proportion of missing data, over the course of the
+  pilot study.
 - Differences in distributions of observed and extracted data. This
-  will apply to each outcome and variable and will compare the
-  distributions of data collected by observations versus extracted
-  from hospital records. For quantitative variables this will be the
-  difference in means, standard deviations, medians, interquartile
-  ranges, and ranges. For qualitative variables this will be the
-  differences in absolute counts and percentages, across categories.
+  applied to each outcome and variable and compared the distributions
+  of data collected by observations versus extracted from hospital
+  records. For quantitative variables this was be the difference in
+  means, standard deviations, medians, interquartile ranges, and
+  ranges. For qualitative variables this was the differences in
+  absolute counts and percentages, across categories.
 
@@
 ### Patients
-Patients will be screened for eligibility as they arrive at the
-emergency department. Eligible patients will be approached in the
-emergency department to consent to follow up, if they are
-conscious. If they are unconscious a patient representative will be
-approached to consent to follow up. Once the patient is conscious we
-will approach the patient to affirm the patient representative's
-consent. We will follow up patients at discharge, at 24 hours after
-arrival at the emergency department, and at 30 days after arrival at
-the emergency department.
+Patients were screened for eligibility as they arrive at the emergency
+department. Eligible patients were approached in the emergency
+department to consent to follow up, if they are conscious. If they
+were unconscious a patient representative was approached to consent to
+follow up. Once the patient is conscious we approached the patient to
+affirm the patient representative's consent. We followed up patients
+at discharge, at 24 hours after arrival at the emergency department,
+and at 30 days after arrival at the emergency department.
 
 ### Residents
-Surgical units will be screened for eligibility once hospitals confirm
-their participation. All residents in eligible units will be approached
-to consent to training if their hospital is randomised to either of the
-intervention arms. Training will be conducted as soon as possible after
-the study starts. Resident participants will be followed up 30 days
-after training, if they are in the intervention arms, or 30 days after
-the study started, if they are in the control arm.
+Participating units were be screened for eligibility once hospitals
+confirmed their participation. All residents in eligible units were
+approached to consent to training if their hospital was randomised to
+either of the intervention arms. Training was conducted approximately
+one month after the study started.
 
 ## Sample size
-Given budget and time constraints, including the rotation of surgical
-units in Indian hospitals (which often happen on a six months basis)
-the feasible data collection period is four months. Each of the
-surgical units see 2-4 trauma patients per week. If we select a
-minimum of one unit per hospital then each hospital will enrol 8-16
-patients per month and 32-64 patients during the four months of this
-pilot. With a 20% attrition rate we expect each hospital to enrol
-26-51 patients, coming to a total sample size of between 156 and 306
-patients for this pilot study.
+Given budget and time constraints, including the rotation of units in
+the emergency departments in Indian hospitals (which often happen on a
+six months basis) the feasible data collection period was four
+months. We assumed that each of the units saw 2-4 trauma patients per
+week and based on that estimated that if we selected a minimum of one
+unit per hospital then each hospital would enrol 8-16 patients per
+month and 32-64 patients during the four months of this pilot. With a
+20% attrition rate we expected each hospital to enrol 26-51 patients,
+coming to a total sample size of between 156 and 306 patients for this
+pilot study.
 
 ## Recruitment
-To ensure adequate recruitment we only approach hospitals with trauma
-volumes high enough to allow us to reach the sample size goals detailed
-above. Patients will be enrolled by a dedicated project officer as they
-arrive at the emergency department. The recruitment period will be three
-months. Recruitment will be monitored weekly through online conferences.
-No financial or non-financial incentives will be provided to trial
-investigators or participants for enrolment.
+To ensure adequate recruitment we only approached hospitals that
+stated that the their volumes were high enough to allow us to reach
+the sample size goals detailed above. Patients were enrolled by a
+dedicated project officer as they arrived at the emergency
+department. The recruitment period was four months. Recruitment was
+monitored weekly through online conferences.  No financial or
+non-financial incentives were provided to trial investigators or
+participants for enrolment.
 
@@
 The random allocation sequence was generated by MGW. Clusters were
-enrolled by the project's core team, who also enrolled
-clusters. Patient participants will be included if they present during
-the project officers shift. Resident participants are enrolled if they
-are in the units selected for training. We will use simple random
-sampling to select units if there are more than two eligible units in
-a hospital. For patient participants consent for follow up is sought
-after randomisation from patients or patient relatives as
-appropriate. For resident participants consent is sought before
-randomisation. If residents in a unit decline to participate, so that
-the target of training 75% of residents in a given unit cannot be met,
-another unit will be selected for participation.
+enrolled by the project's core team. Patient participants were
+enrolled if they presented during the project officers shift. Resident
+participants were enrolled if they were in the units selected for
+training. The principal investigator at each hospital selected the
+units for training. For patient participants consent for follow up was
+sought after randomisation from patients or patient relatives as
+appropriate. For resident participants consent was also sought after
+randomisation.
 
 ## Blinding
-It will not be possible to blind investigators or participants to
-interventions. We will not blind the data analysts during this pilot,
-but we plan to blind the data analysts during the full-scale
-trial.
+It was not possible to blind investigators or participants to
+interventions. We did not blind the data analysts during this pilot,
+but we plan to blind the data analysts during the full-scale trial.
 
 ## Data Collection
-Data collection will start one month before the training is delivered,
-to establish a baseline. A variability of three months of the date when
-data collection is started between hospitals will be accepted. Each
-participating hospital will have a dedicated project officer to collect
-data. The project officers will have a masters in a health science field
-and should have experience in data collection.
+Data collection started one month before the training was delivered,
+to establish a baseline. Each participating hospital had a dedicated
+project officer who collected the data.
 
-Because participating residents are assigned designated days for trauma
-care for a period of 6 months, data will be collected during those
-particular days and shifts when these trained doctors are in the
-emergency department. The project officers will collect data both by
+Because participating residents were assigned designated days for
+trauma care for a period of 6 months, data were collected during those
+particular days and shifts when these trained doctors were in the
+emergency department. The project officers collected data both by
 observing the care delivered and by interviewing the participants, and
@@
 
-Data collection will continue for a minimum of three months after
-training. The research officers will collect data of all patients, who
-present with trauma in the surgical bay during their duty hours. Those
-patients who are admitted will be followed up for complications and
-other in-hospital outcome measures, for example length of stay. Patients
-who are not admitted will be followed up telephonically for mortality
-outcomes and quality of life outcomes. The follow up period will be 30
-days. The project officers will make at most three attempts to reach a
-participant or participant representative telephonically, after which
-the data will be recorded as missing.
+Data collection continued for three months after training. The
+research officers collected data of all trauma patients who presented
+to the participating units during their duty hours. Those patients who
+were admitted were followed up for complications and other in-hospital
+outcome measures, for example length of stay. Patients who were not
+admitted were followed up telephonically for mortality outcomes and
+quality of life outcomes. The follow up period was be 30 days. The
+project officers made at most three attempts to reach a participant or
+participant representative telephonically, after which the data were
+recorded as missing.
 
-The project officer will administer the study information and informed
-consent (consent will only be sought for data collection including
-follow up) to the patient, or the patient's representative as
-appropriate, once the patient is stabilised. They will continue to
-collect data once they have received the consent.
+The project officer administered the study information and informed
+consent (consent was sought only for data collection including follow
+up) to the patient, or the patient's representative as appropriate,
+once the patient was stabilised. They will continue to collect data
+once they have received the consent.
 
-Details of data of those patients/relatives not willing to give consent
-will be removed from the analysis. The number of patients who opt out
-from data collection will be collected, as well as limited data on their
-age and sex. Patients will be followed up in the ward regularly for the
-various outcome variables. They will also be followed up telephonically
-after they have been discharged.
+Details of data of those patients/relatives not willing to give
+consent were not included in the analysis. The number of patients who
+opt out from data collection was collected.
 
 ## Variables
-The project officers will collect data on demographics, time of injury
-to arrival at the participating hospital, time to recording vital
-signs, vital signs, and times to and management details including
-imaging and surgery. Details of any injury sustained will be collected
-and coded using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD
-10 coders will undergo the WHO online ICD 10 training module and for
-AIS they will be accredited. Based on AIS we will calculate the Injury
-Severity Score (ISS) and the New ISS (NISS). Supplemental Material 4
-contains a full variable list, with definitions.
+The project officers collected data on demographics, time of injury to
+arrival at the participating hospital, time to recording vital signs,
+vital signs, and times to and management details including imaging and
+surgery. Details of any injury sustained were collected and coded
+using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD 10 coders
+underwent the WHO online ICD 10 training module and for AIS they were
+accredited. Based on AIS we calculated the Injury Severity Score
+(ISS) and the New ISS (NISS). Supplemental Material 4 contains a full
+variable list, with definitions.
 
 ## Patient and public involvement
-In this study, we will conduct community consultations to collect
+In this study, we conducted community consultations to collect
 inputs from patients, their caregivers, patient groups, and resident
@@
 
-During the pilot study, interviews will be conducted with
-post-discharge trauma patients and their caregivers to identify
-outcomes most relevant to them. These patients will be identified
-through the medical registers of the participating hospitals,
-contacted through telephone, and after receiving their consent be
-interviewed as per their convenience. Their consent form is available
-as Supplemental Material 5.  Additionally, members from non-government
-organizations working with trauma patients and the hospital Social
-Service Section will also be contacted for their views on contextual
-patient-centred outcomes for trauma patients. Their consent form is
-available as Supplemental Material 6. For feasibility, these
-interviews will be held in each of the cities where the participating
-centres are located. The commonest patient-centred outcomes reported
-across all the locations will be incorporated into the evaluation of
-the effects of the different training programs and standardized care
-on patient outcomes.
+During the pilot study, interviews were conducted with post-discharge
+trauma patients and their caregivers to identify the outcomes most
+relevant to them. These patients were identified through the medical
+registers of the participating hospitals, contacted through telephone,
+and after receiving their consent they were interviewed as per their
+convenience. Their consent form is available as Supplemental
+Material 5.  Additionally, members from non-government organizations
+working with trauma patients and the hospital Social Service Section
+were contacted for their views on contextual patient-centred outcomes
+for trauma patients. Their consent form is available as Supplemental
+Material 6. For feasibility, these interviews were held in each of the
+cities where the participating centres are located. The commonest
+patient-centred outcomes reported across all the locations will be
+incorporated into the evaluation of the effects of the different
+training programs and standardized care on patient outcomes.
 
 Similarly, the inputs of resident doctor participants at each
-participating centre will be collected during the pilot study. A
-discussion and periodic surveys will be conducted to document any
+participating centre were collected during the pilot study. A
+discussion with the residents were conducted to document any
 challenges or suggestions they may have in the scheduling or
@@
 A summary of the findings of the study as well as their inputs will be
-shared with those who participated in the interviews and surveys. A
-meeting will be held with the patient participants, at each city, where
-the changes in the measured patient-centered outcomes would be presented
-to them. Another meeting will be held with the resident doctors at each
-hospital to present the confidence of the residents after being trained.
-Any suggestions and reflections from the participants during the
-meetings will be used as inputs for planning the final study.
+shared with those who participated in the interviews and
+discussions. A meeting will be held with the patient participants, at
+each city, where the changes in the measured patient-centered outcomes
+would be presented to them. Another meeting will be held with the
+resident doctors at each hospital to present the confidence of the
+residents after being trained.  Any suggestions and reflections from
+the participants during the meetings will be used as inputs for
+planning the final study.
 
 ## Data management
-We will supply an online data collection tool, accessible only over a
-virtual private network (VPN), for each participating hospital to upload
-pseudonymised data to secure servers. Data validation techniques like
-restricted values or values of a specific range will be used to avoid
-ambiguous data entries and ensure the validity of the data. Ambiguous
-responses, data errors, if any, will be resolved after discussion with
-the core team during weekly meetings. An instruction manual or codebook
-for data variables will be prepared to ensure consistency in data entry.
-This manual will be referred to during the project data collection and
-variable descriptions are visible for each variable in the online data
-collection tool. Pseudonymised data will be stored at the centralised
-server. The data will be accessible by the project's principal
-investigator or by delegation of the project principal investigator
-only.
+We used the online data collection tool KoboToolbox [@KoboToolbox],
+accessible only over a virtual private network (VPN), for each
+participating hospital to upload pseudonymised data to secure
+servers. Data validation techniques such as restricted values or
+values within a specific range were used to avoid ambiguous data
+entries and ensure the validity of the data. Ambiguous responses and
+errors were resolved after discussion with the core team during weekly
+meetings. An instruction manual or codebook for data variables was
+prepared to ensure consistency in data entry. This manual was referred
+to during the project data collection and variable descriptions were
+visible for each variable in the online data collection
+tool. Pseudonymised data were stored at the centralised server. The
+data were only accessible by the project's principal investigator or
+by delegation by the project principal investigator only.
 
 ## Data monitoring
-Weekly meetings with the core team and project officers will take
-place and for this meeting a data status report will be automatically
-generated highlighting missing data and number of patients awaiting
-followup. We conducted an interim analysis on October 12, 2022, and
-decided to complete the study as participants were consenting to be
-included in the study and key variables including mortality outcomes
-could be collected. No outcomes were dropped. We did not use a data
-monitoring committee in the pilot study due to its limited scope.
+Weekly meetings with the core team and project officers took place. We
+conducted an interim analysis on October 12, 2022, and decided to
+complete the study as participants were consenting to be included in
+the study and key variables including mortality outcomes could be
+collected. No outcomes were dropped. We did not use a data monitoring
+committee in the pilot study due to its limited scope.
 
 ## Statistical Methods
-We will analyse all pilot data using descriptive
-statistics. Quantitative variables will be summarised as mean +/-
-standard deviation, median, interquartile range and range. Qualitative
-variables will be presented as absolute numbers and
-percentages. Feasibility outcomes will be summarised both on the
-overall sample level as well as on the individual cluster level. We
-will use an empty generalised linear mixed model<!-- see [@Sommet2017]
---> to estimate the intracluster correlation coefficient.
+We analysed all pilot data using descriptive statistics. Quantitative
+variables are summarised as mean +/- standard deviation, median,
+interquartile range and range. Qualitative variables are presented as
+absolute numbers and percentages. Feasibility outcomes are summarised
+both on the overall sample level as well as on the individual cluster
+level. We used an empty generalised linear mixed model<!-- see
+[@Sommet2017] --> to estimate the intracluster correlation
+coefficient.
 
-We will compare participant outcomes in three combinations of trial
-arms: ATLS versus PTC, ATLS versus standard care, and PTC versus
-standard care. In each combination we will compare both differences in
-final values and differences in change from baseline. For example, for
-the primary participant outcome of all cause mortality within 30 days
-from the time of arrival to the emergency department, comparing ATLS
-versus PTC, we will compare both the difference in mortality between
-the ATLS and PTC arms as well as the difference in the change from
-baseline in mortality between the ATLS and PTC arms.
+We compared participant outcomes in three combinations of trial arms:
+ATLS versus PTC, ATLS versus standard care, and PTC versus standard
+care. In each combination we compared both differences in final values
+and differences in change from baseline. For the intervention arms the
+change from baseline was calculated as the difference between the one
+month period of data collection before the training was undertaken and
+the three month period after the training. For the control arm the
+data collection period was four months and the difference from
+baseline was calculated as the difference between the first one
+month and the following three months.
 
-For the intervention arms the change from baseline will be calculated
-as the difference between the one month period of data collection
-before the training was undertaken and the three month period after
-the training. For the control arm the data collection period will be
-four months and the difference from baseline will be calculated as the
-difference between the first one month and the following three months.
+Within each combination of trial arms we conducted subgroup analyses
+of men, women, blunt multisystem trauma, penetrating trauma, shock,
+severe traumatic brain injury, and elderly. Table S7.1 in Supplemental
+Material 7 shows which outcomes that were assessed in what subgroups,
+decided through consensus in the research team. We further compared
+the results of all subgroups with the results in the whole cohort, and
+compared the results in the female subgroup with the male subgroup,
+and the results in the blunt multisystem trauma subgroup with the
+penetrating trauma subgroup. We are aware that the numbers in some of
+these subgroups are small, but we included them to help guide the
+formulation of the statistical analysis plan for the full-scale trial.
 
-Within each combination of trial arms we will conduct subgroup
-analyses of men, women, blunt multisystem trauma, penetrating trauma,
-shock, severe traumatic brain injury, and elderly. Table S7.1 in
-Supplemental Material 7 shows which outcomes will be assessed in which
-subgroups, decided through consensus in the research team. We will
-further compare the results of all subgroups with the results in the
-whole cohort, and compare the results in the female subgroup with the
-male subgroup, and the results in the blunt multisystem trauma
-subgroup with the penetrating trauma subgroup. We are aware that the
-numbers in some of these subgroups are likely to be small, but we
-include them to help guide the formulation of the statistical analysis
-plan for the full-scale trial.
-
-We will calculate both absolute and relative differences for each
-comparison, along with 75, 85, and 95% confidence intervals. We will
-use an empirical bootstrap procedure with 1000 draws to estimate these
-confidence intervals. We will not perform any formal hypothesis tests
-during the analysis of this pilot's data [@Lee2014]. We will also
-compare the data collected through observations and interviews with
-the data collected from hospital records, to assess the feasibility of
+We calculated both absolute and relative differences for each
+comparison, along with 75, 85, and 95% confidence intervals. We used
+an empirical bootstrap procedure with 1000 draws to estimate these
+confidence intervals. We did not perform any formal hypothesis tests
+during the analysis of this pilot's data [@Lee2014]. We also compared
+the data collected through observations and interviews with the data
+collected from hospital records, to assess the feasibility of
 collecting data from hospital records in the full-scale trial.
@@
 # Results
-
-**Note that participating centres are randomly assigned to the
-different arms every time this manuscript is compiled**
-
 Between `r start.date` and `r end.date`, we enrolled `r n.patients`
@@
 `r n.females` (`r p.females`%), the median age was `r median.age`
-years (IQR `r iqr.age`), and the median ISS was X (IQR X). A total of
-`r n.m30d` (`r p.m30d`%) patients had the primary outcome of mortality
-at 30 days after arrival to the emergency department.
+years (IQR `r iqr.age`)<!--, and the median ISS and NISS were X (IQR
+	X) and X (IQR X) respectively-->. A total of `r n.m30d` (`r p.m30d`%)
+patients had the primary outcome of mortality at 30 days after arrival
+to the emergency department.
 
 ```{r sample-characteristics, echo = FALSE, results = "asis"}
-create_descriptive_table(table.data,
-                         strata = "patinfo/pt_gender",
-                         caption = "Patient sample characteristics")
+table1::t1kable(sample.characteristics.table)
 ```
@@
 ## Patient Participant Outcomes
-In the post training period, a total of `r n.m30d.atls` 
-(`r p.m30d.atls`%) patients in the ATLS arm had the primary outcome,
+In the post training period, a total of `r n.m30d.atls` (
+`r p.m30d.atls`%) patients in the ATLS arm had the primary outcome,
 compared to `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm,
@@
 care arm. The absolute risk difference between the ATLS and PTC arm
-was X (95% CI X - X), between the ATLS and standard care arms X 
-(95% CI X - X), and between the PTC and standard care arms X 
-(95% CI X - X).
+was `r arr.atls.ptc`% units <!--(95% CI X - X)-->, between the ATLS and 
+standard care arms `r arr.atls.control`% units <!--(95% CI X - X)-->, and 
+between the PTC and standard care arms `r arr.ptc.control`% units 
+<!--(95% CI X -X)-->. The relative risk between the ATLS and 
+PTC arm was `r rr.atls.ptc` <!--(95% CI X - X)-->, between the ATLS and
+standard care arms `r rr.atls.control` <!--(95% CI X - X)-->, and 
+between the PTC and standard care arms `r rr.ptc.control` <!--(95% CI X - X)-->.
 
+<!--
 The change from baseline for the primary outcome was X in the ATLS
@@
 arms with 75, 85, and 95% CIs. 
+-->
 
 Differences in secondary outcomes between trial arms and subgroups are
-available as Supplementary material X.
+available as Supplementary material \@ref(tab:secondary-outcomes).
 
 ## Resident Participant Outcomes
-A total of X residents were trained during this pilot. Overall their
-confidence in managing trauma patients was X (IQR X - X) on a 10 point
-Likert scale with 10 being most confident. In the post training
-period, the confidence in the ATLS arm was X (IQR X - X), compared to
-X (IQR X - X) in the PTC arm, and X (IQR X - X) in the standard care
-arm.
+A total of `r n.residents` residents were trained during this
+pilot. Overall their median confidence in managing trauma patients was
+`r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert
+scale with 10 being most confident. <!--In the post training period, the
+confidence in the ATLS arm was X (IQR X - X), compared to X (IQR X -
+X) in the PTC arm, and X (IQR X - X) in the standard care arm.-->
 
 ## Feasibility Outcomes
-Out of X potentially eligible patients, X (X%) patients were enrolled
-in the study. Out of X potentially eligible residents, X (X%) were
-enrolled in the study. The loss to follow up rate was X, with X out of
-X patients not completing 30 day follow up. Among residents the pass
-rate, after two attempts, was X (X%). The missing data rate ranged
-from X to X, with details per variable and outcome in Table X. The
-differences in distributions between observed and extracted data, for
-variables that were collected through observation or interview, are
-shown in Table X.
+<!--Out of X potentially eligible patients, X (X%) patients were enrolled
+in the study-->. Out of `r n.eligible.residents` potentially eligible
+residents, `r n.residents` (`r recruitment.rate.residents`%) were
+enrolled in the study. The lost to follow up rate was 
+`r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of 
+`r n.patients` patients not completing 30 day follow up. Among residents
+the pass rate, after two attempts, was `r rate.pass`%. The missing
+data rate ranged from `r min.missing.data` to `r max.missing.data`,
+with details per variable and outcome in Table X. <!--The differences in
+distributions between observed and extracted data, for variables that
+were collected through observation or interview, are shown in Table X.-->
 
@@
 The primary patient participant outcome, all-cause 30-day mortality,
-was missing in X% of patients. This may appear high, especially
-compared to for example the CRASH-2 and REACT-2 trials, which report
-missing primary outcome in less than 0.01% of patients [@Roberts2013;
-@Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2
-used in-hospital mortality as their primary outcome measure, whereas
-we attempted to follow up patients after discharge. The missing data
-rate for in-hospital mortality was only X%, comparable to previous
-trials.
+was missing in `r rate.lost.to.follow.up`% of patients. This may
+appear high, especially compared to for example the CRASH-2 and
+REACT-2 trials, which report missing primary outcome in less than
+0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma
+trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
+primary outcome measure, whereas we attempted to follow up patients
+after discharge. Our missing data rate for in-hospital mortality was
+only `r p.missing.in.hospital.mortality`%, comparable to previous trials.
 
@@
 # Contributorship statement
+<!--
 MGW conceived of the study. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH,
@@
 prospective participating hospitals.
+-->
 
@@
 # Data Sharing Statement
-The final anonymized dataset and code for analysis will be released publicly.
+The final anonymized dataset and code for analysis are released publicly.
 
@@
 
+## Follow up of residents
+We stated that resident participants would be followed up 30 days after training, if
+they are in the intervention arms, or 30 days after the study started,
+if they are in the control arm, but the intervention period was three months.
+
 ## Data collection from records
@@
 
+## Selection of units for training
+We planned to use simple random sampling to select units if there were
+more than two eligible units in a hospital but instead the hospital
+principal investigator decided which units to train.
+
+## Timing of resident consent
+We had initially planned to ask residents for consent before
+randomisation, but the units were only finalised after the hospitals
+had been randomised, and residents were therefore approached for
+consent afterwards.
+
+\pagebreak
+# Supplementary material
+
+```{r secondary-outcomes, echo = FALSE, results = "asis"}
+table1::t1kable(secondary.outcomes.table, longtable = TRUE) %>%
+    kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)
+
+```
+
 # References

### 2022-12-02 pilot-publication/functions/create_descriptive_table.R

@@ -25,7 +25,7 @@ create_descriptive_table <- function(table.data,
                                      caption = NULL) {
     ## Check arguments
     assertthat::assert_that(is.data.frame(table.data))
-    assertthat::assert_that(is.character(variables))
+    assertthat::assert_that(is.character(variables) | is.null(variables))
     assertthat::assert_that(is.character(strata) | is.null(strata))
     assertthat::assert_that(is.logical(include.overall))
     assertthat::assert_that(is.logical(use.labels))
@@ -53,9 +53,10 @@ create_descriptive_table <- function(table.data,
         })
         labels <- list(variables = labels)
     }
-    descriptive.table <- table1::t1kable((table1::table1(strata.list,
-                                                         labels = labels,
-                                                         caption = caption,
-                                                         render.missing=table1::render.missing.default)))
+    descriptive.table <- table1::table1(strata.list,
+                                        labels = labels,
+                                        caption = caption,
+                                        droplevels = TRUE,
+                                        render.missing = table1::render.missing.default)
     return(descriptive.table)
 }

### 2022-12-02 pilot-publication/functions/prepare_data.R

@@ -24,9 +24,15 @@ prepare_data <- function(data, codebook = NULL) {
     ## Replace with missing
     prepared.data <- prepared.data %>%
         naniar::replace_with_na_if(.predicate = is.character,
-                                   condition = ~ .x == "999") %>%
+                                   condition = ~ .x %in% c("999", "unknown")) %>%
         naniar::replace_with_na_if(.predicate = is.numeric,
                                    condition = ~ .x == 999)
+    ## Deal with edge cases
+    variable <- prepared.data$complications__failure_of_conservative_management
+    variable[variable == "0" | variable == "NO" | variable == "no"] <- "No"
+    variable[variable == "Yes - Surgery on the 3rd day"] <- "Yes"
+    variable <- as.factor(variable)
+    prepared.data$complications__failure_of_conservative_management <- variable
     ## Label variables
     prepared.data[] <- lapply(names(prepared.data), function(column.name) {
         column.data <- prepared.data %>%
@@ -47,8 +53,11 @@ label_variable <- function(variable.data, name, codebook) {
         type.index <- grep(gsub("select_one ", "", type), codebook$choices$list_name)
         levels <- codebook$choices$name[type.index]
         labels <- codebook$choices$label[type.index]
-        if ()
         relabelled.data <- factor(variable.data, levels, labels)
+        if (any(name == names(binary_outcomes()))) {
+            logical.data <- relabelled.data == binary_outcomes()[[name]]$positive.class
+            relabelled.data <- factor(logical.data, levels = c(TRUE, FALSE), labels = c("Yes", "No"))
+        }
     }
     if (!is.null(type))
         labelled::var_label(relabelled.data) <- variable.label

### 2022-12-02 pilot-publication/main.R

@@ -39,8 +39,6 @@ pre.post.break.points <- list("11542" = "2022-04-24",
 pre.post.break.points <- lapply(pre.post.break.points, as.Date)
 n.patients <- nrow(data)
 n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre
-n.atls.residents.passed.first.attempt <- 2 + 2 # The number of residents who had passed ATLS after the first attempt, per centre
-n.atls.residents.passed.second.attempt <- 4 + 2 # The number of residents who had passed ATLS after the second attempt, per centre
 n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
 n.residents <- n.atls.residents + n.ptc.residents
 centre.ids <- data %>% pull(id__reg_hospital_id) %>% unique() 
@@ -53,33 +51,70 @@ atls.data <- bind_rows(centre.data[as.character(atls.centres)]) %>% labelled::co
 atls.data$arm <-"ATLS"
 ptc.data <- bind_rows(centre.data[as.character(ptc.centres)]) %>% labelled::copy_labels_from(data)
 ptc.data$arm <- "PTC"
-control.data <- bind_rows(centre.data[as.character(control.centres)])  %>% labelled::copy_labels_from(data)
-controldata$arm <- "Standard care"
+control.data <- bind_rows(centre.data[as.character(control.centres)]) %>% labelled::copy_labels_from(data)
+control.data$arm <- "Standard care"
 arms.data.list <- list(atls = atls.data, ptc = ptc.data, control = control.data)
+data <- bind_rows(arms.data.list) %>% labelled::copy_labels_from(data)
 n.atls <- nrow(atls.data)
 n.ptc <- nrow(ptc.data)
 n.control <- nrow(control.data)
 n.females <- with(data, sum(patinfo__pt_gender == "Female"))
 p.females <- round(n.females/nrow(data) * 100)
 median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
-iqr.age <- quantile(data$patinfo__pt_age, probs = c(0.25, 0.75), na.rm = TRUE) %>% paste0(collapse = "-")
+iqr.age <- quantile(data$patinfo__pt_age, probs = c(0.25, 0.75), na.rm = TRUE) %>%
+    paste0(collapse = "-")
 
 ## Patient participant outcomes
-n.outcomes <- lapply(data[, grep("outcomes__", names(data))])
[…]

### 2022-12-06 pilot-publication/manuscript.Rmd
Update analysis with 235 patients

@@
 neither ATLS, PTC, nor any other trauma life support training program
-is routinely taught. The seven hospitals were King Edward Memorial
-Seth G. S. Medical College and K.E.M. Hospital, Mumbai; Lokmanya Tilak
-Municipal Medical College and General Hospital, Mumbai; HBT Medical
-College And Dr. R N Cooper Municipal General Hospital, Mumbai; Medica
-Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata;
-Sir Nil Ratan Sircar Medical College & Hospital, Kolkata; Postgraduate
-Institute of Medical Education & Research, Chandigarh.
+is routinely taught. The seven hospitals were Seth G. S. Medical
+College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
+College and General Hospital, Mumbai; HBT Medical College And Dr. R N
+Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
+Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
+Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
+Medical Education & Research, Chandigarh.
 
@@
 ### Patient Participants
-Adults (15 years or older) who present to the emergency department at
+Adults (15 years or older) who presented to the emergency department at
 participating hospitals with a history of trauma. History of trauma is
@@
 ### Clusters
-Indian tertiary care hospitals that admit 400-800 adult patients with
-trauma each year. We randomised on the cluster (hospital) level to
-avoid contamination between intervention and control arms. To be
+Indian tertiary care hospitals that admit more than 400 adult patients
+with trauma each year. We randomised on the cluster (hospital) level
+to avoid contamination between intervention and control arms. To be
 eligible for inclusion hospitals had to provide the following services
@@
 ## Patient Participant Outcomes
-In the post training period, a total of `r n.m30d.atls` (
-`r p.m30d.atls`%) patients in the ATLS arm had the primary outcome,
-compared to `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm,
-and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard
-care arm. The absolute risk difference between the ATLS and PTC arm
-was `r arr.atls.ptc`% units <!--(95% CI X - X)-->, between the ATLS and 
-standard care arms `r arr.atls.control`% units <!--(95% CI X - X)-->, and 
-between the PTC and standard care arms `r arr.ptc.control`% units 
-<!--(95% CI X -X)-->. The relative risk between the ATLS and 
-PTC arm was `r rr.atls.ptc` <!--(95% CI X - X)-->, between the ATLS and
-standard care arms `r rr.atls.control` <!--(95% CI X - X)-->, and 
-between the PTC and standard care arms `r rr.ptc.control` <!--(95% CI X - X)-->.
+A total of `r n.m30d.atls` ( `r p.m30d.atls`%) patients in the ATLS
+arm had the primary outcome, compared to `r n.m30d.ptc` (
+`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (
+`r p.m30d.control`%) patients in the standard care arm. The absolute risk
+difference between the ATLS and PTC arm was `r arr.atls.ptc`% units
+<!--(95% CI X - X)-->, between the ATLS and standard care arms 
+`r arr.atls.control`% units <!--(95% CI X - X)-->, and between the PTC
+and standard care arms `r arr.ptc.control`% units <!--(95% CI X
+-X)-->. The relative risk between the ATLS and PTC arm was 
+`r rr.atls.ptc` <!--(95% CI X - X)-->, between the ATLS and standard care
+arms `r rr.atls.control` <!--(95% CI X - X)-->, and between the PTC
+and standard care arms `r rr.ptc.control` <!--(95% CI X - X)-->.
 
@@
 
-Differences in secondary outcomes between trial arms and subgroups are
-available as Supplementary material \@ref(tab:secondary-outcomes).
+Differences in secondary outcomes between trial arms<!-- and subgroups--> are
+available as Supplementary material table \@ref(tab:secondary-outcomes).
 
@@
 data rate ranged from `r min.missing.data` to `r max.missing.data`,
-with details per variable and outcome in Table X. <!--The differences in
+with details per variable and outcome in Table 
+\ref@(tab:sample-characteristics) and Supplementary material table
+\@ref(tab:secondary-outcomes). <!--The differences in
 distributions between observed and extracted data, for variables that

### 2023-01-16 pilot-publication/manuscript.Rmd
Add function to calculate iqr and add ISS and NISS to manuscript

@@
 `r n.females` (`r p.females`%), the median age was `r median.age`
-years (IQR `r iqr.age`)<!--, and the median ISS and NISS were X (IQR
-	X) and X (IQR X) respectively-->. A total of `r n.m30d` (`r p.m30d`%)
-patients had the primary outcome of mortality at 30 days after arrival
-to the emergency department.
+years (IQR `r iqr.age`), and the median ISS and NISS were `r median.iss`
+(IQR `r iqr.iss`) and `median.niss` (IQR `iqr.niss`) respectively. A 
+total of `r n.m30d` (`r p.m30d`%) patients had the primary outcome of 
+mortality at 30 days after arrival to the emergency department.
 

### 2023-01-16 pilot-publication/functions/prepare_data.R

@@ -74,6 +74,8 @@ prepare_data <- function(data, codebook = NULL) {
             label_variable(name = column.name, codebook = codebook)
         return (column.data)
     })
+    labelled::var_label(prepared.data$riss) <- "Injury Severity Score"
+    labelled::var_label(prepared.data$niss) <- "New Injury Severity Score"
     return (prepared.data)
 }
 

### 2023-01-16 pilot-publication/main.R

@@ -62,8 +62,11 @@ n.control <- nrow(control.data)
 n.females <- with(data, sum(patinfo__pt_gender == "Female"))
 p.females <- round(n.females/nrow(data) * 100)
 median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
-iqr.age <- quantile(data$patinfo__pt_age, probs = c(0.25, 0.75), na.rm = TRUE) %>%
-    paste0(collapse = "-")
+iqr.age <- get_iqr(data$patinfo__pt_age)
+median.iss <- median(data$riss, na.rm = TRUE)
+iqr.iss <- get_iqr(data$riss)
+median.niss <- median(data$niss, na.rm = TRUE)
+iqr.niss <- get_iqr(data$niss)
 
 ## Patient participant outcomes
 
@@ -72,7 +75,7 @@ table.variables <- c("patinfo__pt_age", "patinfo__pt_gender",
                      "incident__dominating_injury_type",
                      "patvitals__ed_rr", "patvitals__ed_sat",
                      "patvitals__ed_hr", "patvitals__ed_sbp",
-                     "outcomes__alive_after_30_days",
+                     "riss", "niss", "outcomes__alive_after_30_days",
                      "arm")
 table.data <- data[, table.variables]
 sample.characteristics.table <- create_descriptive_table(table.data,

### 2023-01-23 pilot-publication/functions/estimate_icc.R
Draft presentation for meeting on 2023-02-01 and estimate composite outcome

@@ -4,12 +4,14 @@
 #' @param outcome Character. The name of the outcome variable. No default.
 #' @param cluster.variable Character. The name of the variable defining clusters. No default.
 #' @param data A data.frame. No default.
+#' @param digits Numeric. The number of significant digits. Defaults to 3.
 #' @export
-estimate_icc <- function(outcome, cluster.variable, data) {
+estimate_icc <- function(outcome, cluster.variable, data, digits = 3) {
     ## Check arguments
     for (argument in c(outcome, cluster.variable))
          assertthat::assert_that(is.character(argument) & length(argument) == 1)
     assertthat::assert_that(is.data.frame(data))
+    assertthat::assert_that(is.numeric(digits))
 
     ## Build null model
     arguments <- list(as.formula(paste0(outcome, " ~ ", "( 1 | ", cluster.variable, ")")),
@@ -22,6 +24,6 @@ estimate_icc <- function(outcome, cluster.variable, data) {
     ## logistic distribution, that is, the assumed level-1 variance
     ## component: We take this assumed value, as the logistic
     ## regression model does not include level-1 residual"
-    icc <- M0@theta[1]^2/ (M0@theta[1]^2 + (3.14159^2/3)) 
+    icc <- round(M0@theta[1]^2/ (M0@theta[1]^2 + (3.14159^2/3)), digits = digits)
     return (icc)
 }

### 2023-01-23 pilot-publication/main.R

@@ -70,8 +70,6 @@ iqr.niss <- get_iqr(data$niss)
 n.admitted <- with(data, sum(interventions__admitted == "Yes"))
 p.admitted <- round(n.admitted/nrow(data) * 100)
 
-## Patient participant outcomes
-
 ## Create table of sample characteristics
 table.variables <- c("patinfo__pt_age", "patinfo__pt_gender",
                      "incident__dominating_injury_type",
@@ -114,6 +112,34 @@ rr.atls.ptc <- round(p.m30d.atls/p.m30d.ptc, 2)
 rr.atls.control <- round(p.m30d.atls/p.m30d.control, 2)
 rr.ptc.control <- round(p.m30d.ptc/p.m30d.control, 2)
 p.missing.in.hospital.mortality <- round(sum(is.na(data$outcomes__discharge_alive))/n.patients * 100)
+
+## Estimate composite outcome
+in.hospital.mortality <- data$outcomes__discharge_alive == "Yes"
+data$in.hospital.mortality <- in.hospital.mortality
+icc.in.hospital.mortality <- estimate_icc("in.hospital.mortality", "id__reg_hospital_id", data)
+labelled::var_label(data$in.hospital.mortality) <- "In-hospital mortality"
+confined.to.bed <- data$outcomes__eq5dm == "I am confined to bed"
+data$confined.to.bed <- confined.to.bed
+labelled::var_label(data$confined.to.bed) <- "Confined to bed"
+extreme.pain.discomfort <- data$outcomes__eq5dpd == "I have extreme pain or discomfort"
+data$extreme.pain.discomfort <- extreme.pain.discomfort
+labelled::var_label(data$extreme.pain.discomfort) <- "Extreme pain or discomfort"
+unable.bath.dress <- data$outcomes__eq5dsc == "I am unable to bathe or dress myself"
+data$unable.bath.dress <- unable.bath.dress
+labelled::var_label(data$unable.bath.dress) <- "Unable to bathe or dress oneself"
+unable.usual.activities <- data$outcomes__eq5dua == "I am unable to perform my usual activities"
+data$unable.usual.activities <- unable.usual.activities
+labelled::var_label(data$unable.usual.activities) <- "Unable to perform usual activities"
+composite.outcome <- in.hospital.mortality |
+    confined.to.bed |
+    extreme.pain.discomfort |
+    unable.bath.dress |
+    unable.usual.activities
+data$composite.outcome <- composite.outcome
[…]

### 2023-01-31 pilot-publication/manuscript.Rmd
Draft presentation for meeting on 2023-02-01 and estimate composite outcome
Update Global Burden of Disease citation

@@
 the body's associated response, causes 4.5 millions deaths every year
-[@Naghavi2017]. Almost 10% of the global burden of disease is due
+[@GBD2018]. Almost 10% of the global burden of disease is due
 to trauma and trauma is the top contributor to the burden of disease in
@@
 years (IQR `r iqr.age`), and the median ISS and NISS were `r median.iss`
-(IQR `r iqr.iss`) and `median.niss` (IQR `iqr.niss`) respectively. A 
+(IQR `r iqr.iss`) and `median.niss` (IQR `iqr.niss`) respectively. The 
+number of admitted patients was `r n.admitted` (`r p.admitted`%). A 
 total of `r n.m30d` (`r p.m30d`%) patients had the primary outcome of 

### 2023-06-23 pilot-publication/lab-book.Rmd
Start working on generating synthethic pilot data

# Lab book for the TERN pilot publication

This file includes code and scripts used to experiment and explore with the TERN pilot data.

## Generate synthetic data

Because developing on the server is generally slow, a first step to speed up development was to generate a synthetic dataset that could be safely used locally.

```{r, synthetic-data-generation}
install.packages("synthpop")
codebook.syn(data)

```

### 2023-06-23 pilot-publication/main.R

@@ -52,7 +52,10 @@ data.with.post.indicator <- do.call(rbind, lapply(split(data, data$id__reg_hospi
     centre.data$post.training <- centre.data$incident__date_of_arrival > pre.post.break.points[[centre.id]][1]
     return(centre.data)
 })) %>% labelled::copy_labels_from(data)
+
+## This is the data that should be used as the basis for the synthetic data
 data <- data.with.post.indicator
+
 icc <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
 n.patients <- nrow(data)
 n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre

### 2023-06-27 pilot-publication/lab-book.Rmd
Work on creating synthetic data for local development

@@
 Because developing on the server is generally slow, a first step to speed up development was to generate a synthetic dataset that could be safely used locally.
+Check [main.R](./main.R) to make sure that `data` is in the correct state before creating the synthetic data using the code below.
 
 ```{r, synthetic-data-generation}
-install.packages("synthpop")
-codebook.syn(data)
+## Generate synthetic data. For each column, sample from the same distribution as the original data, but do not keep correlations present in the orginal data.
+library(pbapply)
+synth.data <- lapply(data, function(column) {
+    if (is.factor(column)) {
+        return (as.factor(sample(levels(column), size = length(column), replace = TRUE)))
+    } else if (is.character(column)) {
+        return (as.character(sample(column, size = length(column), replace = TRUE)))
+    } else if (is.numeric(column)) {
+        return (rnorm(n = length(column), mean = mean(column, na.rm = TRUE), sd = sd(column, na.rm = TRUE)))
+    } else if (is.logical(column)) {
+        return (sample(c(TRUE, FALSE), size = length(column), replace = TRUE))
+    } else if (is.POSIXct(column) | is.POSIXt(column)) {
+        return (as.POSIXct(sample(as.numeric(column), size = length(column), replace = TRUE), origin = "1970-01-01"))
+    } else {
+        return (rep(NA, length(column)))
+    }
+}) %>% as.data.frame()
+
+## Replace values in columns with character values with random strings
+synth.data[, sapply(synth.data, is.character)] <- lapply(synth.data[, sapply(synth.data, is.character)], function(column) {
+    n.unique <- length(unique(column))
+    new.values <- replicate(n.unique, paste(sample(letters, size = 10, replace = TRUE), collapse = ""))
+    new.values <- rep(new.values, length.out = length(column))
+    return (new.values)
+})
+
+## Replace values in any variable with id in the name random numbers
+id.vars <- c("X_id", "id__reg_hospital_id", "id__assigned_patient_id", "meta__instanceID", "meta__deprecatedID", "formhub__uuid", "X_xform_id_string", "X_uuid")
+synth.data[, id.vars] <- lapply(synth.data[, id.vars], function(column) {
+    n.unique <- length(unique(column))
+    new.values <- rep(sample(1:1e6, size = n.unique, replace = FALSE), length.out = length(column))
+    return (new.values)
+})
+
+## Check that no rows in the original data are present in the synthetic data
+stopifnot(!any(pbapply(data, 1, function(row) {
+    any(apply(synth.data, 1, function(synth.row) {
+        all(row == synth.row, na.rm = TRUE)
+    }))
+})))
+
+## Write synthetic data to file
+saveRDS(synth.data, file = "data/synth-data.rds")
+
+## Write codebook to file
+saveRDS(codebook, file = "data/codebook.rds")
 

### 2023-06-27 pilot-publication/functions/prepare_data.R

@@ -5,10 +5,14 @@
 #' @param data A data.frame. The data to be prepared. No default.
 #' @param codebook A list. A list describing the columns of data or
 #'     NULL. Defaults to NULL.
-prepare_data <- function(data, codebook = NULL) {
+#' @param pre.post.break.points A list. A list of pre and post break points. Defauls
+#'    to NULL.
+prepare_data <- function(data, codebook = NULL, pre.post.break.points = NULL) {
     ## Check arguments
     assertthat::assert_that(is.data.frame(data))
     assertthat::assert_that(is.list(codebook) | is.null(codebook))
+    assertthat::assert_that(is.list(pre.post.break.points) | is.null(pre.post.break.points))
+
     ## Modify codebook
     variable.names <- codebook$survey$name
     codebook$survey$type <- setNames(codebook$survey$type, nm = variable.names)
@@ -16,11 +20,14 @@ prepare_data <- function(data, codebook = NULL) {
     for (binary.outcome in binary_outcomes()) {
         codebook$survey$label[unlist(strsplit(binary.outcome$name, "__"))[2]] <- binary.outcome$label
     }
+
     ## Rename variables
     names(data) <- gsub("/", "__", names(data), fixed = TRUE)
+
     ## Prepare data
     prepared.data <- data
     prepared.data$`patinfo/pt_age` <- as.numeric(prepared.data$patinfo__pt_age)
+
     ## Replace with missing
     prepared.data <- prepared.data %>%
         naniar::replace_with_na_if(.predicate = is.character,
@@ -33,6 +40,7 @@ prepare_data <- function(data, codebook = NULL) {
     variable[variable == "Yes - Surgery on the 3rd day"] <- "Yes"
     variable <- as.factor(variable)
     prepared.data$complications__failure_of_conservative_management <- variable
+
     ## Generate AIS codes
     icd10.data <- prepared.data[, c("interventions__injury_first_surg_icd10",
                                     "interventions__injury_xray_icd10",
@@ -67,6 +75,21 @@ prepare_data <- function(data, codebook = NULL) {
                                                                         "riss",
                                                                         "niss")]
     prepared.data <- cbind(prepared.data, iss.data)
+
+    ## Classify mechanism of injury (incident__moi and incident__moi_001) as factors
+    prepared.data$incident__moi <- as.factor(prepared.data$incident__moi)
+    prepared.data$incident__moi_001 <- as.factor(prepared.data$incident__moi_001)
+
[…]

### 2023-06-27 pilot-publication/main.R

@@ -4,6 +4,8 @@ library(dplyr)
 library(tibble)
 library(lubridate)
 library(gtsummary)
+library(naniar)
+library(icdpicr)
 
 ## Source functions
 noacsr::source_all_functions()
@@ -16,8 +18,21 @@ codebook.arguments <- lapply(c("URL", "UID", "USERNAME", "PASSWORD"),
                              function(x) Sys.getenv(paste0("KOBO_", x)))
 codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
 
+## The pre-post breakt points are the dates to use when comparing
+## before training to after training. For intervention centres these
+## are the dates when the training happened. For standard care centres
+## these are one month after data collection started.  
+pre.post.break.points <- list("11542" = ymd(centre.start.dates["11542"]) + months(1), # standard care
+                              "44805" = c("2022-05-30", "2022-06-20"), # atls, two dates because two students did not pass the first time
+                              "55356" = "2022-09-02", # ptc
+                              "78344" = "2022-06-03", # atls
+                              "95846" = "2022-09-01", # ptc
+                              "88456" = ymd(centre.start.dates["88456"]) + months(1), # standard care 
+                              "10263" = ymd(centre.start.dates["10263"]) + months(1)) # standard care 
+pre.post.break.points <- lapply(pre.post.break.points, ymd)
+
 ## Prepare data
-data <- prepare_data(data, codebook)
+data <- prepare_data(data, codebook, pre.post.break.points)
 
 ## Define basic results
 arrival.dates <- data %>% select(incident__date_of_arrival, id__reg_hospital_id) %>% arrange(incident__date_of_arrival)
@@ -35,27 +50,6 @@ n.no.consent <-  list("11542" = 40,
                       "95846" = 9, 
                       "88456" = 0, # To be updated
                       "10263" = 2)
-## The pre-post breakt points are the dates to use when comparing
-## before training to after training. For intervention centres these
-## are the dates when the training happened. For standard care centres
-## these are one month after data collection started.  
-pre.post.break.points <- list("11542" = ymd(centre.start.dates["11542"]) + months(1), # standard care
-                              "44805" = c("2022-05-30", "2022-06-20"), # atls
-                              "55356" = "2022-09-02", # ptc
-                              "78344" = "2022-06-03", # atls
-                              "95846" = "2022-09-01", # ptc
[…]

### 2023-06-27 pilot-publication/lab-book.Rmd
Start adding encryption

@@
 
+## Encrypt synthetic data before writing to file
+encrypted.synth.data <-
+
 ## Write synthetic data to file
@@
 
-## Write codebook to file
+## Write codebook to file, but first remove the settings entry
+codebook$settings <- NULL
 saveRDS(codebook, file = "data/codebook.rds")
+```
+
+To develop locally both the data and the codebook need to be loaded.
+
+```{r load-data}
+data <- readRDS("data/synth-data.rds")
+codebook <- readRDS("data/codebook.rds")
 

### 2023-06-27 pilot-publication/functions/prepare_data.R

@@ -5,13 +5,10 @@
 #' @param data A data.frame. The data to be prepared. No default.
 #' @param codebook A list. A list describing the columns of data or
 #'     NULL. Defaults to NULL.
-#' @param pre.post.break.points A list. A list of pre and post break points. Defauls
-#'    to NULL.
-prepare_data <- function(data, codebook = NULL, pre.post.break.points = NULL) {
+prepare_data <- function(data, codebook = NULL) {
     ## Check arguments
     assertthat::assert_that(is.data.frame(data))
     assertthat::assert_that(is.list(codebook) | is.null(codebook))
-    assertthat::assert_that(is.list(pre.post.break.points) | is.null(pre.post.break.points))
 
     ## Modify codebook
     variable.names <- codebook$survey$name
@@ -80,7 +77,27 @@ prepare_data <- function(data, codebook = NULL, pre.post.break.points = NULL) {
     prepared.data$incident__moi <- as.factor(prepared.data$incident__moi)
     prepared.data$incident__moi_001 <- as.factor(prepared.data$incident__moi_001)
 
-    ## Apply pre-post break points
+    ## Define the post intervention period
+
+    ## The pre-post breakt points are the dates to use when comparing
+    ## before training to after training. For intervention centres these
+    ## are the dates when the training happened. For standard care centres
+    ## these are one month after data collection started.
+    arrival.dates <- data %>% select(incident__date_of_arrival, id__reg_hospital_id) %>% arrange(incident__date_of_arrival)
+    format_date <- function(date) paste0(month(date[1], label = TRUE, abbr = FALSE), " ", year(date[1]))
+    centre.start.dates <- arrival.dates %>%
+        group_by(id__reg_hospital_id) %>%
+        summarise(start_date = format(min(incident__date_of_arrival), "%Y-%m-%d")) %>%
+        deframe()
+    pre.post.break.points <- list("11542" = ymd(centre.start.dates["11542"]) + months(1), # standard care
+                                  "44805" = c("2022-05-30", "2022-06-20"), # atls, two dates because two students did not pass the first time
+                                  "55356" = "2022-09-02", # ptc
+                                  "78344" = "2022-06-03", # atls
+                                  "95846" = "2022-09-01", # ptc
+                                  "88456" = ymd(centre.start.dates["88456"]) + months(1), # standard care 
+                                  "10263" = ymd(centre.start.dates["10263"]) + months(1)) # standard care
[…]

### 2023-06-27 pilot-publication/main.R

@@ -18,29 +18,10 @@ codebook.arguments <- lapply(c("URL", "UID", "USERNAME", "PASSWORD"),
                              function(x) Sys.getenv(paste0("KOBO_", x)))
 codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
 
-## The pre-post breakt points are the dates to use when comparing
-## before training to after training. For intervention centres these
-## are the dates when the training happened. For standard care centres
-## these are one month after data collection started.  
-pre.post.break.points <- list("11542" = ymd(centre.start.dates["11542"]) + months(1), # standard care
-                              "44805" = c("2022-05-30", "2022-06-20"), # atls, two dates because two students did not pass the first time
-                              "55356" = "2022-09-02", # ptc
-                              "78344" = "2022-06-03", # atls
-                              "95846" = "2022-09-01", # ptc
-                              "88456" = ymd(centre.start.dates["88456"]) + months(1), # standard care 
-                              "10263" = ymd(centre.start.dates["10263"]) + months(1)) # standard care 
-pre.post.break.points <- lapply(pre.post.break.points, ymd)
-
 ## Prepare data
-data <- prepare_data(data, codebook, pre.post.break.points)
+data <- prepare_data(data, codebook)
 
 ## Define basic results
-arrival.dates <- data %>% select(incident__date_of_arrival, id__reg_hospital_id) %>% arrange(incident__date_of_arrival)
-format_date <- function(date) paste0(month(date[1], label = TRUE, abbr = FALSE), " ", year(date[1]))
-centre.start.dates <- arrival.dates %>%
-    group_by(id__reg_hospital_id) %>%
-    summarise(start_date = format(min(incident__date_of_arrival), "%Y-%m-%d")) %>%
-    deframe()
 start.date <- arrival.dates %>% pull(incident__date_of_arrival) %>% min() %>% format_date()
 end.date <- arrival.dates %>% pull(incident__date_of_arrival) %>% max() %>% format_date()
 n.no.consent <-  list("11542" = 40,

### 2023-06-28 pilot-publication/lab-book.Rmd
Add encrypted synthetic data

--- 2023-06-27 pilot-publication/lab-book.Rmd
@@
 This file includes code and scripts used to experiment and explore with the TERN pilot data.
+
+```{r setup}
+library(pbapply)
+library(openssl)
+library(dotenv)
+```
 
@@
 ## Generate synthetic data. For each column, sample from the same distribution as the original data, but do not keep correlations present in the orginal data.
-library(pbapply)
 synth.data <- lapply(data, function(column) {
@@
 
+## Encrypt synthetic data before writing to file
+key <- sha256(charToRaw(Sys.getenv("SYNTH_DATA_KEY")))
+encrypted.synth.data <- aes_cbc_encrypt(serialize(synth.data, NULL), key = key)
+
 ## Write synthetic data to file
-saveRDS(synth.data, file = "data/synth-data.rds")
+saveRDS(encrypted.synth.data, file = "data/synth-data.rds")
 
-## Write codebook to file
-saveRDS(codebook, file = "data/codebook.rds")
+## Encrypt and write codebook to file, but first remove the settings entry
+codebook$settings <- NULL
+encrypted.codebook <- aes_cbc_encrypt(serialize(codebook, NULL), key = key)
+saveRDS(encrypted.codebook, file = "data/codebook.rds")
+```
 
+To develop locally both the data and the codebook need to be loaded.
+
+```{r load-data}
+key <- sha256(charToRaw(Sys.getenv("SYNTH_DATA_KEY")))
+data <- unserialize(aes_cbc_decrypt(readRDS("data/synth-data.rds"), key = key))
+codebook <- unserialize(aes_cbc_decrypt(readRDS("data/codebook.rds"), key = key))
 ```

### 2023-07-09 pilot-publication/lab-book.Rmd
Write function to define basic results in the lab book

@@
 ```
+
+Probably a good idea to wrap the generation of these results into a function, and introduce a test argument, to prevent us from introducing local results in the main manuscript.
+
+```{r basic-results}
+get_basic_results <- function(test = FALSE) {
+    results <- list()
+    ## Define basic results
+    arrival.dates <- data %>% pull(incident__date_of_arrival) %>% as.Date()
+    results$start.date <- arrival.dates %>% min() %>% format_date()
+    results$end.date <- arrival.dates %>% max() %>% format_date()
+    n.no.consent <-  list("11542" = 40,
+                          "44805" = 10,
+                          "55356" = 43,
+                          "78344" = 3,
+                          "95846" = 9, 
+                          "88456" = 0, # To be updated
+                          "10263" = 2)
+    results$icc <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
+    results$n.patients <- nrow(data)
+    results$n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre
+    results$n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
+    results$n.residents <- with(results, n.atls.residents + n.ptc.residents)
+    results$n.centres <-  data %>% pull(id__reg_hospital_id) %>% unique() %>% length()
+    results$n.atls <- sum(data$arm == "ATLS")
+    results$n.ptc <- sum(data$arm == "PTC")
+    results$n.control <- sum(data$arm == "Standard care")
+    results$n.females <- with(data, sum(patinfo__pt_gender == "Female"))
+    results$p.females <- round(results$n.females/nrow(data) * 100)
+    results$median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
+    results$iqr.age <- get_iqr(data$patinfo__pt_age)
+    results$median.iss <- median(data$riss, na.rm = TRUE)
+    results$iqr.iss <- get_iqr(data$riss)
+    results$median.niss <- median(data$niss, na.rm = TRUE)
+    results$iqr.niss <- get_iqr(data$niss)
+    results$n.admitted <- with(data, sum(interventions__admitted == "Yes"))
+    results$p.admitted <- round(results$n.admitted/nrow(data) * 100)
+    return (results)
+}
+```

### 2023-07-09 pilot-publication/functions/get_basic_results.R
Move basic results to own function

@@ -1,0 +1,41 @@
+#' Get basic results
+#'
+#' Function to calculate basic results.
+#' @export
+get_basic_results <- function() {
+    results <- list()
+    arrival.dates <- data %>% pull(incident__date_of_arrival) %>% as.Date()
+    results$start.date <- arrival.dates %>% min() %>% format_date()
+    results$end.date <- arrival.dates %>% max() %>% format_date()
+
+    ## This vector of n.no.consent is not used here yet, but is
+    ## probably used somewhere else. Keeping it for now.
+    n.no.consent <-  list("11542" = 40,
+                          "44805" = 10,
+                          "55356" = 43,
+                          "78344" = 3,
+                          "95846" = 9, 
+                          "88456" = 0, # To be updated
+                          "10263" = 2)
+    
+    results$icc <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
+    results$n.patients <- nrow(data)
+    results$n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre
+    results$n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
+    results$n.residents <- with(results, n.atls.residents + n.ptc.residents)
+    results$n.centres <-  data %>% pull(id__reg_hospital_id) %>% unique() %>% length()
+    results$n.atls <- sum(data$arm == "ATLS")
+    results$n.ptc <- sum(data$arm == "PTC")
+    results$n.control <- sum(data$arm == "Standard care")
+    results$n.females <- with(data, sum(patinfo__pt_gender == "Female"))
+    results$p.females <- round(results$n.females/nrow(data) * 100)
+    results$median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
+    results$iqr.age <- get_iqr(data$patinfo__pt_age)
+    results$median.iss <- median(data$riss, na.rm = TRUE)
+    results$iqr.iss <- get_iqr(data$riss)
+    results$median.niss <- median(data$niss, na.rm = TRUE)
+    results$iqr.niss <- get_iqr(data$niss)
+    results$n.admitted <- with(data, sum(interventions__admitted == "Yes"))
+    results$p.admitted <- round(results$n.admitted/nrow(data) * 100)
+    return (results)
+}

### 2023-07-09 pilot-publication/main.R
Work to create and compare outcome tables

@@ -37,23 +37,136 @@ sample.characteristics.table <- create_descriptive_table(table.data,
                                                          caption = "Patient sample characteristics",
                                                          include.overall = TRUE)
 
-## Create table comparing outcomes
+## Create tables comparing outcomes
 outcomes <- c(names(binary_outcomes()),
               categorical_outcomes(),
               quantitative_outcomes())
 outcomes.data <- data[, c(outcomes, "arm", "post.training")]
+outcomes.data.list <- list(
+    overall = outcomes.data %>% select(-post.training),
+    pre.training = outcomes.data %>% filter(!post.training) %>% select(-post.training),
+    post.training = outcomes.data %>% filter(post.training) %>% select(-post.training))
+outcomes.tables <- lapply(outcomes.data.list, create_outcomes_table)
+outcomes.row.names <- outcomes.tables$overall[, 1]
 
-## This creates a table with summary data for all outcomes
-outcomes.table.overall <- outcomes.data %>%
-    select(-post.training) %>%
-    tbl_summary(by = "arm",
-                type = all_dichotomous() ~ "categorical",
-                statistic = list(
-                    all_continuous() ~ "{median}",
-                    all_categorical() ~ "{p}"
-                ),
-                missing_text = "Missing") %>%
-    as_tibble() 
+create_outcomes_table <- function(data) {
+    outcomes.table <- data %>%
+        tbl_summary(by = "arm",
+                    type = all_dichotomous() ~ "categorical",
+                    statistic = list(
+                        all_continuous() ~ "{median}",
+                        all_categorical() ~ "{p}"
+                    ),
+                    missing_text = "Missing") %>%
+        as_tibble()
+    return (outcomes.table)
+}
+
+## Convert all outcome data in tables to numeric
+outcomes.tables <- lapply(outcomes.tables, convert_table_data_to_numeric)
+
+convert_table_data_to_numeric <- function(table.object) {
+    ## This function converts the three last columns of the table to numeric
+    ## and returns the table
+    table.object[, 2:4] <- lapply(table.object[, 2:4], function(column)
+        as.numeric(gsub(",", "", column)))
+    return (table.object)
+}
+
[…]

### 2023-08-06 pilot-publication/lab-book.Rmd
Add labels to synthetic data
Work to create and compare outcome tables
Update synthetic data

@@
 library(dotenv)
+library(DataExplorer)
 ```
@@
 })))
+
+## Put labels back
+synth.data <- synth.data %>% labelled::copy_labels_from(data)
+
+## Create data report
+DataExplorer::create_report(synth.data, report_title = "Synthetic data report", output_file = "synth-data-report.html")
 

### 2023-08-07 pilot-publication/lab-book.Rmd
Separate functions to generate results

@@
 ```
+
+Cutting code from the main script and putting it here for future reference:
+
+```{r}
+## Calculate outcome results
+n.m30d <- with(data, sum(outcomes__alive_after_30_days == "Yes", na.rm = TRUE))
+p.m30d <- round(n.m30d/nrow(data) * 100)
+arms.n.m30d.list <- lapply(setNames(arms.data.list, nm = paste0("m30d.", names(arms.data.list))),
+                           function(arms.data)
+                               with(arms.data,
+                                    sum(outcomes__alive_after_30_days == "Yes", na.rm = TRUE)))
+arms.p.m30d.list <- mapply(arms.n.m30d.list, arms.data.list,
+                           FUN = function(count, arms.data) list(round(count/nrow(arms.data) * 100)))
+names(arms.n.m30d.list) <- paste0("n.", names(arms.n.m30d.list))
+names(arms.p.m30d.list) <- paste0("p.", names(arms.p.m30d.list))
+attach(arms.n.m30d.list)
+attach(arms.p.m30d.list)
+arr.atls.ptc <- p.m30d.atls - p.m30d.ptc
+arr.atls.control <- p.m30d.atls - p.m30d.control
+arr.ptc.control <- p.m30d.ptc - p.m30d.control
+rr.atls.ptc <- round(p.m30d.atls/p.m30d.ptc, 2)
+rr.atls.control <- round(p.m30d.atls/p.m30d.control, 2)
+rr.ptc.control <- round(p.m30d.ptc/p.m30d.control, 2)
+p.missing.in.hospital.mortality <- round(sum(is.na(data$outcomes__discharge_alive))/n.patients * 100)
+
+## Estimate composite outcome
+in.hospital.mortality <- data$outcomes__discharge_alive == "Yes"
+data$in.hospital.mortality <- in.hospital.mortality
+icc.in.hospital.mortality <- estimate_icc("in.hospital.mortality", "id__reg_hospital_id", data)
+labelled::var_label(data$in.hospital.mortality) <- "In-hospital mortality"
+confined.to.bed <- data$outcomes__eq5dm == "I am confined to bed"
+data$confined.to.bed <- confined.to.bed
+labelled::var_label(data$confined.to.bed) <- "Confined to bed"
+extreme.pain.discomfort <- data$outcomes__eq5dpd == "I have extreme pain or discomfort"
+data$extreme.pain.discomfort <- extreme.pain.discomfort
+labelled::var_label(data$extreme.pain.discomfort) <- "Extreme pain or discomfort"
+unable.bath.dress <- data$outcomes__eq5dsc == "I am unable to bathe or dress myself"
+data$unable.bath.dress <- unable.bath.dress
+labelled::var_label(data$unable.bath.dress) <- "Unable to bathe or dress oneself"
+unable.usual.activities <- data$outcomes__eq5dua == "I am unable to perform my usual activities"
+data$unable.usual.activities <- unable.usual.activities
+labelled::var_label(data$unable.usual.activities) <- "Unable to perform usual activities"
+composite.outcome <- in.hospital.mortality |
+    confined.to.bed |
+    extreme.pain.discomfort |
+    unable.bath.dress |
+    unable.usual.activities
+data$composite.outcome <- composite.outcome
+labelled::var_label(data$composite.outcome) <- "Composite endpoint"
+n.composite.outcome <- sum(composite.outcome, na.rm = TRUE)
+p.composite.outcome <- round(n.composite.outcome/nrow(data) * 100)
+icc.composite.outcome <- estimate_icc("composite.outcome", "id__reg_hospital_id", data)
+
+## Resident outcomes
+median.confidence <- median(as.numeric(data$resident__res_comfort), na.rm = TRUE)
+iqr.confidence <- quantile(as.numeric(data$resident__res_comfort), probs = c(0.25, 0.75), na.rm = TRUE) %>%
+    paste0(collapse = "-")
+
+## Feasibility outcomes
+
+## List of potentially eligible patient participants per site, replace
+## IDs with actual IDs once known and replace figures with actual
+## figures once known
+n.potentially.eligible = list(n.potentially.eligible.centre.1 = 999,
+                              n.potentially.eligible.centre.2 = 999,
+                              n.potentially.eligible.centre.3 = 999,
+                              n.potentially.eligible.centre.4 = 999,
+                              n.potentially.eligible.centre.5 = 999,
+                              n.potentially.eligible.centre.6 = 999,
+                              n.potentially.eligible.centre.7 = 999)
+recruitment.rate.patients <- round(n.patients/do.call(sum, n.potentially.eligible) * 100)
+n.eligible.residents <- 21
+recruitment.rate.residents <- round(n.residents/n.eligible.residents * 100)
+n.lost.to.follow.up <- sum(is.na(data$outcomes__alive_after_30_days))
+rate.lost.to.follow.up <- round(n.lost.to.follow.up/n.patients * 100)
+n.atls.residents.passed.first.attempt <- 2 + 2 # The number of residents who had passed ATLS after the first attempt, per centre
+n.atls.residents.passed.second.attempt <- 4 + 2 # The number of residents who had passed ATLS after the second attempt, per centre
+rate.pass <- round((n.atls.residents.passed.second.attempt + n.ptc.residents)/n.residents * 100)
+rates.missing.data <- unlist(lapply(data, function(column) round(sum(is.na(column)/length(column) * 100))))
+min.missing.data <- min(rates.missing.data)
+max.missing.data <- max(rates.missing.data)
+
+## Calculate varying cluster sizes
+cluster.sizes <- sapply(centre.ids, function(centre.id) nrow(data[data$id__reg_hospital_id == centre.id, ]))
+sd.cluster.size <- sd(cluster.sizes)
+mean.cluster.size <- mean(cluster.sizes)
+ratio.sd.mean.cluster.size <- sd.cluster.size/mean.cluster.size
+```

### 2023-08-07 pilot-publication/functions/compare_outcomes_between_arms.R

@@ -1,0 +1,31 @@
+#' Compare outcomes between two arms in a clinical trial
+#'
+#' This function takes the combination of two arms in a clinical trial, the name of the outcomes table, and a list of outcomes tables as input.
+#' It calculates the absolute and relative differences between the outcomes of the two arms and returns the comparison as a tibble.
+#'
+#' @param arm.combination A vector of length 2, specifying the names of the two arms to compare.
+#' @param table.name A character string specifying the name of the outcomes table to be used for comparison.
+#' @param outcomes.tables A list containing the outcomes tables, each stored as a data frame.
+#'
+#' @return A tibble containing the absolute and relative differences between the outcomes of the two arms.
+#'
+#' @importFrom dplyr select add_column as_tibble
+#' @importFrom assertthat assert_that is.list is.data.frame
+#' @export
+compare_outcomes_between_arms <- function(arm.combination, table.name, outcomes.tables) {
+    assertthat::assert_that(is.list(outcomes.tables))
+    assertthat::assert_that(is.data.frame(outcomes.tables[[table.name]]))
+    outcomes.table <- outcomes.tables[[table.name]] %>%
+        select(contains(arm.combination[1]), contains(arm.combination[2]))
+    ## Calculate the difference between the two arms
+    outcomes.comparison <- data.frame(
+        absolute.difference = outcomes.table[, 1] - outcomes.table[, 2],
+        relative.difference = outcomes.table[, 1]/outcomes.table[, 2]
+    )
+    colnames(outcomes.comparison) <- c("Absolute difference", "Relative difference")
+    ## Convert outcomes comparison to tibble
+    outcomes.comparison <- as_tibble(outcomes.comparison) %>%
+        add_column(outcomes.row.names, .before = 1)
+    ## Return the table
+    return (outcomes.comparison)
+}

### 2023-08-07 pilot-publication/functions/convert_table_data_to_numeric.R

@@ -1,0 +1,19 @@
+#' Convert table data to numeric
+#'
+#' This function takes a table object as input and converts the data in the last three columns of the table to numeric format.
+#' It removes any commas from the data before converting it to numeric and returns the modified table.
+#'
+#' @param table.object The table object to be converted to numeric.
+#'
+#' @return The modified table object with the last three columns converted to numeric format.
+
+#' @importFrom base gsub as.numeric
+#'
+#' @export
+convert_table_data_to_numeric <- function(table.object) {
+    ## This function converts the three last columns of the table to numeric
+    ## and returns the table
+    table.object[, 2:4] <- lapply(table.object[, 2:4], function(column)
+        as.numeric(gsub(",", "", column)))
+    return (table.object)
+}

### 2023-08-07 pilot-publication/functions/convert_tibble_to_named_vector.R

@@ -1,0 +1,37 @@
+#' Convert a tibble to a named vector
+#'
+#' This function takes a tibble as input and converts it to a named vector by concatenating the row and column names.
+#'
+#' @param tibble A tibble to be converted to a named vector.
+#'
+#' @return A named vector with concatenated row and column names.
+#'
+#' @examples
+#' tib <- tibble::tribble(
+#'     ~name, ~age, ~city,
+#'     "Alice", 28, "New York",
+#'     "Bob", 32, "Los Angeles"
+#' )
+#' named_vector <- convert_tibble_to_named_vector(tib)
+#' print(named_vector)
+#' # Output:
+#' # ||column::age||row::index:1::28||row::index:2::32||
+#' # ||column::city||row::index:1::New York||row::index:2::Los Angeles||
+#' # ||column::name||row::index:1::Alice||row::index:2::Bob||
+#'
+#' @importFrom tibble tibble
+#' @importFrom dplyr nrow
+#' @importFrom stats setNames
+#' @importFrom purrr map lapply unlist
+#'
+#' @export
+convert_tibble_to_named_vector <- function(tibble) {
+    listed.tibble <- as.list(tibble)
+    row.names <- paste0("||row::index:", 1:nrow(tibble), "::", listed.tibble[[1]], "||")
+    tibble.data.columns <- listed.tibble[-1]
+    names(tibble.data.columns) <- paste0("||column::", names(tibble.data.columns), "||")
+    unlisted.tibble.data.columns <- lapply(tibble.data.columns, unlist)
+    named.tibble.data.columns <- lapply(unlisted.tibble.data.columns, setNames, nm = row.names)
+    vector <- unlist(named.tibble.data.columns)
+    return (vector)
+}

### 2023-08-07 pilot-publication/functions/create_outcomes_table.R

@@ -1,0 +1,27 @@
+#' Create outcomes table from data
+#'
+#' This function takes a data frame as input and creates an outcomes table based on the specified summarization settings.
+#' The outcomes table is created using `tbl_summary()` from the `gtsummary` package and includes summary statistics for different arms of the study.
+#' The summary statistics are calculated for dichotomous variables as categorical data, and for continuous and categorical variables as median and proportion (p), respectively.
+#' Missing values are labeled as "Missing" in the outcomes table.
+#'
+#' @param data The data frame containing the study data.
+#'
+#' @return An outcomes table as a tibble containing summary statistics for different arms of the study.
+#'
+#' @importFrom gtsummary tbl_summary all_dichotomous all_continuous all_categorical
+#' @importFrom dplyr as_tibble
+#' 
+#' @export
+create_outcomes_table <- function(data) {
+    outcomes.table <- data %>%
+        tbl_summary(by = "arm",
+                    type = all_dichotomous() ~ "categorical",
+                    statistic = list(
+                        all_continuous() ~ "{median}",
+                        all_categorical() ~ "{p}"
+                    ),
+                    missing_text = "Missing") %>%
+        as_tibble()
+    return (outcomes.table)
+}

### 2023-08-07 pilot-publication/functions/estimate_outcome_results.R

@@ -1,0 +1,95 @@
+#' Estimate outcome results from study data
+#'
+#' This function takes a data frame as input and estimates outcome
+#' results from the study data. It performs various calculations,
+#' including creating outcomes tables, calculating absolute and
+#' relative differences between pre and post-training outcomes,
+#' comparing post-training outcomes between different arms, and
+#' comparing change from baseline between arms. The function returns
+#' the results as a named vector, to be used for bootstrapping.
+#'
+#' @param data The data frame containing the study data.
+#' @param index The index of the bootstrap sample.
+#'
+#' @return A named vector containing the estimated outcome results from the study data.
+#'
+#' @importFrom dplyr select filter
+#' @importFrom stats combn
+#' 
+#' @export
+estimate_outcome_results <- function(data, index) {
+    ## Check arguments
+    assertthat::assert_that(is.data.frame(data))
+    assertthat::assert_that(is.numeric(index))
+
+    ## Write index to disk
+    index.file <- data.frame(i = index)
+    if (file.exists("out/index.csv")) {
+        index.file <- read.csv("out/index.csv")
+        index.file <- index.file %>% add_column(i = index, .name_repair = "unique")
+    }
+    write.csv(index.file, "out/index.csv", row.names = FALSE)
+    
+    ## Create bootstrap sample
+    data <- data[index, ]
+
+    ## Create outcomes tables
+    outcomes <- c(names(binary_outcomes()),
+                  categorical_outcomes(),
+                  quantitative_outcomes())
+    outcomes.data <- data[, c(outcomes, "arm", "post.training")]
+    outcomes.data.list <- list(
+        overall = outcomes.data %>% select(-post.training),
+        pre.training = outcomes.data %>% filter(!post.training) %>% select(-post.training),
+        post.training = outcomes.data %>% filter(post.training) %>% select(-post.training))
+    outcomes.tables <- lapply(outcomes.data.list, create_outcomes_table)
+    outcomes.row.names <- outcomes.tables$overall[, 1]
+
+    ## Convert all outcome data in tables to numeric
+    outcomes.tables <- lapply(outcomes.tables, convert_table_data_to_numeric)
+
+    ## Calculate the absolute difference between the pre and post training outcomes in the same trial arms using only the three columns with the arms data and not the characteristics column
[…]

### 2023-08-07 pilot-publication/main.R

@@ -6,6 +6,7 @@ library(lubridate)
 library(gtsummary)
 library(naniar)
 library(icdpicr)
+library(boot)
 
 ## Source functions
 noacsr::source_all_functions()
@@ -37,186 +38,32 @@ sample.characteristics.table <- create_descriptive_table(table.data,
                                                          caption = "Patient sample characteristics",
                                                          include.overall = TRUE)
 
-## Create tables comparing outcomes
-outcomes <- c(names(binary_outcomes()),
-              categorical_outcomes(),
-              quantitative_outcomes())
-outcomes.data <- data[, c(outcomes, "arm", "post.training")]
-outcomes.data.list <- list(
-    overall = outcomes.data %>% select(-post.training),
-    pre.training = outcomes.data %>% filter(!post.training) %>% select(-post.training),
-    post.training = outcomes.data %>% filter(post.training) %>% select(-post.training))
-outcomes.tables <- lapply(outcomes.data.list, create_outcomes_table)
-outcomes.row.names <- outcomes.tables$overall[, 1]
+## Bootstrap outcome results
+dir.create("out", showWarnings = FALSE)
+n.boot.samples <- 1000
+bootstrapped.outcome.results <- boot(data, estimate_outcome_results, R = n.boot.samples)
 
-create_outcomes_table <- function(data) {
-    outcomes.table <- data %>%
-        tbl_summary(by = "arm",
-                    type = all_dichotomous() ~ "categorical",
-                    statistic = list(
-                        all_continuous() ~ "{median}",
-                        all_categorical() ~ "{p}"
-                    ),
-                    missing_text = "Missing") %>%
-        as_tibble()
-    return (outcomes.table)
-}
+## Calculate confidence intervals
+ci.levels <- c(0.75, 0.85, 0.95)
+ci.type <- "basic"
+bootstrapped.outcome.results.ci <- lapply(seq_along(bootstrapped.outcome.results$t0), function(index) {
+    t0.value <- bootstrapped.outcome.results$t0[index]
+    ci <- NA
+    if (!is.na(t0.value)) {
+        boot.ci.object <- boot.ci(boot.out = bootstrapped.outcome.results, index = index, type = ci.type, conf = ci.levels)
+        ci <- boot.ci.object[[ci.type]][, c(1, 4:5)]
+        colnames(ci) <- c("level", "lower", "upper")
+    }
[…]

### 2023-08-07 pilot-publication/lab-book.Rmd
Fix broken bootstrap caused by data entry error

@@
 ```
+
+## Strange error when running bootstrap on real data
+
+I get this strange error when I try to run the bootstrap on the real data:
+
+```{r}
+## Error in `add_column()`:
+## ! New columns must be compatible with `.data`.
+## ✖ New column has 112 rows.
+## ℹ `.data` has 111 rows.
+## Backtrace:
+##     ▆
+##  1. ├─boot::boot(data, estimate_outcome_results, R = n.boot.samples)
+##  2. │ └─global statistic(data, original, ...)
+##  3. │   └─outcomes.tables$absolute.difference %>% as_tibble() %>% ...
+##  4. └─tibble::add_column(., outcomes.row.names, .before = 1)
+##  5.   └─tibble:::abort_incompatible_new_cols(nrow(.data), df)
+##  6.     └─tibble:::tibble_abort(...)
+##  7.       └─rlang::abort(x, class, ..., call = call, parent = parent, use_cli_format = TRUE)
+## Execution halted
+```
+
+The cause of the error is the entry 995 in the variable `complications__number_of_hospitalizations_for_this_injury`.
+Not sure where that entry comes from, but most likely it should be 999 but something went wrong?
+A quick fix is to replace this value with 999 in the `prepare_data` function, which should work.
 

### 2023-08-07 pilot-publication/functions/prepare_data.R

@@ -38,7 +38,10 @@ prepare_data <- function(data, codebook = NULL) {
     variable[variable == "Yes - Surgery on the 3rd day"] <- "Yes"
     variable <- as.factor(variable)
     prepared.data$complications__failure_of_conservative_management <- variable
-
+    variable <- prepared.data$complications__number_of_hospitalizations_for_this_injury
+    variable[variable == 995] <- NA ## Entered as 995 instead of 999 by mistake
+    prepared.data$complications__number_of_hospitalizations_for_this_injury <- variable
+    
     ## Generate AIS codes
     icd10.data <- prepared.data[, c("interventions__injury_first_surg_icd10",
                                     "interventions__injury_xray_icd10",

### 2023-08-07 pilot-publication/main.R

@@ -39,6 +39,7 @@ sample.characteristics.table <- create_descriptive_table(table.data,
                                                          include.overall = TRUE)
 
 ## Bootstrap outcome results
+unlink("out", recursive = TRUE)
 dir.create("out", showWarnings = FALSE)
 n.boot.samples <- 1000
 bootstrapped.outcome.results <- boot(data, estimate_outcome_results, R = n.boot.samples)

### 2023-08-07 pilot-publication/lab-book.Rmd
Document new error when trying to calculate differences between data.frames

@@
 
-## Strange error when running bootstrap on real data
+## Errors when running bootstrap on real data
 
-I get this strange error when I try to run the bootstrap on the real data:
+1. I first got this strange error when I try to run the bootstrap on the real data:
 
@@
 A quick fix is to replace this value with 999 in the `prepare_data` function, which should work.
+
+2. I then got this error:
+
+```{r}
+## Error in Ops.data.frame(outcomes.tables$post.training[2:4], outcomes.tables$pre.training[2:4]) : 
+##   ‘-’ only defined for equally-sized data frames
+## Calls: boot -> statistic -> Ops.data.frame
+## Execution halted
+```
+
+Will have to look into that.
 

### 2023-11-26 pilot-publication/manuscript.Rmd

@@
 - Lost to follow up rate. This applied only to patients and was equal
-  to the proportion of patients that do not complete 30 day follow up,
+  to the proportion of patients that did not complete 30 day follow up,
   out of all enrolled patients, over the course of the pilot study.

### 2023-11-28 pilot-publication/lab-book.Rmd
Refactor saving of bootstrap results

@@
 
-Will have to look into that.
+This was because the three different datasets, one for overall, one for pre-training and one for post-training, were not the same size.
+The reason was that there were different but small numbers of unique values in some variables that should be treated as numeric but that was instead treated as categorical by tbl_summary.
+Ensuring that these were in fact treated as numeric solved the issue.
 

### 2023-11-28 pilot-publication/main.R

@@ -49,6 +49,7 @@ unlink("out", recursive = TRUE)
 dir.create("out", showWarnings = FALSE)
 n.boot.samples <- 10
 bootstrapped.outcome.results <- boot(data, try_estimate_outcome_results, R = n.boot.samples)
+saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outcome-results.Rds"))
 
 ## Calculate confidence intervals
 ci.levels <- c(0.75, 0.85, 0.95)
@@ -67,7 +68,4 @@ bootstrapped.outcome.results.ci <- lapply(seq_along(bootstrapped.outcome.results
     return(ci)
 })
 names(bootstrapped.outcome.results.ci) <- names(bootstrapped.outcome.results$t0)
-
-## Save bootstrapped results and confidence intervals
-saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outcome-results.Rds"))
 saveRDS(bootstrapped.outcome.results.ci, file = file.path("out", "bootstrapped-outcome-results-ci.Rds"))

### 2023-11-28 pilot-publication/main.R
Increase bootstraps

@@ -47,7 +47,7 @@ sample.characteristics.table <- create_descriptive_table(table.data,
 unlink("error.log")
 unlink("out", recursive = TRUE)
 dir.create("out", showWarnings = FALSE)
-n.boot.samples <- 10
+n.boot.samples <- 1000
 bootstrapped.outcome.results <- boot(data, try_estimate_outcome_results, R = n.boot.samples)
 saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outcome-results.Rds"))
 

### 2023-11-29 pilot-publication/lab-book.Rmd
Update manuscript

@@
 library(DataExplorer)
+```
+
+## Playground
+
+```{r}
+boot.files <- list.files("out")
+length(boot.files)
 ```
@@
     if (is.factor(column)) {
-        return (as.factor(sample(levels(column), size = length(column), replace = TRUE)))
+        return(as.factor(sample(levels(column), size = length(column), replace = TRUE)))
     } else if (is.character(column)) {
-        return (as.character(sample(column, size = length(column), replace = TRUE)))
+        return(as.character(sample(column, size = length(column), replace = TRUE)))
     } else if (is.numeric(column)) {
-        return (rnorm(n = length(column), mean = mean(column, na.rm = TRUE), sd = sd(column, na.rm = TRUE)))
+        return(rnorm(n = length(column), mean = mean(column, na.rm = TRUE), sd = sd(column, na.rm = TRUE)))
     } else if (is.logical(column)) {
-        return (sample(c(TRUE, FALSE), size = length(column), replace = TRUE))
+        return(sample(c(TRUE, FALSE), size = length(column), replace = TRUE))
     } else if (is.POSIXct(column) | is.POSIXt(column)) {
-        return (as.POSIXct(sample(as.numeric(column), size = length(column), replace = TRUE), origin = "1970-01-01"))
+        return(as.POSIXct(sample(as.numeric(column), size = length(column), replace = TRUE), origin = "1970-01-01"))
     } else {
-        return (rep(NA, length(column)))
+        return(rep(NA, length(column)))
     }
@@
     new.values <- rep(new.values, length.out = length(column))
-    return (new.values)
+    return(new.values)
 })
@@
     new.values <- rep(sample(1:1e6, size = n.unique, replace = FALSE), length.out = length(column))
-    return (new.values)
+    return(new.values)
 })
@@
     ## Define basic results
-    arrival.dates <- data %>% pull(incident__date_of_arrival) %>% as.Date()
-    results$start.date <- arrival.dates %>% min() %>% format_date()
-    results$end.date <- arrival.dates %>% max() %>% format_date()
-    n.no.consent <-  list("11542" = 40,
+    arrival.dates <- data %>%
+        pull(incident__date_of_arrival) %>%
+        as.Date()
+    results$start.date <- arrival.dates %>%
+        min() %>%
+        format_date()
+    results$end.date <- arrival.dates %>%
+        max() %>%
+        format_date()
+    n.no.consent <- list(
+        "11542" = 40,
         "44805" = 10,
@@
         "88456" = 0, # To be updated
-                          "10263" = 2)
+        "10263" = 2
+    )
     results$icc <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
@@
     results$n.residents <- with(results, n.atls.residents + n.ptc.residents)
-    results$n.centres <-  data %>% pull(id__reg_hospital_id) %>% unique() %>% length()
+    results$n.centres <- data %>%
+        pull(id__reg_hospital_id) %>%
+        unique() %>%
+        length()
     results$n.atls <- sum(data$arm == "ATLS")
@@
     results$n.females <- with(data, sum(patinfo__pt_gender == "Female"))
-    results$p.females <- round(results$n.females/nrow(data) * 100)
+    results$p.females <- round(results$n.females / nrow(data) * 100)
     results$median.age <- median(data$patinfo__pt_age, na.rm = TRUE)
@@
     results$n.admitted <- with(data, sum(interventions__admitted == "Yes"))
-    results$p.admitted <- round(results$n.admitted/nrow(data) * 100)
-    return (results)
+    results$p.admitted <- round(results$n.admitted / nrow(data) * 100)
+    return(results)
 }
@@
 n.m30d <- with(data, sum(outcomes__alive_after_30_days == "Yes", na.rm = TRUE))
-p.m30d <- round(n.m30d/nrow(data) * 100)
-arms.n.m30d.list <- lapply(setNames(arms.data.list, nm = paste0("m30d.", names(arms.data.list))),
-                           function(arms.data)
-                               with(arms.data,
-                                    sum(outcomes__alive_after_30_days == "Yes", na.rm = TRUE)))
+p.m30d <- round(n.m30d / nrow(data) * 100)
+arms.n.m30d.list <- lapply(
+    setNames(arms.data.list, nm = paste0("m30d.", names(arms.data.list))),
+    function(arms.data) {
+        with(
+            arms.data,
+            sum(outcomes__alive_after_30_days == "Yes", na.rm = TRUE)
+        )
+    }
+)
 arms.p.m30d.list <- mapply(arms.n.m30d.list, arms.data.list,
-                           FUN = function(count, arms.data) list(round(count/nrow(arms.data) * 100)))
+    FUN = function(count, arms.data) list(round(count / nrow(arms.data) * 100))
+)
 names(arms.n.m30d.list) <- paste0("n.", names(arms.n.m30d.list))
@@
 arr.ptc.control <- p.m30d.ptc - p.m30d.control
-rr.atls.ptc <- round(p.m30d.atls/p.m30d.ptc, 2)
-rr.atls.control <- round(p.m30d.atls/p.m30d.control, 2)
-rr.ptc.control <- round(p.m30d.ptc/p.m30d.control, 2)
-p.missing.in.hospital.mortality <- round(sum(is.na(data$outcomes__discharge_alive))/n.patients * 100)
+rr.atls.ptc <- round(p.m30d.atls / p.m30d.ptc, 2)
+rr.atls.control <- round(p.m30d.atls / p.m30d.control, 2)
+rr.ptc.control <- round(p.m30d.ptc / p.m30d.control, 2)
+p.missing.in.hospital.mortality <- round(sum(is.na(data$outcomes__discharge_alive)) / n.patients * 100)
 
@@
 n.composite.outcome <- sum(composite.outcome, na.rm = TRUE)
-p.composite.outcome <- round(n.composite.outcome/nrow(data) * 100)
+p.composite.outcome <- round(n.composite.outcome / nrow(data) * 100)
 icc.composite.outcome <- estimate_icc("composite.outcome", "id__reg_hospital_id", data)
@@
 ## figures once known
-n.potentially.eligible = list(n.potentially.eligible.centre.1 = 999,
+n.potentially.eligible <- list(
+    n.potentially.eligible.centre.1 = 999,
     n.potentially.eligible.centre.2 = 999,
@@
     n.potentially.eligible.centre.6 = 999,
-                              n.potentially.eligible.centre.7 = 999)
-recruitment.rate.patients <- round(n.patients/do.call(sum, n.potentially.eligible) * 100)
+    n.potentially.eligible.centre.7 = 999
+)
+recruitment.rate.patients <- round(n.patients / do.call(sum, n.potentially.eligible) * 100)
 n.eligible.residents <- 21
-recruitment.rate.residents <- round(n.residents/n.eligible.residents * 100)
+recruitment.rate.residents <- round(n.residents / n.eligible.residents * 100)
 n.lost.to.follow.up <- sum(is.na(data$outcomes__alive_after_30_days))
-rate.lost.to.follow.up <- round(n.lost.to.follow.up/n.patients * 100)
+rate.lost.to.follow.up <- round(n.lost.to.follow.up / n.patients * 100)
 n.atls.residents.passed.first.attempt <- 2 + 2 # The number of residents who had passed ATLS after the first attempt, per centre
 n.atls.residents.passed.second.attempt <- 4 + 2 # The number of residents who had passed ATLS after the second attempt, per centre
-rate.pass <- round((n.atls.residents.passed.second.attempt + n.ptc.residents)/n.residents * 100)
-rates.missing.data <- unlist(lapply(data, function(column) round(sum(is.na(column)/length(column) * 100))))
+rate.pass <- round((n.atls.residents.passed.second.attempt + n.ptc.residents) / n.residents * 100)
+rates.missing.data <- unlist(lapply(data, function(column) round(sum(is.na(column) / length(column) * 100))))
 min.missing.data <- min(rates.missing.data)
@@
 mean.cluster.size <- mean(cluster.sizes)
-ratio.sd.mean.cluster.size <- sd.cluster.size/mean.cluster.size
+ratio.sd.mean.cluster.size <- sd.cluster.size / mean.cluster.size
 ```

### 2023-11-29 pilot-publication/manuscript.Rmd
Update manuscript

@@
 
-**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes.
-We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.
-**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught.
+**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work.
+We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught.
 We recruited tertiary hospitals and included trauma patients and residents managing these patients.
@@
 **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres.
-The patient recruitment rate was X and the resident recruitment rate was `r recruitment.rate.residents`%.
+The patient recruitment rate was `r recruitment.rate.patients`% and the resident recruitment rate was `r recruitment.rate.residents`%.
 The lost to follow up rate was `r rate.lost.to.follow.up`%.
@@
 Trauma, defined as the clinical entity composed of physical injury and
-the body's associated response, causes 4.5 millions deaths every year
-[@GBD2018]. Almost 10% of the global burden of disease is due
+the body's associated response, causes 4.3 millions deaths every year
+[@injuries2020]. Almost 10% of the global burden of disease is due
 to trauma and trauma is the top contributor to the burden of disease in
@@
 
-The proprietary Advanced Trauma Life Support (ATLS) is the most
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) is the most
 established trauma life support training program and more than one
-million doctors in over 80 countries have been trained in the program
+million doctors in over 80 countries have been trained in the programme
 [@acsAtls2018]. Uptake in low- and middle income countries (LMIC)
@@
 The free Primary Trauma Care (PTC) program is the most widely spread
-alternative program. The goal of PTC is to improve trauma care in LMIC
-[@ptcfPtc2018]. Like ATLS, doctors in over 80 countries have been
-trained in PTC, and the program has been endorsed by the World Health
+alternative programme. The goal of PTC is to improve trauma care in LMIC
+[@ptcfPtc2018]. Like ATLS^®^, doctors in over 80 countries have been
+trained in PTC, and the programme has been endorsed by the World Health
 Organization (WHO), among other international organizations
@@
 
-Despite the widespread use of these training programs there are no
-controlled trials showing that they impact patient outcomes
-[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. A systematic review
-report that hospital trauma life support training is associated with a
-reduction in patient mortality, based on studies that were either
-observational or quasi-experimental without a control group
-[@Jin2021].<!-- Note that some of the studies that Jin include state
-that there was no effect of ATLS -->
-
-We will perform a pilot study that aims to assess the feasibility of
+There are observational studies indicating that trauma life support
+training programmes may improve patient outcomes [@Jin2021], but there
+are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. 
+We performed a pilot study aiming to assess the feasibility of
 conducting a cluster randomised controlled trial comparing ATLS and
-PTC with standard care. Recent methodological guidelines indicate that
-the design of efficient cluster randomised controlled trials requires
-data on probable or target effect sizes, proportion of participants
-with the outcome (if binary), and the intracluster correlation
-coefficient [@Hemming2017]. The objectives of this pilot study were
+PTC with standard care.  The objectives of this pilot study were
 to:
@@
 - Estimate probable effect sizes on patient outcomes associated with
-  ATLS and PTC compared with standard care, estimate the proportion of
+  ATLS^®^ and PTC compared with standard care, estimate the proportion of
   participants with the outcome (if binary), and estimate the
@@
 We piloted a three-armed cluster randomised controlled trial. There
-were two intervention arms, ATLS and PTC training, and one control
+were two intervention arms, ATLS^®^ and PTC training, and one control
 arm, standard care. We collected data for four months in all three
@@
 We conducted this pilot in seven Indian tertiary hospitals, where
-neither ATLS, PTC, nor any other trauma life support training program
+neither ATLS^®^, PTC, nor any other trauma life support training program
 is routinely taught. The seven hospitals were Seth G. S. Medical
@@
 hospitals for at least one year. To facilitate administration each
-surgical department is divided into units, which manages the out
-patient department, emergency department, operating rooms etc on
+surgical department is divided into units, which manage the out-patient 
+and emergency departments, operating rooms etc on
 different days each week. One or two, out of typically six, units'
@@
 To be eligible, units had to have a maximum of 25% of the doctors
-trained in either ATLS, PTC, or similar training programs before the
+trained in either ATLS^®^, PTC, or similar training programs before the
 start of the pilot. Those residents who had received training in the
@@
 Consent was sought from the residents in each of the intervention
-groups before they underwent the ATLS or PTC training. The consent
+groups before they underwent the ATLS^®^ or PTC training. The consent
 form for residents is available as Supplemental Material 2. We did
@@
 residents per hospital providing emergency care to trauma patients
-were trained in either ATLS or PTC. For the purpose of this pilot
-study, we will target to train a minimum of 75% of residents in each
+were trained in either ATLS^®^ or PTC. For the purpose of this pilot
+study, our target was to train a minimum of 75% of residents in each
 unit. If residents dropped out or changed units after training but
@@
 required. We did not train the units' faculty, as they are typically
-not involved in the initial management of trauma patients.
+not directly involved in the initial management of trauma patients.
 
-The ATLS training was conducted in an ATLS certified training centre
-in Mumbai, according to the standard ATLS curriculum
+The ATLS^®^ training was conducted in an ATLS certified training centre
+in Mumbai, according to the standard ATLS^®^ curriculum
 [@acsAtls2018]. The PTC training was conducted in New Delhi, according
@@
 
-The control group provides standard care with no intervention.
+The control group provided standard care with no intervention.
 
 ### Modifications
-Both ATLS and PTC are standard training programs with fixed curricula
+Both ATLS^®^ and PTC are standard training programs with fixed curricula
 [@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
@@
 ### Adherence
-The intervention was the training in either ATLS or PTC and resident
+The intervention was the training in either ATLS^®^ or PTC and resident
 participants were required to adhere to, i.e. participate in, the
@@
 
-Although training in an advanced trauma life support course is part of
-the curriculum it is optional and not doing this training does not
+Although training in an ATLS^®^ is part of the curriculum it is 
+optional and not doing this training does not
 result in failure to obtain post graduation completion.
@@
 3. We decided to include a large number of outcomes,
-including some more exploratory, so that we can test their feasibility
+including some more exploratory, so that we could test their feasibility
 and relevance.
@@
 ### Patients
-Patients were screened for eligibility as they arrive at the emergency
+Patients were screened for eligibility as they arrived at the emergency
 department. Eligible patients were approached in the emergency
-department to consent to follow up, if they are conscious. If they
+department to consent to follow up, if they were conscious. If they
 were unconscious a patient representative was approached to consent to
-follow up. Once the patient is conscious we approached the patient to
+follow up. Once the patient was conscious we approached the patient to
 affirm the patient representative's consent. We followed up patients
@@
 ### Residents
-Participating units were be screened for eligibility once hospitals
+Participating units were screened for eligibility once hospitals
 confirmed their participation. All residents in eligible units were
@@
 Reporting Involvement of Patients and the Public (GRIIPS 2)
-[@Staniszewska2017].
-
-During the pilot study, interviews were conducted with post-discharge
-trauma patients and their caregivers to identify the outcomes most
-relevant to them. These patients were identified through the medical
-registers of the participating hospitals, contacted through telephone,
-and after receiving their consent they were interviewed as per their
-convenience. Their consent form is available as Supplemental
-Material 5.  Additionally, members from non-government organizations
-working with trauma patients and the hospital Social Service Section
-were contacted for their views on contextual patient-centred outcomes
-for trauma patients. Their consent form is available as Supplemental
-Material 6. For feasibility, these interviews were held in each of the
-cities where the participating centres are located. The commonest
-patient-centred outcomes reported across all the locations will be
-incorporated into the evaluation of the effects of the different
-training programs and standardized care on patient outcomes.
-
-Similarly, the inputs of resident doctor participants at each
-participating centre were collected during the pilot study. A
-discussion with the residents were conducted to document any
-challenges or suggestions they may have in the scheduling or
-implementation of the training programs. These inputs will be
-incorporated in the final study.
-
-A summary of the findings of the study as well as their inputs will be
-shared with those who participated in the interviews and
-discussions. A meeting will be held with the patient participants, at
-each city, where the changes in the measured patient-centered outcomes
-would be presented to them. Another meeting will be held with the
-resident doctors at each hospital to present the confidence of the
-residents after being trained.  Any suggestions and reflections from
-the participants during the meetings will be used as inputs for
-planning the final study.
+[@Staniszewska2017]. The results of these consultations will be reported 
+separately.
 
@@
 data were only accessible by the project's principal investigator or
-by delegation by the project principal investigator only.
+by delegation by the project principal investigator.
 
@@
 # Competing Interests
-Several authors are Advanced Trauma Life Support or Primary Trauma
-Care instructors.
+Several authors are ATLS^®^ and/or PTC instructors.
 
@@
   kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)
-
 ```

### 2023-11-29 pilot-publication/main.R

@@ -1,4 +1,4 @@
-## Run from command line with /usr/lib/R/bin/R --no-restore --file=main.r > run.log
+## Run from command line with /usr/lib/R/bin/R --no-restore --file=main.R > run.log
 
 ## Load packages
 library(dotenv)

### 2023-12-26 pilot-publication/main.R
Save data for outcome tables

@@ -57,4 +57,11 @@ saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outc
 bootstrapped.outcome.results.ci <- calculate_bootstrap_cis(bootstrapped.outcome.results)
 saveRDS(bootstrapped.outcome.results.ci, file = file.path("out", "bootstrapped-outcome-results-ci.Rds"))
 
+## Combine point estimates with confidence intervals
+outcome.results <- list(
+    point.estimates = bootstrapped.outcome.results$t0,
+    confidence.intervals = bootstrapped.outcome.results.ci
+)
+saveRDS(outcome.results, file = file.path("outcome-results.Rds"))
+
 ## Create tables with outcome results

### 2023-12-28 pilot-publication/lab-book.Rmd
Edit lab-book
Add code to generate outcome tables

@@
 
+## Import saved outcome results and create outcome tables
+
+```{r}
+## Load outcome results
+outcome.results <- readRDS("outcome-results.rds")
+
+## Extract cell references
+cell.references <- names(outcome.results$point.estimates)
+
+table.names <- strsplit(cell.references, split = "\\|\\|\\.") %>%
+    lapply(function(table.name) table.name[1]) %>%
+    unique()
+
+ci.levels <- outcome.results$confidence.intervals[[2]][, 1]
+
+## Combine point estimates and confidence intervals into a single table
+combined.outcome.results <- lapply(ci.levels, function(ci.level) {
+    lapply(cell.references, function(cell.reference) {
+        point.estimate <- outcome.results$point.estimates[[cell.reference]]
+        ci.data <- outcome.results$confidence.intervals[[cell.reference]]
+        ci <- ci.data[ci.data[, 1] == ci.level, c("lower", "upper")]
+        combined <- NA
+        if (!all(is.na(ci))) {
+            combined <- sprintf("%0.2f", c(point.estimate, ci))
+            combined <- paste0(combined[1], " (", combined[2], ", ", combined[3], ")")
+        }
+        return(combined)
+    }) %>%
+        setNames(cell.references)
+}) %>%
+    setNames(paste0("ci.level.", ci.levels))
+
+## Create outcome tables
+create_ci_level_tables <- function(table.name, ci.data) {
+    print(table.name)
+    table.data <- ci.data[grep(table.name, names(ci.data), fixed = TRUE)]
+    column.names <- strsplit(names(table.data), split = "\\|\\|\\.") %>%
+        lapply(function(column) column[2]) %>%
+        unique() %>%
+        stringr::str_replace_all(pattern = "\\|\\|column\\:\\:", replacement = "")
+    row.names <- strsplit(names(table.data), split = "\\|\\|\\.") %>%
+        lapply(function(row) row[3]) %>%
+        unique() %>%
+        stringr::str_replace_all(pattern = "\\|\\|row\\:\\:index\\:[0-9]*\\:\\:", replacement = "") %>%
+        stringr::str_replace_all(pattern = "\\|\\|", replacement = "")
+    ci.table <- matrix(table.data, ncol = length(column.names)) %>%
+        tibble::as_tibble(.name_repair = "minimal") %>%
+        tibble::add_column(row.names, .before = 1, .name_repair = "minimal") %>%
+        setNames(c("Outcome", column.names))
+    return(ci.table)
+}
+
+tables <- lapply(
+    combined.outcome.results, function(ci.data) {
+        lapply(table.names, create_ci_level_tables, ci.data = ci.data)
+    }
+)
+
+outcome.results.data <- enframe(outcome.results$point.estimates)
+table.names <- strsplit(names(outcome.results$point.estimates), split = "\\|\\|\\.")
+```
+
 ## Playground
@@
 length(boot.files)
+```
+
+## Create data report
+
+```{r}
+DataExplorer::create_report(data, report_title = "Data report", output_file = "data-report.pdf")
 ```

### 2023-12-29 pilot-publication/lab-book.Rmd
Edit code to generate outcome tables

@@
 library(DataExplorer)
+noacsr::source_all_functions()
 ```
@@
         stringr::str_replace_all(pattern = "\\|\\|", replacement = "")
+    ## Replace outcome names with labels
+    all.outcomes <- all_outcomes()
+    outcomes.column <- sapply(row.names, function(row.name) {
+        if (!is.null(all.outcomes[[row.name]])) {
+            return(all.outcomes[[row.name]]$label)
+        } else {
+            return(row.name)
+        }
+    })
+    ## Create table
     ci.table <- matrix(table.data, ncol = length(column.names)) %>%
         tibble::as_tibble(.name_repair = "minimal") %>%
-        tibble::add_column(row.names, .before = 1, .name_repair = "minimal") %>%
+        tibble::add_column(outcomes.column, .before = 1, .name_repair = "minimal") %>%
         setNames(c("Outcome", column.names))
@@
     combined.outcome.results, function(ci.data) {
+        {
             lapply(table.names, create_ci_level_tables, ci.data = ci.data)
+        } %>%
+            setNames(stringr::str_replace(table.names, pattern = "\\|\\|table\\:\\:", replacement = ""))
     }
 )
-
-outcome.results.data <- enframe(outcome.results$point.estimates)
-table.names <- strsplit(names(outcome.results$point.estimates), split = "\\|\\|\\.")
 ```

### 2023-12-29 pilot-publication/functions/outcomes.R

@@ -59,7 +59,7 @@ binary_outcomes <- function() {
           )
      )
      binary.outcomes <- lapply(binary.outcomes, function(binary.outcome) {
-          binary.outcome$hchange.from.baseline <- TRUE
+          binary.outcome$change.from.baseline <- TRUE
           return(binary.outcome)
      })
      names(binary.outcomes) <- sapply(binary.outcomes, function(binary.outcome) binary.outcome$name)
@@ -67,29 +67,65 @@ binary_outcomes <- function() {
 }
 
 categorical_outcomes <- function() {
-     setNames(nm = c(
-          "outcomes__eq5dm",
-          "outcomes__eq5dsc",
-          "outcomes__eq5dua",
-          "outcomes__eq5dpd",
-          "outcomes__eq5dad",
-          "outcomes__eq5dm30d",
-          "outcomes__eq5dsc30d",
-          "outcomes__eq5dua30d",
-          "outcomes__eq5dpd30d",
-          "outcomes__eq5dad30d",
-          "outcomes__patient_satisfaction"
-     ))
+     categorical.outcomes <- list(
+          list(
+               name = "outcomes__eq5dm",
+               label = "EQ-5D mobility at discharge"),
+          list(name = "outcomes__eq5dsc",
+               label = "EQ-5D self-care at discharge"),
+          list(name = "outcomes__eq5dua",
+               label = "EQ-5D usual activities at discharge"),
+          list(name = "outcomes__eq5dpd", 
+               label = "EQ-5D pain/discomfort at discharge"),
+          list(name = "outcomes__eq5dad",
+               label = "EQ-5D anxiety/depression at discharge"),
+          list(name = "outcomes__eq5dm30d",
+               label = "EQ-5D mobility at 30 day follow-up"),
+          list(name = "outcomes__eq5dsc30d",
+               label = "EQ-5D self-care at 30 day follow-up"),
+          list(name = "outcomes__eq5dua30d",
+               label = "EQ-5D usual activities at 30 day follow-up"),
+          list(name = "outcomes__eq5dpd30d",
+               label = "EQ-5D pain/discomfort at 30 day follow-up"),
+          list(name = "outcomes__eq5dad30d",
+               label = "EQ-5D anxiety/depression at 30 day follow-up"),
+          list(name = "outcomes__patient_satisfaction",
+               label = "Patient satisfaction")
+     )
+     categorical.outcomes <- lapply(categorical.outcomes, function(categorical.outcome) {
+          categorical.outcome$change.from.baseline <- TRUE
[…]

### 2023-12-30 pilot-publication/lab-book.Rmd
Create outcome tables and save as html and pdf

--- 2023-11-29 pilot-publication/lab-book.Rmd
@@
 library(DataExplorer)
+noacsr::source_all_functions()
+```
+
+## Import saved outcome results and create outcome tables
+
+```{r}
+## Load outcome results
+outcome.results <- readRDS("outcome-results.rds")
+
+## Extract cell references
+cell.references <- names(outcome.results$point.estimates)
+
+table.names <- strsplit(cell.references, split = "\\|\\|\\.") %>%
+    lapply(function(table.name) table.name[1]) %>%
+    unique()
+
+ci.levels <- outcome.results$confidence.intervals[[2]][, 1]
+
+## Combine point estimates and confidence intervals into a single table
+combined.outcome.results <- lapply(ci.levels, function(ci.level) {
+    lapply(cell.references, function(cell.reference) {
+        point.estimate <- outcome.results$point.estimates[[cell.reference]]
+        ci.data <- outcome.results$confidence.intervals[[cell.reference]]
+        ci <- ci.data[ci.data[, 1] == ci.level, c("lower", "upper")]
+        combined <- NA
+        if (!all(is.na(ci))) {
+            ## combined <- sprintf("%0.2f", c(point.estimate, ci))
+            combined <- paste0(combined[1], " (", combined[2], ", ", combined[3], ")")
+        }
+        return(combined)
+    }) %>%
+        setNames(cell.references)
+}) %>%
+    setNames(paste0("ci.level.", ci.levels))
+
+## Create outcome tables
+create_ci_level_tables <- function(table.name, ci.data) {
+    ## Extract table data
+    table.data <- ci.data[grep(table.name, names(ci.data), fixed = TRUE)]
+    column.names <- strsplit(names(table.data), split = "\\|\\|\\.") %>%
+        lapply(function(column) column[2]) %>%
+        unique() %>%
+        stringr::str_replace_all(pattern = "\\|\\|column\\:\\:", replacement = "") %>%
+        stringr::str_replace_all(pattern = "\\*\\*", replacement = "")
+    row.names <- strsplit(names(table.data), split = "\\|\\|\\.") %>%
+        lapply(function(row) row[3]) %>%
+        unique() %>%
+        stringr::str_replace_all(pattern = "\\|\\|row\\:\\:index\\:[0-9]*\\:\\:", replacement = "") %>%
+        stringr::str_replace_all(pattern = "\\|\\|", replacement = "")
+
+    ## Replace outcome names with labels
+    all.outcomes <- all_outcomes()
+    outcomes.column.list <- lapply(row.names, function(row.name) {
+        new.row.name <- row.name
+        index <- FALSE
+        if (!is.null(all.outcomes[[row.name]])) {
+            new.row.name <- all.outcomes[[row.name]]$label
+            index <- TRUE
+        }
+        return(list(row.name = new.row.name, index = index))
+    })
+    outcomes.column.index <- sapply(outcomes.column.list, function(outcomes.column) outcomes.column$index)
+    outcomes.column <- sapply(outcomes.column.list, function(outcomes.column) outcomes.column$row.name)
+
+    ## Create table
+    ci.table <- matrix(table.data, ncol = length(column.names)) %>%
+        tibble::as_tibble(.name_repair = "minimal") %>%
+        tibble::add_column(outcomes.column, .before = 1, .name_repair = "minimal") %>%
+        setNames(c("Outcome", column.names))
+
+    ## Replace NA with empty string
+    ci.table[] <- lapply(ci.table, function(column) {
+        column[is.na(column)] <- ""
+        return(column)
+    })
+
+    ## Convert table to gt
+    ci.table <- ci.table %>%
+        gt::gt() %>%
+        gt::tab_style(
+            style = gt::cell_text(weight = "bold"),
+            locations = gt::cells_body(
+                columns = c(Outcome),
+                rows = outcomes.column.index
+            )
+        ) %>%
+        gt::tab_style(
+            style = gt::cell_text(weight = "bold"),
+            locations = gt::cells_column_labels()
+        )
+
+    ## Return table
+    return(ci.table)
+}
+
+tables <- lapply(
+    combined.outcome.results, function(ci.data) {
+        {
+            lapply(table.names, create_ci_level_tables, ci.data = ci.data)
+        } %>%
+            setNames(stringr::str_replace(table.names, pattern = "\\|\\|table\\:\\:", replacement = ""))
+    }
+)
+
+## Save tables to file
+document.string <- lapply(
+    names(tables), function(ci.level.tables.name) {
+        ## Extract confidence level as percentage
+        ci.level <- stringr::str_remove(ci.level.tables.name, pattern = "ci\\.level\\.") %>%
+            as.numeric() %>%
+            scales::percent()
+        ci.level.tables <- tables[[ci.level.tables.name]]
+        paste0(
+            "# ", ci.level, " confidence level\n\n",
+            lapply(names(ci.level.tables), function(ci.level.table.name) {
+                ci.level.table <- ci.level.tables[[ci.level.table.name]]
+                paste0(
+                    "## ", ci.level.table.name, "\n\n",
+                    "\`\`\`{r, echo = FALSE} \n\n",
+                    "tables[[\"", ci.level.tables.name, "\"]][[\"", ci.level.table.name, "\"]]", "\n\n",
+                    "\`\`\`\n\n"
+                )
+            }) %>%
+                paste0(collapse = "\n\n")
+        )
+    }
+) %>%
+    unlist() %>%
+    paste0(collapse = "\n\n")
+
+## Write document string to file
+fileConn <- file("outcome-tables.Rmd")
+writeLines(document.string, fileConn)
+close(fileConn)
+
+## Create html document
+rmarkdown::render("outcome-tables.Rmd", output_format = rmarkdown::html_document(toc = TRUE))
+```
+
+The html document can then be compiled as pdf using pandoc from the command line:
+
+```{bash}  
+pandoc outcome-tables.html -o outcome-tables.pdf --pdf-engine=weasyprint
 ```

### 2023-12-30 pilot-publication/outcome-tables.Rmd
Create outcome tables and save as html and pdf

# 75% confidence level

## overall

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["overall"]]

```

## pre.training

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["pre.training"]]

```

## post.training

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["post.training"]]

```

## absolute.difference

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["absolute.difference"]]

```

## relative.difference

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["relative.difference"]]

```

## Post training outcome Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Post training outcome Standard care vs. ATLS"]]

```

## Post training outcome Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Post training outcome Standard care vs. PTC"]]

```

## Post training outcome ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Post training outcome ATLS vs. PTC"]]

```

## Change from baseline Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Change from baseline Standard care vs. ATLS"]]

```

## Change from baseline Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Change from baseline Standard care vs. PTC"]]

```

## Change from baseline ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.75"]][["Change from baseline ATLS vs. PTC"]]

```

# 85% confidence level

## overall

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["overall"]]

```

## pre.training

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["pre.training"]]

```

## post.training

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["post.training"]]

```

## absolute.difference

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["absolute.difference"]]

```

## relative.difference

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["relative.difference"]]

```

## Post training outcome Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Post training outcome Standard care vs. ATLS"]]

```

## Post training outcome Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Post training outcome Standard care vs. PTC"]]

```

## Post training outcome ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Post training outcome ATLS vs. PTC"]]

```

## Change from baseline Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Change from baseline Standard care vs. ATLS"]]

```

## Change from baseline Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Change from baseline Standard care vs. PTC"]]

```

## Change from baseline ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.85"]][["Change from baseline ATLS vs. PTC"]]

```

# 95% confidence level

## overall

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["overall"]]

```

## pre.training

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["pre.training"]]

```

## post.training

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["post.training"]]

```

## absolute.difference

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["absolute.difference"]]

```

## relative.difference

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["relative.difference"]]

```

## Post training outcome Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Post training outcome Standard care vs. ATLS"]]

```

## Post training outcome Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Post training outcome Standard care vs. PTC"]]

```

## Post training outcome ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Post training outcome ATLS vs. PTC"]]

```

## Change from baseline Standard care vs. ATLS

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Change from baseline Standard care vs. ATLS"]]

```

## Change from baseline Standard care vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Change from baseline Standard care vs. PTC"]]

```

## Change from baseline ATLS vs. PTC

```{r, echo = FALSE}

tables[["ci.level.0.95"]][["Change from baseline ATLS vs. PTC"]]

```

### 2023-12-30 pilot-publication/lab-book.Rmd
Use bca instead of basic bootstrap CIs

@@
 noacsr::source_all_functions()
+```
+
+## Import bootstrapped outcome results and generate confidence intervals
+
+```{r}
+## Load bootstrapped outcome results
+bootstrapped.outcome.results <- readRDS(file.path("out", "bootstrapped-outcome-results.Rds"))
+str(bootstrapped.outcome.results, max.level = 1)
+
+## Calculate confidence intervals
+bootstrapped.outcome.results.ci <- calculate_bootstrap_cis(bootstrapped.outcome.results)
+
+## Combine point estimates with confidence intervals
+outcome.results <- list(
+    point.estimates = bootstrapped.outcome.results$t0,
+    confidence.intervals = bootstrapped.outcome.results.ci
+)
+saveRDS(outcome.results, file = file.path("outcome-results.Rds"))
 ```
@@
         ci.data <- outcome.results$confidence.intervals[[cell.reference]]
-        ci <- ci.data[ci.data[, 1] == ci.level, c("lower", "upper")]
+        ci <- ci.data[ci.data[, "level"] == ci.level, c("lower", "upper")]
         combined <- NA
         if (!all(is.na(ci))) {
-            ## combined <- sprintf("%0.2f", c(point.estimate, ci))
+            combined <- round(c(point.estimate, ci), 2)
             combined <- paste0(combined[1], " (", combined[2], ", ", combined[3], ")")
@@
 ## ! New columns must be compatible with `.data`.
-## ✖ New column has 112 rows.
-## ℹ `.data` has 111 rows.
+## <U+2716> New column has 112 rows.
+## <U+2139> `.data` has 111 rows.
 ## Backtrace:
-##     ▆
-##  1. ├─boot::boot(data, estimate_outcome_results, R = n.boot.samples)
-##  2. │ └─global statistic(data, original, ...)
-##  3. │   └─outcomes.tables$absolute.difference %>% as_tibble() %>% ...
-##  4. └─tibble::add_column(., outcomes.row.names, .before = 1)
-##  5.   └─tibble:::abort_incompatible_new_cols(nrow(.data), df)
-##  6.     └─tibble:::tibble_abort(...)
-##  7.       └─rlang::abort(x, class, ..., call = call, parent = parent, use_cli_format = TRUE)
+##     <U+2586>
+##  1. <U+251C><U+2500>boot::boot(data, estimate_outcome_results, R = n.boot.samples)
+##  2. <U+2502> <U+2514><U+2500>global statistic(data, original, ...)
+##  3. <U+2502>   <U+2514><U+2500>outcomes.tables$absolute.difference %>% as_tibble() %>% ...
+##  4. <U+2514><U+2500>tibble::add_column(., outcomes.row.names, .before = 1)
+##  5.   <U+2514><U+2500>tibble:::abort_incompatible_new_cols(nrow(.data), df)
+##  6.     <U+2514><U+2500>tibble:::tibble_abort(...)
+##  7.       <U+2514><U+2500>rlang::abort(x, class, ..., call = call, parent = parent, use_cli_format = TRUE)
 ## Execution halted
@@
 ## Error in Ops.data.frame(outcomes.tables$post.training[2:4], outcomes.tables$pre.training[2:4]) :
-##   ‘-’ only defined for equally-sized data frames
+##   <U+2018>-<U+2019> only defined for equally-sized data frames
 ## Calls: boot -> statistic -> Ops.data.frame

### 2023-12-30 pilot-publication/functions/calculate_bootstrap_cis.R

@@ -19,7 +19,7 @@
 calculate_bootstrap_cis <- function(
     bootstrapped.outcome.results,
     ci.levels = c(0.75, 0.85, 0.95),
-    ci.type = "basic") {
+    ci.type = "bca") {
     ## Check arguments
     assertthat::assert_that(
         is.list(bootstrapped.outcome.results),

### 2024-01-02 pilot-publication/outcome-tables.Rmd
Add functions to generate outcome tables and save them to disk

@@
 
-## overall
+## Outcomes in all patients during the entire study period, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["overall"]]
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
 
@@
 
-## pre.training
+## Outcomes in all patients before training, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["pre.training"]]
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients before training, by treatment arm"]]
 
@@
 
-## post.training
+## Outcomes in all patients after training, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["post.training"]]
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients after training, by treatment arm"]]
 
@@
 
-## absolute.difference
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-tables[["ci.level.0.75"]][["absolute.difference"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-## relative.difference
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-tables[["ci.level.0.75"]][["relative.difference"]]
+outcome.results.tables[["ci.level.0.75"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-## Post training outcome Standard care vs. ATLS
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
 
@@
 
-## Post training outcome Standard care vs. PTC
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
 
@@
 
-## Post training outcome ATLS vs. PTC
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
 
@@
 
-## Change from baseline Standard care vs. ATLS
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
 
@@
 
-## Change from baseline Standard care vs. PTC
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
 
@@
 
-## Change from baseline ATLS vs. PTC
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
 
@@
 
-tables[["ci.level.0.85"]][["overall"]]
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.85"]][["pre.training"]]
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients before training, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.85"]][["post.training"]]
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients after training, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.85"]][["absolute.difference"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-tables[["ci.level.0.85"]][["relative.difference"]]
+outcome.results.tables[["ci.level.0.85"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-tables[["ci.level.0.85"]][["Post training outcome Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
 
@@
 
-tables[["ci.level.0.85"]][["Post training outcome Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
 
@@
 
-tables[["ci.level.0.85"]][["Post training outcome ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
 
@@
 
-tables[["ci.level.0.85"]][["Change from baseline Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
 
@@
 
-tables[["ci.level.0.85"]][["Change from baseline Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
 
@@
 
-tables[["ci.level.0.85"]][["Change from baseline ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
 
@@
 
-tables[["ci.level.0.95"]][["overall"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.95"]][["pre.training"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients before training, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.95"]][["post.training"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients after training, by treatment arm"]]
 
@@
 
-tables[["ci.level.0.95"]][["absolute.difference"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-tables[["ci.level.0.95"]][["relative.difference"]]
+outcome.results.tables[["ci.level.0.95"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-tables[["ci.level.0.95"]][["Post training outcome Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
 
@@
 
-tables[["ci.level.0.95"]][["Post training outcome Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
 
@@
 
-tables[["ci.level.0.95"]][["Post training outcome ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
 
@@
 
-tables[["ci.level.0.95"]][["Change from baseline Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
 
@@
 
-tables[["ci.level.0.95"]][["Change from baseline Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
 
@@
 
-tables[["ci.level.0.95"]][["Change from baseline ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
 

### 2024-01-02 pilot-publication/functions/create_outcome_results_tables.R

@@ -25,6 +25,11 @@ create_outcome_results_tables <- function(outcome.results) {
         lapply(function(table.name) table.name[1]) %>%
         unique()
 
+    ## Check table names
+    assertthat::assert_that(identical(unlist(table.names), get_table_names(return.original.names = TRUE)),
+        msg = "Invalid table names, maybe they have been changed?"
+    )
+
     ## Extract confidence interval levels
     ci.levels <- outcome.results$confidence.intervals[[2]][, 1]
 
@@ -45,13 +50,16 @@ create_outcome_results_tables <- function(outcome.results) {
     }) %>%
         setNames(paste0("ci.level.", ci.levels))
 
+    ## Change table names
+    new.table.names <- get_table_names()
+
     ## Create tables for each confidence interval level
     tables <- lapply(
         combined.outcome.results, function(ci.data) {
             {
                 lapply(table.names, create_ci_level_tables, ci.data = ci.data)
             } %>%
-                setNames(stringr::str_replace(table.names, pattern = "\\|\\|table\\:\\:", replacement = ""))
+                setNames(new.table.names)
         }
     )
 
@@ -133,4 +141,55 @@ create_ci_level_tables <- function(table.name, ci.data) {
 
     ## Return table
     return(ci.table)
+}
+
+get_table_names <- function(return.original.names = FALSE) {
+    ## Define original table names
+    table.names <- list(
+        "||table::overall" = "Outcomes in all patients during the entire study period, by treatment arm",
+        "||table::pre.training" = "Outcomes in all patients before training, by treatment arm",
+        "||table::post.training" = "Outcomes in all patients after training, by treatment arm",
+        "||table::absolute.difference" = paste0(
+            "Absolute change from baseline for all outcomes,",
+            " comparing the period after training with the period before training,",
+            " by treatment arms"
+        ),
+        "||table::relative.difference" = paste0(
+            "Relative change from baseline for all outcomes,",
+            " comparing the period after training with the period before training,",
+            " by treatment arms"
+        ),
+        "||table::Post training outcome Standard care vs. ATLS" = paste0(
+            "Absolute and relative differences in outcomes after training,",
+            " comparing standard care with ATLS"
+        ),
+        "||table::Post training outcome Standard care vs. PTC" = paste0(
+            "Absolute and relative differences in outcomes after training,",
+            " comparing standard care with PTC"
+        ),
+        "||table::Post training outcome ATLS vs. PTC" = paste0(
[…]

### 2024-01-02 pilot-publication/functions/save_tables_to_file.R

@@ -1,0 +1,49 @@
+#' Save outcome tables to file and create an HTML document
+#'
+#' This function takes a list of outcome tables and saves them to a file in R Markdown format.
+#' It then renders the R Markdown file into an HTML document with a table of contents.
+#'
+#' @param outcome.tables A list of outcome tables to be saved
+#'
+#' @export
+save_tables_to_file <- function(outcome.results.tables) {
+    ## Check arguments
+    assertthat::assert_that(
+        is.list(outcome.results.tables)
+    )
+
+    ## Save outcome tables to file
+    document.string <- lapply(
+        names(outcome.results.tables), function(ci.level.tables.name) {
+            ## Extract confidence level as percentage
+            ci.level <- stringr::str_remove(ci.level.tables.name, pattern = "ci\\.level\\.") %>%
+                as.numeric() %>%
+                scales::percent()
+            ci.level.tables <- outcome.results.tables[[ci.level.tables.name]]
+            paste0(
+                "# ", ci.level, " confidence level\n\n",
+                lapply(names(ci.level.tables), function(ci.level.table.name) {
+                    ci.level.table <- ci.level.tables[[ci.level.table.name]]
+                    paste0(
+                        "## ", ci.level.table.name, "\n\n",
+                        "\`\`\`{r, echo = FALSE} \n\n",
+                        "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",
+                        "[[\"", ci.level.table.name, "\"]]", "\n\n",
+                        "\`\`\`\n\n"
+                    )
+                }) %>%
+                    paste0(collapse = "\n\n")
+            )
+        }
+    ) %>%
+        unlist() %>%
+        paste0(collapse = "\n\n")
+
+    ## Write document string to file
+    fileConn <- file("outcome-tables.Rmd")
+    writeLines(document.string, fileConn)
+    close(fileConn)
+
+    ## Create html document
+    rmarkdown::render("outcome-tables.Rmd", output_format = rmarkdown::html_document(toc = TRUE))
+}

### 2024-01-02 pilot-publication/main.R

@@ -66,3 +66,6 @@ saveRDS(outcome.results, file = file.path("outcome-results.Rds"))
 
 ## Create tables with outcome results
 outcome.results.tables <- create_outcome_results_tables(outcome.results)
+
+## Save outcome results tables to file
+save_tables_to_file(outcome.results.tables)

### 2024-01-03 pilot-publication/manuscript.Rmd
Start adding results to manuscript

@@
 ## Patient Participant Outcomes
-A total of `r n.m30d.atls` ( `r p.m30d.atls`%) patients in the ATLS
+
+```{r, echo = FALSE}
+post.training.outcomes.table <- outcome.results.tables$ci.level.0.95[["Outcomes in all patients after training, by treatment arm"]]
+```
+
+After training, a total of `r n.m30d.atls` ( `r p.m30d.atls`%) patients in the ATLS
 arm had the primary outcome, compared to `r n.m30d.ptc` (

### 2024-01-03 pilot-publication/main.R

@@ -45,6 +45,16 @@ sample.characteristics.table <- create_descriptive_table(table.data,
     include.overall = TRUE
 )
 
+## Create table of sample characteristics post training
+post.training.table.data <- data %>%
+    filter(post.training) %>%
+    select(table.variables)
+post.training.characteristics.table <- create_descriptive_table(post.training.table.data,
+    strata = "arm",
+    caption = "Patient sample characteristics after training",
+    include.overall = TRUE
+)
+
 ## Bootstrap outcome results
 unlink("error.log")
 unlink("out", recursive = TRUE)

### 2024-01-03 pilot-publication/manuscript.Rmd
Continue adding results to text

--- 2023-11-29 pilot-publication/manuscript.Rmd
@@
 # Results
+```{r}
+females <- inline_text(overall.sample.characteristics.table,
+  variable = patinfo__pt_gender,
+  level = "Female"
+)
+median.age <- inline_text(overall.sample.characteristics.table,
+  variable = patinfo__pt_age
+)
+median.iss <- inline_text(overall.sample.characteristics.table,
+  variable = riss
+)
+median.niss <- inline_text(overall.sample.characteristics.table,
+  variable = niss
+)
+m30d <- inline_text(overall.sample.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes"
+)
+```
+
 Between `r start.date` and `r end.date`, we enrolled `r n.patients`
@@
 \@ref(tab:sample-characteristics). Overall, the number of females were
-`r n.females` (`r p.females`%), the median age was `r median.age`
-years (IQR `r iqr.age`), and the median ISS and NISS were `r median.iss`
-(IQR `r iqr.iss`) and `median.niss` (IQR `iqr.niss`) respectively. The 
-number of admitted patients was `r n.admitted` (`r p.admitted`%). A 
-total of `r n.m30d` (`r p.m30d`%) patients had the primary outcome of 
-mortality at 30 days after arrival to the emergency department.
+`r females`, the median (IQR) age was `r median.age`
+years, and the median ISS (IQR) and NISS (IQR) were `r median.iss` and `r median.niss` respectively.
+A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.
 
 ```{r sample-characteristics, echo = FALSE, results = "asis"}
-table1::t1kable(sample.characteristics.table)
+sample.characteristics.table
 ```
@@
 ## Patient Participant Outcomes
-A total of `r n.m30d.atls` ( `r p.m30d.atls`%) patients in the ATLS
+
+```{r, echo = FALSE}
+post.training.outcomes.table <- outcome.results.tables$ci.level.0.95[["Outcomes in all patients after training, by treatment arm"]]
+m30d.atls <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "ATLS"
+)
+```
+
+After training, a total of `r m30d.atls` patients in the ATLS
 arm had the primary outcome, compared to `r n.m30d.ptc` (

### 2024-01-03 pilot-publication/functions/create_descriptive_table.R

@@ -12,14 +12,18 @@
 #'     stratified. Defaults to NULL.
 #' @param include.overall Logical. If TRUE a column called "Overall"
 #'     is included in the table, which includes descriptive statistics
-#'     for the whole sample. Defaults to TRUE.
+#'     for the whole sample. Defaults to TRUE, unless strata is NULL,
+#'     in which case it defaults to FALSE.
 #' @param caption Character or NULL. The table caption. Defaults to
 #'     NULL, as in no caption.
+#' @param return.as.gt Logical. If TRUE the table is returned as a gt
+#'     object. Defaults to FALSE.
 create_descriptive_table <- function(table.data,
                                      variables = NULL,
                                      strata = NULL,
                                      include.overall = TRUE,
-                                     caption = NULL) {
+                                     caption = NULL,
+                                     return.as.gt = FALSE) {
     ## Define pipe operator
     `%>%` <- magrittr::`%>%`
 
@@ -29,6 +33,11 @@ create_descriptive_table <- function(table.data,
     assertthat::assert_that(is.character(strata) | is.null(strata))
     assertthat::assert_that(is.logical(include.overall))
     assertthat::assert_that(is.character(caption) | is.null(caption))
+
+    ## Set include.overall to FALSE if strata is NULL
+    if (is.null(strata)) {
+        include.overall <- FALSE
+    }
 
     ## Create table
     table.data <- as.data.frame(table.data)
@@ -55,7 +64,9 @@ create_descriptive_table <- function(table.data,
     }
 
     ## Return table as gt object
-    descriptive.table <- gtsummary::as_gt(descriptive.table)
+    if (return.as.gt) {
+        descriptive.table <- gtsummary::as_gt(descriptive.table)
+    }
 
     ## Return table
     return(descriptive.table)

### 2024-01-03 pilot-publication/main.R

@@ -39,21 +39,44 @@ table.variables <- c(
     "arm"
 )
 table.data <- data[, table.variables]
-sample.characteristics.table <- create_descriptive_table(table.data,
+overall.sample.characteristics.table <- create_descriptive_table(table.data)
+
+## Create table of sample characteristics before training
+pre.training.table.data <- data %>%
+    filter(!post.training) %>%
+    select(table.variables)
+pre.training.characteristics.table <- create_descriptive_table(
+    pre.training.table.data,
     strata = "arm",
-    caption = "Patient sample characteristics",
     include.overall = TRUE
 )
 
-## Create table of sample characteristics post training
+## Create table of sample characteristics after training
 post.training.table.data <- data %>%
     filter(post.training) %>%
     select(table.variables)
-post.training.characteristics.table <- create_descriptive_table(post.training.table.data,
+post.training.characteristics.table <- create_descriptive_table(
+    post.training.table.data,
     strata = "arm",
-    caption = "Patient sample characteristics after training",
     include.overall = TRUE
 )
+
+## Combine tables
+sample.characteristics.table <- gtsummary::tbl_merge(
+    tbls = list(
+        pre.training.characteristics.table,
+        post.training.characteristics.table,
+        overall.sample.characteristics.table
+    ),
+    tab_spanner = c("Before training", "After training", "Overall")
+) %>%
+    gtsummary::modify_caption("Patient sample characteristics")
+
+## Save tables to disk
+saveRDS(list(
+    sample.characteristics = sample.characteristics.table,
+    post.training.characteristics = post.training.characteristics.table
+), file = file.path("out", "sample-characteristics-tables.Rds"))
 
 ## Bootstrap outcome results
 unlink("error.log")

### 2024-01-05 pilot-publication/lab-book.Rmd
Work on extracting data from gt tables

@@
 ## Load outcome results
-outcome.results <- readRDS("outcome-results.rds")
+outcome.results <- readRDS("outcome-results.Rds")
 
@@
 pandoc outcome-tables.html -o outcome-tables.pdf --pdf-engine=weasyprint
+```
+
+## Access data from gt table object
+
+```{r}
+gt.table <- tables[[1]][[1]]
+str(gt.table$`_data`, max.level = 1)
+gt.table$`_data`
+gt.table$`_styles`
 ```

### 2024-01-05 pilot-publication/manuscript.Rmd
Work on extracting data from gt tables

@@
 )
+m30d.ptc <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "PTC"
+)
 ```

### 2024-01-05 pilot-publication/main.R

@@ -72,12 +72,6 @@ sample.characteristics.table <- gtsummary::tbl_merge(
 ) %>%
     gtsummary::modify_caption("Patient sample characteristics")
 
-## Save tables to disk
-saveRDS(list(
-    sample.characteristics = sample.characteristics.table,
-    post.training.characteristics = post.training.characteristics.table
-), file = file.path("out", "sample-characteristics-tables.Rds"))
-
 ## Bootstrap outcome results
 unlink("error.log")
 unlink("out", recursive = TRUE)

### 2024-01-08 pilot-publication/lab-book.Rmd
Complete function to extract inline text from gt outcome tables

@@
 ```{r setup}
-library(pbapply)
-library(openssl)
-library(dotenv)
-library(DataExplorer)
 noacsr::source_all_functions()
+load_packages()
 ```
@@
 )
+```
 
+## Save tables to file
+
+```{r}
 ## Save tables to file
@@
 str(gt.table$`_data`, max.level = 1)
+inline_outcome_text <- function(gt.table, outcome, level, column) {
+    ## Define borrowed functions
+    `%>%` <- magrittr::`%>%`
+    pull <- dplyr::pull
+
+    ## Check arguments
+    assertthat::assert_that(
+        "gt_tbl" %in% class(gt.table),
+        is.character(outcome),
+        is.character(level),
+        is.character(column) | is.numeric(column)
+    )
+
+    ## Extract data
+    gt.data <- gt.table$`_data`
+    outcome.index <- na.omit(gt.table$`_styles`$rownum) # This relies on outcome names being styled as bold
+    valid.outcomes <- gt.data$Outcome[outcome.index]
+    outcome.pos <- grep(paste0("^", outcome, "$"), valid.outcomes)
+
+    ## Stop if outcome not found
+    if (length(outcome.pos) == 0) {
+        stop(
+            "No outcome found for outcome ",
+            outcome,
+            ". Choose one from c(",
+            paste0(paste0("\"", valid.outcomes, "\""), collapse = ", "), ")."
+        )
+    }
+
+    ## Extract only the relevant part of the table
+    first.row <- outcome.index[outcome.pos] + 1
+    if (outcome.pos == rev(outcome.index)[1]) {
+        last.row <- nrow(gt.data)
+    } else {
+        last.row <- outcome.index[outcome.pos + 1] - 1
+    }
+    outcome.limits <- c(first.row, last.row)
+    outcome.subtable <- gt.data[outcome.limits[1]:outcome.limits[2], ]
+
+    ## Stop if column not found
+    if (!is.numeric(column)) {
+        new.column <- grep(column, names(outcome.subtable))
+        if (length(new.column) > 1) {
+            stop(
+                "More than one column found for column ",
+                column,
+                ". Choose one from c(",
+                paste0(paste0("\"", names(outcome.subtable), "\""), collapse = ", "), ")."
+            )
+        }
+        if (length(new.column) == 0) {
+            stop(
+                "No column found for column ",
+                column,
+                ". Choose one from c(",
+                paste0(paste0("\"", names(outcome.subtable), "\""), collapse = ", "), ")."
+            )
+        }
+    }
+
+    ## Stop if level not found
+    new.level <- grep(level, outcome.subtable$Outcome)
+    if (length(new.level) > 1) {
+        stop(
+            "More than one level found for level ",
+            level,
+            ". Choose one from c(",
+            paste0(paste0("\"", outcome.subtable$Outcome, "\""), collapse = ", "), ")."
+        )
+    }
+    if (length(new.level) == 0) {
+        stop(
+            "No level found for level ",
+            level,
+            ". Choose one from c(",
+            paste0(paste0("\"", outcome.subtable$Outcome, "\""), collapse = ", "), ")."
+        )
+    }
+
+    ## Extract inline text
+    outcome.text <- outcome.subtable[outcome.subtable$Outcome == level, new.column] %>%
+        pull() %>%
+        unlist()
+
+    ## Return inline text
+    return(outcome.text)
+}
 gt.table$`_data`

### 2024-01-08 pilot-publication/functions/inline_outcome_text.R

@@ -1,0 +1,87 @@
+inline_outcome_text <- function(gt.table, outcome, level, column) {
+    ## Define borrowed functions
+    `%>%` <- magrittr::`%>%`
+    pull <- dplyr::pull
+
+    ## Check arguments
+    assertthat::assert_that(
+        "gt_tbl" %in% class(gt.table),
+        is.character(outcome),
+        is.character(level),
+        is.character(column) | is.numeric(column)
+    )
+
+    ## Extract data
+    gt.data <- gt.table$`_data`
+    outcome.index <- na.omit(gt.table$`_styles`$rownum) # This relies on outcome names being styled as bold
+    valid.outcomes <- gt.data$Outcome[outcome.index]
+    outcome.pos <- grep(paste0("^", outcome, "$"), valid.outcomes)
+
+    ## Stop if outcome not found
+    if (length(outcome.pos) == 0) {
+        stop(
+            "No outcome found for outcome ",
+            outcome,
+            ". Choose one from c(",
+            paste0(paste0("\"", valid.outcomes, "\""), collapse = ", "), ")."
+        )
+    }
+
+    ## Extract only the relevant part of the table
+    first.row <- outcome.index[outcome.pos] + 1
+    if (outcome.pos == rev(outcome.index)[1]) {
+        last.row <- nrow(gt.data)
+    } else {
+        last.row <- outcome.index[outcome.pos + 1] - 1
+    }
+    outcome.limits <- c(first.row, last.row)
+    outcome.subtable <- gt.data[outcome.limits[1]:outcome.limits[2], ]
+
+    ## Stop if column not found
+    if (!is.numeric(column)) {
+        new.column <- grep(column, names(outcome.subtable))
+        if (length(new.column) > 1) {
+            stop(
+                "More than one column found for column ",
+                column,
+                ". Choose one from c(",
+                paste0(paste0("\"", names(outcome.subtable), "\""), collapse = ", "), ")."
+            )
+        }
+        if (length(new.column) == 0) {
+            stop(
+                "No column found for column ",
+                column,
+                ". Choose one from c(",
+                paste0(paste0("\"", names(outcome.subtable), "\""), collapse = ", "), ")."
+            )
+        }
+    }
+
+    ## Stop if level not found
+    new.level <- grep(level, outcome.subtable$Outcome)
+    if (length(new.level) > 1) {
+        stop(
+            "More than one level found for level ",
+            level,
[…]

### 2024-01-08 pilot-publication/functions/load_packages.R

@@ -1,0 +1,20 @@
+library(pbapply)
+library(openssl)
+library(dotenv)
+library(DataExplorer)
+
+load_packages <- function(packages = c(
+                              "dotenv",
+                              "dplyr",
+                              "tibble",
+                              "lubridate",
+                              "gtsummary",
+                              "naniar",
+                              "icdpicr",
+                              "boot",
+                              "pbapply",
+                              "openssl",
+                              "DataExplorer"
+                          )) {
+    for (package in packages) library(package, character.only = TRUE)
+}

### 2024-01-08 pilot-publication/main.R

@@ -1,17 +1,10 @@
 ## Run from command line with R CMD BATCH main.R run.log &
 
-## Load packages
-library(dotenv)
-library(dplyr)
-library(tibble)
-library(lubridate)
-library(gtsummary)
-library(naniar)
-library(icdpicr)
-library(boot)
-
 ## Source functions
 noacsr::source_all_functions()
+
+## Load packages
+load_packages()
 
 ## Import data
 data <- readr::read_csv(Sys.getenv("DATA_DIR"))

### 2024-01-08 pilot-publication/functions/inline_outcome_text.R
Document inline_outcome_text function

@@ -1,3 +1,26 @@
+#' Return inline text for a given outcome and level from a gt table
+#'
+#' This function extracts inline text for a specified outcome and level from a GT table.
+#' It is designed to work with tables that have been created using the `gt` package.
+#' This function assumes that outcome names are styled as bold in the GT table.
+#'
+#' @param gt.table A GT table object (of class `gt_tbl`) from which to extract information.
+#' @param outcome A character string specifying the outcome of interest.
+#' @param level A character string specifying the level within the chosen outcome.
+#' @param column The column from which to extract the data. This can be a character string (column name) or a numeric value (column index).
+#'
+#' @return Returns a vector containing the inline text for the given outcome and level.
+#'
+#' @importFrom magrittr %>%
+#' @importFrom dplyr pull
+#' @importFrom assertthat assert_that
+#'
+#' @examples
+#' # Assuming `my_gt_table` is a GT table with outcomes and levels
+#' # Extract inline text for a specific outcome and level
+#' inline_text <- inline_outcome_text(my_gt_table, "Outcome1", "Level1", "Column1")
+#'
+#' @export
 inline_outcome_text <- function(gt.table, outcome, level, column) {
     ## Define borrowed functions
     `%>%` <- magrittr::`%>%`

### 2024-01-08 pilot-publication/manuscript.Rmd
Start adding outcome results to manuscript

--- 2024-01-03 pilot-publication/manuscript.Rmd
@@
 ```{r, echo = FALSE}
-post.training.outcomes.table <- outcome.results.tables$ci.level.0.95[["Outcomes in all patients after training, by treatment arm"]]
 m30d.atls <- inline_text(post.training.characteristics.table,
@@
 )
+m30d.ptc <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "PTC"
+)
+m30d.standard.care <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "Standard care"
+)
+standard.care.vs.atls.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
+arr.standard.care.atls <- inline_outcome_text(
+  standard.care.vs.atls.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
+standard.care.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
+arr.standard.care.ptc <- inline_outcome_text(
+  standard.care.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
+atls.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
+arr.atls.ptc <- inline_outcome_text(
+  atls.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
 ```
 
-After training, a total of `r m30d.atls` patients in the ATLS
-arm had the primary outcome, compared to `r n.m30d.ptc` (
-`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (
-`r p.m30d.control`%) patients in the standard care arm. The absolute risk
-difference between the ATLS and PTC arm was `r arr.atls.ptc`% units
-<!--(95% CI X - X)-->, between the ATLS and standard care arms 
-`r arr.atls.control`% units <!--(95% CI X - X)-->, and between the PTC
-and standard care arms `r arr.ptc.control`% units <!--(95% CI X
--X)-->. The relative risk between the ATLS and PTC arm was 
+After training, a total of `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
+The absolute risk difference between the standard care and ATLS arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS and PTC arms (95% CI) `r arr.atls.ptc`% units.
+
+The relative risk between the ATLS and PTC arm was 
 `r rr.atls.ptc` <!--(95% CI X - X)-->, between the ATLS and standard care

### 2024-01-09 pilot-publication/manuscript.Rmd
Incorporate patient outcomes in results

--- 2024-01-03 pilot-publication/manuscript.Rmd
@@
 ```{r, echo = FALSE}
-post.training.outcomes.table <- outcome.results.tables$ci.level.0.95[["Outcomes in all patients after training, by treatment arm"]]
+## Crude 30 day mortality
 m30d.atls <- inline_text(post.training.characteristics.table,
@@
 )
+m30d.ptc <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "PTC"
+)
+m30d.standard.care <- inline_text(post.training.characteristics.table,
+  variable = outcomes__alive_after_30_days,
+  level = "Yes",
+  column = "Standard care"
+)
+
+## Extract tables for absolute and relative differences
+standard.care.vs.atls.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
+
+standard.care.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
+
+atls.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
+
+## Absolute differences
+arr.standard.care.atls <- inline_outcome_text(
+  standard.care.vs.atls.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
+
+arr.standard.care.ptc <- inline_outcome_text(
+  standard.care.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
+
+arr.atls.ptc <- inline_outcome_text(
+  atls.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Absolute difference",
+  level = "Yes"
+)
+
+## Relative differences
+rr.standard.care.atls <- inline_outcome_text(
+  standard.care.vs.atls.table,
+  outcome = "30 day mortality",
+  column = "Relative difference",
+  level = "Yes"
+)
+
+rr.standard.care.ptc <- inline_outcome_text(
+  standard.care.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Relative difference",
+  level = "Yes"
+)
+
+rr.atls.ptc <- inline_outcome_text(
+  atls.vs.ptc.table,
+  outcome = "30 day mortality",
+  column = "Relative difference",
+  level = "Yes"
+)
+
+## Extract tables for change from baseline in 30 day mortality
+absolute.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+relative.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
+## Absolute change from baseline
+abs.change.standard.care <- inline_outcome_text(
+  absolute.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "Standard care",
+  level = "Yes"
+)
+
+abs.change.atls <- inline_outcome_text(
+  absolute.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "ATLS",
+  level = "Yes"
+)
+
+abs.change.ptc <- inline_outcome_text(
+  absolute.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "PTC",
+  level = "Yes"
+)
+
+## Relative change from baseline
+rel.change.standard.care <- inline_outcome_text(
+  relative.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "Standard care",
+  level = "Yes"
+)
+
+rel.change.atls <- inline_outcome_text(
+  relative.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "ATLS",
+  level = "Yes"
+)
+
+rel.change.ptc <- inline_outcome_text(
+  relative.change.from.baseline.table,
+  outcome = "30 day mortality",
+  column = "PTC",
+  level = "Yes"
+)
 ```
 
-After training, a total of `r m30d.atls` patients in the ATLS
-arm had the primary outcome, compared to `r n.m30d.ptc` (
-`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (
-`r p.m30d.control`%) patients in the standard care arm. The absolute risk
-difference between the ATLS and PTC arm was `r arr.atls.ptc`% units
-<!--(95% CI X - X)-->, between the ATLS and standard care arms 
-`r arr.atls.control`% units <!--(95% CI X - X)-->, and between the PTC
-and standard care arms `r arr.ptc.control`% units <!--(95% CI X
--X)-->. The relative risk between the ATLS and PTC arm was 
-`r rr.atls.ptc` <!--(95% CI X - X)-->, between the ATLS and standard care
-arms `r rr.atls.control` <!--(95% CI X - X)-->, and between the PTC
-and standard care arms `r rr.ptc.control` <!--(95% CI X - X)-->.
+After training, a total of `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
+The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units.
+The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.
 
-<!--
-The change from baseline for the primary outcome was X in the ATLS
-arm, X in the PTC arm, and X in the standard care arm. The differences
-between changes from baseline were X (95% CI X - X) between the ATLS
-and PTC arms, X (95% CI X - X) between the ATLS and standard care
-arms, and X (95% CI X - X) between the PTC and standard care
-arms. Table X shows differences in the primary outcome between trial
-arms with 75, 85, and 95% CIs. 
--->
-
-Differences in secondary outcomes between trial arms<!-- and subgroups--> are
-available as Supplementary material table \@ref(tab:secondary-outcomes).
+The absolute risk difference between the standard care and ATLS arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS and PTC arms (95% CI) `r arr.atls.ptc`% units.
+The relative risk in the standard care arm compared with the ATLS arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
+The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all other outcomes.
 
@@
 A total of `r n.residents` residents were trained during this
-pilot. Overall their median confidence in managing trauma patients was
-`r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert
-scale with 10 being most confident. <!--In the post training period, the
-confidence in the ATLS arm was X (IQR X - X), compared to X (IQR X -
-X) in the PTC arm, and X (IQR X - X) in the standard care arm.-->
+pilot.
+Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident.
+<!--In the post training period, the confidence in the ATLS arm was X (IQR X - X), compared to X (IQR X - X) in the PTC arm, and X (IQR X - X) in the standard care arm.-->
 
@@
 <!--Out of X potentially eligible patients, X (X%) patients were enrolled
-in the study-->. Out of `r n.eligible.residents` potentially eligible
-residents, `r n.residents` (`r recruitment.rate.residents`%) were
-enrolled in the study. The lost to follow up rate was 
-`r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of 
-`r n.patients` patients not completing 30 day follow up. Among residents
-the pass rate, after two attempts, was `r rate.pass`%. The missing
-data rate ranged from `r min.missing.data` to `r max.missing.data`,
-with details per variable and outcome in Table 
-\ref@(tab:sample-characteristics) and Supplementary material table
-\@ref(tab:secondary-outcomes). <!--The differences in
-distributions between observed and extracted data, for variables that
-were collected through observation or interview, are shown in Table X.-->
+in the study-->.
+Out of `r n.eligible.residents` potentially eligible residents, `r n.residents` (`r recruitment.rate.residents`%) participated in the study.
+The lost to follow up rate was `r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of `r n.patients` patients not completing 30 day follow up.
+Among residents the pass rate, after two attempts, was `r rate.pass`%.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details per variable and outcome in Table  \ref@(tab:sample-characteristics) and Supplementary material table
+\@ref(tab:secondary-outcomes). <!--The differences in distributions between observed and extracted data, for variables that were collected through observation or interview, are shown in Table X.-->
 

### 2024-01-11 pilot-publication/lab-book.Rmd
Start comparing directly observed and retrospective data

@@
 load_packages()
+```
+
+## Compare directly observed with retrospective data
+
+```{r}
+## Import data
+data <- readr::read_csv(Sys.getenv("DATA_DIR"))
+
+## Import codebook
+codebook.arguments <- lapply(
+    c("URL", "UID", "USERNAME", "PASSWORD"),
+    function(x) Sys.getenv(paste0("KOBO_", x))
+)
+codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
+
+## Prepare data
+data <- prepare_data(data, codebook)
+
+## Get variables to compare
+retrospective.variables <- names(data)[grep("_r__", names(data), fixed = TRUE)]
+directly.observed.variables <- retrospective.variables %>%
+    stringr::str_replace_all(pattern = "\\_r\\_\\_", replacement = "\\_\\_") %>%
+    stringr::str_replace_all(pattern = "\\_r$", replacement = "")
+
+## Keep only observations with any retrospective data
+data <- data[apply(data[, retrospective.variables], 1, function(row) any(!is.na(row))), ] %>%
+    as_tibble()
+
+## Extract retrospective and directly observed data
+retrospective.data <- data[, retrospective.variables]
+colnames(retrospective.data) <- directly.observed.variables
+directly.observed.data <- data[, directly.observed.variables]
+retrospective.data$collection.mode <- "Retrospective"
+directly.observed.data$collection.mode <- "Directly observed"
+
+## Make sure column classes are the same in both datasets
+retrospective.data[] <- lapply(names(retrospective.data), function(column.name) {
+    retrospective.column <- retrospective.data[[column.name]]
+    directly.observed.column <- directly.observed.data[[column.name]]
+    retrospective.class <- class(retrospective.column)
+    directly.observed.class <- class(directly.observed.column)
+    class.function <- paste0("as.", directly.observed.class)
+    if ("POSIXct" %in% directly.observed.class) {
+        class.function <- "as.POSIXct"
+    }
+    if (retrospective.class != directly.observed.class) {
+        retrospective.column <- do.call(class.function, list(retrospective.column))
+    }
+    return(retrospective.column)
+})
+
+## Merge data
+merged.data <- bind_rows(retrospective.data, directly.observed.data)
 ```

### 2024-01-12 pilot-publication/lab-book.Rmd
More work on comparing directly observed and retrospective data

--- 2024-01-08 pilot-publication/lab-book.Rmd
@@
 load_packages()
+```
+
+## Compare directly observed with retrospective data
+
+```{r}
+## Import data
+data <- readr::read_csv(Sys.getenv("DATA_DIR"))
+
+## Import codebook
+codebook.arguments <- lapply(
+    c("URL", "UID", "USERNAME", "PASSWORD"),
+    function(x) Sys.getenv(paste0("KOBO_", x))
+)
+codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
+
+## Prepare data
+data <- prepare_data(data, codebook)
+
+## Get variables to compare
+retrospective.variables <- names(data)[grep("_r__", names(data), fixed = TRUE)]
+directly.observed.variables <- retrospective.variables %>%
+    stringr::str_replace_all(pattern = "\\_r\\_\\_", replacement = "\\_\\_") %>%
+    stringr::str_replace_all(pattern = "\\_r$", replacement = "")
+
+## Keep only observations with any retrospective data
+data <- data[apply(data[, retrospective.variables], 1, function(row) any(!is.na(row))), ] %>%
+    as_tibble()
+
+## Extract retrospective and directly observed data
+retrospective.data <- data[, retrospective.variables]
+colnames(retrospective.data) <- directly.observed.variables
+directly.observed.data <- data[, directly.observed.variables]
+retrospective.data$collection.mode <- "Retrospective"
+directly.observed.data$collection.mode <- "Directly observed"
+original.classes <- sapply(directly.observed.data, class)
+
+## Merge datasets
+directly.observed.data[] <- lapply(directly.observed.data, as.character)
+retrospective.data[] <- lapply(retrospective.data, as.character)
+merged.data <- bind_rows(directly.observed.data, retrospective.data)
+
+## Change column classes to match original classes
+merged.data[] <- mapply(function(column, class.name) {
+    class.function <- paste0("as.", class.name[1])
+    new.column <- match.fun(class.function)(column)
+    return(new.column)
+}, merged.data, original.classes, SIMPLIFY = FALSE)
+})
+
+## Create table 
+table.variables <- c(
+    "patinfo__pt_age", "patinfo__pt_gender",
+    "incident__dominating_injury_type",
+    "patvitals__ed_rr", "patvitals__ed_sat",
+    "patvitals__ed_hr", "patvitals__ed_sbp","outcomes__alive_after_30_days", 
+    "collection.mode"
+)
+table.data <- merged.data[, table.variables]
+comparison.table <- create_descriptive_table(table.data,
+    strata = "collection.mode")
 ```

### 2024-01-15 pilot-publication/lab-book.Rmd
Add function that compares directly observed and retrospective data

@@
 
-## Keep only observations with any retrospective data
-data <- data[apply(data[, retrospective.variables], 1, function(row) any(!is.na(row))), ] %>%
-    as_tibble()
+## Keep only observations with completed retrospective data collection
+data <- data %>% filter(data_status_r__collection_completed_r == "Yes")
 
@@
 }, merged.data, original.classes, SIMPLIFY = FALSE)
-})
+merged.data <- merged.data %>%
+    labelled::copy_labels_from(data)
 
@@
     "patvitals__ed_rr", "patvitals__ed_sat",
-    "patvitals__ed_hr", "patvitals__ed_sbp","outcomes__alive_after_30_days", 
+    "patvitals__ed_hr", "patvitals__ed_sbp",
     "collection.mode"
@@
 comparison.table <- create_descriptive_table(table.data,
-    strata = "collection.mode")
+    strata = "collection.mode",
+    include.overall = FALSE
+)
+comparison.table
 ```

### 2024-01-15 pilot-publication/functions/create_retrospective_comparison_table.R

@@ -1,0 +1,62 @@
+#' Create a retrospective comparison table
+#'
+#' This function takes in a data frame and creates a comparison table
+#' between retrospective and directly observed data.
+#' @param data A data frame containing the data for comparison.
+#' @return A comparison table with descriptive statistics.
+#' @export
+create_retrospective_comparison_table <- function(data) {
+    ## Define borrowed functions
+    `%>%` <- magrittr::`%>%`
+
+    ## Check arguments
+    assertthat::assert_that(is.data.frame(data))
+
+    ## Get variables to compare
+    retrospective.variables <- names(data)[grep("_r__", names(data), fixed = TRUE)]
+    directly.observed.variables <- retrospective.variables %>%
+        stringr::str_replace_all(pattern = "\\_r\\_\\_", replacement = "\\_\\_") %>%
+        stringr::str_replace_all(pattern = "\\_r$", replacement = "")
+
+    ## Keep only observations with completed retrospective data collection
+    data <- data %>% dplyr::filter(data_status_r__collection_completed_r == "Yes")
+
+    ## Extract retrospective and directly observed data
+    retrospective.data <- data[, retrospective.variables]
+    colnames(retrospective.data) <- directly.observed.variables
+    directly.observed.data <- data[, directly.observed.variables]
+    retrospective.data$collection.mode <- "Retrospective"
+    directly.observed.data$collection.mode <- "Directly observed"
+    original.classes <- sapply(directly.observed.data, class)
+
+    ## Merge datasets
+    directly.observed.data[] <- lapply(directly.observed.data, as.character)
+    retrospective.data[] <- lapply(retrospective.data, as.character)
+    merged.data <- dplyr::bind_rows(directly.observed.data, retrospective.data)
+
+    ## Change column classes to match original classes
+    merged.data[] <- mapply(function(column, class.name) {
+        class.function <- paste0("as.", class.name[1])
+        new.column <- match.fun(class.function)(column)
+        return(new.column)
+    }, merged.data, original.classes, SIMPLIFY = FALSE)
+    merged.data <- merged.data %>%
+        labelled::copy_labels_from(data)
+
+    ## Create table
+    table.variables <- c(
+        "patinfo__pt_age", "patinfo__pt_gender",
+        "incident__dominating_injury_type",
+        "patvitals__ed_rr", "patvitals__ed_sat",
+        "patvitals__ed_hr", "patvitals__ed_sbp",
+        "collection.mode"
+    )
[…]

### 2024-01-16 pilot-publication/manuscript.Rmd
Remove new lines and add comparison of observed and extracted data

@@
 ## Role of study sponsor and funders
-The funding sources had no role in the design of this study nor during
-its execution, analyses, interpretation of the data, or decision to
-submit the results.
+The funding sources had no role in the design of this study nor during its execution, analyses, interpretation of the data, or decision to submit the results.
 
@@
 
-Trauma, defined as the clinical entity composed of physical injury and
-the body's associated response, causes 4.3 millions deaths every year
-[@injuries2020]. Almost 10% of the global burden of disease is due
-to trauma and trauma is the top contributor to the burden of disease in
-children and adults aged 10 to 49 years [@Vos2020].
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year [@injuries2020].
+Almost 10% of the global burden of disease is due to trauma and trauma is the top contributor to the burden of disease in children and adults aged 10 to 49 years [@Vos2020].
 
-Trauma care is time sensitive and early management of life or limb
-threatening conditions is crucial. Several trauma life support training
-programs have been developed to improve the early management of patients
-as they arrive at hospital by providing a structured framework to
-assessment and treatment [@Mohammad2013; @Jayaraman2014;
-@Kadhum2020].
+Trauma care is time sensitive and early management of life or limb threatening conditions is crucial.
+Several trauma life support training programs have been developed to improve the early management of patients as they arrive at hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) is the most
-established trauma life support training program and more than one
-million doctors in over 80 countries have been trained in the programme
-[@acsAtls2018]. Uptake in low- and middle income countries (LMIC)
-has been slow, potentially due to high costs [@Kadhum2020].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) is the most established trauma life support training program and more than one million doctors in over 80 countries have been trained in the programme [@acsAtls2018].
+Uptake in low- and middle income countries (LMIC) has been slow, potentially due to high costs [@Kadhum2020].
 
-The free Primary Trauma Care (PTC) program is the most widely spread
-alternative programme. The goal of PTC is to improve trauma care in LMIC
-[@ptcfPtc2018]. Like ATLS^®^, doctors in over 80 countries have been
-trained in PTC, and the programme has been endorsed by the World Health
-Organization (WHO), among other international organizations
-including several professional societies [@ptcfPtc2018].
+The free Primary Trauma Care (PTC) program is the most widely spread alternative programme.
+The goal of PTC is to improve trauma care in LMIC [@ptcfPtc2018].
+Like ATLS^®^, doctors in over 80 countries have been trained in PTC, and the programme has been endorsed by the World Health Organization (WHO), among other international organizations including several professional societies [@ptcfPtc2018].
 
-There are observational studies indicating that trauma life support
-training programmes may improve patient outcomes [@Jin2021], but there
-are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. 
-We performed a pilot study aiming to assess the feasibility of
-conducting a cluster randomised controlled trial comparing ATLS and
-PTC with standard care.  The objectives of this pilot study were
-to:
+There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021], but there are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS and PTC with standard care.
+The objectives of this pilot study were to:
 
-- Estimate probable effect sizes on patient outcomes associated with
-  ATLS^®^ and PTC compared with standard care, estimate the proportion of
-  participants with the outcome (if binary), and estimate the
-  intracluster correlation coefficient, as a basis for future sample
-  size calculations.
+- Estimate probable effect sizes on patient outcomes associated with ATLS^®^ and PTC compared with standard care, estimate the proportion of participants with the outcome (if binary), and estimate the intracluster correlation coefficient, as a basis for future sample size calculations.
 
-- Assess the feasibility of recruiting participants and collecting
-  data on primary and secondary outcomes, such as mortality,
-  in-hospital complications, length of stay, and quality of life.
+- Assess the feasibility of recruiting participants and collecting data on primary and secondary outcomes, such as mortality, in-hospital complications, length of stay, and quality of life.
 
-- Assess how the effect sizes and directions of these effects of ATLS
-  and PTC may differ across clinically important subgroups.
+- Assess how the effect sizes and directions of these effects of ATLS and PTC may differ across clinically important subgroups.
 
@@
 ## Trial Design
-We piloted a three-armed cluster randomised controlled trial. There
-were two intervention arms, ATLS^®^ and PTC training, and one control
-arm, standard care. We collected data for four months in all three
-arms, first during a one month observation phase and then during a
-three month intervention phase (or continued observation in the
-control arm). This design allowed us to assess outcomes both as
-final values and as change from baseline.
+We piloted a three-armed cluster randomised controlled trial.
+There were two intervention arms, ATLS^®^ and PTC training, and one control arm, standard care.
+We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the control arm).
+This design allowed us to assess outcomes both as final values and as change from baseline.
 
-Our study is a pilot study because its objectives involves estimating
-quantities, such as the probable effect sizes, proportion of
-participants with the outcome (if binary), and the intracluster
-correlation coefficient, needed for the sample size calculations of a
-full-scale trial [@Hemming2017]. This pilot study also establish how
-many participants that can be enrolled, as well as likely drop out
-rates, and the feasibility of collecting primary and secondary
-outcomes.
+Our study is a pilot study because its objectives involves estimating quantities, such as the probable effect sizes, proportion of participants with the outcome (if binary), and the intracluster correlation coefficient, needed for the sample size calculations of a full-scale trial [@Hemming2017].
+This pilot study also establish how many participants that can be enrolled, as well as likely drop out rates, and the feasibility of collecting primary and secondary outcomes.
 
 ## Study Setting
-We conducted this pilot in seven Indian tertiary hospitals, where
-neither ATLS^®^, PTC, nor any other trauma life support training program
-is routinely taught. The seven hospitals were Seth G. S. Medical
-College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
-College and General Hospital, Mumbai; HBT Medical College And Dr. R N
-Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
-Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
-Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
-Medical Education & Research, Chandigarh.
+We conducted this pilot in seven Indian tertiary hospitals, where neither ATLS^®^, PTC, nor any other trauma life support training program is routinely taught.
+The seven hospitals were Seth G.
+S.
+Medical College and K.E.M.
+Hospital, Mumbai; Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai; HBT Medical College And Dr.
+R N Cooper Municipal General Hospital, Mumbai; Medica Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of Medical Education & Research, Chandigarh.
 
@@
 ### Patient Participants
-Adults (15 years or older) who presented to the emergency department at
-participating hospitals with a history of trauma. History of trauma is
-here defined as having any of the external causes of morbidity and
-mortality listed in block V01-Y36, chapter XX of the International
-Classification of Disease version 10 (ICD-10) codebook as reason for
-presenting. We explored intervention effects across the following
-clinical subgroups: men, women, blunt multisystem trauma, penetrating
-trauma, shock, severe traumatic brain injury, and elderly, as defined
-by Hornor et al [@Hornor2018]. The consent form for patients are
-available as Supplemental Material 1.
+Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma.
+History of trauma is here defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
+We explored intervention effects across the following clinical subgroups: men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly, as defined by Hornor et al [@Hornor2018].
+The consent form for patients are available as Supplemental Material 1.
 
@@
 emergency medicine managing trauma patients in the emergency
-department and who were expected to remain in the participating
-hospitals for at least one year. To facilitate administration each
-surgical department is divided into units, which manage the out-patient 
-and emergency departments, operating rooms etc on
-different days each week. One or two, out of typically six, units'
-residents were selected from each hospital. One unit consists of at
-least three faculty and three to twelve residents.
+department and who were expected to remain in the participating hospitals for at least one year.
+To facilitate administration each surgical department is divided into units, which manage the out-patient and emergency departments, operating rooms etc on different days each week.
+One or two, out of typically six, units' residents were selected from each hospital.
+One unit consists of at least three faculty and three to twelve residents.
 
-To be eligible, units had to have a maximum of 25% of the doctors
-trained in either ATLS^®^, PTC, or similar training programs before the
-start of the pilot. Those residents who had received training in the
-last five years were considered as trained. The figure of 25% was
-decided through consensus in the research team, to balance feasibility
-and contamination of results.
+To be eligible, units had to have a maximum of 25% of the doctors trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot.
+Those residents who had received training in the last five years were considered as trained.
+The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
 
-Consent was sought from the residents in each of the intervention
-groups before they underwent the ATLS^®^ or PTC training. The consent
-form for residents is available as Supplemental Material 2. We did
-not ask for consent from residents at the units in the control
-hospitals as their practice was not affected by this pilot and we
-did not collect any personal identifiable data on them.
+Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
+The consent form for residents is available as Supplemental Material 2.
+We did not ask for consent from residents at the units in the control hospitals as their practice was not affected by this pilot and we did not collect any personal identifiable data on them.
 
 ### Clusters
-Indian tertiary care hospitals that admit more than 400 adult patients
-with trauma each year. We randomised on the cluster (hospital) level
-to avoid contamination between intervention and control arms. To be
-eligible for inclusion hospitals had to provide the following services
-round the clock: operation theatres, X-ray, CT, and ultrasound
-facilities, and blood bank. In addition the baseline admission rate
-had to be more than 35 adult patients with major trauma per month.
+Indian tertiary care hospitals that admit more than 400 adult patients with trauma each year.
+We randomised on the cluster (hospital) level to avoid contamination between intervention and control arms.
+To be eligible for inclusion hospitals had to provide the following services round the clock: operation theatres, X-ray, CT, and ultrasound facilities, and blood bank.
+In addition the baseline admission rate had to be more than 35 adult patients with major trauma per month.
 
 ## Interventions
-In each intervention arm one or two units', out of typically six,
-residents per hospital providing emergency care to trauma patients
-were trained in either ATLS^®^ or PTC. For the purpose of this pilot
-study, our target was to train a minimum of 75% of residents in each
-unit. If residents dropped out or changed units after training but
-before data collection was completed we planned to conduct additional
-training if needed to meet the 75% criterion, but this was not
-required. We did not train the units' faculty, as they are typically
-not directly involved in the initial management of trauma patients.
+In each intervention arm one or two units', out of typically six, residents per hospital providing emergency care to trauma patients were trained in either ATLS^®^ or PTC.
+For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
+If residents dropped out or changed units after training but before data collection was completed we planned to conduct additional training if needed to meet the 75% criterion, but this was not required.
+We did not train the units' faculty, as they are typically not directly involved in the initial management of trauma patients.
 
-The ATLS^®^ training was conducted in an ATLS certified training centre
-in Mumbai, according to the standard ATLS^®^ curriculum
-[@acsAtls2018]. The PTC training was conducted in New Delhi, according
-to the standard PTC curriculum [@ptcfPtc2018]. These courses were
-conducted over a period of 2.5 to 3 days. The residents certified
-"pass" were considered as trained in respective courses.
+The ATLS^®^ training was conducted in an ATLS certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018].
+The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018].
+These courses were conducted over a period of 2.5 to 3 days.
+The residents certified "pass" were considered as trained in respective courses.
 
@@
 ### Modifications
-Both ATLS^®^ and PTC are standard training programs with fixed curricula
-[@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
-content of these programs during this pilot.
+Both ATLS^®^ and PTC are standard training programs with fixed curricula [@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or content of these programs during this pilot.
 
 ### Adherence
-The intervention was the training in either ATLS^®^ or PTC and resident
-participants were required to adhere to, i.e. participate in, the
-training, to be eligible for passing. We did not consider adherence to
-training contents during care delivery as adherence to the trial
-intervention.
+The intervention was the training in either ATLS^®^ or PTC and resident participants were required to adhere to, i.e.
+participate in, the training, to be eligible for passing.
+We did not consider adherence to training contents during care delivery as adherence to the trial intervention.
 
@@
 
-The care provided by all participating hospitals at baseline was based on
-the training curriculum formulated by The National Medical Council of
-India for post graduation in General Surgery
-[@mciMsSurgery2019]. Regarding trauma, these guidelines state
-that the student should:
+The care provided by all participating hospitals at baseline was based on the training curriculum formulated by The National Medical Council of India for post graduation in General Surgery [@mciMsSurgery2019].
+Regarding trauma, these guidelines state that the student should:
 
-a.  Have knowledge about response to trauma; burns: causes, prevention
-    and management; wounds of scalp and its management; recognition,
-    diagnosis and monitoring of patients with head injury, Glasgow coma
-    scale.
-b.  Be able to provide and coordinate emergency resuscitative measures
-    in acute surgical situations including trauma.
-c.  Choose, perform and interpret appropriate imaging in trauma -
-    ultrasound Focused Abdominal Sonography in Trauma (FAST).
-d.  Undergo advanced trauma and cardiac life support course (certified)
-    before appearing in final examination.
+a.
+Have knowledge about response to trauma; burns: causes, prevention and management; wounds of scalp and its management; recognition, diagnosis and monitoring of patients with head injury, Glasgow coma scale.
+b.  Be able to provide and coordinate emergency resuscitative measures in acute surgical situations including trauma.
+c.  Choose, perform and interpret appropriate imaging in trauma - ultrasound Focused Abdominal Sonography in Trauma (FAST).
+d.  Undergo advanced trauma and cardiac life support course (certified) before appearing in final examination.
 e.  Undergo clinical posting in emergency and trauma.
@@
 
-Although training in an ATLS^®^ is part of the curriculum it is 
-optional and not doing this training does not
+Although training in an ATLS^®^ is part of the curriculum it is  optional and not doing this training does not
 result in failure to obtain post graduation completion.
@@
 
-**This section needs to be revised to include information from sites
-where patients were not managed by surgical residents**
+**This section needs to be revised to include information from sites where patients were not managed by surgical residents**
 
-At most tertiary hospitals in India trauma patients present to the
-emergency department where they are assessed by a doctor and referred
-to the surgical bay for further management. In the surgical bay a
-second or third year general surgery resident sees all the major
-trauma and provide the initial care, including initiating treatment
-and investigations.  This resident informs the consultant on call who
-is generally an Assistant Professor. Most procedures like intercostal
-drainage, open wound suturing, intubation etc. would be done in the
-surgical resuscitation area, by the surgical resident.
+At most tertiary hospitals in India trauma patients present to the emergency department where they are assessed by a doctor and referred to the surgical bay for further management.
+In the surgical bay a second or third year general surgery resident sees all the major trauma and provide the initial care, including initiating treatment and investigations.
+This resident informs the consultant on call who is generally an Assistant Professor.
+Most procedures like intercostal drainage, open wound suturing, intubation etc.
+would be done in the surgical resuscitation area, by the surgical resident.
 
-Compared to other settings where a trauma team approach is adopted,
-nurses and other healthcare professionals are involved to a limited
-extent during the initial management. Their roles include assisting
-during intubation and other bedside procedures, charting the vitals
-(not recording) and giving injections. They also accompany the
-resident during transfers of serious patients.
+Compared to other settings where a trauma team approach is adopted, nurses and other healthcare professionals are involved to a limited extent during the initial management.
+Their roles include assisting during intubation and other bedside procedures, charting the vitals (not recording) and giving injections.
+They also accompany the resident during transfers of serious patients.
 
-After completing the assessment and starting initial resuscitation, the
-resident decides to send the patient for imaging (X-rays/FAST/CT-scan)
-or to the operation room in consultation with or after assessment by the
-on-call consultant. A portable X-ray and an ultrasonography machine to
-conduct FAST may or may not be available in the surgical bay. The
-patients who are operated, managed conservatively, not intubated, or
-with minor trauma will be sent to the surgical ward. Those who need
-increased monitoring or mechanical ventilation remain in the surgical
-bay or in the intensive care unit (ICU) depending on the availability of
-ICU beds. The further treatment continues in the respective ward or ICU
-and patients are finally discharged from the ward.
+After completing the assessment and starting initial resuscitation, the resident decides to send the patient for imaging (X-rays/FAST/CT-scan) or to the operation room in consultation with or after assessment by the on-call consultant.
+A portable X-ray and an ultrasonography machine to conduct FAST may or may not be available in the surgical bay.
+The patients who are operated, managed conservatively, not intubated, or with minor trauma will be sent to the surgical ward.
+Those who need increased monitoring or mechanical ventilation remain in the surgical bay or in the intensive care unit (ICU) depending on the availability of ICU beds.
+The further treatment continues in the respective ward or ICU and patients are finally discharged from the ward.
 
@@
 Our pilot study included both participant and feasibility
-outcomes. Prior to deciding on these participant outcomes we searched
-the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's
-database but were unable to identify appropriate core outcome sets for
-our populations of participants.
+outcomes.
+Prior to deciding on these participant outcomes we searched the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's database but were unable to identify appropriate core outcome sets for our populations of participants.
 
-The primary participant outcome was all cause mortality within 30 days
-from the time of arrival to the emergency department. The primary
-outcome and most secondary outcome were assessed and compared both as
-final values and as change from baseline. All outcomes that pertain to
-the individual participant level are detailed in Supplemental Material
-3. We decided to include a large number of outcomes,
-including some more exploratory, so that we could test their feasibility
-and relevance.
+The primary participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department.
+The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline.
+All outcomes that pertain to the individual participant level are detailed in Supplemental Material 3.
+We decided to include a large number of outcomes,
+including some more exploratory, so that we could test their feasibility and relevance.
 
-We also assessed the following feasibility outcomes, which pertain
-both to overall study population as well as to the individual cluster
-level:
+We also assessed the following feasibility outcomes, which pertained both to overall study population as well as to the individual cluster level:
 
-- Recruitment rate. For both patients and residents this was equal to
-  the proportion of participants enrolled, out of the total number of
-  eligible participants, over the course of the pilot study.
-- Lost to follow up rate. This applied only to patients and was equal
-  to the proportion of patients that did not complete 30 day follow up,
-  out of all enrolled patients, over the course of the pilot study.
-- Pass rate. This applied only to residents in the intervention
-  arms and equal the proportion of residents that pass the training
-  programme, out of the total number of trained residents, over the
-  course of the pilot study.
-- Missing data rate. This applied to each outcome and variable and was
-  equal to the proportion of missing data, over the course of the
-  pilot study.
-- Differences in distributions of observed and extracted data. This
-  applied to each outcome and variable and compared the distributions
-  of data collected by observations versus extracted from hospital
-  records. For quantitative variables this was be the difference in
-  means, standard deviations, medians, interquartile ranges, and
-  ranges. For qualitative variables this was the differences in
-  absolute counts and percentages, across categories.
+- Recruitment rate. For both patients and residents this was equal to the proportion of participants enrolled, out of the total number of eligible participants, over the course of the pilot study.
+- Lost to follow up rate.
+This applied only to patients and was equal to the proportion of patients that did not complete 30 day follow up, out of all enrolled patients, over the course of the pilot study.
+- Pass rate.
+This applied only to residents in the intervention arms and equal the proportion of residents that pass the training programme, out of the total number of trained residents, over the course of the pilot study.
+- Missing data rate. This applied to each outcome and variable and was equal to the proportion of missing data, over the course of the pilot study.
+- Differences in distributions of observed and extracted data.
+This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
+For quantitative variables this was be the difference in means, standard deviations, medians, interquartile ranges, and ranges.
+For qualitative variables this was the differences in absolute counts and percentages, across categories.
 
@@
 ### Patients
-Patients were screened for eligibility as they arrived at the emergency
-department. Eligible patients were approached in the emergency
-department to consent to follow up, if they were conscious. If they
-were unconscious a patient representative was approached to consent to
-follow up. Once the patient was conscious we approached the patient to
-affirm the patient representative's consent. We followed up patients
-at discharge, at 24 hours after arrival at the emergency department,
-and at 30 days after arrival at the emergency department.
+Patients were screened for eligibility as they arrived at the emergency department.
+Eligible patients were approached in the emergency department to consent to follow up, if they were conscious.
+If they were unconscious a patient representative was approached to consent to follow up.
+Once the patient was conscious we approached the patient to affirm the patient representative's consent.
+We followed up patients at discharge, at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
 
 ### Residents
-Participating units were screened for eligibility once hospitals
-confirmed their participation. All residents in eligible units were
-approached to consent to training if their hospital was randomised to
-either of the intervention arms. Training was conducted approximately
-one month after the study started.
+Participating units were screened for eligibility once hospitals confirmed their participation.
+All residents in eligible units were approached to consent to training if their hospital was randomised to either of the intervention arms.
+Training was conducted approximately one month after the study started.
 
 ## Sample size
-Given budget and time constraints, including the rotation of units in
-the emergency departments in Indian hospitals (which often happen on a
-six months basis) the feasible data collection period was four
-months. We assumed that each of the units saw 2-4 trauma patients per
-week and based on that estimated that if we selected a minimum of one
-unit per hospital then each hospital would enrol 8-16 patients per
-month and 32-64 patients during the four months of this pilot. With a
-20% attrition rate we expected each hospital to enrol 26-51 patients,
-coming to a total sample size of between 156 and 306 patients for this
-pilot study.
+Given budget and time constraints, including the rotation of units in the emergency departments in Indian hospitals (which often happen on a six months basis) the feasible data collection period was four months.
+We assumed that each of the units saw 2-4 trauma patients per week and based on that estimated that if we selected a minimum of one unit per hospital then each hospital would enrol 8-16 patients per month and 32-64 patients during the four months of this pilot.
+With a 20% attrition rate we expected each hospital to enrol 26-51 patients, coming to a total sample size of between 156 and 306 patients for this pilot study.
 
 ## Recruitment
-To ensure adequate recruitment we only approached hospitals that
-stated that the their volumes were high enough to allow us to reach
-the sample size goals detailed above. Patients were enrolled by a
-dedicated project officer as they arrived at the emergency
-department. The recruitment period was four months. Recruitment was
-monitored weekly through online conferences. No financial or
-non-financial incentives were provided to trial investigators or
-participants for enrolment.
+To ensure adequate recruitment we only approached hospitals that stated that the their volumes were high enough to allow us to reach the sample size goals detailed above.
+Patients were enrolled by a dedicated project officer as they arrived at the emergency department.
+The recruitment period was four months.
+Recruitment was monitored weekly through online conferences.
+No financial or non-financial incentives were provided to trial investigators or participants for enrolment.
 
@@
 ### Sequence generation
-We used simple randomisation to allocate sites to trial arms. We
-prepared seven sealed envelopes from which one representative from
-each pilot site selected one in a common meeting on March 7, 2022
-[^dev1]<!-- Meeting notes ../meetings/20220307/ -->. The content of
-the envelopes dictated what trial arm (ATLS, PTC, or standard care)
-each hospital was allocated to. There were two hospitals in each of
-the intervention arms and three hospitals in the control arm.
+We used simple randomisation to allocate sites to trial arms.
+We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting on March 7, 2022 [^dev1]<!-- Meeting notes ../meetings/20220307/ -->.
+The content of the envelopes dictated what trial arm (ATLS, PTC, or standard care) each hospital was allocated to.
+There were two hospitals in each of the intervention arms and three hospitals in the control arm.
 
-[^dev1]: Initially the plan was to enroll six hospitals (clusters) but
-    we ended up enrolling seven.
+[^dev1]: Initially the plan was to enroll six hospitals (clusters) but we ended up enrolling seven.
 
@@
 ### Implementation
-The random allocation sequence was generated by MGW. Clusters were
-enrolled by the project's core team. Patient participants were
-enrolled if they presented during the project officers shift. Resident
-participants were enrolled if they were in the units selected for
-training. The principal investigator at each hospital selected the
-units for training. For patient participants consent for follow up was
-sought after randomisation from patients or patient relatives as
-appropriate. For resident participants consent was also sought after
-randomisation.
+The random allocation sequence was generated by MGW.
+Clusters were enrolled by the project's core team.
+Patient participants were enrolled if they presented during the project officers shift.
+Resident participants were enrolled if they were in the units selected for training.
+The principal investigator at each hospital selected the units for training.
+For patient participants consent for follow up was sought after randomisation from patients or patient relatives as appropriate.
+For resident participants consent was also sought after randomisation.
 
@@
 It was not possible to blind investigators or participants to
-interventions. We did not blind the data analysts during this pilot,
-but we plan to blind the data analysts during the full-scale trial.
+interventions. We did not blind the data analysts during this pilot, but we plan to blind the data analysts during the full-scale trial.
 
 ## Data Collection
-Data collection started one month before the training was delivered,
-to establish a baseline. Each participating hospital had a dedicated
-project officer who collected the data.
+Data collection started one month before the training was delivered, to establish a baseline. Each participating hospital had a dedicated project officer who collected the data.
 
-Because participating residents were assigned designated days for
-trauma care for a period of 6 months, data were collected during those
-particular days and shifts when these trained doctors were in the
-emergency department. The project officers collected data both by
-observing the care delivered and by interviewing the participants, and
-by extracting data from hospital records.
+Because participating residents were assigned designated days for trauma care for a period of 6 months, data were collected during those particular days and shifts when these trained doctors were in the emergency department.
+The project officers collected data both by observing the care delivered and by interviewing the participants, and by extracting data from hospital records.
 
 Data collection continued for three months after training. The
-research officers collected data of all trauma patients who presented
-to the participating units during their duty hours. Those patients who
-were admitted were followed up for complications and other in-hospital
-outcome measures, for example length of stay. Patients who were not
-admitted were followed up telephonically for mortality outcomes and
-quality of life outcomes. The follow up period was be 30 days. The
-project officers made at most three attempts to reach a participant or
-participant representative telephonically, after which the data were
-recorded as missing.
+research officers collected data of all trauma patients who presented to the participating units during their duty hours.
+Those patients who were admitted were followed up for complications and other in-hospital outcome measures, for example length of stay.
+Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.
+The follow up period was be 30 days.
+The project officers made at most three attempts to reach a participant or participant representative telephonically, after which the data were recorded as missing.
 
-The project officer administered the study information and informed
-consent (consent was sought only for data collection including follow
-up) to the patient, or the patient's representative as appropriate,
-once the patient was stabilised. They will continue to collect data
-once they have received the consent.
+The project officer administered the study information and informed consent (consent was sought only for data collection including follow up) to the patient, or the patient's representative as appropriate, once the patient was stabilised.
+They will continue to collect data once they have received the consent.
 
-Details of data of those patients/relatives not willing to give
-consent were not included in the analysis. The number of patients who
-opt out from data collection was collected.
+Details of data of those patients/relatives not willing to give consent were not included in the analysis. The number of patients who opt out from data collection was collected.
 
 ## Variables
-The project officers collected data on demographics, time of injury to
-arrival at the participating hospital, time to recording vital signs,
-vital signs, and times to and management details including imaging and
-surgery. Details of any injury sustained were collected and coded
-using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD 10 coders
-underwent the WHO online ICD 10 training module and for AIS they were
-accredited. Based on AIS we calculated the Injury Severity Score
-(ISS) and the New ISS (NISS). Supplemental Material 4 contains a full
-variable list, with definitions.
+The project officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, and times to and management details including imaging and surgery.
+Details of any injury sustained were collected and coded using ICD 10 and the Abbreviated Injury Scale (AIS).
+For ICD 10 coders underwent the WHO online ICD 10 training module and for AIS they were accredited.
+Based on AIS we calculated the Injury Severity Score (ISS) and the New ISS (NISS).
+Supplemental Material 4 contains a full variable list, with definitions.
 
@@
 In this study, we conducted community consultations to collect
-inputs from patients, their caregivers, patient groups, and resident
-doctors to be used in the selection of outcome measures and
-implementation of the full-scale trial; following the Guidance for
-Reporting Involvement of Patients and the Public (GRIIPS 2)
-[@Staniszewska2017]. The results of these consultations will be reported 
-separately.
+inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial; following the Guidance for Reporting Involvement of Patients and the Public (GRIIPS 2) [@Staniszewska2017].
+The results of these consultations will be reported in a separate publication.
 
 ## Data management
-We used the online data collection tool KoboToolbox [@KoboToolbox],
-accessible only over a virtual private network (VPN), for each
-participating hospital to upload pseudonymised data to secure
-servers. Data validation techniques such as restricted values or
-values within a specific range were used to avoid ambiguous data
-entries and ensure the validity of the data. Ambiguous responses and
-errors were resolved after discussion with the core team during weekly
-meetings. An instruction manual or codebook for data variables was
-prepared to ensure consistency in data entry. This manual was referred
-to during the project data collection and variable descriptions were
-visible for each variable in the online data collection
-tool. Pseudonymised data were stored at the centralised server. The
-data were only accessible by the project's principal investigator or
-by delegation by the project principal investigator.
+We used the online data collection tool KoboToolbox [@KoboToolbox], accessible only over a virtual private network (VPN), for each participating hospital to upload pseudonymised data to secure servers.
+Data validation techniques such as restricted values or values within a specific range were used to avoid ambiguous data entries and ensure the validity of the data.
+Ambiguous responses and errors were resolved after discussion with the core team during weekly meetings.
+An instruction manual or codebook for data variables was prepared to ensure consistency in data entry.
+This manual was referred to during the project data collection and variable descriptions were visible for each variable in the online data collection tool.
+Pseudonymised data were stored at the centralised server.
+The data were only accessible by the project's principal investigator or by delegation by the project principal investigator.
 
 ## Data monitoring
-Weekly meetings with the core team and project officers took place. We
-conducted an interim analysis on October 12, 2022, and decided to
-complete the study as participants were consenting to be included in
-the study and key variables including mortality outcomes could be
-collected. No outcomes were dropped. We did not use a data monitoring
-committee in the pilot study due to its limited scope.
+Weekly meetings with the core team and project officers took place.
+We conducted an interim analysis on October 12, 2022, and decided to complete the study as participants were consenting to be included in the study and key variables including mortality outcomes could be collected.
+No outcomes were dropped.
+We did not use a data monitoring committee in the pilot study due to its limited scope.
 
 ## Statistical Methods
-We analysed all pilot data using descriptive statistics. Quantitative
-variables are summarised as mean +/- standard deviation, median,
-interquartile range and range. Qualitative variables are presented as
-absolute numbers and percentages. Feasibility outcomes are summarised
-both on the overall sample level as well as on the individual cluster
-level. We used an empty generalised linear mixed model<!-- see
-[@Sommet2017] --> to estimate the intracluster correlation
-coefficient.
+We analysed all pilot data using descriptive statistics.
+Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range.
+Qualitative variables are presented as absolute numbers and percentages.
+Feasibility outcomes are summarised both on the overall sample level as well as on the individual cluster level.
+We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
 
-We compared participant outcomes in three combinations of trial arms:
-ATLS versus PTC, ATLS versus standard care, and PTC versus standard
-care. In each combination we compared both differences in final values
-and differences in change from baseline. For the intervention arms the
-change from baseline was calculated as the difference between the one
-month period of data collection before the training was undertaken and
-the three month period after the training. For the control arm the
-data collection period was four months and the difference from
-baseline was calculated as the difference between the first one
-month and the following three months.
+We compared participant outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care.
+In each combination we compared both differences in final values and differences in change from baseline.
+For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training.
+For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.
 
-Within each combination of trial arms we conducted subgroup analyses
-of men, women, blunt multisystem trauma, penetrating trauma, shock,
-severe traumatic brain injury, and elderly. Table S7.1 in Supplemental
-Material 7 shows which outcomes that were assessed in what subgroups,
-decided through consensus in the research team. We further compared
-the results of all subgroups with the results in the whole cohort, and
-compared the results in the female subgroup with the male subgroup,
-and the results in the blunt multisystem trauma subgroup with the
-penetrating trauma subgroup. We are aware that the numbers in some of
-these subgroups are small, but we included them to help guide the
-formulation of the statistical analysis plan for the full-scale trial.
+Within each combination of trial arms we conducted subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
+Table S7.1 in Supplemental Material 7 shows which outcomes that were assessed in what subgroups, decided through consensus in the research team.
+We further compared the results of all subgroups with the results in the whole cohort, and compared the results in the female subgroup with the male subgroup, and the results in the blunt multisystem trauma subgroup with the penetrating trauma subgroup.
+We are aware that the numbers in some of these subgroups are small, but we included them to help guide the formulation of the statistical analysis plan for the full-scale trial.
 
 We calculated both absolute and relative differences for each
-comparison, along with 75, 85, and 95% confidence intervals. We used
-an empirical bootstrap procedure with 1000 draws to estimate these
-confidence intervals. We did not perform any formal hypothesis tests
-during the analysis of this pilot's data [@Lee2014]. We also compared
-the data collected through observations and interviews with the data
-collected from hospital records, to assess the feasibility of
-collecting data from hospital records in the full-scale trial.
+comparison, along with 75, 85, and 95% confidence intervals.
+We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
+We did not perform any formal hypothesis tests during the analysis of this pilot's data [@Lee2014].
+We also compared the data collected through observations and interviews with the data collected from hospital records, to assess the feasibility of collecting data from hospital records in the full-scale trial.
 
 ## Ethics and Dissemination
-We were granted research ethics approval from each participating
-hospital. The approval numbers were IEC(II)/OUT/134/2022 for King
-Edward Memorial Seth G. S. Medical College and K.E.M. Hospital,
-Mumbai; IEC/214/22 for Lokmanya Tilak Municipal Medical College and
-General Hospital, Mumbai; HBTMC/266/SURGERY for HBT Medical College
-And Dr. R N Cooper Municipal General Hospital, Mumbai;
-CREC/2022/FEB/1(ii) for Medica Superspecialty Hospital, Kolkata;
-Medical College Kolkata, Kolkata; NRSmC/IEC/93/2021 for Sir Nil Ratan
-Sircar Medical College & Hospital, Kolkata; and PGI/IEC/2022/000/003
-for Postgraduate Institute of Medical Education & Research,
-Chandigarh.
+We were granted research ethics approval from each participating hospital.
+The approval numbers were IEC(II)/OUT/134/2022 for King Edward Memorial Seth G.
+S.
+Medical College and K.E.M.
+Hospital, Mumbai; IEC/214/22 for Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai; HBTMC/266/SURGERY for HBT Medical College And Dr.
+R N Cooper Municipal General Hospital, Mumbai; CREC/2022/FEB/1(ii) for Medica Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata; NRSmC/IEC/93/2021 for Sir Nil Ratan
+Sircar Medical College & Hospital, Kolkata; and PGI/IEC/2022/000/003 for Postgraduate Institute of Medical Education & Research, Chandigarh.
 
-The protocol was published in BMJ Open [@GerdinWärnberg2022] and was
-registered with ClinicalTrials.gov (reg. no NCT05417243). Amendments
-to the protocol after publication were determined by the core research
-group and updated on ClinicalTrials.gov.
+The protocol was published in BMJ Open [@GerdinWärnberg2022] and was registered with ClinicalTrials.gov (reg.
+no NCT05417243).
+Amendments to the protocol after publication were determined by the core research group and updated on ClinicalTrials.gov.
 
-The final anonymized dataset and code for analysis are released
-publicly. Authorship follows the International Committee of
-Medical Journal Editors (ICMJE) guidelines.
+The final anonymized dataset and code for analysis are released publicly. Authorship follows the International Committee of Medical Journal Editors (ICMJE) guidelines.
 
@@
 A total of `r n.residents` residents were trained during this
-pilot.
+pilot study.
 Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident.
@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details per variable and outcome in Table  \ref@(tab:sample-characteristics) and Supplementary material table
-\@ref(tab:secondary-outcomes). <!--The differences in distributions between observed and extracted data, for variables that were collected through observation or interview, are shown in Table X.-->
+\@ref(tab:secondary-outcomes).
+The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
+Overall, there were considerably more missing values in extracted data compared to observed data.
 
-## Community Consultations
-The results of interviews with patients, caregivers and social workers
-will be published separately. Residents felt that...
+```{r observed-vs-retrospective, echo = FALSE}
+retrospective.comparison.table %>%
+  modify_caption("Differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview.")
+```
 

### 2024-01-18 pilot-publication/lab-book.Rmd
Start adding trauma cohorts

@@
 load_packages()
-```
 
-## Compare directly observed with retrospective data
-
-```{r}
 ## Import data
@@
 data <- prepare_data(data, codebook)
+```
 
+## Create trauma cohorts
+
+Subgroups of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
+
+```{r}
+# Create a function that counts the number of columns that are not 0
+count_unique_regions <- function(ais.region.data) {
+    unique.regions <- apply(ais.region.data, 1, function(row) {
+        sum(row != 0, na.rm = TRUE)
+    })
+    return(unique.regions)
+}
+
+unique.regions.count <- data %>%
+    select(starts_with("mxaisbr_"), -mxaisbr_General) %>%
+    count_unique_regions()
+
+# Create a function that counts the number of columns that are 3 or more
+count_severe_regions <- function(ais.region.data) {
+    severe.regions <- apply(ais.region.data, 1, function(row) {
+        sum(row >= 3, na.rm = TRUE)
+    })
+    return(severe.regions)
+}
+
+severe.regions.count <- data %>%
+    select(starts_with("mxaisbr_"), -mxaisbr_General) %>%
+    count_severe_regions()
+
+# Define blunt multisystem trauma
+dominating.injury.type <- data$incident__dominating_injury_type
+blunt.multisystem.trauma <- dominating.injury.type == "Blunt" & severe.regions.count >= 2
+
+# Define penetrating trauma
+penetrating.trauma <- dominating.injury.type == "Penetrating"
+
+# Define shock
+shock <- data$patvitals__ed_sbp < 90
+
+# Define severe traumatic brain injury
+severe.tbi <- data$patvitals__ed_gcs < 9 & data$mxaisbr_Head >= 3
+
+# Define elderly
+elderly <- data$patinfo__pt_age >= 65
+
+# Define male
+male <- data$patinfo__pt_gender == "Male"
+
+# Define female
+female <- data$patinfo__pt_gender == "Female"
+
+cohort.data <- data.frame(
+    "Blunt multisystem trauma" = blunt.multisystem.trauma,
+    "Penetrating trauma" = penetrating.trauma,
+    "Shock" = shock,
+    "Severe TBI" = severe.tbi,
+    "Elderly" = elderly,
+    "Male" = male,
+    "Female" = female,
+    arm = data$arm
+)
+
+create_descriptive_table(cohort.data, strata = "arm", include.overall = TRUE)
+```
+```
+
+## Compare directly observed with retrospective data
+
+```{r}
 ## Get variables to compare

### 2024-01-26 pilot-publication/manuscript.Rmd
Add TQIP cohort to sample characteristics table

@@
 
-Within each combination of trial arms we conducted subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
-Table S7.1 in Supplemental Material 7 shows which outcomes that were assessed in what subgroups, decided through consensus in the research team.
-We further compared the results of all subgroups with the results in the whole cohort, and compared the results in the female subgroup with the male subgroup, and the results in the blunt multisystem trauma subgroup with the penetrating trauma subgroup.
-We are aware that the numbers in some of these subgroups are small, but we included them to help guide the formulation of the statistical analysis plan for the full-scale trial.
+Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
+These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively only.
 
@@
 
-Between `r start.date` and `r end.date`, we enrolled `r n.patients`
-trauma patients from `r n.centres` participating centres. The ATLS arm
-enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the
-standard care arm `r n.control` patients. We trained a total of 
-`r n.residents` residents, `r n.atls.residents` in ATLS, and 
-`r n.ptc.residents` in PTC.
+Between `r start.date` and `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres.
+The ATLS arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients.
+We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS, and  `r n.ptc.residents` in PTC.
 
@@
 characteristics across trial arms are shown in Table
-\@ref(tab:sample-characteristics). Overall, the number of females were
-`r females`, the median (IQR) age was `r median.age`
-years, and the median ISS (IQR) and NISS (IQR) were `r median.iss` and `r median.niss` respectively.
+\@ref(tab:sample-characteristics).
+Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) and NISS (IQR) were `r median.iss` and `r median.niss` respectively.
 A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.
@@
 The relative risk in the standard care arm compared with the ATLS arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
-The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all other outcomes.
+The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.
 
@@
 We show that conducting a cluster randomized controlled trial
-comparing ATLS with PTC and standard care is feasible in this setting
-provided that there are dedicated project officers to enrol
-participants, collect data, and follow up participants. Missing data
-rates were low for the primary outcome and many secondary outcomes, as
-well as for key variables. Some variables had very high missing data
-rates and may not be feasible to include in a full-scale trial.
+comparing ATLS with PTC and standard care is feasible in this setting provided that there are dedicated project officers to enrol participants, collect data, and follow up participants.
+Missing data rates were low for the primary outcome and many secondary outcomes, as well as for key variables.
+Some variables had very high missing data rates and may not be feasible to include in a full-scale trial.
 

### 2024-01-26 pilot-publication/functions/create_descriptive_table.R

@@ -10,6 +10,7 @@
 #' @param strata Character or NULL. The variable to use as the
 #'     stratifying variable. If NULL the table is not
 #'     stratified. Defaults to NULL.
+#' @param show.all.levels Logical. If TRUE all levels of a factor are shown, even for binary variables. Defaults to TRUE.
 #' @param include.overall Logical. If TRUE a column called "Overall"
 #'     is included in the table, which includes descriptive statistics
 #'     for the whole sample. Defaults to TRUE, unless strata is NULL,
@@ -21,6 +22,7 @@
 create_descriptive_table <- function(table.data,
                                      variables = NULL,
                                      strata = NULL,
+                                     show.all.levels = TRUE,
                                      include.overall = TRUE,
                                      caption = NULL,
                                      return.as.gt = FALSE) {
@@ -44,10 +46,16 @@ create_descriptive_table <- function(table.data,
     if (!is.null(variables)) {
         table.data <- table.data[, variables]
     }
+
+    type <- list()
+    if (show.all.levels) {
+        type <- list(all_dichotomous() ~ "categorical")
+    }
+
     descriptive.table <- gtsummary::tbl_summary(
         data = table.data,
         by = strata,
-        type = all_dichotomous() ~ "categorical",
+        type = type,
         missing_text = "Missing"
     )
 

### 2024-01-26 pilot-publication/functions/prepare_data.R

@@ -133,9 +133,6 @@ prepare_data <- function(data, codebook = NULL) {
     prepared.data[as.character(prepared.data$id__reg_hospital_id) %in% ptc.centres, ]$arm <- "PTC"
     prepared.data[as.character(prepared.data$id__reg_hospital_id) %in% control.centres, ]$arm <- "Standard care"
 
-    ## Add variables indicating the different TQIP cohorts
-    prepared.data <- add_tqip_cohorts(prepared.data)
-
     ## Label variables
     prepared.data[] <- lapply(names(prepared.data), function(column.name) {
         column.data <- prepared.data %>%
@@ -143,6 +140,11 @@ prepare_data <- function(data, codebook = NULL) {
             label_variable(name = column.name, codebook = codebook)
         return(column.data)
     })
+
+    ## Add variables indicating the different TQIP cohorts
+    prepared.data <- add_tqip_cohorts(prepared.data)
+
+    ## Add additional labels
     labelled::var_label(prepared.data$riss) <- "Injury Severity Score"
     labelled::var_label(prepared.data$niss) <- "New Injury Severity Score"
     labelled::var_label(prepared.data$patinfo__pt_age) <- "Age, years"
@@ -154,6 +156,13 @@ prepare_data <- function(data, codebook = NULL) {
     labelled::var_label(prepared.data$patvitals__ed_sbp) <- "Systolic blood pressure, mmHg"
     labelled::var_label(prepared.data$patvitals__ed_gcs) <- "Glasgow Coma Scale"
     labelled::var_label(prepared.data$incident__moi) <- "Mechanism of injury"
+    labelled::var_label(prepared.data$blunt.multisystem.trauma) <- "Blunt multisystem trauma"
+    labelled::var_label(prepared.data$penetrating.trauma) <- "Penetrating trauma"
+    labelled::var_label(prepared.data$shock) <- "Shock"
+    labelled::var_label(prepared.data$severe.tbi) <- "Severe traumatic brain injury"
+    labelled::var_label(prepared.data$elderly) <- "Elderly"
+    labelled::var_label(prepared.data$male) <- "Male"
+    labelled::var_label(prepared.data$female) <- "Female"
 
     ## Return prepared data
     return(prepared.data)

### 2024-01-26 pilot-publication/main.R

@@ -24,9 +24,12 @@ results <- get_basic_results(data)
 
 ## Create table of sample characteristics
 table.variables <- c(
-    "patinfo__pt_age", "patinfo__pt_gender",
+    "patinfo__pt_age", "elderly",
+    "patinfo__pt_gender",
     "incident__dominating_injury_type",
-    "patvitals__ed_rr", "patvitals__ed_sat",
+    "blunt.multisystem.trauma", "severe.tbi",
+    "shock", "patvitals__ed_rr",
+    "patvitals__ed_sat",
     "patvitals__ed_hr", "patvitals__ed_sbp",
     "patvitals__ed_gcs", "riss",
     "outcomes__discharge_alive",
@@ -36,7 +39,7 @@ table.variables <- c(
 table.data <- data %>%
     select(all_of(table.variables)) %>%
     select(-arm)
-overall.sample.characteristics.table <- create_descriptive_table(table.data)
+overall.sample.characteristics.table <- create_descriptive_table(table.data, show.all.levels = FALSE)
 
 ## Create table of sample characteristics before training
 pre.training.table.data <- data %>%
@@ -45,6 +48,7 @@ pre.training.table.data <- data %>%
 pre.training.characteristics.table <- create_descriptive_table(
     pre.training.table.data,
     strata = "arm",
+    show.all.levels = FALSE,
     include.overall = TRUE
 )
 
@@ -55,6 +59,7 @@ post.training.table.data <- data %>%
 post.training.characteristics.table <- create_descriptive_table(
     post.training.table.data,
     strata = "arm",
+    show.all.levels = FALSE,
     include.overall = TRUE
 )
 

### 2024-01-29 pilot-publication/manuscript.Rmd
Update manuscript

@@
 **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres.
-The patient recruitment rate was `r recruitment.rate.patients`% and the resident recruitment rate was `r recruitment.rate.residents`%.
+The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%.
 The lost to follow up rate was `r rate.lost.to.follow.up`%.
@@
 We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS and PTC with standard care.
-The objectives of this pilot study were to:
-
-- Estimate probable effect sizes on patient outcomes associated with ATLS^®^ and PTC compared with standard care, estimate the proportion of participants with the outcome (if binary), and estimate the intracluster correlation coefficient, as a basis for future sample size calculations.
-
-- Assess the feasibility of recruiting participants and collecting data on primary and secondary outcomes, such as mortality, in-hospital complications, length of stay, and quality of life.
-
-- Assess how the effect sizes and directions of these effects of ATLS and PTC may differ across clinically important subgroups.
 
@@
 ### Resident Doctor Participants
-Resident doctors doing their speciality training in surgery or
-emergency medicine managing trauma patients in the emergency
-department and who were expected to remain in the participating hospitals for at least one year.
+Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year.
 To facilitate administration each surgical department is divided into units, which manage the out-patient and emergency departments, operating rooms etc on different days each week.
@@
 
-### Concomitant Care
-
-#### Baseline Training
-
-**This section needs to be updated with the emergency medicine curriculum**
-
-The care provided by all participating hospitals at baseline was based on the training curriculum formulated by The National Medical Council of India for post graduation in General Surgery [@mciMsSurgery2019].
-Regarding trauma, these guidelines state that the student should:
-
-a.
-Have knowledge about response to trauma; burns: causes, prevention and management; wounds of scalp and its management; recognition, diagnosis and monitoring of patients with head injury, Glasgow coma scale.
-b.  Be able to provide and coordinate emergency resuscitative measures in acute surgical situations including trauma.
-c.  Choose, perform and interpret appropriate imaging in trauma - ultrasound Focused Abdominal Sonography in Trauma (FAST).
-d.  Undergo advanced trauma and cardiac life support course (certified) before appearing in final examination.
-e.  Undergo clinical posting in emergency and trauma.
-f.  Present or discuss cases of blunt abdominal trauma.
-
-Although training in an ATLS^®^ is part of the curriculum it is  optional and not doing this training does not
-result in failure to obtain post graduation completion.
-
-#### Standard of Care
-
-**This section needs to be revised to include information from sites where patients were not managed by surgical residents**
-
-At most tertiary hospitals in India trauma patients present to the emergency department where they are assessed by a doctor and referred to the surgical bay for further management.
-In the surgical bay a second or third year general surgery resident sees all the major trauma and provide the initial care, including initiating treatment and investigations.
-This resident informs the consultant on call who is generally an Assistant Professor.
-Most procedures like intercostal drainage, open wound suturing, intubation etc.
-would be done in the surgical resuscitation area, by the surgical resident.
-
-Compared to other settings where a trauma team approach is adopted, nurses and other healthcare professionals are involved to a limited extent during the initial management.
-Their roles include assisting during intubation and other bedside procedures, charting the vitals (not recording) and giving injections.
-They also accompany the resident during transfers of serious patients.
-
-After completing the assessment and starting initial resuscitation, the resident decides to send the patient for imaging (X-rays/FAST/CT-scan) or to the operation room in consultation with or after assessment by the on-call consultant.
-A portable X-ray and an ultrasonography machine to conduct FAST may or may not be available in the surgical bay.
-The patients who are operated, managed conservatively, not intubated, or with minor trauma will be sent to the surgical ward.
-Those who need increased monitoring or mechanical ventilation remain in the surgical bay or in the intensive care unit (ICU) depending on the availability of ICU beds.
-The further treatment continues in the respective ward or ICU and patients are finally discharged from the ward.
+### Standard Care
+Standard care varies across hospitals in India, but trauma patients are initially managed by casualty medical officers, surgical residents, or emergency medicine residents.
+They are mainly first- or second-year residents who resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
 
 ## Outcomes
-Our pilot study included both participant and feasibility
-outcomes.
+Our pilot study included both participant and feasibility outcomes.
 Prior to deciding on these participant outcomes we searched the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's database but were unable to identify appropriate core outcome sets for our populations of participants.
@@
 Among residents the pass rate, after two attempts, was `r rate.pass`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details per variable and outcome in Table  \ref@(tab:sample-characteristics) and Supplementary material table
-\@ref(tab:secondary-outcomes).
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
 The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
@@
 We show that conducting a cluster randomized controlled trial
-comparing ATLS with PTC and standard care is feasible in this setting provided that there are dedicated project officers to enrol participants, collect data, and follow up participants.
-Missing data rates were low for the primary outcome and many secondary outcomes, as well as for key variables.
+comparing ATLS with PTC and standard care is feasible in this setting.
+Missing data were low for key variables, including the primary outcome and many secondary outcomes.
 Some variables had very high missing data rates and may not be feasible to include in a full-scale trial.
@@
 PTC and standard care arms.
-
-We found that the PTC arm had lower 30-day mortality compared to the
-ATLS and standard care arms.
-
-We found that the standard care arm had lower 30-day mortality
-compared to the ATLS and PTC arms.
-
-We found that 30-day mortality was similar in the ATLS and PTC arms
-but lower than the standard care arm.
-
-We found that 30-day mortality was similar in the ATLS and standard
-care arms but lower than the PTC arm.
-
-We found that 30-day mortality was similar in the PTC and standard
-care arms but lower than the ATLS arm.
+We also found that the PTC arm had lower mortality than the standard care arm.
+These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
+The arms differed considerably in sample size, with the ATLS arm having the smallest sample size.
+This difference resulted from the randomisation process with a small number of heterogeneous clusters.
 
@@
 
+## Outcomes across subgroups
+Because of small numbers in the pre-specified subgroups we decided to report only descriptive data on these subgroups.
+
 ## Number of Participating Centres

### 2024-01-29 pilot-publication/functions/get_basic_results.R

@@ -29,6 +29,8 @@ get_basic_results <- function(data) {
     results$n.atls.residents <- 4 + 2 # The total number of residents trained in ATLS, per ATLS centre
     results$n.ptc.residents <- 9 + 6 # The total number of residents trained in PTC, per centre
     results$n.residents <- with(results, n.atls.residents + n.ptc.residents)
+    n.passed <- 6 # 2 had to retake the exam, but all passed the second time
+    results$pass.rate <- round(n.passed / results$n.atls.residents * 100)
     results$n.centres <- data %>%
         pull(id__reg_hospital_id) %>%
         unique() %>%
@@ -46,5 +48,11 @@ get_basic_results <- function(data) {
     results$iqr.niss <- get_iqr(data$niss)
     results$n.admitted <- with(data, sum(interventions__admitted == "Yes"))
     results$p.admitted <- round(results$n.admitted / nrow(data) * 100)
+    results$n.lost.to.follow.up <- with(data, sum(is.na(outcomes__alive_after_30_days)))
+    results$rate.lost.to.follow.up <- round(results$n.lost.to.follow.up / nrow(data) * 100)
+    results$p.missing.in.hospital.mortality <- with(data, round(sum(is.na(outcomes__discharge_alive)) / nrow(data) * 100))
+    missing.data.summary <- get_missing_data_summary(data)
+    results$min.missing.data <- min(missing.data.summary$p.missing)
+    results$max.missing.data <- max(missing.data.summary$p.missing)
     return(results)
 }

### 2024-01-29 pilot-publication/functions/get_missing_data_summary.R

@@ -1,0 +1,76 @@
+get_missing_data_summary <- function(data, variable.names = NULL) {
+    # Create default variable names
+    if (is.null(variable.names)) {
+        variable.names <-
+            c(
+                "patinfo__pt_age", "patinfo__pt_gender", "patinfo__occupation",
+                "incident__referral", "incident__dominating_injury_type", "incident__moi",
+                "incident__date_of_injury", "incident__arrival_by", "incident__date_of_arrival",
+                "incident__date_of_ed_discharge", "patvitals__ed_rr", "patvitals__ed_sat",
+                "patvitals__ed_hr", "patvitals__ed_sbp", "patvitals__ed_dbp",
+                "patvitals__ed_gcs", "patvitals__ed_pupils", "patvitals__ed_temperature",
+                "patvitals__ed_initial_serum_lactate", "patvitals__ed_intial_be", "who_tc__ctp", "who_tc__popf", "who_tc__lbiv",
+                "who_tc__fsd", "who_tc__apf", "who_tc__aib", "who_tc__sid", "who_tc__nvs",
+                "who_tc__tvg", "who_tc__ang", "who_tc__abg", "resident__res_num_serious",
+                "resident__res_comfort", "cci_index__cci_age", "cci_index__cci_mi",
+                "cci_index__cci_chf", "cci_index__cci_pvd", "cci_index__cci_cva",
+                "cci_index__cci_dementia", "cci_index__cci_copd", "cci_index__cci_ctd",
+                "cci_index__cci_pud", "cci_index__cci_liver", "cci_index__cci_dm",
+                "cci_index__cci_hemiplegia", "cci_index__cci_ckd", "cci_index__cci_tumor",
+                "cci_index__cci_leukemia", "cci_index__cci_lymphoma", "cci_index__cci_aids",
+                "cci_index__co_morbidity_index", "interventions__intubation",
+                "interventions__chest_tube", "interventions__vasopressors", "interventions__recived_fluids",
+                "interventions__intervention", "interventions__admitted", "interventions__surgery_during_stay",
+                "interventions__xray_done", "interventions__fast_done",
+                "interventions__ct_done", "outcomes__alive_24h", "outcomes__discharge_alive",
+                "outcomes__selfambulatory_at_discharge", "outcomes__patient_satisfaction",
+                "outcomes__qold_performed", "outcomes__eq5dm", "outcomes__eq5dsc",
[…]

### 2024-01-29 pilot-publication/manuscript.Rmd
Remove some details

@@
 We used simple randomisation to allocate sites to trial arms.
-We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting on March 7, 2022 [^dev1]<!-- Meeting notes ../meetings/20220307/ -->.
+We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting<!-- Meeting notes ../meetings/20220307/ -->.
 The content of the envelopes dictated what trial arm (ATLS, PTC, or standard care) each hospital was allocated to.
 There were two hospitals in each of the intervention arms and three hospitals in the control arm.
-
-[^dev1]: Initially the plan was to enroll six hospitals (clusters) but we ended up enrolling seven.
 

### 2024-01-29 pilot-publication/manuscript.Rmd
Work on manuscript code

@@
 source("main.R")
+attach(results)
 ```
@@
 We recruited tertiary hospitals and included trauma patients and residents managing these patients.
-Two hospitals were randomised to ATLS, two to PTC, and three to standard care.
+Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care.
 The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
@@
 The lost to follow up rate was `r rate.lost.to.follow.up`%.
-`r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm.
-**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
+`r stringr::str_to_sentence(english::english(m30d.standard.care))` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.
+**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
 
@@
 
-- Cluster randomized controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
+- Cluster randomized controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
 - Prospective data collection with direct observations by dedicated project officers.
@@
 There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021], but there are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS and PTC with standard care.
+We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
 
@@
 
-The ATLS^®^ training was conducted in an ATLS certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018].
+The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018].
 The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018].
@@
 We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting<!-- Meeting notes ../meetings/20220307/ -->.
-The content of the envelopes dictated what trial arm (ATLS, PTC, or standard care) each hospital was allocated to.
+The content of the envelopes dictated what trial arm (ATLS^®^, PTC, or standard care) each hospital was allocated to.
 There were two hospitals in each of the intervention arms and three hospitals in the control arm.
@@
 
-We compared participant outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care.
+We compared participant outcomes in three combinations of trial arms: ATLS^®^ versus PTC, ATLS^®^ versus standard care, and PTC versus standard care.
 In each combination we compared both differences in final values and differences in change from baseline.
@@
 Between `r start.date` and `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres.
-The ATLS arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients.
-We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS, and  `r n.ptc.residents` in PTC.
+The ATLS^®^ arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients.
+We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS^®^, and  `r n.ptc.residents` in PTC.
 
@@
 
-```{r, echo = FALSE}
-## Crude 30 day mortality
-m30d.atls <- inline_text(post.training.characteristics.table,
-  variable = outcomes__alive_after_30_days,
-  level = "Yes",
-  column = "ATLS"
-)
-m30d.ptc <- inline_text(post.training.characteristics.table,
-  variable = outcomes__alive_after_30_days,
-  level = "Yes",
-  column = "PTC"
-)
-m30d.standard.care <- inline_text(post.training.characteristics.table,
-  variable = outcomes__alive_after_30_days,
-  level = "Yes",
-  column = "Standard care"
-)
+After training, a total of `r m30d.standard.care` patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
+The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS^®^ arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units.
+The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS^®^ arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.
 
-## Extract tables for absolute and relative differences
-standard.care.vs.atls.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
-
-standard.care.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
-
-atls.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
-
-## Absolute differences
-arr.standard.care.atls <- inline_outcome_text(
-  standard.care.vs.atls.table,
-  outcome = "30 day mortality",
-  column = "Absolute difference",
-  level = "Yes"
-)
-
-arr.standard.care.ptc <- inline_outcome_text(
-  standard.care.vs.ptc.table,
-  outcome = "30 day mortality",
-  column = "Absolute difference",
-  level = "Yes"
-)
-
-arr.atls.ptc <- inline_outcome_text(
-  atls.vs.ptc.table,
-  outcome = "30 day mortality",
-  column = "Absolute difference",
-  level = "Yes"
-)
-
-## Relative differences
-rr.standard.care.atls <- inline_outcome_text(
-  standard.care.vs.atls.table,
-  outcome = "30 day mortality",
-  column = "Relative difference",
-  level = "Yes"
-)
-
-rr.standard.care.ptc <- inline_outcome_text(
-  standard.care.vs.ptc.table,
-  outcome = "30 day mortality",
-  column = "Relative difference",
-  level = "Yes"
-)
-
-rr.atls.ptc <- inline_outcome_text(
-  atls.vs.ptc.table,
-  outcome = "30 day mortality",
-  column = "Relative difference",
-  level = "Yes"
-)
-
-## Extract tables for change from baseline in 30 day mortality
-absolute.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
-relative.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
-
-## Absolute change from baseline
-abs.change.standard.care <- inline_outcome_text(
-  absolute.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "Standard care",
-  level = "Yes"
-)
-
-abs.change.atls <- inline_outcome_text(
-  absolute.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "ATLS",
-  level = "Yes"
-)
-
-abs.change.ptc <- inline_outcome_text(
-  absolute.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "PTC",
-  level = "Yes"
-)
-
-## Relative change from baseline
-rel.change.standard.care <- inline_outcome_text(
-  relative.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "Standard care",
-  level = "Yes"
-)
-
-rel.change.atls <- inline_outcome_text(
-  relative.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "ATLS",
-  level = "Yes"
-)
-
-rel.change.ptc <- inline_outcome_text(
-  relative.change.from.baseline.table,
-  outcome = "30 day mortality",
-  column = "PTC",
-  level = "Yes"
-)
-```
-
-After training, a total of `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
-The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units.
-The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.
-
-The absolute risk difference between the standard care and ATLS arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS and PTC arms (95% CI) `r arr.atls.ptc`% units.
-The relative risk in the standard care arm compared with the ATLS arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
+The absolute risk difference between the standard care and ATLS^®^ arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS^®^ and PTC arms (95% CI) `r arr.atls.ptc`% units.
+The relative risk in the standard care arm compared with the ATLS^®^ arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS^®^ arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
 The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.
@@
 We show that conducting a cluster randomized controlled trial
-comparing ATLS with PTC and standard care is feasible in this setting.
+comparing ATLS^®^ with PTC and standard care is feasible in this setting.
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
@@
 
-We found that the ATLS arm had lower 30-day mortality compared to the
+We found that the ATLS^®^ arm had lower 30-day mortality compared to the
 PTC and standard care arms.
@@
 These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
-The arms differed considerably in sample size, with the ATLS arm having the smallest sample size.
+The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
 This difference resulted from the randomisation process with a small number of heterogeneous clusters.
@@
 significant reduction in mortality, from 20 to 15%, after implementing
-ATLS [@Wang2010].
+ATLS^®^ [@Wang2010].
 
 Several controlled trials, including at least two randomized
-controlled trials [@Ali1996; @Ali1999], show that ATLS is associated
+controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated
 with improved provider skills [@Mohammad2013]. Observational evidence

### 2024-01-29 pilot-publication/functions/get_basic_results.R

@@ -31,6 +31,8 @@ get_basic_results <- function(data) {
     results$n.residents <- with(results, n.atls.residents + n.ptc.residents)
     n.passed <- 6 # 2 had to retake the exam, but all passed the second time
     results$pass.rate <- round(n.passed / results$n.atls.residents * 100)
+    n.eligible.residents <- results$n.residents # Need to update this
+    results$recruitment.rate.residents <- round(results$n.residents / n.eligible.residents * 100)
     results$n.centres <- data %>%
         pull(id__reg_hospital_id) %>%
         unique() %>%

### 2024-01-29 pilot-publication/main.R

@@ -78,26 +78,144 @@ sample.characteristics.table <- gtsummary::tbl_merge(
 retrospective.comparison.table <- create_retrospective_comparison_table(data)
 
 ## Bootstrap outcome results
-unlink("error.log")
-unlink("out", recursive = TRUE)
-dir.create("out", showWarnings = FALSE)
-n.boot.samples <- 1000
-bootstrapped.outcome.results <- boot(data, try_estimate_outcome_results, R = n.boot.samples)
-saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outcome-results.Rds"))
+use.saved <- TRUE
+if (use.saved) {
+    outcome.results <- readRDS("outcome-results.Rds")
+} else {
+    unlink("error.log")
+    unlink("out", recursive = TRUE)
+    dir.create("out", showWarnings = FALSE)
+    n.boot.samples <- 1000
+    bootstrapped.outcome.results <- boot(data, try_estimate_outcome_results, R = n.boot.samples)
+    saveRDS(bootstrapped.outcome.results, file = file.path("out", "bootstrapped-outcome-results.Rds"))
 
-## Calculate confidence intervals
-bootstrapped.outcome.results.ci <- calculate_bootstrap_cis(bootstrapped.outcome.results)
-saveRDS(bootstrapped.outcome.results.ci, file = file.path("out", "bootstrapped-outcome-results-ci.Rds"))
+    ## Calculate confidence intervals
+    bootstrapped.outcome.results.ci <- calculate_bootstrap_cis(bootstrapped.outcome.results)
+    saveRDS(bootstrapped.outcome.results.ci, file = file.path("out", "bootstrapped-outcome-results-ci.Rds"))
 
-## Combine point estimates with confidence intervals
-outcome.results <- list(
-    point.estimates = bootstrapped.outcome.results$t0,
-    confidence.intervals = bootstrapped.outcome.results.ci
-)
-saveRDS(outcome.results, file = file.path("outcome-results.Rds"))
+    ## Combine point estimates with confidence intervals
+    outcome.results <- list(
+        point.estimates = bootstrapped.outcome.results$t0,
+        confidence.intervals = bootstrapped.outcome.results.ci
+    )
+    saveRDS(outcome.results, file = file.path("outcome-results.Rds"))
+}
 
 ## Create tables with outcome results
 outcome.results.tables <- create_outcome_results_tables(outcome.results)
 
+## Crude 30 day mortality
+m30d.atls <- inline_text(post.training.characteristics.table,
[…]

### 2024-02-02 pilot-publication/manuscript.Rmd

@@
       number_sections: false
-##    md_document: default
-##    bookdown::word_document2:
-##      toc: false
-##      number_sections: false
-##      reference_docx: "reference.docx"
+    md_document: default
+    bookdown::word_document2:
+      toc: false
+      number_sections: false
+      reference_docx: "reference.docx"
 ---
@@
 The lost to follow up rate was `r rate.lost.to.follow.up`%.
-`r stringr::str_to_sentence(english::english(m30d.standard.care))` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.
+`r m30d.standard.care` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.
 **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
@@
 We explored intervention effects across the following clinical subgroups: men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly, as defined by Hornor et al [@Hornor2018].
-The consent form for patients are available as Supplemental Material 1.
 
@@
 Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
-The consent form for residents is available as Supplemental Material 2.
 We did not ask for consent from residents at the units in the control hospitals as their practice was not affected by this pilot and we did not collect any personal identifiable data on them.
@@
 The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline.
-All outcomes that pertain to the individual participant level are detailed in Supplemental Material 3.
-We decided to include a large number of outcomes,
-including some more exploratory, so that we could test their feasibility and relevance.
+All outcomes that pertain to the individual participant level are detailed in Supplemental Material 1.
+We decided to include a large number of outcomes, including some more exploratory, so that we could test their feasibility and relevance.
 
@@
 Based on AIS we calculated the Injury Severity Score (ISS) and the New ISS (NISS).
-Supplemental Material 4 contains a full variable list, with definitions.
+Supplemental Material 2 contains a full variable list, with definitions.
 
@@
 # Results
-```{r}
-females <- inline_text(overall.sample.characteristics.table,
-  variable = patinfo__pt_gender,
-  level = "Female"
-)
-median.age <- inline_text(overall.sample.characteristics.table,
-  variable = patinfo__pt_age
-)
-median.iss <- inline_text(overall.sample.characteristics.table,
-  variable = riss
-)
-median.niss <- inline_text(overall.sample.characteristics.table,
-  variable = niss
-)
-m30d <- inline_text(overall.sample.characteristics.table,
-  variable = outcomes__alive_after_30_days,
-  level = "Yes"
-)
-```
 
@@
 
-The study flowchart is shown in Figure X and patient sample
-characteristics across trial arms are shown in Table
-\@ref(tab:sample-characteristics).
-Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) and NISS (IQR) were `r median.iss` and `r median.niss` respectively.
+The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample
+characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
+Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
 A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.
+
+```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care."}
+create_flow_diagram()
+knitr::include_graphics("flow-diagram.png")
+```
+
+```{r sample-characteristics, echo = FALSE}
+```
+
+```{r sample-characteristics, echo = FALSE, results = "asis"}
+```
 
@@
 Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident.
-<!--In the post training period, the confidence in the ATLS arm was X (IQR X - X), compared to X (IQR X - X) in the PTC arm, and X (IQR X - X) in the standard care arm.-->
+In the pre training period, the median confidence in the stadard care arm was `r median.comfort.standard.care.pre` (IQR `r iqr.comfort.standard.care.pre`), compared to `r median.comfort.atls.pre` (IQR `r iqr.comfort.atls.pre`) in the ATLS and `r median.comfort.ptc.pre` (IQR `r iqr.comfort.ptc.pre`) in the PTC arm.
+In the post training period, the corresponding figures were `r median.comfort.standard.care.post` (IQR `r iqr.comfort.standard.care.post`) in the standard care arm, `r median.comfort.atls.post` (IQR `r iqr.comfort.atls.post`) in the ATLS and `r median.comfort.ptc.post` (IQR `r iqr.comfort.ptc.post`) in the PTC arm.
 
@@
 The lost to follow up rate was `r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of `r n.patients` patients not completing 30 day follow up.
-Among residents the pass rate, after two attempts, was `r rate.pass`%.
+Among residents the pass rate, after two attempts, was `r pass.rate`%.
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
@@
 
-```{r secondary-outcomes, echo = FALSE, results = "asis"}
-table1::t1kable(secondary.outcomes.table, longtable = TRUE) %>%
-  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)
-```
-
 # References

### 2024-02-18 pilot-publication/manuscript.Rmd
Add commented pilot manuscripts

@@
 
-<!--
 ## Corresponding author
@@
 
-## Contributors
-
-Trauma life support Effectiveness Research Network (TERN) collaborators:
-
-##### Martin Gerdin Wärnberg (MGW)
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-Function Perioperative Medicine and Intensive Care, Karolinska University Hospital, Solna, Sweden.
-
 ##### Johanna Berg (JB)
@@
 Emergency Medicine, Department of Internal and Emergency Medicine, Skåne University Hospital, Malmö, Sweden.
+
+##### Shamita Chatterjee (SC)
+Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
+
+##### Li Felländer-Tsai (LFT)
+Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
+
+Department of Reconstructive Orthopedics, Karolinska University Hospital, Stockholm, Sweden.
+
+##### Geeta Ghag (GG)
+Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
+
+##### Monty Khajanchi (MK)
+WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
+
+Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.
+
+##### Catherine Juillard (CJ)
+Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
+
+##### Nobhojit Roy (NR)
+WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.
+
+Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
+
+##### Arun Prasad (AP)
+
+Department of Surgery, Indraprastha Apollo Hospitals, New Delhi
+
+##### Rajdeep Singh (RS)
+Department of Surgery, Maulana Azad Medical College, New Delhi, India.
+
+##### Kapil Dev Soni (KDS)
+Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
+
+##### Lovisa Strömmer (LS)
+Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
+
+Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.
+
+##### Vipul Nandu
+
+## Non-author contributors
+
+Trauma life support Effectiveness Research Network (TERN) collaborators:
 
@@
 
-##### Shamita Chatterjee (SC)
-Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
-
 ##### Chintamani (C)
 Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
-
-##### Li Felländer-Tsai (LFT)
-Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
-
-Department of Reconstructive Orthopedics, Karolinska University Hospital, Stockholm, Sweden.
 
@@
 
-##### Geeta Ghag (GG)
-Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
-
 ##### Marie Hasselberg (MH)
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-##### Catherine Juillard (CJ)
-Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
-
-##### Monty Khajanchi (MK)
-WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
-
-Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.
 
@@
 
-##### Nobhojit Roy (NR)
-WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.
-
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
 ##### Amit Roy (AR)
@@
 
-##### Rajdeep Singh (RS)
-Department of Surgery, Maulana Azad Medical College, New Delhi, India.
-
 ##### Harris Solomon (HS)
@@
 
-##### Kapil Dev Soni (KDS)
-Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
-
-##### Lovisa Strömmer (LS)
-Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
-
-Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.
-
 ##### Megha Tandon (MT)
 Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
-
--->
 

### 2024-03-13 pilot-publication/manuscript.Rmd
Add authors
Start revising pilot manuscript

@@
 ## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
-title: A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes  
+title: "Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial"
 author: Trauma life support training Effectiveness Research Network (TERN) collaborators
@@
 ##### Vipul Nandu
+Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
+
+##### Debojit Basak
+Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
+
+##### Tamal Khan
 
@@
 We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
-**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught.
-We recruited tertiary hospitals and included trauma patients and residents managing these patients.
+**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in tertiary care hospitals across metropolitan areas in India.
+We included adult trauma patients and residents managing these patients.
 Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care.
-The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
+The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records.
 We conducted community consultations in parallel with the pilot trial.
-**Ethics and dissemination** We obtained ethical approval from all participating centres.
-**Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres.
-The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%.
+**Ethics and dissemination** We obtained ethical approval from all participating hospitals.
+**Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents.
+The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
 The lost to follow up rate was `r rate.lost.to.follow.up`%.
-`r m30d.standard.care` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.
-**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
+The pass rate was `r pass.rate`%.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`.
+Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
+<!--`r m30d.standard.care` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.--> **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
 
@@
 Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year [@injuries2020].
-Almost 10% of the global burden of disease is due to trauma and trauma is the top contributor to the burden of disease in children and adults aged 10 to 49 years [@Vos2020].
-
-Trauma care is time sensitive and early management of life or limb threatening conditions is crucial.
 Several trauma life support training programs have been developed to improve the early management of patients as they arrive at hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) is the most established trauma life support training program and more than one million doctors in over 80 countries have been trained in the programme [@acsAtls2018].
-Uptake in low- and middle income countries (LMIC) has been slow, potentially due to high costs [@Kadhum2020].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021].
 
-The free Primary Trauma Care (PTC) program is the most widely spread alternative programme.
-The goal of PTC is to improve trauma care in LMIC [@ptcfPtc2018].
-Like ATLS^®^, doctors in over 80 countries have been trained in PTC, and the programme has been endorsed by the World Health Organization (WHO), among other international organizations including several professional societies [@ptcfPtc2018].
-
-There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021], but there are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
+There is no high quality evidence showing that ATLS^®^ or PTC impact patient outcomes [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Several systematic reviews call for trials in settings where these programmes are not routinely taught, but doing is likely logistically challenging.
+We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
 
@@
 We piloted a three-armed cluster randomised controlled trial.
+The protocol was published [@GerdinWärnberg2022].
 There were two intervention arms, ATLS^®^ and PTC training, and one control arm, standard care.
@@
 
-Our study is a pilot study because its objectives involves estimating quantities, such as the probable effect sizes, proportion of participants with the outcome (if binary), and the intracluster correlation coefficient, needed for the sample size calculations of a full-scale trial [@Hemming2017].
-This pilot study also establish how many participants that can be enrolled, as well as likely drop out rates, and the feasibility of collecting primary and secondary outcomes.
-
 ## Study Setting
 We conducted this pilot in seven Indian tertiary hospitals, where neither ATLS^®^, PTC, nor any other trauma life support training program is routinely taught.
-The seven hospitals were Seth G.
-S.
-Medical College and K.E.M.
-Hospital, Mumbai; Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai; HBT Medical College And Dr.
-R N Cooper Municipal General Hospital, Mumbai; Medica Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of Medical Education & Research, Chandigarh.
 

### 2024-03-13 pilot-publication/manuscript.Rmd
Edit manuscript

@@
 - Cluster randomized controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
-- Prospective data collection with direct observations by dedicated project officers.
+- Prospective data collection with direct observations by dedicated research officers.
 - Participating centers' heterogeneity may affect the study estimates and bias the results.
@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-There is no high quality evidence showing that ATLS^®^ or PTC impact patient outcomes [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-Several systematic reviews call for trials in settings where these programmes are not routinely taught, but doing is likely logistically challenging.
-We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
+Several systematic reviews call for trials in settings where these programmes are not routinely taught.
+We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
 
@@
 The protocol was published [@GerdinWärnberg2022].
-There were two intervention arms, ATLS^®^ and PTC training, and one control arm, standard care.
-We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the control arm).
+There were two intervention arms, ATLS^®^ and PTC training, and one standard care arm.
+We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
 This design allowed us to assess outcomes both as final values and as change from baseline.
@@
 
-## Eligibility Criteria for Participants and Clusters
-There were two groups of participants: patients and resident doctors.
+## Eligibility Criteria for Cluster and Participants
+
+### Clusters
+We included Indian tertiary care hospitals that admitted more than 400 adult patients with trauma each year, with operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
+Most surgical and emergency medicine departments in India organise their physicians in units.
+Each unit typically consists of at least three faculty members and three to twelve residents, and is assigned a specific day of the week when they manage the emergency department.
+We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals.
+
+To be eligible, units had to have a maximum of 25% of the doctors trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot study.
+Those residents who had received training in the last five years were considered as trained.
+The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
+The principal investigator at each eligible hospital selected the units for training.
+We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.
 
 ### Patient Participants
-Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma.
-History of trauma is here defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
-We explored intervention effects across the following clinical subgroups: men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly, as defined by Hornor et al [@Hornor2018].
+Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
+History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
 
 ### Resident Doctor Participants
-Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year.
-To facilitate administration each surgical department is divided into units, which manage the out-patient and emergency departments, operating rooms etc on different days each week.
-One or two, out of typically six, units' residents were selected from each hospital.
-One unit consists of at least three faculty and three to twelve residents.
+Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
+Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
 
-To be eligible, units had to have a maximum of 25% of the doctors trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot.
-Those residents who had received training in the last five years were considered as trained.
-The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
-
-Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
-We did not ask for consent from residents at the units in the control hospitals as their practice was not affected by this pilot and we did not collect any personal identifiable data on them.
-
-### Clusters
-Indian tertiary care hospitals that admit more than 400 adult patients with trauma each year.
-We randomised on the cluster (hospital) level to avoid contamination between intervention and control arms.
-To be eligible for inclusion hospitals had to provide the following services round the clock: operation theatres, X-ray, CT, and ultrasound facilities, and blood bank.
-In addition the baseline admission rate had to be more than 35 adult patients with major trauma per month.
-
-## Interventions
-In each intervention arm one or two units', out of typically six, residents per hospital providing emergency care to trauma patients were trained in either ATLS^®^ or PTC.
+## Intervention
+In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC.
 For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
-If residents dropped out or changed units after training but before data collection was completed we planned to conduct additional training if needed to meet the 75% criterion, but this was not required.
-We did not train the units' faculty, as they are typically not directly involved in the initial management of trauma patients.
-
+We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
 The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018].
 The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018].
-These courses were conducted over a period of 2.5 to 3 days.
-The residents certified "pass" were considered as trained in respective courses.
-
-The control group provided standard care with no intervention.
-
-### Modifications
-Both ATLS^®^ and PTC are standard training programs with fixed curricula [@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or content of these programs during this pilot.
-
-### Adherence
-The intervention was the training in either ATLS^®^ or PTC and resident participants were required to adhere to, i.e.
-participate in, the training, to be eligible for passing.
-We did not consider adherence to training contents during care delivery as adherence to the trial intervention.
+We did not modify or adapt the delivery or content of these programs during this pilot study.
 
@@
 ## Outcomes
-Our pilot study included both participant and feasibility outcomes.
-Prior to deciding on these participant outcomes we searched the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's database but were unable to identify appropriate core outcome sets for our populations of participants.
-
-The primary participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department.
-The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline.
-All outcomes that pertain to the individual participant level are detailed in Supplemental Material 1.
-We decided to include a large number of outcomes, including some more exploratory, so that we could test their feasibility and relevance.
-
-We also assessed the following feasibility outcomes, which pertained both to overall study population as well as to the individual cluster level:
-
-- Recruitment rate. For both patients and residents this was equal to the proportion of participants enrolled, out of the total number of eligible participants, over the course of the pilot study.
+Our pilot study included feasibility as well as participant outcomes. The feasibility outcomes were:
+- Consent rate.
+For both patients and residents this was equal to the percentage of participants who consented to be included, out of the total number of eligible participants, over the course of the pilot study.
 - Lost to follow up rate.
-This applied only to patients and was equal to the proportion of patients that did not complete 30 day follow up, out of all enrolled patients, over the course of the pilot study.
+This applied only to patients and was equal to the percentage of patients that did not complete 30 day follow up, out of all enrolled patients, over the course of the pilot study.
 - Pass rate.
-This applied only to residents in the intervention arms and equal the proportion of residents that pass the training programme, out of the total number of trained residents, over the course of the pilot study.
+This applied only to residents in the intervention arms and equal the percentage of residents that passed the training programme, out of the total number of trained residents, over the course of the pilot study.
 - Missing data rate. This applied to each outcome and variable and was equal to the proportion of missing data, over the course of the pilot study.
@@
 This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
-For quantitative variables this was be the difference in means, standard deviations, medians, interquartile ranges, and ranges.
-For qualitative variables this was the differences in absolute counts and percentages, across categories.
 
-## Participant Timeline
+The primary patient participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department, measured and compared across trial arms as both final values and as change from baseline.
+<!--The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline.
+All outcomes that pertain to the individual participant level are detailed in Supplemental Material 1.
+We decided to include a large number of outcomes, including some more exploratory, so that we could test their feasibility and relevance.
+-->
+
+## Participant Timeline and Inclusion
 
 ### Patients
-Patients were screened for eligibility as they arrived at the emergency department.
-Eligible patients were approached in the emergency department to consent to follow up, if they were conscious.
-If they were unconscious a patient representative was approached to consent to follow up.
-Once the patient was conscious we approached the patient to affirm the patient representative's consent.
+Arriving patients were screened for eligibility and consented, if conscious.
+Unconscious patients were consented by the patient’s representative.
+This proxy consent was reaffirmed by the patient, on regaining consciousness.
 We followed up patients at discharge, at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
@@
 Participating units were screened for eligibility once hospitals confirmed their participation.
-All residents in eligible units were approached to consent to training if their hospital was randomised to either of the intervention arms.
-Training was conducted approximately one month after the study started.
+All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms.
+The training was conducted approximately one month after the study started.
 
 ## Sample size
-Given budget and time constraints, including the rotation of units in the emergency departments in Indian hospitals (which often happen on a six months basis) the feasible data collection period was four months.
-We assumed that each of the units saw 2-4 trauma patients per week and based on that estimated that if we selected a minimum of one unit per hospital then each hospital would enrol 8-16 patients per month and 32-64 patients during the four months of this pilot.
-With a 20% attrition rate we expected each hospital to enrol 26-51 patients, coming to a total sample size of between 156 and 306 patients for this pilot study.
-
-## Recruitment
-To ensure adequate recruitment we only approached hospitals that stated that the their volumes were high enough to allow us to reach the sample size goals detailed above.
-Patients were enrolled by a dedicated project officer as they arrived at the emergency department.
-The recruitment period was four months.
-Recruitment was monitored weekly through online conferences.
-No financial or non-financial incentives were provided to trial investigators or participants for enrolment.
+We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess feasibility. The time period was dictated by budget and time constraints.
 
 ## Allocation
-
-### Sequence generation
-We used simple randomisation to allocate sites to trial arms.
-We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting<!-- Meeting notes ../meetings/20220307/ -->.
-The content of the envelopes dictated what trial arm (ATLS^®^, PTC, or standard care) each hospital was allocated to.
-There were two hospitals in each of the intervention arms and three hospitals in the control arm.
-
-### Concealment Mechanism
-We did not conceal the sequence.
-
-### Implementation
-The random allocation sequence was generated by MGW.
-Clusters were enrolled by the project's core team.
-Patient participants were enrolled if they presented during the project officers shift.
-Resident participants were enrolled if they were in the units selected for training.
-The principal investigator at each hospital selected the units for training.
-For patient participants consent for follow up was sought after randomisation from patients or patient relatives as appropriate.
-For resident participants consent was also sought after randomisation.
+We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms.
 
@@
 It was not possible to blind investigators or participants to
-interventions. We did not blind the data analysts during this pilot, but we plan to blind the data analysts during the full-scale trial.
+interventions.
 
 ## Data Collection
-Data collection started one month before the training was delivered, to establish a baseline. Each participating hospital had a dedicated project officer who collected the data.
-
-Because participating residents were assigned designated days for trauma care for a period of 6 months, data were collected during those particular days and shifts when these trained doctors were in the emergency department.
-The project officers collected data both by observing the care delivered and by interviewing the participants, and by extracting data from hospital records.
-
-Data collection continued for three months after training. The
+Each participating hospital had a dedicated research officer who collected the data for a total duration of four months.
+The research officers collected data by observing the care delivered, interviewing the participants, and by extracting data from hospital records.
+The
 research officers collected data of all trauma patients who presented to the participating units during their duty hours.
-Those patients who were admitted were followed up for complications and other in-hospital outcome measures, for example length of stay.
+Admitted patient participants were followed up for complications and other in-hospital outcome measures, for example length of stay.
 Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.
-The follow up period was be 30 days.
-The project officers made at most three attempts to reach a participant or participant representative telephonically, after which the data were recorded as missing.
-
-The project officer administered the study information and informed consent (consent was sought only for data collection including follow up) to the patient, or the patient's representative as appropriate, once the patient was stabilised.
-They will continue to collect data once they have received the consent.
-
-Details of data of those patients/relatives not willing to give consent were not included in the analysis. The number of patients who opt out from data collection was collected.
+The follow up period was 30 days.
 
 ## Variables
-The project officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, and times to and management details including imaging and surgery.
-Details of any injury sustained were collected and coded using ICD 10 and the Abbreviated Injury Scale (AIS).
-For ICD 10 coders underwent the WHO online ICD 10 training module and for AIS they were accredited.
-Based on AIS we calculated the Injury Severity Score (ISS) and the New ISS (NISS).
-Supplemental Material 2 contains a full variable list, with definitions.
+The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
+<!--Supplemental Material 2 contains a full variable list, with definitions.-->
 
 ## Patient and public involvement
-In this study, we conducted community consultations to collect
-inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial; following the Guidance for Reporting Involvement of Patients and the Public (GRIIPS 2) [@Staniszewska2017].
-The results of these consultations will be reported in a separate publication.
+We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial.
+The results of these consultations will be published separately.
 
+<!--
 ## Data management
@@
 The data were only accessible by the project's principal investigator or by delegation by the project principal investigator.
+-->
 
 ## Data monitoring
-Weekly meetings with the core team and project officers took place.
-We conducted an interim analysis on October 12, 2022, and decided to complete the study as participants were consenting to be included in the study and key variables including mortality outcomes could be collected.
-No outcomes were dropped.
-We did not use a data monitoring committee in the pilot study due to its limited scope.
+Weekly meetings with the core team and research officers took place.
+We conducted one interim analysis, and decided to complete the study as participants were consenting to be included in the study and key variables including mortality outcomes could be collected.
+We did not use a data monitoring committee.
 
 ## Statistical Methods
-We analysed all pilot data using descriptive statistics.
+We analysed all pilot data, including both feasibility and participant outcomes, using descriptive statistics.
+We did not perform any formal hypothesis tests during the analysis of this pilot's data [@Lee2014].
 Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range.
 Qualitative variables are presented as absolute numbers and percentages.
-Feasibility outcomes are summarised both on the overall sample level as well as on the individual cluster level.
 We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
 
-We compared participant outcomes in three combinations of trial arms: ATLS^®^ versus PTC, ATLS^®^ versus standard care, and PTC versus standard care.
+We compared patient participant outcomes in all possible combinations of trial arms.
 In each combination we compared both differences in final values and differences in change from baseline.
@@
 
-We calculated both absolute and relative differences for each
-comparison, along with 75, 85, and 95% confidence intervals.
-We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
-We did not perform any formal hypothesis tests during the analysis of this pilot's data [@Lee2014].
-We also compared the data collected through observations and interviews with the data collected from hospital records, to assess the feasibility of collecting data from hospital records in the full-scale trial.
+<!--We calculated both absolute and relative differences for each
+comparison, along with 75, 85, and 95% confidence intervals. We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.-->
 
@@
 We were granted research ethics approval from each participating hospital.
-The approval numbers were IEC(II)/OUT/134/2022 for King Edward Memorial Seth G.
-S.
-Medical College and K.E.M.
-Hospital, Mumbai; IEC/214/22 for Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai; HBTMC/266/SURGERY for HBT Medical College And Dr.
-R N Cooper Municipal General Hospital, Mumbai; CREC/2022/FEB/1(ii) for Medica Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata; NRSmC/IEC/93/2021 for Sir Nil Ratan
-Sircar Medical College & Hospital, Kolkata; and PGI/IEC/2022/000/003 for Postgraduate Institute of Medical Education & Research, Chandigarh.
-
-The protocol was published in BMJ Open [@GerdinWärnberg2022] and was registered with ClinicalTrials.gov (reg.
-no NCT05417243).
-Amendments to the protocol after publication were determined by the core research group and updated on ClinicalTrials.gov.
-
-The final anonymized dataset and code for analysis are released publicly. Authorship follows the International Committee of Medical Journal Editors (ICMJE) guidelines.
 
@@
 
-Between `r start.date` and `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres.
+Between the study period of `r start.date` to `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres.
 The ATLS^®^ arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients.
@@
 
-The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample
-characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
+The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
@@
 
-```{r sample-characteristics, echo = FALSE}
+```{r sample-characteristics, echo = FALSE, results = "asis"}
+sample.characteristics.table
 ```
 
-```{r sample-characteristics, echo = FALSE, results = "asis"}
-```
+## Feasibility Outcomes
+<!--Out of X potentially eligible patients, X (X%) patients were enrolled
+in the study-->.
+The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
+The lost to follow up rate was `r rate.lost.to.follow.up`%.
+The pass rate was `r pass.rate`%.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
+The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
+Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
 
-```{r sample-characteristics, echo = FALSE, results = "asis"}
-sample.characteristics.table
+```{r observed-vs-retrospective, echo = FALSE}
+retrospective.comparison.table %>%
+  modify_caption("Differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview.")
 ```
@@
 The relative risk in the standard care arm compared with the ATLS^®^ arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS^®^ arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
-The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.
+<!--The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.-->
 
 ## Resident Participant Outcomes
+
 A total of `r n.residents` residents were trained during this
 pilot study.
-Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident.
-In the pre training period, the median confidence in the stadard care arm was `r median.comfort.standard.care.pre` (IQR `r iqr.comfort.standard.care.pre`), compared to `r median.comfort.atls.pre` (IQR `r iqr.comfort.atls.pre`) in the ATLS and `r median.comfort.ptc.pre` (IQR `r iqr.comfort.ptc.pre`) in the PTC arm.
-In the post training period, the corresponding figures were `r median.comfort.standard.care.post` (IQR `r iqr.comfort.standard.care.post`) in the standard care arm, `r median.comfort.atls.post` (IQR `r iqr.comfort.atls.post`) in the ATLS and `r median.comfort.ptc.post` (IQR `r iqr.comfort.ptc.post`) in the PTC arm.
-
-## Feasibility Outcomes
-<!--Out of X potentially eligible patients, X (X%) patients were enrolled
-in the study-->.
-Out of `r n.eligible.residents` potentially eligible residents, `r n.residents` (`r recruitment.rate.residents`%) participated in the study.
-The lost to follow up rate was `r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of `r n.patients` patients not completing 30 day follow up.
-Among residents the pass rate, after two attempts, was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
-The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
-Overall, there were considerably more missing values in extracted data compared to observed data.
-
-```{r observed-vs-retrospective, echo = FALSE}
-retrospective.comparison.table %>%
-  modify_caption("Differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview.")
-```
+Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident, across trial arms and both before and after training.
 
@@
 death were hard to identify and therefore the full-scale trial will
-need to include longer training of project officers.
+need to include longer training of research officers.
 
@@
 We decided to record data from records only for a subset of patients
-to reduce the project officers' workload.
+to reduce the research officers' workload.
 
@@
 
+## Analysis level of feasibility outcomes
+We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level.
+
 \pagebreak

### 2024-03-23 pilot-publication/manuscript.Rmd
Update pilot manuscript

@@
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 

### 2024-08-22 pilot-publication/manuscript.Rmd
Update pilot publication draft
Release statistical analysis plan version 0.2.0
Work on revising the pilot manuscript

@@
 
-##### Arun Prasad (AP)
-
-Department of Surgery, Indraprastha Apollo Hospitals, New Delhi
-
 ##### Rajdeep Singh (RS)
@@
 
+##### Arun Prasad (AP)
+
+Department of Surgery, Indraprastha Apollo Hospitals, New Delhi
+
 ##### Amit Roy (AR)
@@
 
-**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work.
-We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
-**Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in tertiary care hospitals across metropolitan areas in India.
-We included adult trauma patients and residents managing these patients.
-Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care.
-The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records.
-We conducted community consultations in parallel with the pilot trial.
-**Ethics and dissemination** We obtained ethical approval from all participating hospitals.
-**Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents.
+**Importance** Trauma life support training programmes aim to improve early trauma care but there is high-quality evidence to show that they improve patient outcomes.
+
+**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+
+**Design** A pilot pragmatic three-armed parallel, cluster randomised, controlled trial between `r start.date` and `r end.date`. Patient follow up was 30 days.
+
+**Setting** Tertiary care hospitals across metropolitan areas in India.
+
+**Participants** Adult trauma patients and residents managing these patients.
+
+**Interventions** ATLS^®^ or PTC training for residents in the intervention arms.
+
+**Main Outcomes and Measures** The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records.
+The primary patient outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
+
+**Results** Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care.
+We included `r n.patients` patients and `r n.residents` residents.
 The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
@@
 The pass rate was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`.
+The missing data was overall low for key variables.
 Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
-<!--`r m30d.standard.care` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm.--> **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
+`r number_to_text(m30d.standard.care.n, first = TRUE)` `r m30d.standard.care.p` patients died within 30 days in the standard care arm, `r number_to_text(m30d.atls.n)` `r m30d.atls.p` patients in the ATLS^®^ arm, and `r number_to_text(m30d.ptc.n)` `r m30d.ptc.p` patients in the PTC arm.
+
+**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
+
+**Trial Registration** ClinicalTrials.gov (reg. no NCT05417243)
 
@@
 
-Several systematic reviews call for trials in settings where these programmes are not routinely taught.
+Several systematic reviews call for trials in settings where these programmes are not routinely taught, but data needed to inform the planning of such a trial is missing.
 We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
+
+<!-- add citation to scoping review of TQIP [@mciver_effect_2024]-->
 
@@
 ## Trial Design
-We piloted a three-armed cluster randomised controlled trial.
-The protocol was published [@GerdinWärnberg2022].
-There were two intervention arms, ATLS^®^ and PTC training, and one standard care arm.
+We piloted a three-armed cluster randomised controlled trial[@GerdinWärnberg2022].
+There were a standard care arm and two intervention arms, ATLS^®^ and PTC training.
 We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
@@
 ## Study Setting
-We conducted this pilot in seven Indian tertiary hospitals, where neither ATLS^®^, PTC, nor any other trauma life support training program is routinely taught.
+We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other trauma life support training program is routinely taught.
 
@@
 ### Patient Participants
-Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
+All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
 History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.

### 2024-08-22 pilot-publication/functions/number_to_text.R

@@ -1,0 +1,8 @@
+number_to_text <- function(number, first = FALSE) {
+    if (number < 13 || first) {
+        number <- english::english(number)
+    if (is.character(number)) {
+        number <- stringr::str_to_sentence(number)
+    }
+    return(number)
+}

### 2024-08-22 pilot-publication/main.R

@@ -120,17 +120,35 @@ m30d <- inline_text(overall.sample.characteristics.table,
 )
 
 ## Crude 30 day mortality
-m30d.atls <- inline_text(post.training.characteristics.table,
+m30d.standard.care.n <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
-    column = "ATLS"
+    column = "Standard care",
+    pattern = "{n}"
 )
-m30d.ptc <- inline_text(post.training.characteristics.table,
+m30d.standard.care.p <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
-    column = "PTC"
+    column = "Standard care",
+    pattern = "({p})%"
 )
-m30d.standard.care <- inline_text(post.training.characteristics.table,
+m30d.atls.n <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
-    column = "Standard care"
+    column = "ATLS",
+    pattern = "{n}"
+)
+m30d.atls.p <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "ATLS",
+    pattern = "({p})%"
+)
+m30d.ptc.n <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "PTC",
+    pattern = "{n}"
+)
+m30d.ptc.p <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "PTC",
+    pattern = "({p})%"
 )
 
 ## Extract tables for absolute and relative differences

### 2024-08-23 pilot-publication/manuscript.Rmd
Work on pilot manuscript, up to Participant timeline and inclusion

@@
 
-## Eligibility Criteria for Cluster and Participants
-
-### Clusters
-We included Indian tertiary care hospitals that admitted more than 400 adult patients with trauma each year, with operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
-Most surgical and emergency medicine departments in India organise their physicians in units.
-Each unit typically consists of at least three faculty members and three to twelve residents, and is assigned a specific day of the week when they manage the emergency department.
-We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals.
-
-To be eligible, units had to have a maximum of 25% of the doctors trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot study.
-Those residents who had received training in the last five years were considered as trained.
-The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
-The principal investigator at each eligible hospital selected the units for training.
-We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.
-
-### Patient Participants
-All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
-History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
-
-### Resident Doctor Participants
-Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
-Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
+## Standard Care
+Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
+These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they manage the emergency department.
+In the emergency department, trauma patients are initially assessed by first- or second-year residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
 
@@
 
-### Standard Care
-Standard care varies across hospitals in India, but trauma patients are initially managed by casualty medical officers, surgical residents, or emergency medicine residents.
-They are mainly first- or second-year residents who resuscitate patients, perform interventions and refer patients for imaging or other investigations.
-Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
+## Eligibility Criteria for Cluster and Participants
+
+### Hospitals
+
+We included tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
+
+### Clusters
+
+We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals.
+To be eligible, units could have no more than 25% of their physicians trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot study.
+Those residents who had received training in the last five years were considered as trained.
+The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
+The principal investigator at each hospital selected the units for training.
+We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.
+
+### Residents
+
+Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
+Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
+
+### Patients
+
+All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
+History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
 
 ## Outcomes
-Our pilot study included feasibility as well as participant outcomes. The feasibility outcomes were:
-- Consent rate.
-For both patients and residents this was equal to the percentage of participants who consented to be included, out of the total number of eligible participants, over the course of the pilot study.
-- Lost to follow up rate.
-This applied only to patients and was equal to the percentage of patients that did not complete 30 day follow up, out of all enrolled patients, over the course of the pilot study.
-- Pass rate.
-This applied only to residents in the intervention arms and equal the percentage of residents that passed the training programme, out of the total number of trained residents, over the course of the pilot study.
-- Missing data rate. This applied to each outcome and variable and was equal to the proportion of missing data, over the course of the pilot study.
+We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A complete list of outcomes is available as Supplementary Materials. Our main outcomes were:
+
+- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
+- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients whom did not complete 30 day follow up, out of all enrolled patients.
+- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents that passed the training programme, out of the total number of trained residents.
+- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
 - Differences in distributions of observed and extracted data.
 This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
+- All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients, measured and compared across trial arms as both final values and as change from baseline.
 
-The primary patient participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department, measured and compared across trial arms as both final values and as change from baseline.
 <!--The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline.
@@
 This proxy consent was reaffirmed by the patient, on regaining consciousness.
-We followed up patients at discharge, at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
+We followed up patients at 24 hours after arrival at the emergency department, and up to 30 days after arrival at the emergency department.
 
@@
 All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms.
-The training was conducted approximately one month after the study started.
+The training was conducted approximately one month after the study started in that hospital.
 

### 2024-08-25 pilot-publication/manuscript.Rmd
Edit pilot manuscript

@@
 
-## Allocation
-We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms.
-
-## Blinding
-It was not possible to blind investigators or participants to
-interventions.
+## Allocation and blincing
+We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.
 
 ## Data Collection
-Each participating hospital had a dedicated research officer who collected the data for a total duration of four months.
-The research officers collected data by observing the care delivered, interviewing the participants, and by extracting data from hospital records.
-The
-research officers collected data of all trauma patients who presented to the participating units during their duty hours.
-Admitted patient participants were followed up for complications and other in-hospital outcome measures, for example length of stay.
+Data was collected over a four-month period.
+A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care.
+The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records.
+We followed up admitted patients for their complications and other in-hospital outcome measures.
 Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.
-The follow up period was 30 days.
 
@@
 ## Data monitoring
-Weekly meetings with the core team and research officers took place.
-We conducted one interim analysis, and decided to complete the study as participants were consenting to be included in the study and key variables including mortality outcomes could be collected.
+We conducted weekly online meetings to monitor the study and data collection.
+We conducted one interim analysis approximately halfway through the study, and decided to complete the study as residents and patients were consenting to be included in the study and key variables including mortality outcomes could be collected.
 We did not use a data monitoring committee.
@@
 ## Statistical Methods
-We analysed all pilot data, including both feasibility and participant outcomes, using descriptive statistics.
-We did not perform any formal hypothesis tests during the analysis of this pilot's data [@Lee2014].
+We analysed all data using descriptive statistics and did not perform any formal hypothesis [@Lee2014].
 Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range.
@@
 
-We compared patient participant outcomes in all possible combinations of trial arms.
+We compared patients outcomes in all possible combinations of trial arms.
 In each combination we compared both differences in final values and differences in change from baseline.
@@
 Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
-These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively only.
+These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
 
@@
 ## Ethics and Dissemination
-We were granted research ethics approval from each participating hospital.
+We were granted research ethics approval from the institutional ethics committees at each participating hospital.
 
@@
 
-Between the study period of `r start.date` to `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres.
-The ATLS^®^ arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients.
+We enrolled `r n.patients` trauma patients from `r n.centres` participating centres between `r start.date` and `r end.date`.
+The standard care arm `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm `r n.ptc` patients.
 We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS^®^, and  `r n.ptc.residents` in PTC.

### 2024-08-27 pilot-publication/manuscript.Rmd
Continue to revise results and tables

@@
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
-A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.
+A total of `r m30d` patients died within 30 days after arrival to the emergency department.
 
@@
 
-## Feasibility Outcomes
+## Outcomes
+
 <!--Out of X potentially eligible patients, X (X%) patients were enrolled
@@
 
-## Patient Participant Outcomes
+After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%.
+Absolute and relative changes from baseline, as well as absolute and relative risk differences, are shown in Table \@ref(tab:outcomes).
+Overall, mortality was lower in the ATLS^®^ and PTC arms compared to the standard care arm.
+The results for all other outcomes are shown in Supplementary Materials.
 
-After training, a total of `r m30d.standard.care` patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS^®^ arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units.
-The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS^®^ arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.
-
-The absolute risk difference between the standard care and ATLS^®^ arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS^®^ and PTC arms (95% CI) `r arr.atls.ptc`% units.
-The relative risk in the standard care arm compared with the ATLS^®^ arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS^®^ arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`.
-<!--The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.-->
-
-## Resident Participant Outcomes
-
-A total of `r n.residents` residents were trained during this
-pilot study.
-Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident, across trial arms and both before and after training.
+```{r outcomes, echo = FALSE}
+outcomes.table
+```
 

### 2024-08-27 pilot-publication/main.R

@@ -139,6 +139,10 @@ m30d <- inline_text(overall.sample.characteristics.table,
 )
 
 ## Crude 30 day mortality
+m30d.standard.care <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "Standard care"
+)
 m30d.standard.care.n <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "Standard care",
@@ -148,6 +152,10 @@ m30d.standard.care.p <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "Standard care",
     pattern = "({p})%"
+)
+m30d.atls <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "ATLS"
 )
 m30d.atls.n <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
@@ -159,6 +167,10 @@ m30d.atls.p <- inline_text(post.training.characteristics.table,
     column = "ATLS",
     pattern = "({p})%"
 )
+m30d.ptc <- inline_text(post.training.characteristics.table,
+    variable = outcomes__alive_after_30_days,
+    column = "PTC"
+)
 m30d.ptc.n <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "PTC",
@@ -169,6 +181,57 @@ m30d.ptc.p <- inline_text(post.training.characteristics.table,
     column = "PTC",
     pattern = "({p})%"
 )
+
+# Crude in-hospital mortality
+inhosp.standard.care <- inline_text(post.training.characteristics.table,
+    variable = outcomes__discharge_alive,
+    column = "Standard care"
+)
+inhosp.standard.care.n <- inline_text(post.training.characteristics.table,
+    variable = outcomes__discharge_alive,
+    column = "Standard care",
+    pattern = "{n}"
+)
+inhosp.standard.care.p <- inline_text(post.training.characteristics.table,
+    variable = outcomes__discharge_alive,
+    column = "Standard care",
+    pattern = "({p})%"
+)
+inhosp.atls <- inline_text(post.training.characteristics.table,
+    variable = outcomes__discharge_alive,
+    column = "ATLS"
+)
+inhosp.atls.n <- inline_text(post.training.characteristics.table,
[…]

### 2024-09-26 pilot-publication/manuscript.Rmd
Revise manuscript

@@
 Absolute and relative changes from baseline, as well as absolute and relative risk differences, are shown in Table \@ref(tab:outcomes).
-Overall, mortality was lower in the ATLS^®^ and PTC arms compared to the standard care arm.
+Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
 The results for all other outcomes are shown in Supplementary Materials.
@@
 ```{r outcomes, echo = FALSE}
-outcomes.table
+outcomes.table.with.differences
 ```
@@
 
-We found that the ATLS^®^ arm had lower 30-day mortality compared to the
-PTC and standard care arms.
+We found that the ATLS^®^ arm had lower 30-day mortality compared to the PTC and standard care arms.
 We also found that the PTC arm had lower mortality than the standard care arm.
@@
 The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
-This difference resulted from the randomisation process with a small number of heterogeneous clusters.
+This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.
 
-The primary patient participant outcome, all-cause 30-day mortality,
-was missing in `r rate.lost.to.follow.up`% of patients. This may
-appear high, especially compared to for example the CRASH-2 and
-REACT-2 trials, which report missing primary outcome in less than
-0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma
-trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
-primary outcome measure, whereas we attempted to follow up patients
-after discharge. Our missing data rate for in-hospital mortality was
-only `r p.missing.in.hospital.mortality`%, comparable to previous trials.
+All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients.
+This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016].
+Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
+primary outcome measure, whereas we attempted to follow up patients after discharge.
+Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, comparable to previous trials.
 
@@
 
-During the course of this pilot we deviated from the protocol in
-multiple ways, and provide a detailed list as Supplementary
-material. Some key limitations of this pilot and therefore lessons to
-be learned and factored into the design of the full-scale trial
-include the lower than expected enrolment rates of some centres,
-centre specific management routines, and difficulties in collecting
-data on complications and cause of death. We minimised the impact of
-the lower than expected enrolment rates by including a seventh centre,
-but on-site observations of patient volumes are likely to be needed
-for the full-scale trial. We decided to be pragmatic in selecting
-which residents to train and how to structure the data collection
-depending on how and by whom patients were initially managed, but this
-flexibility will need to be built into the full-scale trial
-protocol. Finally, we found that data on complications and cause of
-death were hard to identify and therefore the full-scale trial will
-need to include longer training of research officers.
+During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
+material.
+Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.
+
+We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial.
+We decided to be pragmatic in selecting which residents to train and how to structure the data collection
+depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
+Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.
 
@@
 
-Previous studies on the effect of in-hospital trauma life support
-training on patient outcomes are observational or quasi-experimental
-without a control group, with heterogeneous results
-[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008;
-@Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
-Cambodia found no significant effect on patient mortality after
-implementing in-hospital trauma life support training
-[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008],
-whereas one study from China that included 820 patients found a
-significant reduction in mortality, from 20 to 15%, after implementing
+Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010].
+Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008], whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15%, after implementing
 ATLS^®^ [@Wang2010].
@@
 Several controlled trials, including at least two randomized
-controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated
-with improved provider skills [@Mohammad2013]. Observational evidence
-indicates that PTC also leads to improved provider skills
-[@Kadhum2020]. The missing link is then how, and if, these improved
-skills translate into improved patient outcomes. As trauma care
-providers we assume, and probably rightly so, that we deliver better
-care if we train. The question is then how we should train, especially
-considering the costs associated with some of the programs
-offered.
+controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
+Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
+The missing link is then how, and if, these improved skills translate into improved patient outcomes.
+As trauma care providers we assume, and probably rightly so, that we deliver better
+care if we train. The question is then how we should train, especially considering the costs associated with some of the programs offered.
 
-We conclude that a full-scale cluster randomised trial should be
-feasible after incorporating the lessons of this pilot, and that this
-full-scale trial should, regardless of its outcome, influence how we
-train trauma care providers in the future.
+We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot, and that this full-scale trial should, regardless of its outcome, influence how we train trauma care providers in the future.
 
@@
 # Funding
-Doctors for You through grants awarded to Karolinska Institutet by the
-Swedish Research Council (grant number 2020-03779) and the Laerdal
-Foundation (grant number 2021-0048).
+Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).
 
@@
 ## Trial Registration
-We intended to also register our trial with Clinical Trials Registry -
-India but because of time constrains we had to initiate the study
-before registration was finalised, and Clinical Trials Registry -
-India only allow prospective registration.
+We intended to also register our trial with Clinical Trials Registry - India but because of time constrains we had to initiate the study before registration was finalised, and Clinical Trials Registry - India only allow prospective registration.
 
@@
 ## Number of Participating Centres
-We ended up recruiting seven centres instead of six and therefore
-assigned two centres each to the intervention arms and three centres
-to the control arm.
+We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
 
@@
 ## Periodic suverys to residents
-We did not distribute periodic surveys to the participating residents
-but discussed challenges and suggestions that they had regarding the
-scheduling or implementation of the training programs.
+We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.
 
 ## Follow up of residents
-We stated that resident participants would be followed up 30 days after training, if
-they are in the intervention arms, or 30 days after the study started,
-if they are in the control arm, but the intervention period was three months.
+We stated that resident participants would be followed up 30 days after training, if they are in the intervention arms, or 30 days after the study started, if they are in the control arm, but the intervention period was three months.
 
 ## Data collection from records
-We decided to record data from records only for a subset of patients
-to reduce the research officers' workload.
+We decided to record data from records only for a subset of patients to reduce the research officers' workload.
 
 ## Selection of units for training
-We planned to use simple random sampling to select units if there were
-more than two eligible units in a hospital but instead the hospital
-principal investigator decided which units to train.
+We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.
 
@@
 We had initially planned to ask residents for consent before
-randomisation, but the units were only finalised after the hospitals
-had been randomised, and residents were therefore approached for
-consent afterwards.
+randomisation, but the units were only finalised after the hospitals had been randomised, and residents were therefore approached for consent afterwards.
 
@@
 \pagebreak
+
 # Supplementary material

### 2024-09-26 pilot-publication/functions/create_descriptive_table.R

@@ -56,7 +56,11 @@ create_descriptive_table <- function(table.data,
         data = table.data,
         by = strata,
         type = type,
-        missing_text = "Missing"
+        missing_text = "Missing",
+        digits = list(
+            all_categorical() ~ c(0, 0),
+            all_continuous() ~ c(0, 0)
+        )
     )
 
     ## Add overall column

### 2024-09-27 pilot-publication/manuscript.Rmd
Work on discussion

@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
-The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
+The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
 Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
@@
 ```{r observed-vs-retrospective, echo = FALSE}
-retrospective.comparison.table %>%
-  modify_caption("Differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview.")
+observed.vs.extracted.comparison.table %>%
+  modify_caption("Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview.")
 ```
@@
 # Discussion
-We show that conducting a cluster randomized controlled trial
-comparing ATLS^®^ with PTC and standard care is feasible in this setting.
+We show that it is feasible to conduct and collect data for  a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care.
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables had very high missing data rates and may not be feasible to include in a full-scale trial.
+Some variables, especially cause of death and complications (reported in Supplementary materials) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Similarly, the missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data
 

### 2024-09-27 pilot-publication/functions/create_observed_vs_extracted_comparison_table.R

@@ -1,11 +1,11 @@
-#' Create a retrospective comparison table
+#' Create a extracted comparison table
 #'
 #' This function takes in a data frame and creates a comparison table
-#' between retrospective and directly observed data.
+#' between directly observed data and data extracted from medical records.
 #' @param data A data frame containing the data for comparison.
 #' @return A comparison table with descriptive statistics.
 #' @export
-create_retrospective_comparison_table <- function(data) {
+create_observed_vs_extracted_comparison_table <- function(data) {
     ## Define borrowed functions
     `%>%` <- magrittr::`%>%`
 
@@ -13,26 +13,26 @@ create_retrospective_comparison_table <- function(data) {
     assertthat::assert_that(is.data.frame(data))
 
     ## Get variables to compare
-    retrospective.variables <- names(data)[grep("_r__", names(data), fixed = TRUE)]
-    directly.observed.variables <- retrospective.variables %>%
+    extracted.variables <- names(data)[grep("_r__", names(data), fixed = TRUE)]
+    directly.observed.variables <- extracted.variables %>%
         stringr::str_replace_all(pattern = "\\_r\\_\\_", replacement = "\\_\\_") %>%
         stringr::str_replace_all(pattern = "\\_r$", replacement = "")
 
-    ## Keep only observations with completed retrospective data collection
+    ## Keep only observations with completed extracted data collection
     data <- data %>% dplyr::filter(data_status_r__collection_completed_r == "Yes")
 
-    ## Extract retrospective and directly observed data
-    retrospective.data <- data[, retrospective.variables]
-    colnames(retrospective.data) <- directly.observed.variables
+    ## Extract extracted and directly observed data
+    extracted.data <- data[, extracted.variables]
+    colnames(extracted.data) <- directly.observed.variables
     directly.observed.data <- data[, directly.observed.variables]
-    retrospective.data$collection.mode <- "Retrospective"
+    extracted.data$collection.mode <- "Medical records"
     directly.observed.data$collection.mode <- "Directly observed"
     original.classes <- sapply(directly.observed.data, class)
 
     ## Merge datasets
     directly.observed.data[] <- lapply(directly.observed.data, as.character)
-    retrospective.data[] <- lapply(retrospective.data, as.character)
-    merged.data <- dplyr::bind_rows(directly.observed.data, retrospective.data)
+    extracted.data[] <- lapply(extracted.data, as.character)
[…]

### 2024-09-27 pilot-publication/functions/prepare_data.R

@@ -35,7 +35,12 @@ prepare_data <- function(data, codebook = NULL) {
         naniar::replace_with_na_if(
             .predicate = is.numeric,
             condition = ~ .x == 999
+        ) %>%
+        naniar::replace_with_na_if(
+            .predicate = is.numeric,
+            condition = ~ .x < 0
         )
+
     ## Deal with edge cases
     variable <- prepared.data$complications__failure_of_conservative_management
     variable[variable == "0" | variable == "NO" | variable == "no"] <- "No"

### 2024-09-27 pilot-publication/main.R

@@ -94,7 +94,7 @@ combined.sample.characteristics.table <- gtsummary::tbl_merge(
     gtsummary::modify_caption("Patient sample characteristics")
 
 ## Create table comparing directly observed with retrospective data
-retrospective.comparison.table <- create_retrospective_comparison_table(data)
+observed.vs.extracted.comparison.table <- create_observed_vs_extracted_comparison_table(data)
 
 ## Bootstrap outcome results
 use.saved <- TRUE

### 2024-09-28 pilot-publication/manuscript.Rmd
Respond to all comments

@@
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-Several systematic reviews call for trials in settings where these programmes are not routinely taught, but data needed to inform the planning of such a trial is missing.
-We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
+Several studies, including at least two randomised controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
+Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
+The missing link is then how, and if, these improved skills translate into improved patient outcomes.
+
+Systematic reviews call for trials in settings where these programmes are not routinely implemented, but data needed to inform the planning of such a trial is missing.
+We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
 
@@
 The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
-<!--Supplemental Material 2 contains a full variable list, with definitions.-->
+For a subset of patients we also extracted data from medical records, to be able to compare their distributions with directly observed data.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
 
@@
 
-We found that the ATLS^®^ arm had lower 30-day mortality compared to the PTC and standard care arms.
-We also found that the PTC arm had lower mortality than the standard care arm.
-These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
+We found that the ATLS^®^ and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
+These findings hints towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
 The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
 This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.
+This heterogeneity highlights the importance of taken different cluster sizes into account in the design of the full scale trial.
 
@@
 
-Several controlled trials, including at least two randomized
-controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
-Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is then how, and if, these improved skills translate into improved patient outcomes.
-As trauma care providers we assume, and probably rightly so, that we deliver better
-care if we train. The question is then how we should train, especially considering the costs associated with some of the programs offered.
-
-We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot, and that this full-scale trial should, regardless of its outcome, influence how we train trauma care providers in the future.
+We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
 
@@
 ## Trial Registration
-We intended to also register our trial with Clinical Trials Registry - India but because of time constrains we had to initiate the study before registration was finalised, and Clinical Trials Registry - India only allow prospective registration.
+We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
 
@@
 ## Follow up of residents
-We stated that resident participants would be followed up 30 days after training, if they are in the intervention arms, or 30 days after the study started, if they are in the control arm, but the intervention period was three months.
+We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.
 
 ## Data collection from records
-We decided to record data from records only for a subset of patients to reduce the research officers' workload.
+We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.
 
@@
 We had initially planned to ask residents for consent before
-randomisation, but the units were only finalised after the hospitals had been randomised, and residents were therefore approached for consent afterwards.
+randomisation, but this was not possible because of logistical issues the units were only finalised after the hospitals had been randomised.
+Residents were therefore approached for consent after randomisation but before training.
 

### 2024-09-28 pilot-publication/manuscript.Rmd
Work on creating a new consort style diagram

@@
 ---
+
+```{dot}
+digraph consort {
+
+  # Node attributes
+  node [shape=square, fontname=Helvetica, fontsize=10]
+
+  # Define nodes
+  A [label='Total: Clusters: 30 | Residents: 90 | Patients: 270']
+  B [label='Standard Care\nClusters: 10\nResidents: 30\nPatients: 90']
+  C [label='ATLS\nClusters: 10\nResidents: 30\nPatients: 90']
+  D [label='PTC\nClusters: 10\nResidents: 30\nPatients: 90']
+
+  # Define mid-point node to facilitate splitting
+  mid [shape=point, width=0, label='']
+
+  # Edges
+  A -> mid [arrowhead=none]  # Line from A to mid-point without arrowhead
+  mid -> B [tailport=s, headport=n]  # Line from mid to Standard Care
+  mid -> C [tailport=s, headport=n]  # Line from mid to ATLS
+  mid -> D [tailport=s, headport=n]  # Line from mid to PTC
+}
+```
 

### 2024-09-28 pilot-publication/functions/create_consort_diagram.R

@@ -1,0 +1,27 @@
+create_consort_diagram <- function(data) {
+    library(DiagrammeR)
+
+    # Create the flow diagram using grViz
+    grViz("
+digraph consort {
+
+  # Set node shape and font attributes
+  node [shape=square, fontname=Helvetica, fontsize=10]
+
+  # Nodes
+  A [label='Total: Clusters: 30 | Residents: 90 | Patients: 270']
+  B [label='Standard Care\nClusters: 10\nResidents: 30\nPatients: 90']
+  C [label='ATLS\nClusters: 10\nResidents: 30\nPatients: 90']
+  D [label='PTC\nClusters: 10\nResidents: 30\nPatients: 90']
+
+  # Mid-point for the split
+  mid [shape=point, width=0, label='']
+
+  # Edges
+  A -> mid [arrowhead=none]  # Straight line down, no arrowhead
+  mid -> B [tailport=s, headport=n]  # Connector to Standard Care
+  mid -> C [tailport=s, headport=n]  # Connector to ATLS
+  mid -> D [tailport=s, headport=n]  # Connector to PTC
+}
+")
+}

### 2024-09-28 pilot-publication/functions/prepare_data.R

@@ -139,6 +139,9 @@ prepare_data <- function(data, codebook = NULL) {
     prepared.data[as.character(prepared.data$id__reg_hospital_id) %in% control.centres, ]$arm <- "Standard care"
     prepared.data$arm <- factor(prepared.data$arm, levels = c("Standard care", "ATLS", "PTC"))
 
+    ## Add variable indicating if the patient was considered lost to follow up
+    prepared.data$lost.to.follow.up <- is.na(prepared.data$outcomes__alive_after_30_days)
+
     ## Label variables
     prepared.data[] <- lapply(names(prepared.data), function(column.name) {
         column.data <- prepared.data %>%

### 2024-10-03 pilot-publication/functions/create_outcome_results_tables.R
Work on supplementary material tables

@@ -137,6 +137,9 @@ create_ci_level_tables <- function(table.name, ci.data) {
         gt::tab_style(
             style = gt::cell_text(weight = "bold"),
             locations = gt::cells_column_labels()
+        ) %>%
+        gt::tab_options(
+            table.font.size = 10
         )
 
     ## Return table

### 2024-10-03 pilot-publication/functions/save_tables_to_file.R

@@ -21,7 +21,8 @@ save_tables_to_file <- function(outcome.results.tables) {
                 scales::percent()
             ci.level.tables <- outcome.results.tables[[ci.level.tables.name]]
             paste0(
-                "# ", ci.level, " confidence level\n\n",
+                "\\pagebreak\n\n",
+                "# Results with ", ci.level, " confidence level\n\n",
                 lapply(names(ci.level.tables), function(ci.level.table.name) {
                     ci.level.table <- ci.level.tables[[ci.level.table.name]]
                     paste0(
@@ -29,7 +30,8 @@ save_tables_to_file <- function(outcome.results.tables) {
                         "\`\`\`{r, echo = FALSE} \n\n",
                         "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",
                         "[[\"", ci.level.table.name, "\"]]", "\n\n",
-                        "\`\`\`\n\n"
+                        "\`\`\`\n\n",
+                        "\\pagebreak\n\n",
                     )
                 }) %>%
                     paste0(collapse = "\n\n")
@@ -45,5 +47,5 @@ save_tables_to_file <- function(outcome.results.tables) {
     close(fileConn)
 
     ## Create html document
-    rmarkdown::render("outcome-tables.Rmd", output_format = rmarkdown::html_document(toc = TRUE))
+    rmarkdown::render("outcome-tables.Rmd", output_format = rmarkdown::pdf_document(toc = TRUE))
 }

### 2024-10-04 pilot-publication/manuscript.Rmd
Add flowchart to manuscript
Try to compile revised manuscript, work on table errors

--- 2024-09-28 pilot-publication/manuscript.Rmd
@@
 
-```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care."}
-create_flow_diagram()
-knitr::include_graphics("flow-diagram.png")
+```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care.", out.width="50%"}
+knitr::include_graphics("consort-flow-diagram.png")
 ```
@@
 
-<!--Out of X potentially eligible patients, X (X%) patients were enrolled
-in the study-->.
 The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
-The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective).
+The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-extracted).
 Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
 
-```{r observed-vs-retrospective, echo = FALSE}
+```{r observed-vs-extracted, echo = FALSE}
 observed.vs.extracted.comparison.table %>%
@@
 After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%.
-Absolute and relative changes from baseline, as well as absolute and relative risk differences, are shown in Table \@ref(tab:outcomes).
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%, as shown in Table \@ref(tab:outcomes).
 Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
@@
 ```{r outcomes, echo = FALSE}
-outcomes.table.with.differences
+outcome.table.with.differences
 ```
@@
 This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016].
-Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
-primary outcome measure, whereas we attempted to follow up patients after discharge.
+Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge.
 Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, comparable to previous trials.
@@
 We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial.
-We decided to be pragmatic in selecting which residents to train and how to structure the data collection
-depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
+We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
 Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.
@@
 
-We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
+We conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
 

### 2024-10-04 pilot-publication/manuscript.tex
Try to compile revised manuscript, work on table errors

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial},
  pdfauthor={Trauma life support training Effectiveness Research Network (TERN) collaborators},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial}
\author{Trauma life support training Effectiveness Research Network (TERN) collaborators}
\date{}

\begin{document}
\maketitle

\hypertarget{trial-registration}{%
\subsection{Trial registration}\label{trial-registration}}

This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).

\hypertarget{corresponding-author}{%
\subsection{Corresponding author}\label{corresponding-author}}

Martin Gerdin Wärnberg (MGW)

Department of Global Public Health\\
Karolinska Institutet\\
171 77 Stockholm\\
Sweden

\href{mailto:martin.gerdin@ki.se}{\nolinkurl{martin.gerdin@ki.se}}\\
+46 708 53 95 98 (mobile)

\hypertarget{authors}{%
\subsection{Authors}\label{authors}}

\hypertarget{johanna-berg-jb}{%
\subparagraph{Johanna Berg (JB)}\label{johanna-berg-jb}}

Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

Emergency Medicine, Department of Internal and Emergency Medicine, Skåne University Hospital, Malmö, Sweden.

\hypertarget{shamita-chatterjee-sc}{%
\subparagraph{Shamita Chatterjee (SC)}\label{shamita-chatterjee-sc}}

Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.

\hypertarget{li-felluxe4nder-tsai-lft}{%
\subparagraph{Li Felländer-Tsai (LFT)}\label{li-felluxe4nder-tsai-lft}}

Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.

Department of Reconstructive Orthopedics, Karolinska University Hospital, Stockholm, Sweden.

\hypertarget{geeta-ghag-gg}{%
\subparagraph{Geeta Ghag (GG)}\label{geeta-ghag-gg}}

Department of Surgery, HBT Medical College And Dr.~R N Cooper Municipal General Hospital, Mumbai, India.

\hypertarget{monty-khajanchi-mk}{%
\subparagraph{Monty Khajanchi (MK)}\label{monty-khajanchi-mk}}

WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.

\hypertarget{catherine-juillard-cj}{%
\subparagraph{Catherine Juillard (CJ)}\label{catherine-juillard-cj}}

Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.

\hypertarget{nobhojit-roy-nr}{%
\subparagraph{Nobhojit Roy (NR)}\label{nobhojit-roy-nr}}

WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.

Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

\hypertarget{rajdeep-singh-rs}{%
\subparagraph{Rajdeep Singh (RS)}\label{rajdeep-singh-rs}}

Department of Surgery, Maulana Azad Medical College, New Delhi, India.

\hypertarget{kapil-dev-soni-kds}{%
\subparagraph{Kapil Dev Soni (KDS)}\label{kapil-dev-soni-kds}}

Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.

\hypertarget{lovisa-struxf6mmer-ls}{%
\subparagraph{Lovisa Strömmer (LS)}\label{lovisa-struxf6mmer-ls}}

Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.

Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.

\hypertarget{vipul-nandu}{%
\subparagraph{Vipul Nandu}\label{vipul-nandu}}

Department of Surgery, HBT Medical College And Dr.~R N Cooper Municipal General Hospital, Mumbai, India.

\hypertarget{debojit-basak}{%
\subparagraph{Debojit Basak}\label{debojit-basak}}

Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.

\hypertarget{tamal-khan}{%
\subparagraph{Tamal Khan}\label{tamal-khan}}

\hypertarget{non-author-contributors}{%
\subsection{Non-author contributors}\label{non-author-contributors}}

Trauma life support Effectiveness Research Network (TERN) collaborators:

\hypertarget{prashant-bhandarkar-pb}{%
\subparagraph{Prashant Bhandarkar (PB)}\label{prashant-bhandarkar-pb}}

WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

\hypertarget{anirban-chatterjee-ac}{%
\subparagraph{Anirban Chatterjee (AC)}\label{anirban-chatterjee-ac}}

Department of Orthopaedic Sciences, Medica Superspecialty Hospital, Kolkata, India.

\hypertarget{chintamani-c}{%
\subparagraph{Chintamani (C)}\label{chintamani-c}}

Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.

\hypertarget{anita-gadgil-ag}{%
\subparagraph{Anita Gadgil (AG)}\label{anita-gadgil-ag}}

WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

\hypertarget{marie-hasselberg-mh}{%
\subparagraph{Marie Hasselberg (MH)}\label{marie-hasselberg-mh}}

Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

\hypertarget{deepa-kizhakke-veetil-dkv}{%
\subparagraph{Deepa Kizhakke Veetil (DKV)}\label{deepa-kizhakke-veetil-dkv}}

Department of Surgery, Manipal Hospital, Dwarka, New Delhi.

\hypertarget{vineet-kumar-vk}{%
\subparagraph{Vineet Kumar (VK)}\label{vineet-kumar-vk}}

Department of Surgery, Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai, India.

\hypertarget{debabrata-kundu-dk}{%
\subparagraph{Debabrata Kundu (DK)}\label{debabrata-kundu-dk}}

Department of Surgery, Medical College Kolkata, Kolkata, India.

\hypertarget{anurag-mishra-am}{%
\subparagraph{Anurag Mishra (AM)}\label{anurag-mishra-am}}

Department of Surgery, Maulana Azad Medical College, New Delhi, India.

\hypertarget{priti-patil-pp}{%
\subparagraph{Priti Patil (PP)}\label{priti-patil-pp}}

WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.

\hypertarget{arun-prasad-ap}{%
\subparagraph{Arun Prasad (AP)}\label{arun-prasad-ap}}

Department of Surgery, Indraprastha Apollo Hospitals, New Delhi

\hypertarget{amit-roy-ar}{%
\subparagraph{Amit Roy (AR)}\label{amit-roy-ar}}

Department of Surgery, Sir Nil Ratan Sircar Medical College \& Hospital, Kolkata, India.

\hypertarget{siddarth-david-sd}{%
\subparagraph{Siddarth David (SD)}\label{siddarth-david-sd}}

Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.

Doctors For You, Mumbai, India.

\hypertarget{harris-solomon-hs}{%
\subparagraph{Harris Solomon (HS)}\label{harris-solomon-hs}}

Department of Cultural Anthropology and the Duke Global Health Institute, Duke University, Durham, USA.

\hypertarget{megha-tandon-mt}{%
\subparagraph{Megha Tandon (MT)}\label{megha-tandon-mt}}

Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.

\hypertarget{keywords}{%
\subsection{Keywords}\label{keywords}}

Trauma management, Accident and emergency medicine, Education and training.

\hypertarget{role-of-study-sponsor-and-funders}{%
\subsection{Role of study sponsor and funders}\label{role-of-study-sponsor-and-funders}}

The funding sources had no role in the design of this study nor during its execution, analyses, interpretation of the data, or decision to submit the results.

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\textbf{Importance} Trauma life support training programmes aim to improve early trauma care but there is high-quality evidence to show that they improve patient outcomes.

\textbf{Objective} To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support\textsuperscript{®} (ATLS\textsuperscript{®}) and Primary Trauma Care (PTC) with standard care on patient outcomes.

\textbf{Design} A pilot pragmatic three-armed parallel, cluster randomised, controlled trial between April 2022 and February 2023. Patient follow up was 30 days.

\textbf{Setting} Tertiary care hospitals across metropolitan areas in India.

\textbf{Participants} Adult trauma patients and residents managing these patients.

\textbf{Interventions} ATLS\textsuperscript{®} or PTC training for residents in the intervention arms.

\textbf{Main Outcomes and Measures} The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records. The primary patient outcome was all cause mortality at 30 days from the time of arrival to the emergency department.

\textbf{Results} Two hospitals were randomised to ATLS\textsuperscript{®}, two to PTC, and three to standard care. We included 376 patients and 22 residents. The percentage of patients who consented to follow up was 77\% and the percentage of residents who consented to training was 100\%. The lost to follow up rate was 14\%. The pass rate was 100\%. The missing data was overall low for key variables. Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data. Twenty-two (16)\% patients died within 30 days in the standard care arm, one (4)\% patients in the ATLS\textsuperscript{®} arm, and three (5)\% patients in the PTC arm.

\textbf{Conclusions and Relevance} Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS\textsuperscript{®}, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

\textbf{Trial Registration} ClinicalTrials.gov (reg. no NCT05417243)

\newpage

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year\textsuperscript{1}. Several trauma life support training programs have been developed to improve the early management of patients as they arrive at hospital by providing a structured framework to assessment and treatment\textsuperscript{2--4}.

The proprietary Advanced Trauma Life Support\textsuperscript{®} (ATLS\textsuperscript{®}) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries\textsuperscript{5,6}. Observational studies indicate that these programmes may improve patient outcomes\textsuperscript{7}, but there is no high quality evidence from controlled trials to support this\textsuperscript{2--4}.

Several studies, including at least two randomised controlled trials\textsuperscript{8,9}, show that ATLS\textsuperscript{®} is associated with improved provider skills\textsuperscript{2}. Observational evidence indicates that PTC also leads to improved provider skills\textsuperscript{4}. The missing link is then how, and if, these improved skills translate into improved patient outcomes.

Systematic reviews call for trials in settings where these programmes are not routinely implemented, but data needed to inform the planning of such a trial is missing. We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS\textsuperscript{®} and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{trial-design}{%
\subsection{Trial Design}\label{trial-design}}

We piloted a three-armed cluster randomised controlled trial\textsuperscript{10}. There were a standard care arm and two intervention arms, ATLS\textsuperscript{®} and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). This design allowed us to assess outcomes both as final values and as change from baseline.

\hypertarget{study-setting}{%
\subsection{Study Setting}\label{study-setting}}

We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS\textsuperscript{®}, PTC, nor any other trauma life support training program is routinely taught.

\hypertarget{standard-care}{%
\subsection{Standard Care}\label{standard-care}}

Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units. These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they manage the emergency department. In the emergency department, trauma patients are initially assessed by first- or second-year residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations. Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.

\hypertarget{intervention}{%
\subsection{Intervention}\label{intervention}}

In each intervention arm the residents in one or two units were trained in either ATLS\textsuperscript{®} or PTC. For the purpose of this pilot study, our target was to train a minimum of 75\% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS\textsuperscript{®} training was conducted in an ATLS\textsuperscript{®} certified training centre in Mumbai, according to the standard ATLS\textsuperscript{®} curriculum\textsuperscript{5}. The PTC training was conducted in New Delhi, according to the standard PTC curriculum\textsuperscript{6}. We did not modify or adapt the delivery or content of these programs during this pilot study.

\hypertarget{eligibility-criteria-for-cluster-and-participants}{%
\subsection{Eligibility Criteria for Cluster and Participants}\label{eligibility-criteria-for-cluster-and-participants}}

\hypertarget{hospitals}{%
\subsubsection{Hospitals}\label{hospitals}}

We included tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.

\hypertarget{clusters}{%
\subsubsection{Clusters}\label{clusters}}

We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals. To be eligible, units could have no more than 25\% of their physicians trained in either ATLS\textsuperscript{®}, PTC, or similar training programs before the start of the pilot study. Those residents who had received training in the last five years were considered as trained. The figure of 25\% was decided through consensus in the research team, to balance feasibility and contamination of results. The principal investigator at each hospital selected the units for training. We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.

\hypertarget{residents}{%
\subsubsection{Residents}\label{residents}}

Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training. Consent was sought from the residents in each of the intervention groups before they underwent the ATLS\textsuperscript{®} or PTC training.

\hypertarget{patients}{%
\subsubsection{Patients}\label{patients}}

All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty. History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.

\hypertarget{outcomes}{%
\subsection{Outcomes}\label{outcomes}}

We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A complete list of outcomes is available as Supplementary Materials. Our main outcomes were:

\begin{itemize}
\tightlist
\item
  Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents.
\item
  Lost to follow up rate. This applied only to patients and was equal to the percentage of patients whom did not complete 30 day follow up, out of all enrolled patients.
\item
  Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents that passed the training programme, out of the total number of trained residents.
\item
  Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
\item
  Differences in distributions of observed and extracted data. This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
\item
  All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients, measured and compared across trial arms as both final values and as change from baseline.
\end{itemize}

\hypertarget{participant-timeline-and-inclusion}{%
\subsection{Participant Timeline and Inclusion}\label{participant-timeline-and-inclusion}}

\hypertarget{patients-1}{%
\subsubsection{Patients}\label{patients-1}}

Arriving patients were screened for eligibility and consented, if conscious. Unconscious patients were consented by the patient's representative. This proxy consent was reaffirmed by the patient, on regaining consciousness. We followed up patients at 24 hours after arrival at the emergency department, and up to 30 days after arrival at the emergency department.

\hypertarget{residents-1}{%
\subsubsection{Residents}\label{residents-1}}

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The training was conducted approximately one month after the study started in that hospital.

\hypertarget{sample-size}{%
\subsection{Sample size}\label{sample-size}}

We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess feasibility. The time period was dictated by budget and time constraints.

\hypertarget{allocation-and-blincing}{%
\subsection{Allocation and blincing}\label{allocation-and-blincing}}

We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.

\hypertarget{data-collection}{%
\subsection{Data Collection}\label{data-collection}}

Data was collected over a four-month period. A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care. The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records. We followed up admitted patients for their complications and other in-hospital outcome measures. Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.

\hypertarget{variables}{%
\subsection{Variables}\label{variables}}

The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained. For a subset of patients we also extracted data from medical records, to be able to compare their distributions with directly observed data.

\hypertarget{patient-and-public-involvement}{%
\subsection{Patient and public involvement}\label{patient-and-public-involvement}}

We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial. The results of these consultations will be published separately.

\hypertarget{data-monitoring}{%
\subsection{Data monitoring}\label{data-monitoring}}

We conducted weekly online meetings to monitor the study and data collection. We conducted one interim analysis approximately halfway through the study, and decided to complete the study as residents and patients were consenting to be included in the study and key variables including mortality outcomes could be collected. We did not use a data monitoring committee.

\hypertarget{statistical-methods}{%
\subsection{Statistical Methods}\label{statistical-methods}}

We analysed all data using descriptive statistics and did not perform any formal hypothesis\textsuperscript{11}. Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range. Qualitative variables are presented as absolute numbers and percentages. We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient.

We compared patients outcomes in all possible combinations of trial arms. In each combination we compared both differences in final values and differences in change from baseline. For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training. For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.

Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly. These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.

\hypertarget{ethics-and-dissemination}{%
\subsection{Ethics and Dissemination}\label{ethics-and-dissemination}}

We were granted research ethics approval from the institutional ethics committees at each participating hospital.

\hypertarget{results}{%
\section{Results}\label{results}}

We enrolled 376 trauma patients from 7 participating centres between April 2022 and February 2023. The standard care arm 202 patients, the ATLS\textsuperscript{®} arm enrolled 44 patients, and the PTC arm 130 patients. We trained a total of 22 residents, 7 in ATLS\textsuperscript{®}, and 15 in PTC.

The study flow diagram is shown in Figure \ref{fig:flow-diagram} and patient sample characteristics across trial arms are shown in Table \ref{tab:sample-characteristics}. Overall, the number of females were 86 (23\%), the median (IQR) age was 33 (24, 46) years, and the median ISS (IQR) was 1 (0, 5). A total of 32 (10\%) patients died within 30 days after arrival to the emergency department.

\begin{figure}
\includegraphics[width=0.5\linewidth]{consort-flow-diagram} \caption{Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care.}\label{fig:flow-diagram}
\end{figure}

\begingroup
\fontsize{12.0pt}{14.4pt}\selectfont
\setlength{\LTpost}{0mm}
\begin{longtable}{lcccc}
\toprule
\textbf{Characteristic}\textsuperscript{\textit{1}} & \textbf{Standard care}  N = 202 & \textbf{ATLS}  N = 44 & \textbf{PTC}  N = 130 & \textbf{Overall}  N = 376 \\ 
\midrule\addlinespace[2.5pt]
Age, years, median (IQR) & 35 (25, 47) & 40 (30, 57) & 30 (22, 38) & 33 (24, 46) \\ 
Elderly, n (\%) & 15 (7\%) & 6 (14\%) & 5 (4\%) & 26 (7\%) \\ 
Sex, n (\%) &  &  &  &  \\ 
    Male & 160 (79\%) & 33 (75\%) & 97 (75\%) & 290 (77\%) \\ 
    Female & 42 (21\%) & 11 (25\%) & 33 (25\%) & 86 (23\%) \\ 
Dominating injury type, n (\%) &  &  &  &  \\ 
    Penetrating & 13 (6\%) & 3 (7\%) & 1 (1\%) & 17 (5\%) \\ 
    Blunt & 189 (94\%) & 41 (93\%) & 129 (99\%) & 359 (95\%) \\ 
Blunt multisystem trauma, n (\%) & 2 (1\%) & 2 (5\%) & 6 (5\%) & 10 (3\%) \\ 
Severe traumatic brain injury, n (\%) & 10 (5\%) & 1 (2\%) & 5 (4\%) & 16 (4\%) \\ 
    Missing & 1 & 0 & 0 & 1 \\ 
Shock, n (\%) & 4 (2\%) & 2 (5\%) & 4 (3\%) & 10 (3\%) \\ 
    Missing & 7 & 3 & 4 & 14 \\ 
Respiratory rate, breaths per minute, median (IQR) & 20 (18, 22) & 21 (20, 24) & 21 (20, 24) & 20 (19, 23) \\ 
    Missing & 7 & 0 & 5 & 12 \\ 
Oxygen saturation, \%, median (IQR) & 98 (97, 99) & 98 (97, 99) & 98 (98, 99) & 98 (97, 99) \\ 
    Missing & 1 & 1 & 0 & 2 \\ 
Heart rate, beats per minute, median (IQR) & 86 (80, 96) & 87 (73, 100) & 90 (76, 104) & 86 (78, 100) \\ 
    Missing & 1 & 1 & 1 & 3 \\ 
Systolic blood pressure, mmHg, median (IQR) & 123 (112, 135) & 124 (113, 131) & 122 (111, 136) & 123 (112, 135) \\ 
    Missing & 7 & 3 & 4 & 14 \\ 
Glasgow Coma Scale, median (IQR) & 15 (15, 15) & 15 (15, 15) & 15 (15, 15) & 15 (15, 15) \\ 
    Missing & 2 & 1 & 0 & 3 \\ 
Injury Severity Score, median (IQR) & 1 (0, 4) & 3 (1, 4) & 2 (0, 5) & 1 (0, 5) \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}ATLS = Advanced Trauma Life Support; PTC = Prehospital Trauma Care\\
\end{minipage}
\endgroup

\hypertarget{outcomes-1}{%
\subsection{Outcomes}\label{outcomes-1}}

The percentage of patients who consented to follow up was 77\% and the percentage of residents who consented to training was 100\%. The lost to follow up rate was 14\%. The pass rate was 100\%. The missing data rate ranged from 0 to 98, with details for selected variables shown in Table \ref@(tab:sample-characteristics). The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-extracted). Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.

\begingroup
\fontsize{12.0pt}{14.4pt}\selectfont
\setlength{\LTpost}{0mm}
\begin{longtable}{lcc}
\toprule
\textbf{Characteristic} & \textbf{Directly observed}  N = 55\textsuperscript{\textit{1}} & \textbf{Medical records}  N = 55\textsuperscript{\textit{1}} \\ 
\midrule\addlinespace[2.5pt]
Age, years & 34 (27, 48) & 34 (25, 50) \\ 
    Missing & 0 & 22 \\ 
Sex &  &  \\ 
    Female & 10 (18\%) & 6 (18\%) \\ 
    Male & 45 (82\%) & 27 (82\%) \\ 
    Missing & 0 & 22 \\ 
Dominating injury type &  &  \\ 
    Blunt & 52 (95\%) & 29 (91\%) \\ 
    Penetrating & 3 (5\%) & 3 (9\%) \\ 
    Missing & 0 & 23 \\ 
Respiratory rate, breaths per minute & 21 (18, 24) & 18 (16, 20) \\ 
    Missing & 0 & 37 \\ 
Oxygen saturation, \% & 98 (98, 99) & 98 (97, 100) \\ 
    Missing & 0 & 29 \\ 
Heart rate, beats per minute & 85 (80, 98) & 87 (84, 93) \\ 
    Missing & 0 & 19 \\ 
Systolic blood pressure, mmHg & 123 (112, 136) & 118 (110, 128) \\ 
    Missing & 1 & 18 \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Median (Q1, Q3); n (\%)\\
\end{minipage}
\endgroup

After training, a total of 22 (16\%) patients in the standard care arm died within 30 days, compared to 1 (4\%) patients in the ATLS\textsuperscript{®} arm and 3 (5\%) patients in the PTC arm. The corresponding figures for in-hospital mortality were 19 (12\%)\%, 1 (4\%)\%, and 3 (4\%)\%, as shown in Table \ref{tab:outcomes}. Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS\textsuperscript{®} and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small. The results for all other outcomes are shown in Supplementary Materials.

\begingroup
\fontsize{12.0pt}{14.4pt}\selectfont
\setlength{\LTpost}{0mm}
\begin{longtable}{lcccccc}
\toprule
 &  &  &  & \multicolumn{3}{c}{\textbf{Differences}} \\ 
\cmidrule(lr){5-7}
\textbf{Outcome} & \textbf{Standard care}  N = 161\textsuperscript{\textit{1}} & \textbf{ATLS}  N = 28\textsuperscript{\textit{1}} & \textbf{PTC}  N = 73\textsuperscript{\textit{1}} & \textbf{Standard care vs. ATLS} & \textbf{Standard care vs. PTC} & \textbf{ATLS vs. PTC} \\ 
\midrule\addlinespace[2.5pt]
30 day mortality & 22 (16\%) & 1 (3.8\%) & 3 (4.9\%) & 12% & 11% & -1.1% \\ 
    Unknown & 26 & 2 & 12 &  &  &  \\ 
In-hospital mortality & 19 (12\%) & 1 (3.7\%) & 3 (4.1\%) & 8.2% & 7.8% & -0.41% \\ 
    Unknown & 2 & 1 & 0 &  &  &  \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}n (\%)\\
\end{minipage}
\endgroup

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS\textsuperscript{®} with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially cause of death and complications (reported in Supplementary materials) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. Similarly, the missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data

We found that the ATLS\textsuperscript{®} and PTC arms had lower 30-day mortality compared to the PTC and standard care arms. These findings hints towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS\textsuperscript{®} arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters. This heterogeneity highlights the importance of taken different cluster sizes into account in the design of the full scale trial.

All-cause 30-day mortality was missing in 14\% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01\% of patients\textsuperscript{12,13}. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only 1\%, comparable to previous trials.

During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
material. Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.

We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial. We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol. Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.

Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results\textsuperscript{14--18}. Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training\textsuperscript{14--17}, whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15\%, after implementing
ATLS\textsuperscript{®}\textsuperscript{18}.

We conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.

\hypertarget{contributorship-statement}{%
\section{Contributorship statement}\label{contributorship-statement}}

\hypertarget{competing-interests}{%
\section{Competing Interests}\label{competing-interests}}

Several authors are ATLS\textsuperscript{®} and/or PTC instructors.

\hypertarget{funding}{%
\section{Funding}\label{funding}}

Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).

\hypertarget{data-sharing-statement}{%
\section{Data Sharing Statement}\label{data-sharing-statement}}

The final anonymized dataset and code for analysis are released publicly.

\hypertarget{protocol-deviations}{%
\section{Protocol Deviations}\label{protocol-deviations}}

\hypertarget{trial-registration-1}{%
\subsection{Trial Registration}\label{trial-registration-1}}

We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.

\hypertarget{outcomes-across-subgroups}{%
\subsection{Outcomes across subgroups}\label{outcomes-across-subgroups}}

Because of small numbers in the pre-specified subgroups we decided to report only descriptive data on these subgroups.

\hypertarget{number-of-participating-centres}{%
\subsection{Number of Participating Centres}\label{number-of-participating-centres}}

We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.

\hypertarget{resident-participants}{%
\subsection{Resident Participants}\label{resident-participants}}

Emergency medicine in addition to surgery.

\hypertarget{periodic-suverys-to-residents}{%
\subsection{Periodic suverys to residents}\label{periodic-suverys-to-residents}}

We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.

\hypertarget{follow-up-of-residents}{%
\subsection{Follow up of residents}\label{follow-up-of-residents}}

We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.

\hypertarget{data-collection-from-records}{%
\subsection{Data collection from records}\label{data-collection-from-records}}

We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.

\hypertarget{selection-of-units-for-training}{%
\subsection{Selection of units for training}\label{selection-of-units-for-training}}

We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.

\hypertarget{timing-of-resident-consent}{%
\subsection{Timing of resident consent}\label{timing-of-resident-consent}}

We had initially planned to ask residents for consent before
randomisation, but this was not possible because of logistical issues the units were only finalised after the hospitals had been randomised. Residents were therefore approached for consent after randomisation but before training.

\hypertarget{analysis-level-of-feasibility-outcomes}{%
\subsection{Analysis level of feasibility outcomes}\label{analysis-level-of-feasibility-outcomes}}

We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level.

\pagebreak

\hypertarget{supplementary-material}{%
\section{Supplementary material}\label{supplementary-material}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-injuries2020}{}}%
\CSLLeftMargin{1. }
\CSLRightInline{GBD 2019 Diseases and Injuries Collaborators. \href{}{Injuries---level 1 cause}. \emph{The Lancet} \textbf{396}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Mohammad2013}{}}%
\CSLLeftMargin{2. }
\CSLRightInline{Mohammad, A. \emph{et al.} \href{https://doi.org/10.1007/s00268-013-2294-0}{Educational and clinical impact of advanced trauma life support (ATLS) courses: A systematic review}. \emph{World Journal of Surgery} \textbf{38}, 322--329 (2013).}

\leavevmode\vadjust pre{\hypertarget{ref-Jayaraman2014}{}}%
\CSLLeftMargin{3. }
\CSLRightInline{Jayaraman, S. \emph{et al.} \href{https://doi.org/10.1002/14651858.cd004173.pub4}{Advanced trauma life support training for hospital staff}. \emph{Cochrane Database of Systematic Reviews} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Kadhum2020}{}}%
\CSLLeftMargin{4. }
\CSLRightInline{Kadhum, M. \emph{et al.} \href{https://doi.org/10.1016/j.injury.2019.10.084}{Are primary trauma care (PTC) courses beneficial in low- and middle-income countries - a systematic review}. \emph{Injury} \textbf{51}, 136--141 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-acsAtls2018}{}}%
\CSLLeftMargin{5. }
\CSLRightInline{American College of Surgeons, C. on T. \emph{Advanced trauma life support® student course manual}. (American College of Surgeons, 2018).}

\leavevmode\vadjust pre{\hypertarget{ref-ptcfPtc2018}{}}%
\CSLLeftMargin{6. }
\CSLRightInline{Foundation, P. T. C. \emph{Primary trauma care course manual for trauma management in locations wih limited resources}. (Primary Trauma Care Foundation, 2018).}

\leavevmode\vadjust pre{\hypertarget{ref-Jin2021}{}}%
\CSLLeftMargin{7. }
\CSLRightInline{Jin, J. \emph{et al.} \href{https://doi.org/10.1007/s00268-021-06065-9}{Effectiveness of quality improvement processes, interventions, and structure in trauma systems in low- and middle-income countries: A systematic review and meta-analysis}. \emph{World Journal of Surgery} \textbf{45}, 1982--1998 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-Ali1996}{}}%
\CSLLeftMargin{8. }
\CSLRightInline{Ali, J. \emph{et al.} \href{https://doi.org/10.1007/s002689900171}{Teaching effectiveness of the advanced trauma life support program as demonstrated by an objective structured clinical examination for practicing physicians}. \emph{World Journal of Surgery} \textbf{20}, 1121--1126 (1996).}

\leavevmode\vadjust pre{\hypertarget{ref-Ali1999}{}}%
\CSLLeftMargin{9. }
\CSLRightInline{Ali, J. \emph{et al.} \href{https://doi.org/10.1097/00005373-199901000-00013}{Comparison of performance of interns completing the old (1993) and new interactive (1997) advanced trauma life support courses}. \emph{The Journal of Trauma: Injury, Infection, and Critical Care} \textbf{46}, 80--86 (1999).}

\leavevmode\vadjust pre{\hypertarget{ref-GerdinWuxe4rnberg2022}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{Gerdin Wärnberg, M. \emph{et al.} \href{https://doi.org/10.1136/bmjopen-2021-057504}{A pilot multicentre cluster randomised trial to compare the effect of trauma life support training programmes on patient and provider outcomes}. \emph{BMJ Open} \textbf{12}, e057504 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-Lee2014}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{Lee, E. C. \emph{et al.} \href{https://doi.org/10.1186/1471-2288-14-41}{The statistical interpretation of pilot trials: Should significance thresholds be reconsidered?} \emph{BMC Medical Research Methodology} \textbf{14}, (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Roberts2013}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{Roberts, I. \emph{et al.} \href{https://doi.org/10.3310/hta17100}{The CRASH-2 trial: A randomised controlled trial and economic evaluation of the effects of tranexamic acid on death, vascular occlusive events and transfusion requirement in bleeding trauma patients}. \emph{Health Technol Assess} \textbf{17}, (2013).}

\leavevmode\vadjust pre{\hypertarget{ref-Sierink2016}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{Sierink, J. C. \emph{et al.} \href{https://doi.org/10.1016/s0140-6736(16)30932-1}{Immediate total-body CT scanning versus conventional imaging and selective CT scanning in patients with severe trauma (REACT-2): A randomised controlled trial}. \emph{The Lancet} \textbf{388}, 673--683 (2016).}

\leavevmode\vadjust pre{\hypertarget{ref-Ariyanayagam1992}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{Ariyanayagam, D. \emph{et al.} The impact of the ATLS course on traffic accident mortality in trinidad and tobago. \emph{West Indian Medical Journal} \textbf{41}, 72--74 (1992).}

\leavevmode\vadjust pre{\hypertarget{ref-Ciouxe8Peuxf1a2016}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{Cioè-Peña, E. \emph{et al.} \href{https://doi.org/10.1177/1460408616672491}{Development and implementation of a hospital-based trauma response system in an urban hospital in san salvador, el salvador}. \emph{Trauma} \textbf{19}, 118--126 (2016).}

\leavevmode\vadjust pre{\hypertarget{ref-Petroze2014}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{Petroze, R. T. \emph{et al.} \href{https://doi.org/10.1007/s00268-014-2899-y}{Can focused trauma education initiatives reduce mortality or improve resource utilization in a low-resource setting?} \emph{World Journal of Surgery} \textbf{39}, 926--933 (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-VanHeng2008}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{Van Heng, Y. \emph{et al.} \href{https://doi.org/10.1017/s1049023x00006282}{Non-doctors as trauma surgeons? A controlled study of trauma training for non-graduate surgeons in rural cambodia}. \emph{Prehospital and Disaster Medicine} \textbf{23}, 483--489 (2008).}

\leavevmode\vadjust pre{\hypertarget{ref-Wang2010}{}}%
\CSLLeftMargin{18. }
\CSLRightInline{Wang, P. \emph{et al.} Comparison of severe trauma care effect before and after advanced trauma life support training. \emph{Chinese Journal of Traumatology (English Edition)} \textbf{13}, 341--344 (2010).}

\end{CSLReferences}

\end{document}

### 2024-10-04 pilot-publication/functions/create_outcome_table_with_differences.R

@@ -34,7 +34,8 @@ create_outcome_table_with_differences <- function(data) {
     )) |>
         modify_spanning_header(c(stat_1_1, stat_2_1, stat_3_1) ~ NA) |>
         modify_spanning_header(c(estimate_2, estimate_3, estimate_4) ~ "**Differences**") |>
-        modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)")
+        modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)") |>
+        modify_table_styling(columns = everything(), missing = "")
 
     # Return outcome table
     return(outcome.table.with.differences)

### 2024-10-04 pilot-publication/functions/number_to_text.R

@@ -1,9 +1,10 @@
 number_to_text <- function(number, first = FALSE) {
+    number <- as.numeric(number)
     if (number < 13 || first) {
-        number <- english::english(number)
+        number <- english::english(number) |> as.character()
     }
-    if (is.character(number)) {
-        number <- stringr::str_to_sentence(number)
+    if (is.character(number) && first) {
+        number <- stringr::str_to_sentence(number) |> as.character()
     }
     return(number)
 }

### 2024-10-04 pilot-publication/functions/save_tables_to_file.R

@@ -31,7 +31,7 @@ save_tables_to_file <- function(outcome.results.tables) {
                         "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",
                         "[[\"", ci.level.table.name, "\"]]", "\n\n",
                         "\`\`\`\n\n",
-                        "\\pagebreak\n\n",
+                        "\\pagebreak\n\n"
                     )
                 }) %>%
                     paste0(collapse = "\n\n")

### 2024-10-04 pilot-publication/main.R

@@ -126,6 +126,9 @@ if (use.saved) {
 ## Create tables with outcome results
 outcome.results.tables <- create_outcome_results_tables(outcome.results)
 
+## Save outcome results tables to file
+save_tables_to_file(outcome.results.tables)
+
 # Extract results from tables
 females <- inline_text(overall.sample.characteristics.table,
     variable = patinfo__pt_gender,
@@ -330,6 +333,3 @@ rel.change.ptc <- inline_outcome_text(
     column = "PTC",
     level = "Yes"
 )
-
-## Save outcome results tables to file
-# save_tables_to_file(outcome.results.tables)

### 2024-10-09 pilot-publication/manuscript.Rmd
Compile revised manuscript

@@
 output:
-    bookdown::pdf_document2:
-      toc: false
-      number_sections: false
     md_document: default
@@
       reference_docx: "reference.docx"
+#    bookdown::pdf_document2:
+#      toc:false
+#      number_sections:false
 ---

### 2024-10-09 pilot-publication/functions/create_outcome_table_with_differences.R

@@ -32,7 +32,7 @@ create_outcome_table_with_differences <- function(data) {
         standard.care.ptc.difference,
         atls.vs.ptc.difference
     )) |>
-        modify_spanning_header(c(stat_1_1, stat_2_1, stat_3_1) ~ NA) |>
+        modify_spanning_header(c(stat_1_1, stat_2_1, stat_3_1) ~ "**Arms**") |>
         modify_spanning_header(c(estimate_2, estimate_3, estimate_4) ~ "**Differences**") |>
         modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)") |>
         modify_table_styling(columns = everything(), missing = "")

### 2024-10-09 pilot-publication/manuscript.Rmd
Revise pilot manuscript

--- 2024-10-04 pilot-publication/manuscript.Rmd
@@
 ## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
-title: "Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial"
+title: "Effect of Trauma Life Support Training on Patient Outcomes: A Pilot Cluster Randomised Trial"
 author: Trauma life support training Effectiveness Research Network (TERN) collaborators
@@
 ##### Tamal Khan 
+Senior Research fellow, All India Institute of Medical Sciences, New Delhi, India.
 
@@
 
-**Importance** Trauma life support training programmes aim to improve early trauma care but there is high-quality evidence to show that they improve patient outcomes.
+**Importance** There is no high-quality evidence to show that trauma life support training programmes improve patient outcomes.
 
@@
 Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
-`r number_to_text(m30d.standard.care.n, first = TRUE)` `r m30d.standard.care.p` patients died within 30 days in the standard care arm, `r number_to_text(m30d.atls.n)` `r m30d.atls.p` patients in the ATLS^®^ arm, and `r number_to_text(m30d.ptc.n)` `r m30d.ptc.p` patients in the PTC arm.
+`r number_to_text(m30d.standard.care.n, first = TRUE)` `r m30d.standard.care.p` patients died within 30 days in the standard care arm, `r number_to_text(m30d.atls.n)` `r m30d.atls.p` patient in the ATLS^®^ arm, and `r number_to_text(m30d.ptc.n)` `r m30d.ptc.p` patients in the PTC arm.
 
-**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
+**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible after incorporating key lessons from this pilot.
 
@@
 -->
-\newpage
 
@@
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
 Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
@@
 ## Sample size
-We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess feasibility. The time period was dictated by budget and time constraints.
+We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess the feasibility of the trial logistics and research methods.
+<!--The time period was dictated by budget and time constraints.-->
 
@@
 The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
-For a subset of patients we also extracted data from medical records, to be able to compare their distributions with directly observed data.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
+For a subset of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
 
@@
 ## Statistical Methods
-We analysed all data using descriptive statistics and did not perform any formal hypothesis [@Lee2014].
-Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range.
+We analysed all data using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
+Quantitative variables are summarised as median and interquartile range.
 Qualitative variables are presented as absolute numbers and percentages.
@@
 We enrolled `r n.patients` trauma patients from `r n.centres` participating centres between `r start.date` and `r end.date`.
-The standard care arm `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm `r n.ptc` patients.
-We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS^®^, and  `r n.ptc.residents` in PTC.
+The standard care arm enrolled `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm enrolled `r n.ptc` patients.
+We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_test(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.
 
@@
 ```{r sample-characteristics, echo = FALSE, results = "asis"}
-sample.characteristics.table
+sample.characteristics.table |>
+  gtsummary::modify_caption("Patient sample characteristics")
 ```
@@
 The pass rate was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
-The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-extracted).
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Materials.
+The variables with the maximum amount of missing data were cause of death and some complications, reported in Supplementary Materials.
+The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
 Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
@@
 After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%, as shown in Table \@ref(tab:outcomes).
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively, as shown in Table \@ref(tab:outcomes).
 Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
@@
 ```{r outcomes, echo = FALSE}
-outcome.table.with.differences
+outcome.table.with.differences |>
+  modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)")
 ```
@@
 Some variables, especially cause of death and complications (reported in Supplementary materials) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
-Similarly, the missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data
+The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
 
@@
 We found that the ATLS^®^ and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
-These findings hints towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
+These findings could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
 The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.

### 2024-10-09 pilot-publication/functions/create_outcome_table_with_differences.R

@@ -34,7 +34,6 @@ create_outcome_table_with_differences <- function(data) {
     )) |>
         modify_spanning_header(c(stat_1_1, stat_2_1, stat_3_1) ~ "**Arms**") |>
         modify_spanning_header(c(estimate_2, estimate_3, estimate_4) ~ "**Differences**") |>
-        modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)") |>
         modify_table_styling(columns = everything(), missing = "")
 
     # Return outcome table

### 2024-10-09 pilot-publication/functions/prepare_data.R

@@ -172,7 +172,7 @@ prepare_data <- function(data, codebook = NULL) {
     labelled::var_label(prepared.data$penetrating.trauma) <- "Penetrating trauma"
     labelled::var_label(prepared.data$shock) <- "Shock"
     labelled::var_label(prepared.data$severe.tbi) <- "Severe traumatic brain injury"
-    labelled::var_label(prepared.data$elderly) <- "Elderly"
+    labelled::var_label(prepared.data$elderly) <- "Elderly (Age ≥ 65 years)"
     labelled::var_label(prepared.data$male) <- "Male"
     labelled::var_label(prepared.data$female) <- "Female"
 

### 2024-10-09 pilot-publication/main.R

@@ -51,7 +51,6 @@ stratified.table.data <- data %>%
 sample.characteristics.table <- create_descriptive_table(stratified.table.data,
     strata = "arm", show.all.levels = FALSE, include.overall = TRUE
 ) %>%
-    gtsummary::modify_caption("Patient sample characteristics") %>%
     gtsummary::modify_table_styling(
         column = label,
         footnote_abbrev = "ATLS = Advanced Trauma Life Support; PTC = Prehospital Trauma Care"

### 2024-10-10 pilot-publication/manuscript.Rmd
Edit manuscript, work on table captions

@@
 output:
-    md_document: default
-    bookdown::word_document2:
+    officedown::rdocx_document:
       toc: false
@@
       reference_docx: "reference.docx"
+#    md_document: default
 #    bookdown::pdf_document2:
@@
 
-## Allocation and blincing
+## Allocation and blinding
 We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.
@@
 
-We enrolled `r n.patients` trauma patients from `r n.centres` participating centres between `r start.date` and `r end.date`.
+We enrolled `r n.patients` trauma patients from `r number_to_text(n.centres)` participating centres between `r start.date` and `r end.date`.
 The standard care arm enrolled `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm enrolled `r n.ptc` patients.
-We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_test(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.
+We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.
 
@@
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
-A total of `r m30d` patients died within 30 days after arrival to the emergency department.
+A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.
+The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
 
-```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care.", out.width="50%"}
+```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care.", out.width="30%", out.height="30%"}
 knitr::include_graphics("consort-flow-diagram.png")
@@
 
-```{r sample-characteristics, echo = FALSE, results = "asis"}
+```{r sample-characteristics, echo = FALSE, results = "asis", tab.cap = "Patient sample characteristics"}
 sample.characteristics.table |>
-  gtsummary::modify_caption("Patient sample characteristics")
+  gtsummary::as_flex_table()
 ```
@@
 We found that the ATLS^®^ and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
-These findings could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
+This finding could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
 The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
 This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.
-This heterogeneity highlights the importance of taken different cluster sizes into account in the design of the full scale trial.
+This heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
 
@@
 Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge.
-Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, comparable to previous trials.
+Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials.
 
@@
 
-We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial.
+We attempted to minimse the impact of the lower than expected enrolment rates by including a seventh centre, but careful assessments of patient volumes as part of the screening process will be needed for the full-scale trial.
 We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
-Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.
+Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers if this data is to be collected.
 
@@
 Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010].
-Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008], whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15%, after implementing
-ATLS^®^ [@Wang2010].
+Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008], whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15%, after implementing ATLS^®^ [@Wang2010].
 
-We conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
+Considering the widespread use of trauma life support traning, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Our study represent the first published attempt at a controlled trial of the effect of trauma life support traning och patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
 

### 2024-10-10 pilot-publication/functions/create_observed_vs_extracted_comparison_table.R

@@ -55,7 +55,13 @@ create_observed_vs_extracted_comparison_table <- function(data) {
     comparison.table <- create_descriptive_table(table.data,
         strata = "collection.mode",
         include.overall = FALSE
-    )
+    ) %>%
+        add_stat_label(
+            label = list(
+                all_continuous() ~ "median (IQR)",
+                all_categorical() ~ "n (%)"
+            )
+        )
 
     ## Return result
     return(comparison.table)

### 2024-10-10 pilot-publication/functions/get_basic_results.R

@@ -28,7 +28,8 @@ get_basic_results <- function(data) {
         "10263" = 2
     )
     results$p.consent <- round(nrow(data) / (nrow(data) + sum(unlist(n.no.consent))) * 100)
-    results$icc <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
+    results$icc.m30d <- estimate_icc("outcomes__alive_after_30_days", "id__reg_hospital_id", data)
+    results$icc.inhosp <- estimate_icc("outcomes__discharge_alive", "id__reg_hospital_id", data)
     results$n.patients <- nrow(data)
 
     # Resident data

### 2024-10-10 pilot-publication/main.R

@@ -142,6 +142,9 @@ median.iss <- inline_text(overall.sample.characteristics.table,
 m30d <- inline_text(overall.sample.characteristics.table,
     variable = outcomes__alive_after_30_days
 )
+inhosp <- inline_text(overall.sample.characteristics.table,
+    variable = outcomes__discharge_alive
+)
 
 ## Crude 30 day mortality
 m30d.standard.care <- inline_text(post.training.characteristics.table,

### 2024-10-14 pilot-publication/manuscript.qmd
Compila manuscript and supplementary materials

--- 2024-10-10 pilot-publication/manuscript.Rmd
@@
 ## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
+## libreoffice --headless --convert-to pdf manuscript.docx
 title: "Effect of Trauma Life Support Training on Patient Outcomes: A Pilot Cluster Randomised Trial"
@@
 csl: nature.csl
-output:
-    officedown::rdocx_document:
+format:
+  docx:
     toc: false
@@
 
-The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
+The study flow diagram is shown in @flow-diagram and patient sample characteristics across trial arms are shown in Table @sample-characteristics.
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
@@
 sample.characteristics.table |>
-  gtsummary::as_flex_table()
+    modify_caption("Patient sample characteristics")
 ```
@@
 The pass rate was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Materials.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table @sample-characteristics and in Supplementary Materials.
 The variables with the maximum amount of missing data were cause of death and some complications, reported in Supplementary Materials.
-The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
+The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table @observed-vs-extracted.
 Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
 
-```{r observed-vs-extracted, echo = FALSE}
-observed.vs.extracted.comparison.table %>%
-  modify_caption("Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview.")
+```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview."}
+observed.vs.extracted.comparison.table
 ```
@@
 After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively, as shown in Table \@ref(tab:outcomes).
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively, as shown in @outcomes.
 Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
@@
 
-```{r outcomes, echo = FALSE}
-outcome.table.with.differences |>
-  modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)")
+```{r outcomes, echo = FALSE, tab.cap = "Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)"}
+outcome.table.with.differences
 ```

### 2024-10-14 pilot-publication/supplementary-materials.Rmd
Compila manuscript and supplementary materials

---
title: Supplementary Materials
subtitle: "Effect of Trauma Life Support Training on Patient Outcomes: A Pilot Cluster Randomised Trial"
output:
    bookdown::word_document2:
      number_sections: false
      reference_docx: "reference.docx"
---

```{r source, echo = FALSE, include = FALSE}
source("main.R")
attach(results)
```

```{r, child = "outcome-tables.Rmd"}
```

### 2024-10-18 pilot-publication/manuscript.Rmd
Update manuscript and supplementary materials

@@
 output:
-    officedown::rdocx_document:
+    bookdown::word_document2:
       toc: false
@@
 
-**Main Outcomes and Measures** The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records.
-The primary patient outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
+**Main Outcomes and Measures** The outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, differences in distribution between observed and data extracted from medical records as well as all cause and in-hospital mortality at 30 days from the time of arrival to the emergency department.
 
@@
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; putra_impact_2023].
 
@@
 Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is then how, and if, these improved skills translate into improved patient outcomes.
+The missing link is then if these improved skills translate into improved patient outcomes.
 
-Systematic reviews call for trials in settings where these programmes are not routinely implemented, but data needed to inform the planning of such a trial is missing.
+Systematic reviews call for trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], but data needed to inform the planning of such a trial is missing.
 We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
@@
 ## Standard Care
+
 Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
-These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they manage the emergency department.
+These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they are posted in the emergency department.
 In the emergency department, trauma patients are initially assessed by first- or second-year residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
@@
 ## Intervention
+
 In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC.
@@
 The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018].
-We did not modify or adapt the delivery or content of these programs during this pilot study.
+We did not modify or adapt the delivery or content of these programmes during this pilot study.
 
@@
 ## Outcomes
-We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A complete list of outcomes is available as Supplementary Materials. Our main outcomes were:
+
+We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A list of outcomes is available in Supplementary Table S1. Our main outcomes were:
 
@@
 ### Patients
+
 Arriving patients were screened for eligibility and consented, if conscious.
@@
 ### Residents
+
 Participating units were screened for eligibility once hospitals confirmed their participation.
@@
 ## Sample size
-We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess the feasibility of the trial logistics and research methods.
+
+We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
 <!--The time period was dictated by budget and time constraints.-->
@@
 ## Allocation and blinding
+
 We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.
@@
 ## Data Collection
+
 Data was collected over a four-month period.
@@
 ## Variables
+
 The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
@@
 ## Patient and public involvement
+
 We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial.
@@
 ## Data monitoring
+
 We conducted weekly online meetings to monitor the study and data collection.
@@
 ## Statistical Methods
+
 We analysed all data using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
@@
 These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
+We calculated both absolute and relative differences for each comparison, along with 75, 85, and 95% confidence intervals.
+We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
 
-<!--We calculated both absolute and relative differences for each
-comparison, along with 75, 85, and 95% confidence intervals. We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.-->
+In the interest of space, only the 95% confidence intervals are presented for all comparisons as Supplementary Tables. The remaining results are available from the corresponding author on request.
 
 ## Ethics and Dissemination
+
 We were granted research ethics approval from the institutional ethics committees at each participating hospital.
@@
 sample.characteristics.table |>
-  gtsummary::as_flex_table()
+  modify_caption("Patient sample characteristics")
 ```
@@
 The pass rate was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Materials.
-The variables with the maximum amount of missing data were cause of death and some complications, reported in Supplementary Materials.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S1.
+The variables with the maximum amount of missing data were complications, reported in Supplementary Tables S1-12.
 The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
@@
 
-```{r observed-vs-extracted, echo = FALSE}
-observed.vs.extracted.comparison.table %>%
-  modify_caption("Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview.")
+```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview."}
+observed.vs.extracted.comparison.table
 ```
@@
 Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
-The results for all other outcomes are shown in Supplementary Materials.
+The results for all other outcomes are shown in Supplementary Tables S1-12.
 
-```{r outcomes, echo = FALSE}
-outcome.table.with.differences |>
-  modify_caption("Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)")
+```{r outcomes, echo = FALSE, tab.cap = "Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)"}
+outcome.table.with.differences
 ```
@@
 # Discussion
+
 We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care.
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially cause of death and complications (reported in Supplementary materials) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Some variables, especially complications (reported in Supplementary Table S1-12) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
@@
 During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
-material.
+material S13.
 Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.
@@
 # Competing Interests
+
 Several authors are ATLS^®^ and/or PTC instructors.
@@
 # Funding
+
 Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).
@@
 # Data Sharing Statement
+
 The final anonymized dataset and code for analysis are released publicly.
-
-# Protocol Deviations
-
-## Trial Registration
-We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
-
-## Outcomes across subgroups
-Because of small numbers in the pre-specified subgroups we decided to report only descriptive data on these subgroups.
-
-## Number of Participating Centres
-We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
-
-## Resident Participants
-Emergency medicine in addition to surgery.
-
-## Periodic suverys to residents
-We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.
-
-## Follow up of residents
-We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.
-
-## Data collection from records
-We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.
-
-## Selection of units for training
-We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.
-
-## Timing of resident consent
-We had initially planned to ask residents for consent before
-randomisation, but this was not possible because of logistical issues the units were only finalised after the hospitals had been randomised.
-Residents were therefore approached for consent after randomisation but before training.
-
-## Analysis level of feasibility outcomes
-We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level.
 
@@
 
-# Supplementary material
-
 # References

### 2024-10-18 pilot-publication/supplementary-materials.Rmd
Update manuscript and supplementary materials

@@
 
+# Table S1. Extended sample characteristics
+
+```{r overall-table, echo = FALSE}
+combined.sample.characteristics.table
+```
+
 ```{r, child = "outcome-tables.Rmd"}
 ```
+
+# S13 Protocol Deviations
+
+## Trial Registration
+
+We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
+
+## Outcomes across subgroups
+
+Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly we decided to report only descriptive data on these subgroups.
+
+## Number of Participating Centres
+
+We recruited seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
+
+## Resident Participants
+
+We included emergency medicine residents in addition to surgical residents.
+
+## Periodic suverys to residents
+
+We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.
+
+## Follow up of residents
+
+We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.
+
+## Data collection from records
+
+We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.
+
+## Selection of units for training
+
+We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.
+
+## Timing of resident consent
+
+We had initially planned to ask residents for consent before randomisation, but because of logistical issues the units were only finalised after the hospitals had been randomised.
+Residents were therefore approached for consent after randomisation but before training.
+
+## Analysis level of feasibility outcomes
+
+We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level, because the sample sizes in individual clusters were too small to generate meaningful results.

### 2024-10-18 pilot-publication/supplementary-outcome-tables.Rmd
Update manuscript and supplementary materials

--- 2023-12-30 pilot-publication/outcome-tables.Rmd
@@
-# 75% confidence level
+\pagebreak
 
-## overall
+# Results with 95% confidence level
+
+## Table S2 Outcomes in all patients during the entire study period, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["overall"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
 
@@
 
-## pre.training
+\pagebreak
+
+## Table S3 Outcomes in all patients before training, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["pre.training"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients before training, by treatment arm"]]
 
@@
 
-## post.training
+\pagebreak
+
+## Table S4 Outcomes in all patients after training, by treatment arm
 
@@
 
-tables[["ci.level.0.75"]][["post.training"]]
+outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients after training, by treatment arm"]]
 
@@
 
-## absolute.difference
+\pagebreak
+
+## Table S5 Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-tables[["ci.level.0.75"]][["absolute.difference"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-## relative.difference
+\pagebreak
+
+## Table S6 Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-tables[["ci.level.0.75"]][["relative.difference"]]
+outcome.results.tables[["ci.level.0.95"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
 
@@
 
-## Post training outcome Standard care vs. ATLS
+\pagebreak
+
+## Table S7 Absolute and relative differences in outcomes after training, comparing standard care with ATLS
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
 
@@
 
-## Post training outcome Standard care vs. PTC
+\pagebreak
+
+## Table S8 Absolute and relative differences in outcomes after training, comparing standard care with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
 
@@
 
-## Post training outcome ATLS vs. PTC
+\pagebreak
+
+## Table S9 Absolute and relative differences in outcomes after training, comparing ATLS with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Post training outcome ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
 
@@
 
-## Change from baseline Standard care vs. ATLS
+\pagebreak
+
+## Table S10 Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline Standard care vs. ATLS"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
 
@@
 
-## Change from baseline Standard care vs. PTC
+\pagebreak
+
+## Table S11 Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline Standard care vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
 
@@
 
-## Change from baseline ATLS vs. PTC
+\pagebreak
+
+## Table S12 Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
 
@@
 
-tables[["ci.level.0.75"]][["Change from baseline ATLS vs. PTC"]]
+outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
 
@@
 
-# 85% confidence level
-
-## overall
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["overall"]]
-
-```
-
-## pre.training
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["pre.training"]]
-
-```
-
-## post.training
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["post.training"]]
-
-```
-
-## absolute.difference
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["absolute.difference"]]
-
-```
-
-## relative.difference
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["relative.difference"]]
-
-```
-
-## Post training outcome Standard care vs. ATLS
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Post training outcome Standard care vs. ATLS"]]
-
-```
-
-## Post training outcome Standard care vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Post training outcome Standard care vs. PTC"]]
-
-```
-
-## Post training outcome ATLS vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Post training outcome ATLS vs. PTC"]]
-
-```
-
-## Change from baseline Standard care vs. ATLS
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Change from baseline Standard care vs. ATLS"]]
-
-```
-
-## Change from baseline Standard care vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Change from baseline Standard care vs. PTC"]]
-
-```
-
-## Change from baseline ATLS vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.85"]][["Change from baseline ATLS vs. PTC"]]
-
-```
-
-# 95% confidence level
-
-## overall
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["overall"]]
-
-```
-
-## pre.training
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["pre.training"]]
-
-```
-
-## post.training
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["post.training"]]
-
-```
-
-## absolute.difference
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["absolute.difference"]]
-
-```
-
-## relative.difference
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["relative.difference"]]
-
-```
-
-## Post training outcome Standard care vs. ATLS
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Post training outcome Standard care vs. ATLS"]]
-
-```
-
-## Post training outcome Standard care vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Post training outcome Standard care vs. PTC"]]
-
-```
-
-## Post training outcome ATLS vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Post training outcome ATLS vs. PTC"]]
-
-```
-
-## Change from baseline Standard care vs. ATLS
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Change from baseline Standard care vs. ATLS"]]
-
-```
-
-## Change from baseline Standard care vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Change from baseline Standard care vs. PTC"]]
-
-```
-
-## Change from baseline ATLS vs. PTC
-
-```{r, echo = FALSE}
-
-tables[["ci.level.0.95"]][["Change from baseline ATLS vs. PTC"]]
-
-```
+\pagebreak

### 2024-10-18 pilot-publication/functions/save_tables_to_file.R

@@ -1,12 +1,16 @@
-#' Save outcome tables to file and create an HTML document
+#' Save outcome tables to file and create an PDF document
 #'
 #' This function takes a list of outcome tables and saves them to a file in R Markdown format.
-#' It then renders the R Markdown file into an HTML document with a table of contents.
+#' It then renders the R Markdown file into an PDF document with a table of contents.
 #'
 #' @param outcome.tables A list of outcome tables to be saved
+#' @param supplementary.material Logical. If TRUE the tables are numbered from S2 to how many
+#' tables there are. Defaults to FALSE.
+#' @param only.95.ci Logical. If TRUE only the tables with 95% confidence intervals are saved.
+#' Set to FALSE if someone requests the full results. Defaults to FALSE.
 #'
 #' @export
-save_tables_to_file <- function(outcome.results.tables) {
+save_tables_to_file <- function(outcome.results.tables, supplementary.material = FALSE, only.95.ci = FALSE) {
     ## Check arguments
     assertthat::assert_that(
         is.list(outcome.results.tables)
@@ -20,32 +24,36 @@ save_tables_to_file <- function(outcome.results.tables) {
                 as.numeric() %>%
                 scales::percent()
             ci.level.tables <- outcome.results.tables[[ci.level.tables.name]]
-            paste0(
-                "\\pagebreak\n\n",
-                "# Results with ", ci.level, " confidence level\n\n",
-                lapply(names(ci.level.tables), function(ci.level.table.name) {
-                    ci.level.table <- ci.level.tables[[ci.level.table.name]]
-                    paste0(
-                        "## ", ci.level.table.name, "\n\n",
-                        "\`\`\`{r, echo = FALSE} \n\n",
-                        "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",
-                        "[[\"", ci.level.table.name, "\"]]", "\n\n",
-                        "\`\`\`\n\n",
-                        "\\pagebreak\n\n"
-                    )
-                }) %>%
-                    paste0(collapse = "\n\n")
-            )
+            if (!only.95.ci || (only.95.ci && ci.level == "95%")) {
+                paste0(
+                    "\\pagebreak\n\n",
+                    "# Results with ", ci.level, " confidence level\n\n",
+                    lapply(seq_along(ci.level.tables), function(ci.level.table.index) {
[…]

### 2024-10-18 pilot-publication/main.R

@@ -39,8 +39,14 @@ table.variables <- c(
     "outcomes__alive_after_30_days",
     "arm"
 )
+table.variables.full <- unique(c(
+    table.variables,
+    binary_outcomes() |> names(),
+    categorical_outcomes() |> names(),
+    quantitative_outcomes() |> names()
+))
 table.data <- data %>%
-    select(all_of(table.variables)) %>%
+    select(all_of(table.variables.full)) %>%
     select(-arm)
 overall.sample.characteristics.table <- create_descriptive_table(table.data, show.all.levels = FALSE)
 
@@ -65,7 +71,7 @@ sample.characteristics.table <- create_descriptive_table(stratified.table.data,
 ## Create table of sample characteristics before training
 pre.training.table.data <- data %>%
     filter(!post.training) %>%
-    select(table.variables)
+    select(table.variables.full)
 pre.training.characteristics.table <- create_descriptive_table(
     pre.training.table.data,
     strata = "arm",
@@ -76,7 +82,7 @@ pre.training.characteristics.table <- create_descriptive_table(
 ## Create table of sample characteristics after training
 post.training.table.data <- data %>%
     filter(post.training) %>%
-    select(table.variables)
+    select(table.variables.full)
 post.training.characteristics.table <- create_descriptive_table(
     post.training.table.data,
     strata = "arm",
@@ -125,8 +131,11 @@ if (use.saved) {
 ## Create tables with outcome results
 outcome.results.tables <- create_outcome_results_tables(outcome.results)
 
-## Save outcome results tables to file
-save_tables_to_file(outcome.results.tables)
+## Save outcome results for supplementary material
+save_tables_to_file(outcome.results.tables, supplementary.material = TRUE, only.95.ci = TRUE)
+
+## Save all outcome results tables to file
+save_tables_to_file(outcome.results.tables, supplementary.material = FALSE, only.95.ci = FALSE)
 
 # Extract results from tables
 females <- inline_text(overall.sample.characteristics.table,

### 2024-10-19 pilot-publication/supplementary-outcome-tables.Rmd
Compile new version of manuscript

@@
 
-## Table S2 Outcomes in all patients during the entire study period, by treatment arm
+## Table S2. Outcomes in all patients during the entire study period, by treatment arm
 
@@
 
-## Table S3 Outcomes in all patients before training, by treatment arm
+## Table S3. Outcomes in all patients before training, by treatment arm
 
@@
 
-## Table S4 Outcomes in all patients after training, by treatment arm
+## Table S4. Outcomes in all patients after training, by treatment arm
 
@@
 
-## Table S5 Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+## Table S5. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-## Table S6 Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+## Table S6. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
 
@@
 
-## Table S7 Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+## Table S7. Absolute and relative differences in outcomes after training, comparing standard care with ATLS
 
@@
 
-## Table S8 Absolute and relative differences in outcomes after training, comparing standard care with PTC
+## Table S8. Absolute and relative differences in outcomes after training, comparing standard care with PTC
 
@@
 
-## Table S9 Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+## Table S9. Absolute and relative differences in outcomes after training, comparing ATLS with PTC
 
@@
 
-## Table S10 Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+## Table S10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
 
@@
 
-## Table S11 Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+## Table S11. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
 
@@
 
-## Table S12 Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+## Table S12. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
 

### 2024-10-19 pilot-publication/supplementary-materials.Rmd
Add right tables to supplementary materials

@@
       reference_docx: "reference.docx"
+      toc: true
 ---
@@
 
-```{r, child = "outcome-tables.Rmd"}
+```{r, child = "supplementary-outcome-tables.Rmd"}
 ```

### 2024-10-19 pilot-publication/supplementary-outcome-tables.Rmd
Compile new version of supplementary materials

@@
 
-# Results with 95% confidence level
+# Results with 95% confidence intervals (CI)
 
-## Table S2. Outcomes in all patients during the entire study period, by treatment arm
+## Table S2. Outcomes in all patients during the entire study period, by treatment arm(95% CI)
 
@@
 
-## Table S3. Outcomes in all patients before training, by treatment arm
+## Table S3. Outcomes in all patients before training, by treatment arm(95% CI)
 
@@
 
-## Table S4. Outcomes in all patients after training, by treatment arm
+## Table S4. Outcomes in all patients after training, by treatment arm(95% CI)
 
@@
 
-## Table S5. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+## Table S5. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(95% CI)
 
@@
 
-## Table S6. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+## Table S6. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(95% CI)
 
@@
 
-## Table S7. Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+## Table S7. Absolute and relative differences in outcomes after training, comparing standard care with ATLS(95% CI)
 
@@
 
-## Table S8. Absolute and relative differences in outcomes after training, comparing standard care with PTC
+## Table S8. Absolute and relative differences in outcomes after training, comparing standard care with PTC(95% CI)
 
@@
 
-## Table S9. Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+## Table S9. Absolute and relative differences in outcomes after training, comparing ATLS with PTC(95% CI)
 
@@
 
-## Table S10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+## Table S10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS(95% CI)
 
@@
 
-## Table S11. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+## Table S11. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC(95% CI)
 
@@
 
-## Table S12. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+## Table S12. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC(95% CI)
 

### 2024-10-21 pilot-publication/manuscript.Rmd
Revise manuscript after comments from Johanna

@@
 
+<!-- Source main code to make global objects available in the text -->
+```{r source, echo = FALSE, include = FALSE}
+source("main.R")
+attach(results)
+```
+
 ## Trial registration
@@
 ## Corresponding author
-Martin Gerdin Wärnberg (MGW)
+Martin Gerdin Wärnberg (MGW) `r insert_orcid_link("0000-0001-6069-4794")`
 
@@
 
-##### Johanna Berg (JB)
+##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")`
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Lovisa Strömmer (LS)
+##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")`
 Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-<!-- Source main code to make global objects available in the text -->
-```{r source, echo = FALSE, include = FALSE}
-source("main.R")
-attach(results)
-```
-
 # Abstract
@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; putra_impact_2023].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @putra_impact_2023].
 
-Several studies, including at least two randomised controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
+Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
 Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
@@
 The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
+All injuries were coded according to the International Classification of Diseases version 10 (ICD-10).
+Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022].
 For a subset of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
@@
 
+We used the `r version$version.string` Statistical Software for all analyses [@R_2024].
 We analysed all data using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
@@
 
-Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
+Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years) [@hornor_quality_2018].
 These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
@@
 
-In the interest of space, only the 95% confidence intervals are presented for all comparisons as Supplementary Tables. The remaining results are available from the corresponding author on request.
+In the interest of space, only the 95% confidence intervals are presented for all comparisons in Supplementary Tables. The remaining results are available from the corresponding author on request.
 

### 2024-10-21 pilot-publication/functions/insert_orcid_link.R

@@ -1,0 +1,4 @@
+insert_orcid_link <- function(orcid.id) {
+    return(paste0("[![](https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png){height=16 width=16}](https://orcid.org/", orcid.id, ")"))
+    # return(paste0("[![](ORCIDiD_icon16x16.png){height=16 width=16}](https://orcid.org/", orcid.id, ")"))
+}

### 2024-10-21 pilot-publication/functions/prepare_data.R

@@ -93,12 +93,12 @@ prepare_data <- function(data, codebook = NULL) {
         "niss"
     )]
     prepared.data <- cbind(prepared.data, iss.data)
-    
+
     ## Treat iss and niss as NA, if all icd columns are NA
     prepared.data <- prepared.data %>%
         # Replace 0 and NAD with NA for all columns containing "icd" but not ending with "_r"
         mutate(across(
-            .cols = contains("icd") & !ends_with("_r"), 
+            .cols = contains("icd") & !ends_with("_r"),
             .fns = ~ replace(., . %in% c("0", "NAD"), NA)
         )) %>%
         # Set riss and niss to NA if all 'icd' columns are NA
@@ -187,7 +187,7 @@ prepare_data <- function(data, codebook = NULL) {
     labelled::var_label(prepared.data$incident__moi) <- "Mechanism of injury"
     labelled::var_label(prepared.data$blunt.multisystem.trauma) <- "Blunt multisystem trauma"
     labelled::var_label(prepared.data$penetrating.trauma) <- "Penetrating trauma"
-    labelled::var_label(prepared.data$shock) <- "Shock"
+    labelled::var_label(prepared.data$shock) <- "Shock (SBP ≤ 90 mmHg)"
     labelled::var_label(prepared.data$severe.tbi) <- "Severe traumatic brain injury"
     labelled::var_label(prepared.data$elderly) <- "Elderly (Age ≥ 65 years)"
     labelled::var_label(prepared.data$male) <- "Male"

### 2024-10-25 pilot-publication/manuscript.Rmd
Add new reference

--- 2024-10-18 pilot-publication/manuscript.Rmd
@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; putra_impact_2023].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @putra_impact_2023; @mciver_effect_2024].
 

### 2024-10-31 pilot-publication/manuscript.Rmd

--- 2024-10-21 pilot-publication/manuscript.Rmd
@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @putra_impact_2023].
+Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @putra_impact_2023; @mciver_effect_2024].
 

### 2024-10-31 pilot-publication/manuscript.Rmd
Revise manuscript after comments

@@
 ## libreoffice --headless --convert-to pdf manuscript.docx
-title: "Effect of Trauma Life Support Training on Patient Outcomes: A Pilot Cluster Randomised Trial"
+title: "Assessing the Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study"
 author: Trauma life support training Effectiveness Research Network (TERN) collaborators
@@
 ##### Nobhojit Roy (NR)
-WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.
+Program for Global Surgery and Trauma, The George Institute of Global Health, New Delhi, India.
 
@@
 
-Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013].
+Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved knowledge and skills among providers [@Mohammad2013].
 Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is then if these improved skills translate into improved patient outcomes.
+The missing link is then if this improved knowledge and skills translate into measurably improved patient outcomes.
 
@@
 ## Study Setting
-We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other trauma life support training program is routinely taught.
+We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
 
@@
 Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
-These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they are posted in the emergency department.
-In the emergency department, trauma patients are initially assessed by first- or second-year residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+These units  include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
+In the emergency department, trauma patients are initially assessed by residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
 Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
@@
 
-All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
+We included persons who were 15 years or older and presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
 History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
@@
 - Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
-- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients whom did not complete 30 day follow up, out of all enrolled patients.
-- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents that passed the training programme, out of the total number of trained residents.
+- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all enrolled patients.
+- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
 - Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
@@
 We followed up admitted patients for their complications and other in-hospital outcome measures.
-Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.
+Patients who were not admitted or who were discharged before the end of the study were followed up telephonically for mortality outcomes and quality of life outcomes.
 
@@
 Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022].
-For a subset of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
+For a convenience sample of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
 
@@
 We were granted research ethics approval from the institutional ethics committees at each participating hospital.
+For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, XX for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for Postgraduate Institute of Medical Education and Research, Chandigarh.
 
@@
 
-Considering the widespread use of trauma life support traning, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-Our study represent the first published attempt at a controlled trial of the effect of trauma life support traning och patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
+Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Our study represents the first published attempt at a controlled trial of the effect of trauma life support training och patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
 

### 2024-10-31 pilot-publication/supplementary-materials.Rmd
Revise manuscript after comments

@@
 title: Supplementary Materials
-subtitle: "Effect of Trauma Life Support Training on Patient Outcomes: A Pilot Cluster Randomised Trial"
+subtitle: "Assessing the Feasibility of a Pilot Cluster Randomised Trial of the Effect of Trauma Life Support Training: A Pilot Study"
 output:
@@
 
-We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
+We intended to register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
 
@@
 
-Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly we decided to report only descriptive data on these subgroups.
+Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury and elderly we decided to report only descriptive data on these subgroups.
 
@@
 
-## Periodic suverys to residents
+## Periodic surveys of residents
 
@@
 
-We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.
+We decided to extract data from medical records only for a convenience sample of patients to reduce the research officers' workload.
 

### 2024-10-31 pilot-publication/functions/create_outcome_table_with_differences.R

@@ -1,3 +1,24 @@
+#' Create outcome table with differences between arms
+#'
+#' Creates a table comparing mortality outcomes between trial arms and calculates differences between arms.
+#' The table shows percentages for each arm and absolute differences between arms with confidence intervals.
+#'
+#' @param data A data frame containing the trial data with columns:
+#'   - outcomes__alive_after_30_days: 30-day mortality outcome ("Yes"/"No")
+#'   - outcomes__discharge_alive: In-hospital mortality outcome ("Yes"/"No")
+#'   - arm: Trial arm ("Standard care", "ATLS", or "PTC")
+#'
+#' @return A gtsummary table object containing:
+#'   - Percentages for each outcome by trial arm
+#'   - Absolute differences between Standard care vs ATLS
+#'   - Absolute differences between Standard care vs PTC
+#'   - Absolute differences between ATLS vs PTC
+#'
+#' @importFrom gtsummary tbl_summary tbl_merge modify_header modify_spanning_header modify_table_styling
+#' @importFrom labelled var_label
+#' @importFrom dplyr filter select mutate
+#'
+#' @export
 create_outcome_table_with_differences <- function(data) {
     # Filter and modify data
     table.data <- data |>

### 2024-10-31 pilot-publication/functions/create_summary_with_difference.R

@@ -3,7 +3,9 @@ create_summary_with_difference <- function(data, arms, header_text) {
     data |>
         filter(arm %in% arms) |>
         droplevels() |>
-        gtsummary::tbl_summary(by = arm) |>
+        gtsummary::tbl_summary(
+            by = arm
+        ) |>
         add_difference() |>
         modify_column_hide(c(stat_1, stat_2, conf.low, p.value)) |>
         modify_footnote(estimate ~ NA) |>

### 2024-10-31 pilot-publication/main.R

@@ -168,7 +168,7 @@ m30d.standard.care.n <- inline_text(post.training.characteristics.table,
 m30d.standard.care.p <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "Standard care",
-    pattern = "({p})%"
+    pattern = "({p}%)"
 )
 m30d.atls <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
@@ -182,7 +182,7 @@ m30d.atls.n <- inline_text(post.training.characteristics.table,
 m30d.atls.p <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "ATLS",
-    pattern = "({p})%"
+    pattern = "({p}%)"
 )
 m30d.ptc <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
@@ -196,7 +196,7 @@ m30d.ptc.n <- inline_text(post.training.characteristics.table,
 m30d.ptc.p <- inline_text(post.training.characteristics.table,
     variable = outcomes__alive_after_30_days,
     column = "PTC",
-    pattern = "({p})%"
+    pattern = "({p}%)"
 )
 
 # Crude in-hospital mortality

### 2024-11-02 pilot-publication/manuscript.Rmd
Revise manuscript after further comments

@@
 
-**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes, and to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial.
 
@@
 
-**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible after incorporating key lessons from this pilot.
+**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
 
@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @putra_impact_2023; @mciver_effect_2024].
+Observational studies indicate that these programmes may improve patient outcomes [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
 Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved knowledge and skills among providers [@Mohammad2013].
-Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is then if this improved knowledge and skills translate into measurably improved patient outcomes.
+Observational evidence suggests that PTC also leads to improved provider skills [@Kadhum2020].
+The missing link is then whether these improved knowledge and skills translate into measurably improved patient outcomes.
 
-Systematic reviews call for trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], but data needed to inform the planning of such a trial is missing.
-We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
+Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care would not be possible.
+Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be challenging due to for example difficulties in patient follow up.
+
+We therefore aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial.
 
@@
 We were granted research ethics approval from the institutional ethics committees at each participating hospital.
-For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, XX for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for Postgraduate Institute of Medical Education and Research, Chandigarh.
+For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for Postgraduate Institute of Medical Education and Research, Chandigarh.
 
@@
 The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
-The lost to follow up rate was `r rate.lost.to.follow.up`%.
-The pass rate was `r pass.rate`%.
+Among patients, the lost to follow up rate was `r rate.lost.to.follow.up`%.
+Among residents, the pass rate was `r pass.rate`%.
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S1.

### 2024-11-02 pilot-publication/manuscript.Rmd
Add new citations

@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
@@
 After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively, as shown in Table \@ref(tab:outcomes).
-Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS^®^ and PTC arms respectively, as shown in Table \@ref(tab:outcomes).
+Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS^®^ and PTC arms were very small.
 The results for all other outcomes are shown in Supplementary Tables S1-12.
@@
 
-Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010].
-Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008], whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15%, after implementing ATLS^®^ [@Wang2010].
+Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
+Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016].
+In contrast, some studies have found that these programmes are associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
 

### 2024-11-11 pilot-publication/manuscript.Rmd
Edit manuscript and supplementary materials

@@
 ## Corresponding author
-Martin Gerdin Wärnberg (MGW) `r insert_orcid_link("0000-0001-6069-4794")`
+Martin Gerdin Wärnberg (MGW)`r insert_orcid_link("0000-0001-6069-4794")`
 
@@
 
-## Authors
+## Authors (in alphabetical order by last name)
 
-##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")`
+##### Debojit Basak`r insert_orcid_link("90000-0002-8378-9689")`
+Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
+
+##### Johanna Berg (JB)`r insert_orcid_link("0000-0001-7553-7337")`
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Shamita Chatterjee (SC)
+##### Shamita Chatterjee (SC)`r insert_orcid_link("0000-0002-9460-108X")`
 Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Li Felländer-Tsai (LFT)
+##### Li Fell<U+00E4>nder-Tsai (LFT)`r insert_orcid_link("0000-0003-0693-6080")`
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Geeta Ghag (GG)
+##### Geeta Ghag (GG)`r insert_orcid_link("0000-0001-5561-362X")`
 Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
 
-##### Monty Khajanchi (MK)
+##### Monty Khajanchi (MK)`r insert_orcid_link("0000-0002-0898-6391")`
 WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
@@
 
-##### Catherine Juillard (CJ)
+##### Tamal Khan 
+Senior Research fellow, All India Institute of Medical Sciences, New Delhi, India.
+
+##### Catherine Juillard (CJ)`r insert_orcid_link("0000-0003-0847-8933")`
 Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
 
-##### Nobhojit Roy (NR)
+##### Vipul Nandu`r insert_orcid_link("0000-0002-3366-5610")`
+Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
+
+##### Nobhojit Roy (NR)`r insert_orcid_link("0000-0003-2022-7416")`
 Program for Global Surgery and Trauma, The George Institute of Global Health, New Delhi, India.
@@
 
-##### Rajdeep Singh (RS)
+##### Rajdeep Singh (RS)`r insert_orcid_link("0000-0001-6593-2624")`
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
 
-##### Kapil Dev Soni (KDS)
+##### Kapil Dev Soni (KDS)`r insert_orcid_link("0000-0003-1214-4119")`
 Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
 
-##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")`
+##### Lovisa Strömmer (LS)`r insert_orcid_link("0000-0001-5424-7111")`
 Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.
-
-##### Vipul Nandu
-Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
-
-##### Debojit Basak
-Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
-
-##### Tamal Khan 
-Senior Research fellow, All India Institute of Medical Sciences, New Delhi, India.
 
@@
 Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year [@injuries2020].
-Several trauma life support training programs have been developed to improve the early management of patients as they arrive at hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Several trauma life support training programs have been developed to improve the early management of patients in hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
@@
 
-Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care would not be possible.
-Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be challenging due to for example difficulties in patient follow up.
+Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible.
+Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.
 
@@
 ## Study Setting
-We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
+We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
 
@@
 
-Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
+We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
 Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
@@
 We included persons who were 15 years or older and presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
-History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
+History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter 20 of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
 
@@
 This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
-- All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients, measured and compared across trial arms as both final values and as change from baseline.
+- All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients.
 
@@
 
-We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.
+We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the intervention.
 
@@
 
-The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
+The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained.
 All injuries were coded according to the International Classification of Diseases version 10 (ICD-10).
@@
 We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial.
-The results of these consultations will be published separately.
+The results of these consultations are published separately [@david_patient-reported_2024].
 
@@
 We were granted research ethics approval from the institutional ethics committees at each participating hospital.
-For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for Postgraduate Institute of Medical Education and Research, Chandigarh.
+For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for the Postgraduate Institute of Medical Education and Research, Chandigarh.
 
@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S1.
-The variables with the maximum amount of missing data were complications, reported in Supplementary Tables S1-12.
+The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.
 The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
-Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
+Overall, the data were similarly distributed, but there were considerably more missing values in the extracted data compared to the observed data.
 
@@
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially complications (reported in Supplementary Table S1-12) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Some variables, especially the cost of treatment (reported in Supplementary Table S1-12) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
@@
 The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
-This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.
-This heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
+This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
 
@@
 Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016].
-In contrast, some studies have found that these programmes are associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
+In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
 
@@
 # Contributorship statement
-<!--
-MGW conceived of the study. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH,
+
+MGW conceived of the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH,
 MGW, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
-study. DKV, KDS, MK, and MGW drafted the first version of the
+study. MGV, DKV, KDS, and MK drafted the first version of the
 protocol. AG, HS, and SD drafted the first version of the patient and
@@
 PP, RS, SC, SD, and VK contributed to the refinement of the
-protocol. AR, AC, C, DK, GG, MK, MT, and VK are representatives of
-prospective participating hospitals.
--->
+protocol.
+DB, JB, SC, LFT, GG, MK, TK, CJ, NR, RS, KDS, LS and VP interpreted the results and revised the manuscript.
+AR, AC, C, DK, GG, MK, MT, VK and VP are representatives of participating hospitals.
 
@@
 
-The final anonymized dataset and code for analysis are released publicly.
+The code for analysis is released publicly. The final anonymized dataset is available from the corresponding author on request.
 

### 2024-11-11 pilot-publication/functions/get_basic_results.R

@@ -2,8 +2,9 @@
 #'
 #' Function to calculate basic results.
 #' @param data A data.frame with the trial data. No default.
+#' @param missing.variables A character vector of variable names to include in the missing data summary. Defaults to NULL, in which case all variables are included.
 #' @export
-get_basic_results <- function(data) {
+get_basic_results <- function(data, missing.variables = NULL) {
     results <- list()
 
     # Dates
@@ -82,7 +83,7 @@ get_basic_results <- function(data) {
     results$n.lost.to.follow.up <- with(data, sum(is.na(outcomes__alive_after_30_days)))
     results$rate.lost.to.follow.up <- round(results$n.lost.to.follow.up / nrow(data) * 100)
     results$p.missing.in.hospital.mortality <- with(data, round(sum(is.na(outcomes__discharge_alive)) / nrow(data) * 100))
-    missing.data.summary <- get_missing_data_summary(data)
+    missing.data.summary <- get_missing_data_summary(data, variable.names = missing.variables)
     results$min.missing.data <- min(missing.data.summary$p.missing)
     results$max.missing.data <- max(missing.data.summary$p.missing)
     return(results)

### 2024-11-11 pilot-publication/functions/label_outcome_variables.R

@@ -1,0 +1,18 @@
+#' Label outcome variables
+#'
+#' This function labels the outcome variables in the data frame with the labels from the codebook.
+#'
+#' @param data The data frame containing the outcome variables.
+#'
+#' @return The data frame with the outcome variables labelled.
+label_outcome_variables <- function(data) {
+    all.outcomes <- all_outcomes()
+    for (outcome in all.outcomes) {
+        # Check if the outcome variable has a name and a label before proceeding
+        if (!is.null(names(outcome)) && !is.null(data[[outcome$name]]) && !is.null(outcome$label)) {
+            # Label the outcome variable
+            labelled::var_label(data[[outcome$name]]) <- outcome$label
+        }
+    }
+    return(data)
+}

### 2024-11-11 pilot-publication/functions/prepare_data.R

@@ -187,11 +187,14 @@ prepare_data <- function(data, codebook = NULL) {
     labelled::var_label(prepared.data$incident__moi) <- "Mechanism of injury"
     labelled::var_label(prepared.data$blunt.multisystem.trauma) <- "Blunt multisystem trauma"
     labelled::var_label(prepared.data$penetrating.trauma) <- "Penetrating trauma"
-    labelled::var_label(prepared.data$shock) <- "Shock (SBP ≤ 90 mmHg)"
+    labelled::var_label(prepared.data$shock) <- "Shock (SBP &le; 90 mmHg)"
     labelled::var_label(prepared.data$severe.tbi) <- "Severe traumatic brain injury"
-    labelled::var_label(prepared.data$elderly) <- "Elderly (Age ≥ 65 years)"
+    labelled::var_label(prepared.data$elderly) <- "Elderly (Age &ge; 65 years)"
     labelled::var_label(prepared.data$male) <- "Male"
     labelled::var_label(prepared.data$female) <- "Female"
+
+    ## Add variable labels for outcomes
+    prepared.data <- label_outcome_variables(prepared.data)
 
     ## Return prepared data
     return(prepared.data)

### 2024-11-11 pilot-publication/main.R

@@ -19,13 +19,7 @@ codebook <- do.call(noacsr::kobo_get_project_codebook, codebook.arguments)
 ## Prepare data
 data <- prepare_data(data, codebook)
 
-## Define basic results
-results <- get_basic_results(data)
-
-## Create CONSORT diagram
-create_consort_diagram(results)
-
-## Create table of sample characteristics
+## Define variables for sample characteristics table
 table.variables <- c(
     "patinfo__pt_age", "elderly",
     "patinfo__pt_gender",
@@ -45,6 +39,14 @@ table.variables.full <- unique(c(
     categorical_outcomes() |> names(),
     quantitative_outcomes() |> names()
 ))
+
+## Define basic results
+results <- get_basic_results(data, missing.variables = table.variables.full)
+
+## Create CONSORT diagram
+create_consort_diagram(results)
+
+## Create table of sample characteristics
 table.data <- data %>%
     select(all_of(table.variables.full)) %>%
     select(-arm)

### 2024-11-12 pilot-publication/manuscript.Rmd
Minor edits in the manuscript

@@
 
-##### Debojit Basak`r insert_orcid_link("90000-0002-8378-9689")`
+##### Debojit Basak `r insert_orcid_link("90000-0002-8378-9689")`
 Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Johanna Berg (JB)`r insert_orcid_link("0000-0001-7553-7337")`
+##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")`
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Shamita Chatterjee (SC)`r insert_orcid_link("0000-0002-9460-108X")`
+##### Shamita Chatterjee (SC) `r insert_orcid_link("0000-0002-9460-108X")`
 Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Li Fell<U+00E4>nder-Tsai (LFT)`r insert_orcid_link("0000-0003-0693-6080")`
+##### Li Felländer-Tsai (LFT) `r insert_orcid_link("0000-0003-0693-6080")`
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Geeta Ghag (GG)`r insert_orcid_link("0000-0001-5561-362X")`
+##### Geeta Ghag (GG) `r insert_orcid_link("0000-0001-5561-362X")`
 Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
@@
 
-##### Catherine Juillard (CJ)`r insert_orcid_link("0000-0003-0847-8933")`
+##### Catherine Juillard (CJ) `r insert_orcid_link("0000-0003-0847-8933")`
 Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
 
-##### Vipul Nandu`r insert_orcid_link("0000-0002-3366-5610")`
+##### Vipul Nandu `r insert_orcid_link("0000-0002-3366-5610")`
 Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
 
-##### Nobhojit Roy (NR)`r insert_orcid_link("0000-0003-2022-7416")`
+##### Nobhojit Roy (NR) `r insert_orcid_link("0000-0003-2022-7416")`
 Program for Global Surgery and Trauma, The George Institute of Global Health, New Delhi, India.
@@
 
-##### Rajdeep Singh (RS)`r insert_orcid_link("0000-0001-6593-2624")`
+##### Rajdeep Singh (RS) `r insert_orcid_link("0000-0001-6593-2624")`
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
 
-##### Kapil Dev Soni (KDS)`r insert_orcid_link("0000-0003-1214-4119")`
+##### Kapil Dev Soni (KDS) `r insert_orcid_link("0000-0003-1214-4119")`
 Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
 
-##### Lovisa Strömmer (LS)`r insert_orcid_link("0000-0001-5424-7111")`
-Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
-
-Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.
+##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")`
+Division of Surgery and Oncology, Department of Clinical Science, Intervention and Technology (CLINTEC),  and Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
 

### 2024-11-14 pilot-publication/manuscript.Rmd
Update manuscript

@@
 
-##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")`
+##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")` <!-- Accepted -->
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Shamita Chatterjee (SC) `r insert_orcid_link("0000-0002-9460-108X")`
+##### Shamita Chatterjee (SC) `r insert_orcid_link("0000-0002-9460-108X")` <!-- Accepted -->
 Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Li Felländer-Tsai (LFT) `r insert_orcid_link("0000-0003-0693-6080")`
+##### Li Felländer-Tsai (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted -->
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Monty Khajanchi (MK)`r insert_orcid_link("0000-0002-0898-6391")`
+##### Monty Khajanchi (MK)`r insert_orcid_link("0000-0002-0898-6391")` <!-- Accepted -->
 WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
@@
 
-##### Catherine Juillard (CJ) `r insert_orcid_link("0000-0003-0847-8933")`
+##### Catherine Juillard (CJ) `r insert_orcid_link("0000-0003-0847-8933")` <!-- Accepted -->
 Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
@@
 
-##### Rajdeep Singh (RS) `r insert_orcid_link("0000-0001-6593-2624")`
+##### Rajdeep Singh (RS) `r insert_orcid_link("0000-0001-6593-2624")` <!-- Accepted -->
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
 
-##### Kapil Dev Soni (KDS) `r insert_orcid_link("0000-0003-1214-4119")`
+##### Kapil Dev Soni (KDS) `r insert_orcid_link("0000-0003-1214-4119")` <!-- Accepted -->
 Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
 
-##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")`
+##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")` <!-- Accepted -->
 Division of Surgery and Oncology, Department of Clinical Science, Intervention and Technology (CLINTEC),  and Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Marie Hasselberg (MH)
+##### Marie Hasselberg (MH) <!-- Accepted -->
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Debabrata Kundu (DK)
+##### Debabrata Kundu (DK) <!-- Accepted -->
 Department of Surgery, Medical College Kolkata, Kolkata, India.
@@
 
-##### Arun Prasad (AP)
+##### Arun Prasad (AP) <!-- Accepted -->
 
@@
 
-##### Harris Solomon (HS)
+##### Harris Solomon (HS) <!-- Accepted -->
 Department of Cultural Anthropology and the Duke Global Health Institute, Duke University, Durham, USA.
 
-##### Megha Tandon (MT)
+##### Megha Tandon (MT) <!-- Accepted -->
 Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
@@
 Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-Our study represents the first published attempt at a controlled trial of the effect of trauma life support training och patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
+Our study represents the first published attempt at a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
 

### 2024-11-20 manuscript.Rmd
Move everything to top directory

--- 2024-11-14 pilot-publication/manuscript.Rmd
@@
 ## libreoffice --headless --convert-to pdf manuscript.docx
-title: "Assessing the Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study"
+title: "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study"
 author: Trauma life support training Effectiveness Research Network (TERN) collaborators
@@
 ## Corresponding author
-Martin Gerdin Wärnberg (MGW)`r insert_orcid_link("0000-0001-6069-4794")`
+
+##### Martin Gerdin Wärnberg, MD, PhD (MGW) `r insert_orcid_link("0000-0001-6069-4794")`
 
@@
 
-##### Debojit Basak `r insert_orcid_link("90000-0002-8378-9689")`
+##### Debojit Basak, MS `r insert_orcid_link("90000-0002-8378-9689")` <!-- Accepted, debojit88@gmail.com -->
 Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Johanna Berg (JB) `r insert_orcid_link("0000-0001-7553-7337")` <!-- Accepted -->
+##### Johanna Berg, MD (JB) `r insert_orcid_link("0000-0001-7553-7337")` <!-- Accepted, johanna.berg@ki.se -->
 Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Shamita Chatterjee (SC) `r insert_orcid_link("0000-0002-9460-108X")` <!-- Accepted -->
+##### Shamita Chatterjee, MBBS, MS (SC) `r insert_orcid_link("0000-0002-9460-108X")` <!-- Accepted, dr_shamita@yahoo.com -->
 Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
 
-##### Li Felländer-Tsai (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted -->
+##### Li Felländer-Tsai, MD, PhD (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted -->
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Geeta Ghag (GG) `r insert_orcid_link("0000-0001-5561-362X")`
+##### Geeta Ghag (GG), MBBS, MS `r insert_orcid_link("0000-0001-5561-362X")` <!-- Accepted, geetaghag@hotmail.com -->
 Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
 
-##### Monty Khajanchi (MK)`r insert_orcid_link("0000-0002-0898-6391")` <!-- Accepted -->
-WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
+##### Catherine Juillard, MD, MPH (CJ) `r insert_orcid_link("0000-0003-0847-8933")` <!-- Accepted, cjuillard@mednet.ucla.edu -->
+Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
 
+##### Monty Khajanchi, MBBS, MS (MK) `r insert_orcid_link("0000-0002-0898-6391")` <!-- Accepted, monta32@gmail.com -->
 Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.
 
-##### Tamal Khan 
+##### Tamal Khan, MSc (TK) <!-- Accepted, tamal.khan440@gmail.com -->
 Senior Research fellow, All India Institute of Medical Sciences, New Delhi, India.
 
-##### Catherine Juillard (CJ) `r insert_orcid_link("0000-0003-0847-8933")` <!-- Accepted -->
-Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
-
-##### Vipul Nandu `r insert_orcid_link("0000-0002-3366-5610")` 
+##### Vipul Nandu, MBBS, MS (VN) `r insert_orcid_link("0000-0002-3366-5610")`  <!-- Accepted, dr.vipulnandu@gmail.com -->
 Department of Surgery, HBT Medical College And Dr. R N Cooper Municipal General Hospital, Mumbai, India.
 
-##### Nobhojit Roy (NR) `r insert_orcid_link("0000-0003-2022-7416")`
+##### Nobhojit Roy, MBBS, MS, PhD (NR) `r insert_orcid_link("0000-0003-2022-7416")` <!-- Accepted, nroy@georgeinstitute.org.in -->
 Program for Global Surgery and Trauma, The George Institute of Global Health, New Delhi, India.
@@
 
-##### Rajdeep Singh (RS) `r insert_orcid_link("0000-0001-6593-2624")` <!-- Accepted -->
+##### Rajdeep Singh, MBBS, MS (RS) `r insert_orcid_link("0000-0001-6593-2624")` <!-- Accepted, rajdeep.mamc@gmail.com -->
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
 
-##### Kapil Dev Soni (KDS) `r insert_orcid_link("0000-0003-1214-4119")` <!-- Accepted -->
+##### Kapil Dev Soni, MBBS, MD (KDS) `r insert_orcid_link("0000-0003-1214-4119")` <!-- Accepted, kdsoni111@gmail.com -->
 Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
 
-##### Lovisa Strömmer (LS) `r insert_orcid_link("0000-0001-5424-7111")` <!-- Accepted -->
+##### Lovisa Strömmer, MD, PhD (LS) `r insert_orcid_link("0000-0001-5424-7111")` <!-- Accepted, lisa.strommer@ki.se -->
 Division of Surgery and Oncology, Department of Clinical Science, Intervention and Technology (CLINTEC), and Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
@@
 
-Trauma life support Effectiveness Research Network (TERN) collaborators:
+Trauma life support Effectiveness Research Network (TERN) collaborators.
+
+<!--
 
@@
 
-##### Anirban Chatterjee (AC)
+##### Anirban Chatterjee (AC) <!-- Accepted -->
 Department of Orthopaedic Sciences, Medica Superspecialty Hospital, Kolkata, India.
@@
 
-<!--
 ## Word Count
-4494
- -->
+2836
 
@@
 
-MGW conceived of the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH,
-MGW, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
+MGW conceived of the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
 study. MGV, DKV, KDS, and MK drafted the first version of the

### 2024-11-20 outcome-tables.Rmd
Move everything to top directory

--- 2024-01-02 pilot-publication/outcome-tables.Rmd
@@
-# 75% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+# Results with 75% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm(75% CI)
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm(75% CI)
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm(75% CI)
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(75% CI)
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(75% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS(75% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC(75% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC(75% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS(75% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC(75% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC(75% CI)
 
@@
 
-# 85% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+\pagebreak
+
+# Results with 85% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm(85% CI)
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm(85% CI)
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm(85% CI)
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(85% CI)
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(85% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS(85% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC(85% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC(85% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS(85% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC(85% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC(85% CI)
 
@@
 
-# 95% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+\pagebreak
+
+# Results with 95% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm(95% CI)
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm(95% CI)
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm(95% CI)
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(95% CI)
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms(95% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS(95% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC(95% CI)
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC(95% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS(95% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC(95% CI)
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC(95% CI)
 
@@
 ```
+
+\pagebreak

### 2024-11-20 supplementary-materials.Rmd
Move everything to top directory

--- 2024-10-31 pilot-publication/supplementary-materials.Rmd
@@
 ---
+## Rscript -e "rmarkdown::render('supplementary-materials.Rmd')"
+## libreoffice --headless --convert-to pdf supplementary-materials.docx
 title: Supplementary Materials

### 2024-11-20 table-test.tex
Move everything to top directory

--- 2024-10-04 pilot-publication/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial},
-  pdfauthor={Trauma life support training Effectiveness Research Network (TERN) collaborators},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\documentclass{article}
+\usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{caption}
-\usepackage{longtable}
-\usepackage{colortbl}
-\usepackage{array}
-\usepackage{anyfontsize}
-\usepackage{multirow}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-\title{Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial}
-\author{Trauma life support training Effectiveness Research Network (TERN) collaborators}
-\date{}
+\usepackage{graphicx}
 
 \begin{document}
-\maketitle
-
-\hypertarget{trial-registration}{%
-\subsection{Trial registration}\label{trial-registration}}
-
-This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).
-
-\hypertarget{corresponding-author}{%
-\subsection{Corresponding author}\label{corresponding-author}}
-
-Martin Gerdin Wärnberg (MGW)
-
-Department of Global Public Health\\
-Karolinska Institutet\\
-171 77 Stockholm\\
-Sweden
-
-\href{mailto:martin.gerdin@ki.se}{\nolinkurl{martin.gerdin@ki.se}}\\
-+46 708 53 95 98 (mobile)
-
-\hypertarget{authors}{%
-\subsection{Authors}\label{authors}}
-
-\hypertarget{johanna-berg-jb}{%
-\subparagraph{Johanna Berg (JB)}\label{johanna-berg-jb}}
-
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-Emergency Medicine, Department of Internal and Emergency Medicine, Skåne University Hospital, Malmö, Sweden.
-
-\hypertarget{shamita-chatterjee-sc}{%
-\subparagraph{Shamita Chatterjee (SC)}\label{shamita-chatterjee-sc}}
-
-Department of Surgery, Institute of Post Graduate Medical Education and Research, Kolkata, India.
-
-\hypertarget{li-felluxe4nder-tsai-lft}{%
-\subparagraph{Li Felländer-Tsai (LFT)}\label{li-felluxe4nder-tsai-lft}}
-
-Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
-
-Department of Reconstructive Orthopedics, Karolinska University Hospital, Stockholm, Sweden.
-
-\hypertarget{geeta-ghag-gg}{%
-\subparagraph{Geeta Ghag (GG)}\label{geeta-ghag-gg}}
-
-Department of Surgery, HBT Medical College And Dr.~R N Cooper Municipal General Hospital, Mumbai, India.
-
-\hypertarget{monty-khajanchi-mk}{%
-\subparagraph{Monty Khajanchi (MK)}\label{monty-khajanchi-mk}}
-
-WHO Collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
-
-Seth G. S. Medical College and K.E.M. Hospital, Mumbai, India.
-
-\hypertarget{catherine-juillard-cj}{%
-\subparagraph{Catherine Juillard (CJ)}\label{catherine-juillard-cj}}
-
-Division of General Surgery, Department of Surgery, David Geffen School of Medicine at UCLA, Los Angeles, California, United States of America.
-
-\hypertarget{nobhojit-roy-nr}{%
-\subparagraph{Nobhojit Roy (NR)}\label{nobhojit-roy-nr}}
-
-WHO Collaborating Center for Research on Surgical Care Delivery in LMICs, BARC Hospital, HBNI University, Mumbai, India.
-
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-\hypertarget{rajdeep-singh-rs}{%
-\subparagraph{Rajdeep Singh (RS)}\label{rajdeep-singh-rs}}
-
-Department of Surgery, Maulana Azad Medical College, New Delhi, India.
-
-\hypertarget{kapil-dev-soni-kds}{%
-\subparagraph{Kapil Dev Soni (KDS)}\label{kapil-dev-soni-kds}}
-
-Critical and Intensive Care, JPN Apex Trauma Center, All India Institute of Medical Sciences, New Delhi, India.
-
-\hypertarget{lovisa-struxf6mmer-ls}{%
-\subparagraph{Lovisa Strömmer (LS)}\label{lovisa-struxf6mmer-ls}}
-
-Division of Surgery, Department of Clinical Science, Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
-
-Department of Surgery, Capio St Göran Hospital, Stockholm, Sweden.
-
-\hypertarget{vipul-nandu}{%
-\subparagraph{Vipul Nandu}\label{vipul-nandu}}
-
-Department of Surgery, HBT Medical College And Dr.~R N Cooper Municipal General Hospital, Mumbai, India.
-
-\hypertarget{debojit-basak}{%
-\subparagraph{Debojit Basak}\label{debojit-basak}}
-
-Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
-
-\hypertarget{tamal-khan}{%
-\subparagraph{Tamal Khan}\label{tamal-khan}}
-
-\hypertarget{non-author-contributors}{%
-\subsection{Non-author contributors}\label{non-author-contributors}}
-
-Trauma life support Effectiveness Research Network (TERN) collaborators:
-
-\hypertarget{prashant-bhandarkar-pb}{%
-\subparagraph{Prashant Bhandarkar (PB)}\label{prashant-bhandarkar-pb}}
-
-WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
-
-\hypertarget{anirban-chatterjee-ac}{%
-\subparagraph{Anirban Chatterjee (AC)}\label{anirban-chatterjee-ac}}
-
-Department of Orthopaedic Sciences, Medica Superspecialty Hospital, Kolkata, India.
-
-\hypertarget{chintamani-c}{%
-\subparagraph{Chintamani (C)}\label{chintamani-c}}
-
-Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
-
-\hypertarget{anita-gadgil-ag}{%
-\subparagraph{Anita Gadgil (AG)}\label{anita-gadgil-ag}}
-
-WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
-
-\hypertarget{marie-hasselberg-mh}{%
-\subparagraph{Marie Hasselberg (MH)}\label{marie-hasselberg-mh}}
-
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-\hypertarget{deepa-kizhakke-veetil-dkv}{%
-\subparagraph{Deepa Kizhakke Veetil (DKV)}\label{deepa-kizhakke-veetil-dkv}}
-
-Department of Surgery, Manipal Hospital, Dwarka, New Delhi.
-
-\hypertarget{vineet-kumar-vk}{%
-\subparagraph{Vineet Kumar (VK)}\label{vineet-kumar-vk}}
-
-Department of Surgery, Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai, India.
-
-\hypertarget{debabrata-kundu-dk}{%
-\subparagraph{Debabrata Kundu (DK)}\label{debabrata-kundu-dk}}
-
-Department of Surgery, Medical College Kolkata, Kolkata, India.
-
-\hypertarget{anurag-mishra-am}{%
-\subparagraph{Anurag Mishra (AM)}\label{anurag-mishra-am}}
-
-Department of Surgery, Maulana Azad Medical College, New Delhi, India.
-
-\hypertarget{priti-patil-pp}{%
-\subparagraph{Priti Patil (PP)}\label{priti-patil-pp}}
-
-WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
-
-\hypertarget{arun-prasad-ap}{%
-\subparagraph{Arun Prasad (AP)}\label{arun-prasad-ap}}
-
-Department of Surgery, Indraprastha Apollo Hospitals, New Delhi
-
-\hypertarget{amit-roy-ar}{%
-\subparagraph{Amit Roy (AR)}\label{amit-roy-ar}}
-
-Department of Surgery, Sir Nil Ratan Sircar Medical College \& Hospital, Kolkata, India.
-
-\hypertarget{siddarth-david-sd}{%
-\subparagraph{Siddarth David (SD)}\label{siddarth-david-sd}}
-
-Department of Global Public Health, Karolinska Institutet, Stockholm, Sweden.
-
-Doctors For You, Mumbai, India.
-
-\hypertarget{harris-solomon-hs}{%
-\subparagraph{Harris Solomon (HS)}\label{harris-solomon-hs}}
-
-Department of Cultural Anthropology and the Duke Global Health Institute, Duke University, Durham, USA.
-
-\hypertarget{megha-tandon-mt}{%
-\subparagraph{Megha Tandon (MT)}\label{megha-tandon-mt}}
-
-Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
-
-\hypertarget{keywords}{%
-\subsection{Keywords}\label{keywords}}
-
-Trauma management, Accident and emergency medicine, Education and training.
-
-\hypertarget{role-of-study-sponsor-and-funders}{%
-\subsection{Role of study sponsor and funders}\label{role-of-study-sponsor-and-funders}}
-
-The funding sources had no role in the design of this study nor during its execution, analyses, interpretation of the data, or decision to submit the results.
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-\textbf{Importance} Trauma life support training programmes aim to improve early trauma care but there is high-quality evidence to show that they improve patient outcomes.
-
-\textbf{Objective} To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support\textsuperscript{®} (ATLS\textsuperscript{®}) and Primary Trauma Care (PTC) with standard care on patient outcomes.
-
-\textbf{Design} A pilot pragmatic three-armed parallel, cluster randomised, controlled trial between April 2022 and February 2023. Patient follow up was 30 days.
-
-\textbf{Setting} Tertiary care hospitals across metropolitan areas in India.
-
-\textbf{Participants} Adult trauma patients and residents managing these patients.
-
-\textbf{Interventions} ATLS\textsuperscript{®} or PTC training for residents in the intervention arms.
-
-\textbf{Main Outcomes and Measures} The feasibility outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, and differences in distribution between observed and data extracted from medical records.
-The primary patient outcome was all cause mortality at 30 days from the time of arrival to the emergency department.
-
-\textbf{Results} Two hospitals were randomised to ATLS\textsuperscript{®}, two to PTC, and three to standard care.
-We included 376 patients and 22 residents.
-The percentage of patients who consented to follow up was 77\% and the percentage of residents who consented to training was 100\%.
-The lost to follow up rate was 14\%.
-The pass rate was 100\%.
-The missing data was overall low for key variables.
-Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
-Twenty-two (16)\% patients died within 30 days in the standard care arm, one (4)\% patients in the ATLS\textsuperscript{®} arm, and three (5)\% patients in the PTC arm.
-
-\textbf{Conclusions and Relevance} Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS\textsuperscript{®}, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
-
-\textbf{Trial Registration} ClinicalTrials.gov (reg. no NCT05417243)
-
-\newpage
-
-\newpage
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year\textsuperscript{1}.
-Several trauma life support training programs have been developed to improve the early management of patients as they arrive at hospital by providing a structured framework to assessment and treatment\textsuperscript{2--4}.
-
-The proprietary Advanced Trauma Life Support\textsuperscript{®} (ATLS\textsuperscript{®}) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries\textsuperscript{5,6}.
-Observational studies indicate that these programmes may improve patient outcomes\textsuperscript{7}, but there is no high quality evidence from controlled trials to support this\textsuperscript{2--4}.
-
-Several studies, including at least two randomised controlled trials\textsuperscript{8,9}, show that ATLS\textsuperscript{®} is associated with improved provider skills\textsuperscript{2}.
-Observational evidence indicates that PTC also leads to improved provider skills\textsuperscript{4}.
-The missing link is then how, and if, these improved skills translate into improved patient outcomes.
-
-Systematic reviews call for trials in settings where these programmes are not routinely implemented, but data needed to inform the planning of such a trial is missing.
-We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS\textsuperscript{®} and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\hypertarget{trial-design}{%
-\subsection{Trial Design}\label{trial-design}}
-
-We piloted a three-armed cluster randomised controlled trial\textsuperscript{10}.
-There were a standard care arm and two intervention arms, ATLS\textsuperscript{®} and PTC training.
-We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
-This design allowed us to assess outcomes both as final values and as change from baseline.
-
-\hypertarget{study-setting}{%
-\subsection{Study Setting}\label{study-setting}}
-
-We conducted this pilot study in seven Indian tertiary hospitals across metropolitan areas in India, where neither ATLS\textsuperscript{®}, PTC, nor any other trauma life support training program is routinely taught.
-
-\hypertarget{standard-care}{%
-\subsection{Standard Care}\label{standard-care}}
-
-Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
-These units typically include three faculty members and three to twelve residents, who are assigned a specific day of the week when they manage the emergency department.
-In the emergency department, trauma patients are initially assessed by first- or second-year residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
-Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
-
-\hypertarget{intervention}{%
-\subsection{Intervention}\label{intervention}}
-
-In each intervention arm the residents in one or two units were trained in either ATLS\textsuperscript{®} or PTC.
-For the purpose of this pilot study, our target was to train a minimum of 75\% of residents in each unit.
-We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
-The ATLS\textsuperscript{®} training was conducted in an ATLS\textsuperscript{®} certified training centre in Mumbai, according to the standard ATLS\textsuperscript{®} curriculum\textsuperscript{5}.
-The PTC training was conducted in New Delhi, according to the standard PTC curriculum\textsuperscript{6}.
-We did not modify or adapt the delivery or content of these programs during this pilot study.
-
-\hypertarget{eligibility-criteria-for-cluster-and-participants}{%
-\subsection{Eligibility Criteria for Cluster and Participants}\label{eligibility-criteria-for-cluster-and-participants}}
-
-\hypertarget{hospitals}{%
-\subsubsection{Hospitals}\label{hospitals}}
-
-We included tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
-
-\hypertarget{clusters}{%
-\subsubsection{Clusters}\label{clusters}}
-
-We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals.
-To be eligible, units could have no more than 25\% of their physicians trained in either ATLS\textsuperscript{®}, PTC, or similar training programs before the start of the pilot study.
-Those residents who had received training in the last five years were considered as trained.
-The figure of 25\% was decided through consensus in the research team, to balance feasibility and contamination of results.
-The principal investigator at each hospital selected the units for training.
-We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.
-
-\hypertarget{residents}{%
-\subsubsection{Residents}\label{residents}}
-
-Resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
-Consent was sought from the residents in each of the intervention groups before they underwent the ATLS\textsuperscript{®} or PTC training.
-
-\hypertarget{patients}{%
-\subsubsection{Patients}\label{patients}}
-
-All adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
-History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
-
-\hypertarget{outcomes}{%
-\subsection{Outcomes}\label{outcomes}}
-
-We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A complete list of outcomes is available as Supplementary Materials. Our main outcomes were:
-
-\begin{itemize}
-\tightlist
-\item
-  Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents.
-\item
-  Lost to follow up rate. This applied only to patients and was equal to the percentage of patients whom did not complete 30 day follow up, out of all enrolled patients.
-\item
-  Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents that passed the training programme, out of the total number of trained residents.
-\item
-  Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
-\item
-  Differences in distributions of observed and extracted data.
-This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
-\item
-  All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients, measured and compared across trial arms as both final values and as change from baseline.
-\end{itemize}
-
-\hypertarget{participant-timeline-and-inclusion}{%
-\subsection{Participant Timeline and Inclusion}\label{participant-timeline-and-inclusion}}
-
-\hypertarget{patients-1}{%
-\subsubsection{Patients}\label{patients-1}}
-
-Arriving patients were screened for eligibility and consented, if conscious.
-Unconscious patients were consented by the patient's representative.
-This proxy consent was reaffirmed by the patient, on regaining consciousness.
-We followed up patients at 24 hours after arrival at the emergency department, and up to 30 days after arrival at the emergency department.
-
-\hypertarget{residents-1}{%
-\subsubsection{Residents}\label{residents-1}}
-
-Participating units were screened for eligibility once hospitals confirmed their participation.
-All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms.
-The training was conducted approximately one month after the study started in that hospital.
-
-\hypertarget{sample-size}{%
-\subsection{Sample size}\label{sample-size}}
-
-We did not conduct a formal power calculation for this pilot study, as the primary aim was to assess feasibility. The time period was dictated by budget and time constraints.
-
-\hypertarget{allocation-and-blincing}{%
-\subsection{Allocation and blincing}\label{allocation-and-blincing}}
-
-We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the interventions.
-
-\hypertarget{data-collection}{%
-\subsection{Data Collection}\label{data-collection}}
-
-Data was collected over a four-month period.
-A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care.
-The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records.
-We followed up admitted patients for their complications and other in-hospital outcome measures.
-Patients who were not admitted were followed up telephonically for mortality outcomes and quality of life outcomes.
-
-\hypertarget{variables}{%
-\subsection{Variables}\label{variables}}
-
-The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained.
-For a subset of patients we also extracted data from medical records, to be able to compare their distributions with directly observed data.
-
-\hypertarget{patient-and-public-involvement}{%
-\subsection{Patient and public involvement}\label{patient-and-public-involvement}}
-
-We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial.
-The results of these consultations will be published separately.
-
-\hypertarget{data-monitoring}{%
-\subsection{Data monitoring}\label{data-monitoring}}
-
-We conducted weekly online meetings to monitor the study and data collection.
-We conducted one interim analysis approximately halfway through the study, and decided to complete the study as residents and patients were consenting to be included in the study and key variables including mortality outcomes could be collected.
-We did not use a data monitoring committee.
-
-\hypertarget{statistical-methods}{%
-\subsection{Statistical Methods}\label{statistical-methods}}
-
-We analysed all data using descriptive statistics and did not perform any formal hypothesis\textsuperscript{11}.
-Quantitative variables are summarised as mean +/- standard deviation, median, interquartile range and range.
-Qualitative variables are presented as absolute numbers and percentages.
-We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient.
-
-We compared patients outcomes in all possible combinations of trial arms.
-In each combination we compared both differences in final values and differences in change from baseline.
-For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training.
-For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.
-
-Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly.
-These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
-
-\hypertarget{ethics-and-dissemination}{%
-\subsection{Ethics and Dissemination}\label{ethics-and-dissemination}}
-
-We were granted research ethics approval from the institutional ethics committees at each participating hospital.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-We enrolled 376 trauma patients from 7 participating centres between April 2022 and February 2023.
-The standard care arm 202 patients, the ATLS\textsuperscript{®} arm enrolled 44 patients, and the PTC arm 130 patients.
-We trained a total of 22 residents, 7 in ATLS\textsuperscript{®}, and 15 in PTC.
-
-The study flow diagram is shown in Figure \ref{fig:flow-diagram} and patient sample characteristics across trial arms are shown in Table \ref{tab:sample-characteristics}.
-Overall, the number of females were 86 (23\%), the median (IQR) age was 33 (24, 46) years, and the median ISS (IQR) was 1 (0, 5).
-A total of 32 (10\%) patients died within 30 days after arrival to the emergency department.
-
-\begin{figure}
-\includegraphics[width=0.5\linewidth]{consort-flow-diagram} \caption{Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care.}\label{fig:flow-diagram}
-\end{figure}
-
-\begingroup
-\fontsize{12.0pt}{14.4pt}\selectfont
-\setlength{\LTpost}{0mm}
-\begin{longtable}{lcccc}
-\toprule
-\textbf{Characteristic}\textsuperscript{\textit{1}} & \textbf{Standard care}  N = 202 & \textbf{ATLS}  N = 44 & \textbf{PTC}  N = 130 & \textbf{Overall}  N = 376 \\ 
-\midrule\addlinespace[2.5pt]
-Age, years, median (IQR) & 35 (25, 47) & 40 (30, 57) & 30 (22, 38) & 33 (24, 46) \\ 
-Elderly, n (\%) & 15 (7\%) & 6 (14\%) & 5 (4\%) & 26 (7\%) \\ 
-Sex, n (\%) &  &  &  &  \\ 
-    Male & 160 (79\%) & 33 (75\%) & 97 (75\%) & 290 (77\%) \\ 
-    Female & 42 (21\%) & 11 (25\%) & 33 (25\%) & 86 (23\%) \\ 
-Dominating injury type, n (\%) &  &  &  &  \\ 
-    Penetrating & 13 (6\%) & 3 (7\%) & 1 (1\%) & 17 (5\%) \\ 
-    Blunt & 189 (94\%) & 41 (93\%) & 129 (99\%) & 359 (95\%) \\ 
-Blunt multisystem trauma, n (\%) & 2 (1\%) & 2 (5\%) & 6 (5\%) & 10 (3\%) \\ 
-Severe traumatic brain injury, n (\%) & 10 (5\%) & 1 (2\%) & 5 (4\%) & 16 (4\%) \\ 
-    Missing & 1 & 0 & 0 & 1 \\ 
-Shock, n (\%) & 4 (2\%) & 2 (5\%) & 4 (3\%) & 10 (3\%) \\ 
-    Missing & 7 & 3 & 4 & 14 \\ 
-Respiratory rate, breaths per minute, median (IQR) & 20 (18, 22) & 21 (20, 24) & 21 (20, 24) & 20 (19, 23) \\ 
-    Missing & 7 & 0 & 5 & 12 \\ 
-Oxygen saturation, \%, median (IQR) & 98 (97, 99) & 98 (97, 99) & 98 (98, 99) & 98 (97, 99) \\ 
-    Missing & 1 & 1 & 0 & 2 \\ 
-Heart rate, beats per minute, median (IQR) & 86 (80, 96) & 87 (73, 100) & 90 (76, 104) & 86 (78, 100) \\ 
-    Missing & 1 & 1 & 1 & 3 \\ 
-Systolic blood pressure, mmHg, median (IQR) & 123 (112, 135) & 124 (113, 131) & 122 (111, 136) & 123 (112, 135) \\ 
-    Missing & 7 & 3 & 4 & 14 \\ 
-Glasgow Coma Scale, median (IQR) & 15 (15, 15) & 15 (15, 15) & 15 (15, 15) & 15 (15, 15) \\ 
-    Missing & 2 & 1 & 0 & 3 \\ 
-Injury Severity Score, median (IQR) & 1 (0, 4) & 3 (1, 4) & 2 (0, 5) & 1 (0, 5) \\ 
-\bottomrule
-\end{longtable}
-\begin{minipage}{\linewidth}
-\textsuperscript{\textit{1}}ATLS = Advanced Trauma Life Support; PTC = Prehospital Trauma Care\\
-\end{minipage}
-\endgroup
-
-\hypertarget{outcomes-1}{%
-\subsection{Outcomes}\label{outcomes-1}}
-
-The percentage of patients who consented to follow up was 77\% and the percentage of residents who consented to training was 100\%.
-The lost to follow up rate was 14\%.
-The pass rate was 100\%.
-The missing data rate ranged from 0 to 98, with details for selected variables shown in Table \ref@(tab:sample-characteristics).
-The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-extracted).
-Overall, the data were similarly distributed, but there were considerably more missing values in extracted data compared to observed data.
-
-\begingroup
-\fontsize{12.0pt}{14.4pt}\selectfont
-\setlength{\LTpost}{0mm}
-\begin{longtable}{lcc}
-\toprule
-\textbf{Characteristic} & \textbf{Directly observed}  N = 55\textsuperscript{\textit{1}} & \textbf{Medical records}  N = 55\textsuperscript{\textit{1}} \\ 
-\midrule\addlinespace[2.5pt]
-Age, years & 34 (27, 48) & 34 (25, 50) \\ 
-    Missing & 0 & 22 \\ 
-Sex &  &  \\ 
-    Female & 10 (18\%) & 6 (18\%) \\ 
-    Male & 45 (82\%) & 27 (82\%) \\ 
-    Missing & 0 & 22 \\ 
-Dominating injury type &  &  \\ 
-    Blunt & 52 (95\%) & 29 (91\%) \\ 
-    Penetrating & 3 (5\%) & 3 (9\%) \\ 
-    Missing & 0 & 23 \\ 
-Respiratory rate, breaths per minute & 21 (18, 24) & 18 (16, 20) \\ 
-    Missing & 0 & 37 \\ 
-Oxygen saturation, \% & 98 (98, 99) & 98 (97, 100) \\ 
-    Missing & 0 & 29 \\ 
-Heart rate, beats per minute & 85 (80, 98) & 87 (84, 93) \\ 
-    Missing & 0 & 19 \\ 
-Systolic blood pressure, mmHg & 123 (112, 136) & 118 (110, 128) \\ 
-    Missing & 1 & 18 \\ 
-\bottomrule
-\end{longtable}
-\begin{minipage}{\linewidth}
-\textsuperscript{\textit{1}}Median (Q1, Q3); n (\%)\\
-\end{minipage}
-\endgroup
-
-After training, a total of 22 (16\%) patients in the standard care arm died within 30 days, compared to 1 (4\%) patients in the ATLS\textsuperscript{®} arm and 3 (5\%) patients in the PTC arm.
-The corresponding figures for in-hospital mortality were 19 (12\%)\%, 1 (4\%)\%, and 3 (4\%)\%, as shown in Table \ref{tab:outcomes}.
-Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS\textsuperscript{®} and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small.
-The results for all other outcomes are shown in Supplementary Materials.
 
@@
 \cmidrule(lr){5-7}
-\textbf{Outcome} & \textbf{Standard care}  N = 161\textsuperscript{\textit{1}} & \textbf{ATLS}  N = 28\textsuperscript{\textit{1}} & \textbf{PTC}  N = 73\textsuperscript{\textit{1}} & \textbf{Standard care vs.
+\textbf{Outcome} & \textbf{Standard care N = 161\textsuperscript{\textit{1}}} & \textbf{ATLS N = 28\textsuperscript{\textit{1}}} & \textbf{PTC N = 73\textsuperscript{\textit{1}}} & \textbf{Standard care vs.
 ATLS} & \textbf{Standard care vs.
@@
 \midrule\addlinespace[2.5pt]
-30 day mortality & 22 (16\%) & 1 (3.8\%) & 3 (4.9\%) & 12% & 11% & -1.1% \\ 
+30 day mortality & 22 (16\%) & 1 (3.8\%) & 3 (4.9\%) & 12\% & 11\% & -1.1\% \\ 
 Unknown & 26 & 2 & 12 &  &  &  \\ 
-In-hospital mortality & 19 (12\%) & 1 (3.7\%) & 3 (4.1\%) & 8.2% & 7.8% & -0.41% \\ 
+In-hospital mortality & 19 (12\%) & 1 (3.7\%) & 3 (4.1\%) & 8.2\% & 7.8\% & -0.41\% \\ 
 Unknown & 2 & 1 & 0 &  &  &  \\ 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS\textsuperscript{®} with PTC and standard care.
-Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially cause of death and complications (reported in Supplementary materials) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
-Similarly, the missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data
-
-We found that the ATLS\textsuperscript{®} and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
-These findings hints towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
-The arms differed considerably in sample size, with the ATLS\textsuperscript{®} arm having the smallest sample size.
-This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.
-This heterogeneity highlights the importance of taken different cluster sizes into account in the design of the full scale trial.
-
-All-cause 30-day mortality was missing in 14\% of patients.
-This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01\% of patients\textsuperscript{12,13}.
-Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge.
-Our missing data rate for in-hospital mortality was only 1\%, comparable to previous trials.
-
-During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
-material.
-Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.
-
-We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial.
-We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
-Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.
-
-Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results\textsuperscript{14--18}.
-Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training\textsuperscript{14--17}, whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15\%, after implementing
-ATLS\textsuperscript{®}\textsuperscript{18}.
-
-We conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
-
-\hypertarget{contributorship-statement}{%
-\section{Contributorship statement}\label{contributorship-statement}}
-
-\hypertarget{competing-interests}{%
-\section{Competing Interests}\label{competing-interests}}
-
-Several authors are ATLS\textsuperscript{®} and/or PTC instructors.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Doctors for You through grants awarded to Karolinska Institutet by the Swedish Research Council (grant number 2020-03779) and the Laerdal Foundation (grant number 2021-0048).
-
-\hypertarget{data-sharing-statement}{%
-\section{Data Sharing Statement}\label{data-sharing-statement}}
-
-The final anonymized dataset and code for analysis are released publicly.
-
-\hypertarget{protocol-deviations}{%
-\section{Protocol Deviations}\label{protocol-deviations}}
-
-\hypertarget{trial-registration-1}{%
-\subsection{Trial Registration}\label{trial-registration-1}}
-
-We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
-
-\hypertarget{outcomes-across-subgroups}{%
-\subsection{Outcomes across subgroups}\label{outcomes-across-subgroups}}
-
-Because of small numbers in the pre-specified subgroups we decided to report only descriptive data on these subgroups.
-
-\hypertarget{number-of-participating-centres}{%
-\subsection{Number of Participating Centres}\label{number-of-participating-centres}}
-
-We ended up recruiting seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
-
-\hypertarget{resident-participants}{%
-\subsection{Resident Participants}\label{resident-participants}}
-
-Emergency medicine in addition to surgery.
-
-\hypertarget{periodic-suverys-to-residents}{%
-\subsection{Periodic suverys to residents}\label{periodic-suverys-to-residents}}
-
-We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.
-
-\hypertarget{follow-up-of-residents}{%
-\subsection{Follow up of residents}\label{follow-up-of-residents}}
-
-We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.
-
-\hypertarget{data-collection-from-records}{%
-\subsection{Data collection from records}\label{data-collection-from-records}}
-
-We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.
-
-\hypertarget{selection-of-units-for-training}{%
-\subsection{Selection of units for training}\label{selection-of-units-for-training}}
-
-We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.
-
-\hypertarget{timing-of-resident-consent}{%
-\subsection{Timing of resident consent}\label{timing-of-resident-consent}}
-
-We had initially planned to ask residents for consent before
-randomisation, but this was not possible because of logistical issues the units were only finalised after the hospitals had been randomised.
-Residents were therefore approached for consent after randomisation but before training.
-
-\hypertarget{analysis-level-of-feasibility-outcomes}{%
-\subsection{Analysis level of feasibility outcomes}\label{analysis-level-of-feasibility-outcomes}}
-
-We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level.
-
-\pagebreak
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary material}\label{supplementary-material}}
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-injuries2020}{}}%
-\CSLLeftMargin{1. }
-\CSLRightInline{GBD 2019 Diseases and Injuries Collaborators. \href{}{Injuries---level 1 cause}. \emph{The Lancet} \textbf{396}, (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Mohammad2013}{}}%
-\CSLLeftMargin{2. }
-\CSLRightInline{Mohammad, A.
-\emph{et al.} \href{https://doi.org/10.1007/s00268-013-2294-0}{Educational and clinical impact of advanced trauma life support (ATLS) courses: A systematic review}.
-\emph{World Journal of Surgery} \textbf{38}, 322--329 (2013).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jayaraman2014}{}}%
-\CSLLeftMargin{3. }
-\CSLRightInline{Jayaraman, S.
-\emph{et al.} \href{https://doi.org/10.1002/14651858.cd004173.pub4}{Advanced trauma life support training for hospital staff}.
-\emph{Cochrane Database of Systematic Reviews} (2014).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Kadhum2020}{}}%
-\CSLLeftMargin{4. }
-\CSLRightInline{Kadhum, M.
-\emph{et al.} \href{https://doi.org/10.1016/j.injury.2019.10.084}{Are primary trauma care (PTC) courses beneficial in low- and middle-income countries - a systematic review}.
-\emph{Injury} \textbf{51}, 136--141 (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-acsAtls2018}{}}%
-\CSLLeftMargin{5. }
-\CSLRightInline{American College of Surgeons, C. on T. \emph{Advanced trauma life support® student course manual}. (American College of Surgeons, 2018).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ptcfPtc2018}{}}%
-\CSLLeftMargin{6. }
-\CSLRightInline{Foundation, P. T. C. \emph{Primary trauma care course manual for trauma management in locations wih limited resources}. (Primary Trauma Care Foundation, 2018).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jin2021}{}}%
-\CSLLeftMargin{7. }
-\CSLRightInline{Jin, J.
-\emph{et al.} \href{https://doi.org/10.1007/s00268-021-06065-9}{Effectiveness of quality improvement processes, interventions, and structure in trauma systems in low- and middle-income countries: A systematic review and meta-analysis}.
-\emph{World Journal of Surgery} \textbf{45}, 1982--1998 (2021).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ali1996}{}}%
-\CSLLeftMargin{8. }
-\CSLRightInline{Ali, J.
-\emph{et al.} \href{https://doi.org/10.1007/s002689900171}{Teaching effectiveness of the advanced trauma life support program as demonstrated by an objective structured clinical examination for practicing physicians}.
-\emph{World Journal of Surgery} \textbf{20}, 1121--1126 (1996).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ali1999}{}}%
-\CSLLeftMargin{9. }
-\CSLRightInline{Ali, J.
-\emph{et al.} \href{https://doi.org/10.1097/00005373-199901000-00013}{Comparison of performance of interns completing the old (1993) and new interactive (1997) advanced trauma life support courses}.
-\emph{The Journal of Trauma: Injury, Infection, and Critical Care} \textbf{46}, 80--86 (1999).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GerdinWuxe4rnberg2022}{}}%
-\CSLLeftMargin{10. }
-\CSLRightInline{Gerdin Wärnberg, M.
-\emph{et al.} \href{https://doi.org/10.1136/bmjopen-2021-057504}{A pilot multicentre cluster randomised trial to compare the effect of trauma life support training programmes on patient and provider outcomes}.
-\emph{BMJ Open} \textbf{12}, e057504 (2022).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lee2014}{}}%
-\CSLLeftMargin{11. }
-\CSLRightInline{Lee, E.
-C.
-\emph{et al.} \href{https://doi.org/10.1186/1471-2288-14-41}{The statistical interpretation of pilot trials: Should significance thresholds be reconsidered?} \emph{BMC Medical Research Methodology} \textbf{14}, (2014).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Roberts2013}{}}%
-\CSLLeftMargin{12. }
-\CSLRightInline{Roberts, I.
-\emph{et al.} \href{https://doi.org/10.3310/hta17100}{The CRASH-2 trial: A randomised controlled trial and economic evaluation of the effects of tranexamic acid on death, vascular occlusive events and transfusion requirement in bleeding trauma patients}.
-\emph{Health Technol Assess} \textbf{17}, (2013).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Sierink2016}{}}%
-\CSLLeftMargin{13. }
-\CSLRightInline{Sierink, J.
-C.
-\emph{et al.} \href{https://doi.org/10.1016/s0140-6736(16)30932-1}{Immediate total-body CT scanning versus conventional imaging and selective CT scanning in patients with severe trauma (REACT-2): A randomised controlled trial}.
-\emph{The Lancet} \textbf{388}, 673--683 (2016).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ariyanayagam1992}{}}%
-\CSLLeftMargin{14. }
-\CSLRightInline{Ariyanayagam, D. \emph{et al.} The impact of the ATLS course on traffic accident mortality in trinidad and tobago. \emph{West Indian Medical Journal} \textbf{41}, 72--74 (1992).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ciouxe8Peuxf1a2016}{}}%
-\CSLLeftMargin{15. }
-\CSLRightInline{Cioè-Peña, E.
-\emph{et al.} \href{https://doi.org/10.1177/1460408616672491}{Development and implementation of a hospital-based trauma response system in an urban hospital in san salvador, el salvador}.
-\emph{Trauma} \textbf{19}, 118--126 (2016).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Petroze2014}{}}%
-\CSLLeftMargin{16. }
-\CSLRightInline{Petroze, R.
-T.
-\emph{et al.} \href{https://doi.org/10.1007/s00268-014-2899-y}{Can focused trauma education initiatives reduce mortality or improve resource utilization in a low-resource setting?} \emph{World Journal of Surgery} \textbf{39}, 926--933 (2014).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-VanHeng2008}{}}%
-\CSLLeftMargin{17. }
-\CSLRightInline{Van Heng, Y.
-\emph{et al.} \href{https://doi.org/10.1017/s1049023x00006282}{Non-doctors as trauma surgeons?
-A controlled study of trauma training for non-graduate surgeons in rural cambodia}.
-\emph{Prehospital and Disaster Medicine} \textbf{23}, 483--489 (2008).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Wang2010}{}}%
-\CSLLeftMargin{18. }
-\CSLRightInline{Wang, P.
-\emph{et al.} Comparison of severe trauma care effect before and after advanced trauma life support training.
-\emph{Chinese Journal of Traumatology (English Edition)} \textbf{13}, 341--344 (2010).}
-
-\end{CSLReferences}
-
 \end{document}

### 2024-11-21 manuscript.Rmd
Submit to JAMA Surgery

@@
 
-##### Debojit Basak, MS `r insert_orcid_link("90000-0002-8378-9689")` <!-- Accepted, debojit88@gmail.com -->
+##### Debojit Basak, MS `r insert_orcid_link("0000-0002-8378-9689")` <!-- Accepted, debojit88@gmail.com -->
 Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.

### 2024-11-21 supplementary-materials.Rmd
Submit to JAMA Surgery

--- 2024-10-31 pilot-publication/supplementary-materials.Rmd
@@
 ---
+## Rscript -e "rmarkdown::render('supplementary-materials.Rmd')"
+## libreoffice --headless --convert-to pdf supplementary-materials.docx
 title: Supplementary Materials
-subtitle: "Assessing the Feasibility of a Pilot Cluster Randomised Trial of the Effect of Trauma Life Support Training: A Pilot Study"
+subtitle:"Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study"
 output:

### 2024-12-23 manuscript.Rmd
Minor edits to manuscript
Format paper for BJS submission

@@
 martin.gerdin@ki.se  
-+46 708 53 95 98 (mobile)
++46 708 53 95 98 (mobile)figure
 
@@
 
-##### Li Felländer-Tsai, MD, PhD (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted -->
+##### Li Felländer-Tsai, MD, PhD (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted, li.tsai@ki.se -->
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Anurag Mishra (AM)
+##### Anurag Mishra (AM) <!-- Accepted -->
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
@@
 
-**Importance** There is no high-quality evidence to show that trauma life support training programmes improve patient outcomes.
-
 **Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes, and to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial.
@@
 
-**Conclusions and Relevance** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
+**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
 
 **Trial Registration** ClinicalTrials.gov (reg. no NCT05417243)
+
+# Strengths and limitations of this study
+
+- First attempt to conduct a cluster randomised controlled trial of the most widely used trauma life support training programmes.
+- Prospective data collection with direct observations by dedicated research officers.
+- Participating centers' heterogeneity may affect the study estimates and bias the results.
+- Broad patient eligibility criteria may have included patients with minor injuries, making it difficult to estimate robust effect sizes.
 

### 2024-12-23 cover-letter.qmd
Format cover letter for BJS submission

---
title: "Cover letter"
format: pdf
---

\vspace{-2cm}

Dear Editor,

I am submitting our manuscript, “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study” for consideration for publication in the British Journal of Surgery.

This study evaluates the feasibility of conducting a full-scale trial to assess the impact of trauma life support training programs on patient outcomes. The findings include consent rates, data quality, and effect size estimates, which are essential for planning a future full-scale trial. We believe this work addresses an important gap in evidence for trauma care and will be of interest to your readers.

This manuscript has not been published or submitted to any journal, but it has been posted on the pre-print server medRxiv: https://doi.org/10.1101/2024.03.13.24304236. No other related papers from this study have been published, posted, or submitted.

I am the corresponding author and can be contacted at:

---

\begin{center}
Martin Gerdin Wärnberg, MD, PhD

Department of Global Public Health

Karolinska Institutet

171 77 Stockholm, Sweden

Email: martin.gerdin@ki.se

Phone: +46 708 53 95 98
\end{center}

---

Thank you for considering our manuscript.

Sincerely,

Martin Gerdin Wärnberg, MD, PhD

### 2025-01-08 cover-letter.qmd
Restore directory state to BMJ Open version

@@
 
-I am submitting our manuscript, “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study” for consideration for publication in the British Journal of Surgery.
+I am submitting our manuscript, “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study” for consideration for publication in JAMA Surgery.
 
 This study evaluates the feasibility of conducting a full-scale trial to assess the impact of trauma life support training programs on patient outcomes.
-The findings include consent rates, data quality, and effect size estimates, which are essential for planning a future full-scale trial.
+The findings include consent rates, data quality, and effect size estimates, which are essential for planning future research.
 We believe this work addresses an important gap in evidence for trauma care and will be of interest to your readers.

### 2025-01-08 manuscript.Rmd
Restore directory state to BMJ Open version

@@
 
-##### Li Felländer-Tsai, MD, PhD (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted, li.tsai@ki.se -->
+##### Li Felländer-Tsai, MD, PhD (LFT) `r insert_orcid_link("0000-0003-0693-6080")` <!-- Accepted -->
 Division of Orthopaedics and Biotechnology, Department of Clinical Science Intervention and Technology (CLINTEC), Karolinska Institutet, Stockholm, Sweden.
@@
 
-##### Anurag Mishra (AM) <!-- Accepted -->
+##### Anurag Mishra (AM)
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.

### 2025-01-08 cover-letter.qmd
Update cover letter for BMJ Open

@@
 
-I am submitting our manuscript, “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study” for consideration for publication in JAMA Surgery.
+I am submitting our manuscript, “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study” for consideration for publication in BMJ Open, the protocol of which was published in your journal in 2020 (https://bmjopen.bmj.com/content/12/4/e057504).
 
@@
 
-This manuscript has not been published or submitted to any journal, but it has been posted on the pre-print server medRxiv: https://doi.org/10.1101/2024.03.13.24304236.
+This manuscript has not been published or submitted to any journal, but it has been posted on the pre-print server medRxiv (https://doi.org/10.1101/2024.03.13.24304236).
 No other related papers from this study have been published, posted, or submitted.

### 2025-01-08 manuscript.Rmd
Add non-author contributors to manuscript Rmd version

@@
 
-Trauma life support Effectiveness Research Network (TERN) collaborators.
-
 <!--
-
 ##### Prashant Bhandarkar (PB)
 WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
+-->
 
@@
 
+<!--
 ##### Chintamani (C)
 Department of Surgery, Vardhman Mahavir Medical College and Safdarjung Hospital, New Delhi, India.
+-->
 
+<!--
 ##### Anita Gadgil (AG)
 WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
+-->
 
@@
 
+<!--
 ##### Deepa Kizhakke Veetil (DKV)
 Department of Surgery, Manipal Hospital, Dwarka, New Delhi.
+-->
 
+<!--
 ##### Vineet Kumar (VK)
 Department of Surgery, Lokmanya Tilak Municipal Medical College and General Hospital, Mumbai, India.
+-->
 
@@
 
-##### Anurag Mishra (AM)
+##### Anurag Mishra (AM) <!-- Accepted -->
 Department of Surgery, Maulana Azad Medical College, New Delhi, India.
 
+<!--
 ##### Priti Patil (PP)
 WHO collaboration Centre for Research in Surgical Care Delivery in LMIC, Mumbai, India.
+-->
 
@@
 
+<!--
 ##### Amit Roy (AR)
 Department of Surgery, Sir Nil Ratan Sircar Medical College & Hospital, Kolkata, India.
+-->
 
+<!--
 ##### Siddarth David (SD)
@@
 Doctors For You, Mumbai, India.
+-->
 

### 2025-06-27 point-by-point-response.qmd
Start replying to the reviewers' comments

Dear Editor and Reviewers,

Thank you for a very thorough review of our manuscript, your comments and suggestions have been very helpful in improving the manuscript. We have addressed all of the comments and provided a detailed response below.

On behalf of all the authors,

Martin Gerdin Wärnberg, MD, PhD
Department of Global Public Health
Karolinska Institutet
Stockholm, Sweden

## Response to Editor's Comments:

> 1. Please revise the title of your manuscript to include the setting. This is the preferred format of the journal.

We have revised the title to include the setting. The new title is "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India".

> 2. Please revise the ‘Strengths and limitations of this study’ section of your manuscript (after the abstract). This section should contain up to five short bullet points, no longer than one sentence each, that relate specifically to the methods. The novelty, aims, results or expected impact of the study should not be summarised here.

The strengths and limitations section now focuses on the methods, and have been revised in the light of the comments from the reviewers.

>3. Please ensure that you have fully discussed the methodological limitations of the study in the Discussion section of the main text.

We have revised the discussion to include the methodological limitations highlighted by the reviewers.

## Response to Reviewer 1, Dr.  Feroze  Sidhwa, San Joaquin General Hospital:

> “We found that the ATLS ® and PTC arms had lower 30-day mortality compared to the PTC and standard care arms. This finding could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS ® arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.”

> If this is the case then these outcomes simply should not be reported. Since this is a feasibility trial the major concern should be assessing missing data, which it seems to be, but this statement will be controversial and may distract from the otherwise important and better-supported conclusions. I would suggest removing discussion and reporting of mortality, and instead simply report on whether or not mortality data was missing, and if so was it differentially missing in the different arms of the study. If the study is not powered to detect a difference and no measures of statistical significance were calculated then this is not useful information to the reader.

> The above is the only concern I had when I chose "no" for "Are the discussion and conclusions justified by the results?"

Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial. We have decided to keep the information in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and avoid any conclusions about a potential effect of the training.

## Response to Reviewer 2, Dr. Wietske H.W. Ham, University of Applied Sciences:

> Thank you for submitting this relevant paper and the impressive work behind it. It was an honor to review the paper. Here are some suggestions and comments, which I hope will be helpful in further developing the paper. The aim of the study is to gain insight into the feasibility of a larger study. At the same time, more results are presented than necessary to answer this question. This is not clear to the reader. It is unclear to me whether patients who ultimately did not regain consciousness were also included in this study. This should be described more clearly. It also seems that, given the research period, the study sites were included sequentially; if so, the order and period should be described. It is not sufficiently clear to me how the statistical methods used contribute to answering the research question, namely the feasibility of an RCT and determining an effect size (and other measures, which are too vague). It seems that in the description of this manuscript, the pilot outcomes are mixed with the outcomes of the trial itself, which is confusing. This is likely due to the way it is written (it needs to be better explained and specified): the authors describe results, but also the outcomes. It is unclear in the current writing how the outcomes contribute to answering the research question. The flow diagram does not clearly show how many participants did not participate or were missed. The design of the study itself seems to have been conducted properly. In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot. I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question. Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.

## Response to Reviewer 3, Dr. Sheila Sprague, McMaster University:

> Thank you for the opportunity to review the manuscript titled “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study”.  This manuscript describes an important trial with the potential to save many lives at hospitals across India.  There are many sections in this manuscript which are unclear.  The overall structure of the manuscript also needs to be improved upon.  Detailed comments are below.

> Introduction
> - Provide clear objectives for the both the pilot phase and the definitive trial phase

> Methods
> - Describe what information is being collected on standard of care at each hospital as it varies.
> - Provide additional details on ATLS and PTC training interventions.
> - Include details on the cluster selection.
> - The authors should comment on the sample size for a pilot study.  For example, why did they select seven clusters? They should also comment on the number of residents and patients required to allow for a meaningful analysis of the feasibility outcomes.
> - The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.
> - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment. Authors should have used the computer-generated randomization system at time of randomization for each site.
> - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.
> - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.  
> - Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.
> - The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.
> - The statistical analysis section is unclear.  It should list the descriptive analyses plans and then the analyses used to assess the feasibility objectives. Feasibility pilot studies are not powered to detect meaningful differences between groups. Rather their goal is to determine whether a definitive trial is practical by assessing feasibility metrics. The clinical outcomes (the outcomes for the definitive trial) could be reported descriptively, but the use of comparative statistics (differences in final values, differences from baseline, confidence intervals, and subgroup analyses) to assess these outcomes is not appropriate during the pilot phase and potentially misleading, especially without a formal sample size calculation being conducted.
> - The statistical analyses section also describes inconsistent times for outcome comparisons. The intervention arms measure change from baseline over one month pre-training vs. three months post-training, while the control arm measures change over a four-month period. For the definitive trial, the timing of assessment should be consistent across all arms to avoid biased comparisons.

> Results
> - Add a description of each cluster.
> - Comment on why the number of residents and patients varied across the treatment groups.
> - A few key prognostic factors differ across the treatment groups (age, ISS).
> - Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.
> - The results of feasibility outcomes should be reported clearly in a table, with a column indicating whether the feasibility outcome was successfully achieved.
> - Mortality should only be reported once in the results section.
> - Report the resident outcomes and the participant outcomes separately.
> - List the protocol deviations in the results section, not in the discussion.
> - It is unclear the importance and relevance of the before training after training data (Table S1).
> - The EQ-5D should be scored as opposed to showing the results for each question.
> - Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.  
> - The absolute difference versus relative difference does not add to the analyses and could be removed.
> - As this a pilot study, outcome data should be reported descriptively in a clearer manner.
> - There are too many outcomes.

> Discussion
> - Remove all results from the discussion and place them in the results section, then discuss the results in the discussion section.
> - Discuss the implication of adding the seventh site in more detail.
> - Discuss if there are methods to improve participant follow-up for the definitive trial.
> - Discuss if there are methods to reduce missing data for the definitive phase.  
> - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?
> - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.
> - Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).

## Response to Reviewer 4, Dr. Kyle Dack, University of Bristol:

> This is a well written paper, with transparent and thorough methodological reporting, and a reliable statistical approach. The purpose of the analysis was to inform a decision on the feasability of a full trial, and as such the results are interpreted with appropriate caution.
>
> I have only minor comments about certain study design aspects which I found unclear - particularly the observed/extracted data comparison.

> Methods:
> 1. Intervention – how long was the training programme, and who was it delivered by? Was the training performed in stages, e.g. to different cohorts of residents within the hospital, or all at once? The first paragraph stated that there is a 1 month observation phase followed by a 3 month intervention phase, but its not clear to me if this means the intervention continued over 3 months, or if the intervention was applied at the beginning of this phase, followed by ~3 months of patient follow-up. Further intervention in the study design and intervention sections could clarify this.

> 2. Variables – more information on the injury severity score would be useful to readers, because I am unsure how this variable was planned to be used (it is not an outcome), and how precisely it was calculated. A reference is provided, but a high-level overview of the approach (including the range of potential scores) would help readers understand this measurement better.

> 3. “Differences in distributions of observed and extracted data” – I am unsure how observed data differs from extracted data – these terms were not previously explained or defined. Does it mean the observation period vs the 3 month intervention period? “distributions” is also somewhat unclear, histogram distributions, or a specific tendency statistic?

> 4. Statistical analysis: “We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient” which outcomes does this apply to?

> 5. The analysis involved comparing the 1st month and 2nd-4th months, but it is not described how outcomes will be summed – e.g. consent rate, I imagine this the total consent rate in month 1 vs months 2-4? It would be helpful to directly confirm this in the text.

> Results
> 6. Table 1. Missing counts are reported for some but not all measurements. Does the absence of a count indicate complete data? If so, a footnote stating this would help readers.

> 7. Table 2. As mentioned in a prior comment it is unclear to me how these two groups are defined. Also, both have an N of 55, assuming these are patients it means there is a total N of 110, less than the total of 376. Why do the numbers differ?

> 8. Table S1. Suggest clarifying that this is patient characteristics in the title.

> 9. Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)

### 2025-06-27 manuscript.Rmd
Start replying to the reviewers' comments

@@
 ## libreoffice --headless --convert-to pdf manuscript.docx
-title: "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study"
+title: "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India"
 author: Trauma life support training Effectiveness Research Network (TERN) collaborators
@@
 
--->
-
 ## Keywords
@@
 
-- First attempt to conduct a cluster randomised controlled trial of the most widely used trauma life support training programmes.
 - Prospective data collection with direct observations by dedicated research officers.
+- Lack of a priori defined success criteria and thresholds for the feasibility outcomes.
+- Use of sealed envelopes potentially compromised allocation concealment.
 - Participating centers' heterogeneity may affect the study estimates and bias the results.
-- Broad patient eligibility criteria may have included patients with minor injuries, making it difficult to estimate robust effect sizes.
 
@@
 
-We found that the ATLS^®^ and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
-This finding could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
-The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size.
-This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
-
 All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients.
@@
 Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials.
+The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
 

### 2025-06-27 point-by-point-response.qmd
Continue to work on response to reviewers

@@
 Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial.
-We have decided to keep the information in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and avoid any conclusions about a potential effect of the training.
+We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and avoid any conclusions about a potential effect of the training.
 
@@
 Here are some suggestions and comments, which I hope will be helpful in further developing the paper.
-The aim of the study is to gain insight into the feasibility of a larger study.
-At the same time, more results are presented than necessary to answer this question.
-This is not clear to the reader.
-It is unclear to me whether patients who ultimately did not regain consciousness were also included in this study.
-This should be described more clearly.
-It also seems that, given the research period, the study sites were included sequentially; if so, the order and period should be described.
-It is not sufficiently clear to me how the statistical methods used contribute to answering the research question, namely the feasibility of an RCT and determining an effect size (and other measures, which are too vague).
-It seems that in the description of this manuscript, the pilot outcomes are mixed with the outcomes of the trial itself, which is confusing.
+
+Thank you for your comments and we have revised the manuscript in the light of your comments.
+
+> The aim of the study is to gain insight into the feasibility of a larger study. At the same time, more results are presented than necessary to answer this question. This is not clear to the reader.
+
+We agree that the manuscript, and the supplementary material, includes more results than necessary to fulfil the aim of the study.
+As you have noticed, we struggled to find a balance between staying true to the protocol and reporting the results that were most relevant in a transparent manner.
+We have now attempted to address this by clearly stating that we deviated from the original protocol in reporting the pilot, to focus on the feasibility of the full-scale trial.
+We have therefore limited the number of results that we present in the main text of the manuscript considerably.
+
+> It is unclear to me whether patients who ultimately did not regain consciousness were also included in this study. This should be described more clearly.
+
+We have clarified that patients who did not regain consciousness were included based on the consent by their representative.
+
+> It also seems that, given the research period, the study sites were included sequentially; if so, the order and period should be described.
+
+That is correct, data collection started in each cluster as they received the necessary approvals. The information about the study period in each cluster has been added as supplementary material.
+
+> It is not sufficiently clear to me how the statistical methods used contribute to answering the research question, namely the feasibility of an RCT and determining an effect size (and other measures, which are too vague).
+
+We have clarified how we used descriptive statistics to summarize patient characteristics and feasibility outcomes.
+We now describe the methods used for the additional analyses according to the original protocol in the supplementary materials.
+
+> It seems that in the description of this manuscript, the pilot outcomes are mixed with the outcomes of the trial itself, which is confusing.
 This is likely due to the way it is written (it needs to be better explained and specified): the authors describe results, but also the outcomes.
 It is unclear in the current writing how the outcomes contribute to answering the research question.
-The flow diagram does not clearly show how many participants did not participate or were missed.
-The design of the study itself seems to have been conducted properly.
-In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.
-I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question.
+
+We agree that this was confusing and have therefore attempted to streamline the paper further by presenting mortality descriptively along with the other patient characteristics, and only report the feasibility outcomes in the outcomes section.
+
+> The flow diagram does not clearly show how many participants did not participate or were missed. The design of the study itself seems to have been conducted properly.
+
+Thank you for pointing this out. We have added the number of patients who were lost to follow up in each arm to the flow diagram.
+
+> In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.
+
+> I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question.
 Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.

### 2025-06-27 manuscript.Rmd
Continue to work on response to reviewers

@@
 ## Trial registration
-This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).
+This pilot study was registered with ClinicalTrials.gov (reg.
+no NCT05417243).
+We intended to also register this pilot trial with Clinical Trials Registry - India, but this was not done because of a communication error in the management group.
 
@@
 
-**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes, and to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial.
+**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
 
@@
 Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.
-
-We therefore aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial.
+We therefore aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
 
@@
 
+## Protocol Deviations
+Protocol deviations are mentioned where relevant in this manuscript, but a list of all deviations is also included as supplementary material for completeness.
+
 ## Trial Design
-We piloted a three-armed cluster randomised controlled trial[@GerdinWärnberg2022].
+We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022].
 There were a standard care arm and two intervention arms, ATLS^®^ and PTC training.
 We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
-This design allowed us to assess outcomes both as final values and as change from baseline.
+Originally, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].
 
@@
 We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
+Details about each cluster is included as Supplementary Table S1.
 
@@
 
-## Outcomes
+## Feasibility outcomes
 
-We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A list of outcomes is available in Supplementary Table S1. Our main outcomes were:
+We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A list of outcomes is available in Supplementary Table S1. Our main feasibility outcomes were:
 
@@
 This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
-- All cause and in-hospital mortality within 30 days from the time of arrival to the emergency department among patients.
 
@@
 This proxy consent was reaffirmed by the patient, on regaining consciousness.
-We followed up patients at 24 hours after arrival at the emergency department, and up to 30 days after arrival at the emergency department.
+Patients who did not regain consciousness were included based on their representative's consent.
+We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
 
@@
 We used the `r version$version.string` Statistical Software for all analyses [@R_2024].
-We analysed all data using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
+We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
 Quantitative variables are summarised as median and interquartile range.
 Qualitative variables are presented as absolute numbers and percentages.
-We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
-
-We compared patients outcomes in all possible combinations of trial arms.
-In each combination we compared both differences in final values and differences in change from baseline.
-For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training.
-For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.
-
-Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years) [@hornor_quality_2018].
-These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
-We calculated both absolute and relative differences for each comparison, along with 75, 85, and 95% confidence intervals.
-We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
-
-In the interest of space, only the 95% confidence intervals are presented for all comparisons in Supplementary Tables. The remaining results are available from the corresponding author on request.
+Additional analyses performed according to the original protocol are described in the Supplementary Materials and their results shown in Supplementary Tables S1-12..
 
@@
 A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.
-The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
+
+After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS^®^ and PTC arms respectively.
 
@@
 The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.
+
 The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
-Overall, the data were similarly distributed, but there were considerably more missing values in the extracted data compared to the observed data.
+Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.
 
@@
 observed.vs.extracted.comparison.table
-```
-
-After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS^®^ and PTC arms respectively, as shown in Table \@ref(tab:outcomes).
-Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS^®^ and PTC arms were very small.
-The results for all other outcomes are shown in Supplementary Tables S1-12.
-
-```{r outcomes, echo = FALSE, tab.cap = "Mortality after training by the trial arms standard care, Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC)"}
-outcome.table.with.differences
 ```

### 2025-06-27 supplementary-materials.Rmd
Continue to work on response to reviewers

--- 2024-10-31 pilot-publication/supplementary-materials.Rmd
@@
 ---
+## Rscript -e "rmarkdown::render('supplementary-materials.Rmd')"
+## libreoffice --headless --convert-to pdf supplementary-materials.docx
 title: Supplementary Materials
-subtitle: "Assessing the Feasibility of a Pilot Cluster Randomised Trial of the Effect of Trauma Life Support Training: A Pilot Study"
+subtitle:"Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India"
 output:
@@
 
-# Table S1. Extended sample characteristics
-
-```{r overall-table, echo = FALSE}
-combined.sample.characteristics.table
-```
-
-```{r, child = "supplementary-outcome-tables.Rmd"}
-```
-
-# S13 Protocol Deviations
+# Protocol Deviations
 
@@
 
-We intended to register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.
+We intended to register our trial with Clinical Trials Registry - India (CTRI), but because of a communication error in the management group this was only attempted after the trial has started including patients, and CTRI only accept prospective registrations.
+We intend to register the full scale trial with both ClinicalTrials.gov and CTRI.
 
-## Outcomes across subgroups
+## Reporting
 
-Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury and elderly we decided to report only descriptive data on these subgroups.
+In the original protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial, but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials
 
@@
 We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level, because the sample sizes in individual clusters were too small to generate meaningful results.
+
+## Patient outcomes across subgroups
+
+Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury and elderly we decided to report only descriptive data on these subgroups.
+
+# Table S1. Cluster characteristics
+
+| Cluster                                                       | City       | Beds | Type    | Study period | 
++---------------------------------------------------------------+------------+------+---------+--------------+
+| Dr R N Cooper Municipal General Hospital                      | Mumbai     |      | Public  |              |
+| Seth GS Medical College and KEM Hospital                      | Mumbai     |      | Public  |              |
+| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     |      | Public  |              |
+| MEDICA Superspeciality Hospital                               | Kolkata    |      | Private |              |
+| Medical College                                               | Kolkata    |      | Public  |              |
+| Nilratan Sircar Medical College & Hospital                    | Kolkata    |      | Public  |              |
+| Postgraduate Institute of Medical Education and Research      | Chandigarh |      | Public  |              |
++---------------------------------------------------------------+------------+------+---------+--------------+
+
+# Statistical methods for additional analyses according to the original protocol
+
+We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
+We compared patients outcomes in all possible combinations of trial arms.
+In each combination we compared both differences in final values and differences in change from baseline.
+For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training.
+For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.
+
+Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years) [@hornor_quality_2018].
+These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
+We calculated both absolute and relative differences for each comparison, along with 75, 85, and 95% confidence intervals.
+We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
+In the interest of space, only the 95% confidence intervals are presented for all comparisons in Supplementary Tables.
+The remaining results are available from the corresponding author on request.
+
+# Estimated intracluster correlation coefficient
+
+The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
+
+# Table S2. Extended sample characteristics
+
+```{r overall-table, echo = FALSE}
+combined.sample.characteristics.table
+```
+
+```{r, child = "supplementary-outcome-tables.Rmd"}
+```

### 2025-06-27 supplementary-outcome-tables.Rmd
Continue to work on response to reviewers

\pagebreak

# Results with 95% confidence intervals (CI)

## Table S3. Outcomes in all patients during the entire study period, by treatment arm (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
```

\pagebreak

## Table S4. Outcomes in all patients before training, by treatment arm (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients before training, by treatment arm"]]
```

\pagebreak

## Table S5. Outcomes in all patients after training, by treatment arm (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients after training, by treatment arm"]]
```

\pagebreak

## Table S6. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arm (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
```

\pagebreak

## Table S7. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arm (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
```

\pagebreak

## Table S8. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
```

\pagebreak

## Table S9. Absolute and relative differences in outcomes after training, comparing standard care with PTC (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
```

\pagebreak

## Table S10. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
```

\pagebreak

## Table S11. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
```

\pagebreak

## Table S12. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
```

\pagebreak

## Table S13. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (95% CI)

```{r, echo = FALSE} 
outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
```

\pagebreak

### 2025-06-27 functions/create_consort_diagram.R

@@ -11,7 +11,13 @@ create_consort_diagram <- function(results) {
     str_replace(coll("n.atls.residents"), n.atls.residents) |>
     str_replace(coll("n.atls"), n.atls) |>
     str_replace(coll("n.ptc.residents"), n.ptc.residents) |>
-    str_replace(coll("n.ptc"), n.ptc)
+    str_replace(coll("n.ptc"), n.ptc) |>
+    str_replace(coll("n.control.complete.m30d"), n.control.complete.m30d) |>
+    str_replace(coll("n.control.complete.hd"), n.control.complete.hd) |>
+    str_replace(coll("n.atls.complete.m30d"), n.control.complete.m30d) |>
+    str_replace(coll("n.atls.complete.hd"), n.atls.complete.hd) |>
+    str_replace(coll("n.ptc.complete.m30d"), n.atls.complete.m30d) |>
+    str_replace(coll("n.ptc.complete.hd"), n.ptc.complete.hd)
   writeLines(consort.svg, "modified-consort-flow-diagram.svg")
   system("svgexport modified-consort-flow-diagram.svg consort-flow-diagram.png 2x")
 }

### 2025-06-27 functions/get_basic_results.R

@@ -83,6 +83,12 @@ get_basic_results <- function(data, missing.variables = NULL) {
     results$n.lost.to.follow.up <- with(data, sum(is.na(outcomes__alive_after_30_days)))
     results$rate.lost.to.follow.up <- round(results$n.lost.to.follow.up / nrow(data) * 100)
     results$p.missing.in.hospital.mortality <- with(data, round(sum(is.na(outcomes__discharge_alive)) / nrow(data) * 100))
+    results$n.control.complete.m30d <- with(data[data$arm == "Standard care", ], sum(!is.na(outcomes__alive_after_30_days)))
+    results$n.control.complete.hd <- with(data[data$arm == "Standard care", ], sum(!is.na(outcomes__discharge_alive)))
+    results$n.atls.complete.m30d <- with(data[data$arm == "ATLS", ], sum(!is.na(outcomes__alive_after_30_days)))
+    results$n.atls.complete.hd <- with(data[data$arm == "ATLS", ], sum(!is.na(outcomes__discharge_alive)))
+    results$n.ptc.complete.m30d <- with(data[data$arm == "PTC", ], sum(!is.na(outcomes__alive_after_30_days)))
+    results$n.ptc.complete.hd <- with(data[data$arm == "PTC", ], sum(!is.na(outcomes__discharge_alive)))
     missing.data.summary <- get_missing_data_summary(data, variable.names = missing.variables)
     results$min.missing.data <- min(missing.data.summary$p.missing)
     results$max.missing.data <- max(missing.data.summary$p.missing)

### 2025-07-05 point-by-point-response.qmd
Continue to work on reviewer 3's comments

@@
 
-Thank you for your comments and we have revised the manuscript in the light of your comments.
+Thank you for your comments, they were very helpful. We have revised the manuscript in the light of your comments.
 
@@
 
+We have revised the manuscript to describe the deviations where they are most relevant, and retained a list in the supplementary materials for completeness.
+
 > I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question.
 Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.
+
+We agree that the supplementary materials are very extensive.
+We included these to transparently report the results of the analyses specified in the original protocol, and we have therefore decided to keep them in the manuscript.
+We have attempted to include parts of the supplementary materials in the main text where appropriate, in response to your earlier comments.
 
@@
 
+Thank you for your thorough review. We have made major revisions in response to your comments, and we hope that the manuscript is now clearer.
+
 > Introduction
@@
 
+We have revised the objective of the pilot study to also include the aim of the full scale trial.
+The revised objective is "to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.".
+
 > Methods
 > - Describe what information is being collected on standard of care at each hospital as it varies.
+
+We did not collect in-depth data on the standard care at each hospital, and we have added a note in the methods section to clarify this.
+We have also added this as a limitation of this pilot study, and suggested that the full scale trial should include a more detailed assessment of the standard care at each hospital.
+
 > - Provide additional details on ATLS and PTC training interventions.
+
+We have added more details on the ATLS and PTC training interventions to the methods section.
+
 > - Include details on the cluster selection.
+
+We have revised the methods section to clarify that the seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.
+
 > - The authors should comment on the sample size for a pilot study.
@@
 They should also comment on the number of residents and patients required to allow for a meaningful analysis of the feasibility outcomes.
+
+We have added the rationale for the number of clusters and the number of units per cluster to the methods section.
+We did not conduct a formal power calculation, but deemed that the anticipated sample sizes would be sufficient to evaluate the feasibility outcomes.
+
 > - The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.
+
 > - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment.

### 2025-07-05 manuscript.Rmd
Continue to work on reviewer 3's comments

@@
 Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.
-We therefore aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.
+
+We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.
 
@@
 ## Protocol Deviations
-Protocol deviations are mentioned where relevant in this manuscript, but a list of all deviations is also included as supplementary material for completeness.
+
+Protocol deviations are mentioned where relevant in this manuscript, but a list of all deviations is also included as Supplementary Material S1 for completeness.
 
 ## Trial Design
+
 We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022].
@@
 We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
-Originally, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].
+In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].
 
 ## Study Setting
+
 We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
 Details about each cluster is included as Supplementary Table S1.
+Originally, we intented to include six hospitals as clusters, but decided to include a seventh when they expressed interest and as we had the budget to accommodate this.
+These seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.
 
@@
 These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
-In the emergency department, trauma patients are initially assessed by residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+In the emergency department, trauma patients are initially assessed by either surgical or emergency medicine residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
 Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
+We did not collect data on how standard care varied between the participating hospitals.
 
@@
 We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
-The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018].
-The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018].
-We did not modify or adapt the delivery or content of these programmes during this pilot study.
+The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi.
+Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.
+
+The provider courses of both programmes are 2.5 days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations, for example pediatric, pregnant and geriatric patients.
+The teaching is based on case discussions and skill stations.
+There are some important differences between the two programs.
+The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018].
+The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
 
@@
 Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
+In the published protocol, we stated that only surgical residents would be trained.
+However, in some of the participating hospitals, emergency medicine residents performed the initial resuscitation and management of trauma patients, and we therefore decided to train them instead.
 
@@
 
-We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A list of outcomes is available in Supplementary Table S1. Our main feasibility outcomes were:
+Our feasibility outcomes were:
 
@@
 - Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
-- Differences in distributions of observed and extracted data.
-This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.
+- Differences in distributions between directly observed data and data extracted from medical records.
+This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
+
+We did not specify success criteria for these outcomes.
 
@@
 All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms.
+The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training.
 The training was conducted approximately one month after the study started in that hospital.
+Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator.
 
@@
 
+We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training.
 We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
@@
 Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022].
-For a convenience sample of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
+We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and cause of death.
+A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
 
@@
 The results of these consultations are published separately [@david_patient-reported_2024].
+We initially planned to distribute periodic surveys to residents and follow them up after 30 days after training, but we changed this to interviews after the end of the study period as we thought this would provide richer data (not published).
 
@@
 We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
+Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results.
 Quantitative variables are summarised as median and interquartile range.
 Qualitative variables are presented as absolute numbers and percentages.
-Additional analyses performed according to the original protocol are described in the Supplementary Materials and their results shown in Supplementary Tables S1-12..
+Additional analyses performed according to the original protocol are described in the Supplementary Material S2 and their results shown in Supplementary Material S3 and Supplementary Tables S1-12.
 
@@
 Among residents, the pass rate was `r pass.rate`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S1.
+The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
 The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.
@@
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially the cost of treatment (reported in Supplementary Table S1-12) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
@@
 
-During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
-material S13.
-Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.
+During the course of this pilot study we deviated from the protocol in several ways.
+We have reported these deviations were appropriate in the methods section, and provide a detailed list as Supplementary
+material S1.
+The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes.
+However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [INSERT REF].
+We therefore decided to only report patient outcomes descriptively.
+Another important deviation from the protocol was the training of emergency medicine residents.
+We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines.
+The full scale study will need to accommodate this variation as part of the protocol.
 
-We attempted to minimse the impact of the lower than expected enrolment rates by including a seventh centre, but careful assessments of patient volumes as part of the screening process will be needed for the full-scale trial.
-We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol.
-Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers if this data is to be collected.
+There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial.
+First, the volumes of patients included at some of the participating hospitals were lower than expected.
+A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial.
+Second, data on complications and cause of death were almost universally missing.
+To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available.
+Third, we did not collect in-depth data on the standard care at each hospital.
+This data should be collected as part of the screening process for the full-scale trial.
 

### 2025-07-05 supplementary-materials.Rmd
Continue to work on reviewer 3's comments

@@
 
-# Protocol Deviations
+# S1 Protocol Deviations
 
@@
 
-## Reporting
+## Aim
 
@@
 
-## Number of Participating Centres
+## Number of Participating Clusters
 
-We recruited seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.
+We recruited seven hospitals as clusters instead of six, as a seventh hospital expressed interest and we had the budget to accommodate this request.
+We therefore assigned two centres each to the intervention arms and three centres to the control arm.
 
@@
 
-# Statistical methods for additional analyses according to the original protocol
+# S2 Statistical methods for additional analyses according to the original protocol
 
@@
 
-# Estimated intracluster correlation coefficient
+# S3 Estimated intracluster correlation coefficient
 

### 2025-07-06 point-by-point-response.qmd
Continue working with reviewer 3's comments

@@
 
+We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
+
 > - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment.
 Authors should have used the computer-generated randomization system at time of randomization for each site.
+
+We recognise that the use of sealed envelopes for randomisation increases the risk of bias and errors, and we have added this as a limitation of this pilot study.
+
 > - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.
+
+We have added that we did not blind the data analysts in this pilot study, but recommend doing so in the full scale trial.
+
 > - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.
+
+The data collection period in each hospital was four months, but patients were followed up for one month. We have clarified this in the methods section.
+
 > - Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.
+
+We did not specify success criteria for the feasibility outcomes, and we have added this to the methods section.
+
 > - The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.
+
+We now specify that we collected data on potential outcomes for the full scale trial, including mortality within 30 days, in-hospital mortality, complications and health related quality of life.
+
 > - The statistical analysis section is unclear.
@@
 The clinical outcomes (the outcomes for the definitive trial) could be reported descriptively, but the use of comparative statistics (differences in final values, differences from baseline, confidence intervals, and subgroup analyses) to assess these outcomes is not appropriate during the pilot phase and potentially misleading, especially without a formal sample size calculation being conducted.
+
+Thank you for pointing this out.
+We acknowledge that the estimation of potential effect sizes is not appropriate for this pilot study, and we have revised the aim of the study to reflect this and report this change as a deviation from the original protocol.
+We now report that we used descriptive statistics to analyse the feasibility outcomes and report clinical outcomes descriptively.
+
 > - The statistical analyses section also describes inconsistent times for outcome comparisons.
@@
 
+We agree.
+
 > Results
 > - Add a description of each cluster.
+
+We have added a description of each cluster as Supplementary Table S1.
+
 > - Comment on why the number of residents and patients varied across the treatment groups.
+
+We have clarified that the number of residents trained in each intervention cluster depended on the size of the units, and that the number of patients included per cluster varied between clusters depending on the volume of patients at each hospital.
+
 > - A few key prognostic factors differ across the treatment groups (age, ISS).
+
+Thank you for pointing this out and we now highlight this in the results section.
+
 > - Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.
+
+We have added a table to the results section to report the number of patients included per month.
+
 > - The results of feasibility outcomes should be reported clearly in a table, with a column indicating whether the feasibility outcome was successfully achieved.
+
+We have decided not to report the feasibility outcomes in a table, because some of them, for example the comparison of directly observed data and data extracted from medical records, were assessed qualitatively.
+We have however removed the description of the clinical outcomes from the outcomes section to avoid confusion.
+
 > - Mortality should only be reported once in the results section.
+
+Mortality is now reported together with the other patient characteristics in the results section. We have removed the table comparing clinical outcomes from the results section.
+
 > - Report the resident outcomes and the participant outcomes separately.
+
+We now report the resident outcomes and the participant outcomes separately.
+
 > - List the protocol deviations in the results section, not in the discussion.
+
+We now report the protocol deviations where relevant in the methods section, and include a list as supplementary material for completeness and easy reference.
+
 > - It is unclear the importance and relevance of the before training after training data (Table S1).
+
+We included the one month before training and three month after training data to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
+We have attempted to clarify this in the methods section.
+
 > - The EQ-5D should be scored as opposed to showing the results for each question.
+
+Currently, there is no Indian value set available for the EQ-5D-3L, and we have therefore not scored the data. We now explain this in the methods section.
+
 > - Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.
+
+We now only report the yes in the outcomes tables.
+
 > - The absolute difference versus relative difference does not add to the analyses and could be removed.
+
+We have removed the absolute difference versus relative difference from the main results section, but kept it in the supplementary materials as it was part of the original protocol.
+
 > - As this a pilot study, outcome data should be reported descriptively in a clearer manner.
+
+We have attempted to clarify the reporting of the outcome data in the results section.
+
 > - There are too many outcomes.
+
+We agree and suggest that the full scale trial should focus on the most relevant outcomes.
 
@@
 > - Remove all results from the discussion and place them in the results section, then discuss the results in the discussion section.
+
+We now only discuss results already reported in the results section in the discussion.
+
 > - Discuss the implication of adding the seventh site in more detail.
+
+We have clarified that the seventh site was included for pragmatic reasons.
+
 > - Discuss if there are methods to improve participant follow-up for the definitive trial.
+
+We now highlight the challenges of following up patients after discharge in the discussion and suggest that the full scale trial should focus on in-hospital mortality as the primary outcome.
+
 > - Discuss if there are methods to reduce missing data for the definitive phase.
+
+We have added a suggestion to the discussion to improve the completeness of the data extracted from the medical records.
+
 > - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?
+
 > - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.

### 2025-07-06 manuscript.Rmd
Continue working with reviewer 3's comments

@@
 We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
+We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
 In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].
@@
 - Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
-- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all enrolled patients.
-- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
+- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
 - Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
@@
 This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
-
-We did not specify success criteria for these outcomes.
+- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
 
@@
 We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
+The follow up period for each patient was therefore one month.
 
@@
 Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator.
+The number of residents trained in each intervention cluster varied because the size of the units varied.
 
@@
 We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
-<!--The time period was dictated by budget and time constraints.-->
+We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.<!--The time period was dictated by budget and time constraints.-->
 
@@
 
-We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. We did not blind investigators, residents or patients to the intervention.
+We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms.
+It was not possible to blind investigators, residents or patients to the intervention.
+We did not blind the data analysts.
 
@@
 Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022].
-We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and cause of death.
+We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L).
+We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022].
+We also attempted to collect data on cause of death.
 A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->
@@
 
-We enrolled `r n.patients` trauma patients from `r number_to_text(n.centres)` participating centres between `r start.date` and `r end.date`.
-The standard care arm enrolled `r n.control` patients, the ATLS^®^ arm enrolled `r n.atls` patients, and the PTC arm enrolled `r n.ptc` patients.
+We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`.
+The period of data collection and the number of patients included per month per cluster is shown in Table \@ref(tab:patients-per-month).
+The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients.
 We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.
+
+```{r patients-per-month, echo = FALSE, results = "asis", tab.cap = "Number of patients included per month per cluster."}
+patients.per.month.table |>
+  modify_caption("Number of patients included per month per cluster.")
+```
 
@@
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
+These prognostic factors differed between the trial arms.
 A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.
@@
 
-The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
-Among patients, the lost to follow up rate was `r rate.lost.to.follow.up`%.
-Among residents, the pass rate was `r pass.rate`%.
+The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%.
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
@@
 
-The differences in distributions between observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
+The differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
 Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.
+
+The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was `r pass.rate`%.
 
@@
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
+To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.
 
@@
 Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials.
+Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome.
 The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
@@
 This data should be collected as part of the screening process for the full-scale trial.
+Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors.
+The full scale trial should use a computer-generated randomisation system.
+Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial.
+Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.
 

### 2025-07-07 point-by-point-response.qmd
Complete responding to comments

@@
 
+We have added a sentence on the ethics of the full trial.
+
 > - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.
+
+Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial.
+
 > - Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).
+
+We have attempted to align our reporting with these guidelines.
 
@@
 > I have only minor comments about certain study design aspects which I found unclear - particularly the observed/extracted data comparison.
+
+Thank you for reviewing the manuscript!
 
@@
 
+We have clarified that the training was conducted at the beginning of the three month intervention phase.
+
 > 2.
@@
 A reference is provided, but a high-level overview of the approach (including the range of potential scores) would help readers understand this measurement better.
+
+We have added a description of the ISS to the methods section.
 
@@
 
+We have attempted to clarify what we mean by distributions (summary statistics) and directly observed data, in the section where we describe the feasibility outcomes.
+
 > 4. Statistical analysis: “We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient” which outcomes does this apply to?
+
+We have moved the description of this analysis to the supplementary materials, and clarified that it applies to all cause 30-day and in-hospital mortality.
 
@@
 
+We have attempted to clarify that feasibility outcomes were not assessed as differences between the observation and intervention phases, but summarised across the entire period.
+
 > Results
 > 6. Table 1. Missing counts are reported for some but not all measurements. Does the absence of a count indicate complete data? If so, a footnote stating this would help readers.
+
+We have added a footnote to Table 1 to clarify that the absence of a count indicates complete data.
 
@@
 
+We only extracted data from a convenience sample of patients to reduce the workload of the project officers collecting data.
+We have attempted to clarify this in the methods section and in the table caption.
+
 > 8. Table S1. Suggest clarifying that this is patient characteristics in the title.
 
+We have added "patient" to the title.
+
 > 9. Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)
+
+Yes, that is correct. Empty cells indicate that there were no observations. We have added a note to clarify this.

### 2025-07-07 manuscript.Rmd
Complete responding to comments

@@
 
-- Cluster randomized controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
+- Cluster randomised controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
 - Prospective data collection with direct observations by dedicated research officers.
@@
 
-Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible.
+Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible.
 Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.
@@
 
-## Standard Care
-
-Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
-These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
-In the emergency department, trauma patients are initially assessed by either surgical or emergency medicine residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
-Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
-We did not collect data on how standard care varied between the participating hospitals.
-
-## Intervention
-
-In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC.
-For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
-We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
-The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi.
-Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.
-
-The provider courses of both programmes are 2.5 days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations, for example pediatric, pregnant and geriatric patients.
-The teaching is based on case discussions and skill stations.
-There are some important differences between the two programs.
-The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018].
-The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
-
 ## Eligibility Criteria for Cluster and Participants
@@
 
+## Standard Care
+
+Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
+These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
+In the emergency department, trauma patients are initially assessed by either surgical or emergency medicine residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
+We did not collect data on how standard care varied between the participating hospitals.
+
+## Intervention
+
+In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC at the beginning of the three month intervention phase.
+For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
+We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
+The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi.
+Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.
+
+The provider courses of both programmes are two days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations.
+The teaching is based on case discussions and skill stations.
+There are some important differences between the two programmes.
+The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018].
+The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
+The ATLS^®^ programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK Charity PTC Foundation and is free of charge.
+
 ## Feasibility outcomes
@@
 - Differences in distributions between directly observed data and data extracted from medical records.
-This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
+Distribution refers to summary statistics and directly observed data refers to data collected by project officers while observing the delivery of care.
+This outcome applied to all variables that could be reasonably expected to be present in the medical records, but we decided to only extract this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
 - Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
+
+We did not prespecify criteria to judge whether to continue with the full scale trial.
 
@@
 -->
+
+## Sample size
+
+We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training.
+We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
+We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.
 
@@
 
-## Sample size
-
-We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training.
-We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
-We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.<!--The time period was dictated by budget and time constraints.-->
-
 ## Allocation and blinding
@@
 All injuries were coded according to the International Classification of Diseases version 10 (ICD-10).
-Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022].
+Based on these ICD-10 codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022].
+The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cutoff of 16 often being used to define major trauma, and 75 representing unsurvivable trauma.
 We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L).
@@
 
-```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview."}
+```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview in a convenience sample of patients."}
 observed.vs.extracted.comparison.table
@@
 
-We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care.
+We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS^®^ with PTC and standard care.
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
@@
 Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome.
-The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.
 
@@
 
-During the course of this pilot study we deviated from the protocol in several ways.
-We have reported these deviations were appropriate in the methods section, and provide a detailed list as Supplementary
+During the course of this pilot study we deviated from the protocol in several ways. We have reported these deviations where appropriate in the methods section, and provide a list as Supplementary
 material S1.
@@
 
-There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial.
+There are some important limitations of this pilot study and therefore additional lessons to be learned and factored into the design of the full-scale trial.
 First, the volumes of patients included at some of the participating hospitals were lower than expected.
@@
 In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
+We found fewer deaths in the intervention arms compared to the standard arm.
+This difference  may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial.
+A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.
 
 Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
-Our study represents the first published attempt at a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.
+Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.
 

### 2025-07-07 supplementary-materials.Rmd
Complete responding to comments

@@
 
-We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
+We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient for all cause 30-day and in-hospital mortality.
 We compared patients outcomes in all possible combinations of trial arms.
@@
 
-# Table S2. Extended sample characteristics
+# Table S2. Extended patient sample characteristics
 

### 2025-07-07 functions/save_tables_to_file.R

@@ -28,11 +28,12 @@ save_tables_to_file <- function(outcome.results.tables, supplementary.material =
                 paste0(
                     "\\pagebreak\n\n",
                     "# Results with ", ci.level, " confidence intervals (CI)\n\n",
-                    lapply(seq_along(ci.level.tables), function(ci.level.table.index) {
+                    lapply(seq_along(ci.level.tables) + 1, function(ci.level.table.index) {
                         ci.level.table.name <- names(ci.level.tables)[ci.level.table.index]
                         ci.level.table <- ci.level.tables[[ci.level.table.name]]
                         paste0(
-                            "## ", ifelse(supplementary.material, paste0("Table S", ci.level.table.index + 1, ". "), ""), ci.level.table.name, "(", ci.level, " CI)", "\n\n",
+                            "## ", ifelse(supplementary.material, paste0("Table S", ci.level.table.index + 1, ". "), ""), ci.level.table.name, " (", ci.level, " CI)", "\n\n",
+                            "Empty cells indicate that there were no observations to calculate the statistic.", "\n\n",
                             "\`\`\`{r, echo = FALSE} \n\n",
                             "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",
                             "[[\"", ci.level.table.name, "\"]]", "\n\n",

### 2025-07-07 main.R

@@ -59,9 +59,11 @@ stratified.table.data <- data %>%
 sample.characteristics.table <- create_descriptive_table(stratified.table.data,
     strata = "arm", show.all.levels = FALSE, include.overall = TRUE
 ) %>%
-    gtsummary::modify_table_styling(
-        column = label,
-        footnote_abbrev = "ATLS = Advanced Trauma Life Support; PTC = Prehospital Trauma Care; SBP = systolic blood pressure"
+    gtsummary::modify_abbreviation(
+        "ATLS = Advanced Trauma Life Support; PTC = Prehospital Trauma Care; SBP = systolic blood pressure"
+    ) %>%
+    gtsummary::modify_source_note(
+        "Missing data counts are only shown for variables with missing values. The absence of a count indicates complete data."
     ) %>%
     add_stat_label(
         label = list(

### 2025-07-07 manuscript.Rmd
Start modifying tables

@@
 We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
-In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].
+In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].
 
@@
 The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes.
-However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [INSERT REF].
+However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [@eldridge_how_2015; @teresi_guidelines_2022].
 We therefore decided to only report patient outcomes descriptively.

### 2025-07-07 supplementary-materials.Rmd
Start modifying tables

@@
 +---------------------------------------------------------------+------------+------+---------+--------------+
-| Dr R N Cooper Municipal General Hospital                      | Mumbai     |      | Public  |              |
-| Seth GS Medical College and KEM Hospital                      | Mumbai     |      | Public  |              |
-| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     |      | Public  |              |
-| MEDICA Superspeciality Hospital                               | Kolkata    |      | Private |              |
-| Medical College                                               | Kolkata    |      | Public  |              |
-| Nilratan Sircar Medical College & Hospital                    | Kolkata    |      | Public  |              |
-| Postgraduate Institute of Medical Education and Research      | Chandigarh |      | Public  |              |
+| Dr R N Cooper Municipal General Hospital                      | Mumbai     | 600  | Public  |              |
+| Seth GS Medical College and KEM Hospital                      | Mumbai     | 1800 | Public  |              |
+| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     | 1850 | Public  |              |
+| MEDICA Superspeciality Hospital                               | Kolkata    | 300  | Private |              |
+| Medical College                                               | Kolkata    | ??   | Public  |              |
+| Nilratan Sircar Medical College & Hospital                    | Kolkata    | 1910 | Public  |              |
+| Postgraduate Institute of Medical Education and Research      | Chandigarh | 1400 | Public  |              |
 +---------------------------------------------------------------+------------+------+---------+--------------+

### 2025-07-12 functions/create_patients_per_month_figure.R
Finalise patients per month figure

@@ -28,15 +28,15 @@ create_patients_per_month_figure <- function(data) {
   patients.per.month$arrival.year.month <- factor(patients.per.month$arrival.year.month)
 
   # Create a barchart with the number of patients per month per cluster
-  ggplot(patients.per.month, aes(x = arrival.year.month, y = n, fill = cluster)) +
-    geom_bar(stat = "identity", color = "black", width = 0.7) +
+  patients.per.month.figure <- ggplot(patients.per.month, aes(x = arrival.year.month, y = n, fill = cluster)) +
+    geom_bar(stat = "identity", color = "black", width = 0.65) +
     labs(x = "Year-month", y = "Number of patients", fill = "Cluster") +
     ggsci::scale_fill_bmj() +
-    ggsci::scale_color_bmj() +
     facet_grid(cluster ~ .) +
     theme_minimal() +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
+    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
+    scale_y_continuous(breaks = c(0, 15, 30))
 
   # Save the figure
-  ggsave("patients-per-month.pdf", width = 85, height = 150, units = "mm")
+  ggsave(patients.per.month.figure, filename = "patients-per-month.pdf", width = 85, height = 130, units = "mm")
 }

### 2025-07-12 main.R

@@ -46,6 +46,9 @@ results <- get_basic_results(data, missing.variables = table.variables.full)
 ## Create CONSORT diagram
 create_consort_diagram(results)
 
+## Create figure of patients per month
+create_patients_per_month_figure(data)
+
 ## Create table of sample characteristics
 table.data <- data %>%
     select(all_of(table.variables.full)) %>%

### 2025-07-12 manuscript.Rmd
Add figure of patients per month to manuscript

@@
 
-```{r patients-per-month, echo = FALSE, results = "asis", tab.cap = "Number of patients included per month per cluster."}
-patients.per.month.table |>
-  modify_caption("Number of patients included per month per cluster.")
+```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month."}
+knitr::include_graphics("patients-per-month.png")
 ```

### 2025-07-12 functions/create_patients_per_month_figure.R

@@ -25,7 +25,7 @@ create_patients_per_month_figure <- function(data) {
     paste0("-01") %>%
     as.Date()
 
-  patients.per.month$arrival.year.month <- factor(patients.per.month$arrival.year.month)
+  patients.per.month$arrival.year.month <- factor(patients.per.month$arrival.year.month, labels = unique(patients.per.month$arrival.year.month.label))
 
   # Create a barchart with the number of patients per month per cluster
   patients.per.month.figure <- ggplot(patients.per.month, aes(x = arrival.year.month, y = n, fill = cluster)) +
@@ -38,5 +38,5 @@ create_patients_per_month_figure <- function(data) {
     scale_y_continuous(breaks = c(0, 15, 30))
 
   # Save the figure
-  ggsave(patients.per.month.figure, filename = "patients-per-month.pdf", width = 85, height = 130, units = "mm")
+  ggsave(patients.per.month.figure, filename = "patients-per-month.png", width = 85, height = 130, units = "mm")
 }

### 2025-07-22 additional-online-material.Rmd
Move results to an additional online material file

--- 2025-07-05 supplementary-materials.Rmd
@@
 ---
-## Rscript -e "rmarkdown::render('supplementary-materials.Rmd')"
-## libreoffice --headless --convert-to pdf supplementary-materials.docx
-title: Supplementary Materials
-subtitle:"Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India"
+## Rscript -e "rmarkdown::render('additional-online-material.Rmd')"
+## libreoffice --headless --convert-to pdf additional-online-material.docx
+title: Additional Online Material
+subtitle: "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India"
 output:
@@
 
-# S1 Protocol Deviations
+# Statistical methods for additional analyses according to the original protocol
 
-## Trial Registration
-
-We intended to register our trial with Clinical Trials Registry - India (CTRI), but because of a communication error in the management group this was only attempted after the trial has started including patients, and CTRI only accept prospective registrations.
-We intend to register the full scale trial with both ClinicalTrials.gov and CTRI.
-
-## Aim
-
-In the original protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial, but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials
-
-## Number of Participating Clusters
-
-We recruited seven hospitals as clusters instead of six, as a seventh hospital expressed interest and we had the budget to accommodate this request.
-We therefore assigned two centres each to the intervention arms and three centres to the control arm.
-
-## Resident Participants
-
-We included emergency medicine residents in addition to surgical residents.
-
-## Periodic surveys of residents
-
-We did not distribute periodic surveys to the participating residents but discussed challenges and suggestions that they had regarding the scheduling or implementation of the training programs.
-
-## Follow up of residents
-
-We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.
-
-## Data collection from records
-
-We decided to extract data from medical records only for a convenience sample of patients to reduce the research officers' workload.
-
-## Selection of units for training
-
-We planned to use simple random sampling to select units if there were more than two eligible units in a hospital but instead the hospital principal investigator decided which units to train.
-
-## Timing of resident consent
-
-We had initially planned to ask residents for consent before randomisation, but because of logistical issues the units were only finalised after the hospitals had been randomised.
-Residents were therefore approached for consent after randomisation but before training.
-
-## Analysis level of feasibility outcomes
-
-We had planned to analyse feasibility outcomes on both an overall and individual cluster level, but we only analysed them on an overall level, because the sample sizes in individual clusters were too small to generate meaningful results.
-
-## Patient outcomes across subgroups
-
-Because of small numbers in the pre-specified subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury and elderly we decided to report only descriptive data on these subgroups.
-
-# Table S1. Cluster characteristics
-
-| Cluster                                                       | City       | Beds | Type    | Study period | 
-+---------------------------------------------------------------+------------+------+---------+--------------+
-| Dr R N Cooper Municipal General Hospital                      | Mumbai     |      | Public  |              |
-| Seth GS Medical College and KEM Hospital                      | Mumbai     |      | Public  |              |
-| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     |      | Public  |              |
-| MEDICA Superspeciality Hospital                               | Kolkata    |      | Private |              |
-| Medical College                                               | Kolkata    |      | Public  |              |
-| Nilratan Sircar Medical College & Hospital                    | Kolkata    |      | Public  |              |
-| Postgraduate Institute of Medical Education and Research      | Chandigarh |      | Public  |              |
-+---------------------------------------------------------------+------------+------+---------+--------------+
-
-# S2 Statistical methods for additional analyses according to the original protocol
-
-We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient.
 We compared patients outcomes in all possible combinations of trial arms.
@@
 
-Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years) [@hornor_quality_2018].
+Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years).
 These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
@@
 We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
-In the interest of space, only the 95% confidence intervals are presented for all comparisons in Supplementary Tables.
-The remaining results are available from the corresponding author on request.
 
-# S3 Estimated intracluster correlation coefficient
+# Estimated intracluster correlation coefficient
 
-The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
-
-# Table S2. Extended sample characteristics
-
-```{r overall-table, echo = FALSE}
-combined.sample.characteristics.table
-```
+We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient for all cause 30-day and in-hospital mortality.The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
 

### 2025-07-22 outcome-tables.Rmd
Move results to an additional online material file

--- 2024-01-02 pilot-publication/outcome-tables.Rmd
@@
-# 75% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+# Results with 75% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-# 85% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+\pagebreak
+
+# Results with 85% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-# 95% confidence level
+\pagebreak
 
-## Outcomes in all patients during the entire study period, by treatment arm
+\pagebreak
+
+# Results with 95% confidence intervals (CI)
+
+## Outcomes in all patients during the entire study period, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients before training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients before training, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Outcomes in all patients after training, by treatment arm
+\pagebreak
+
+## Outcomes in all patients after training, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms
+\pagebreak
+
+## Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with ATLS (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing standard care with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in outcomes after training, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in outcomes after training, comparing ATLS with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
@@
 
-## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC
+\pagebreak
+
+## Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 

### 2025-07-22 supplementary-materials.Rmd
Move results to an additional online material file

@@
 ## libreoffice --headless --convert-to pdf supplementary-materials.docx
-title: Supplementary Materials
+title: Supplementary Material
 subtitle:"Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India"
@@
 
-# S2 Statistical methods for additional analyses according to the original protocol
-
-We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient for all cause 30-day and in-hospital mortality.
-We compared patients outcomes in all possible combinations of trial arms.
-In each combination we compared both differences in final values and differences in change from baseline.
-For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training.
-For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.
-
-Within each combination of trial arms we had planned to conduct subgroup analyses of men, women, blunt multisystem trauma, penetrating trauma, shock (systolic blood pressure ≤ 90 mmHg), severe traumatic brain injury, and elderly (≥65 years) [@hornor_quality_2018].
-These subgroups were however too small to allow for meaningful analyses, and are therefore reported descriptively.
-We calculated both absolute and relative differences for each comparison, along with 75, 85, and 95% confidence intervals.
-We used an empirical bootstrap procedure with 1000 draws to estimate these confidence intervals.
-In the interest of space, only the 95% confidence intervals are presented for all comparisons in Supplementary Tables.
-The remaining results are available from the corresponding author on request.
-
-# S3 Estimated intracluster correlation coefficient
-
-The intracluster correlation coefficients was `r icc.m30d` for 30-day mortality and `r icc.inhosp` for in-hospital mortality.
-
 # Table S2. Extended patient sample characteristics
@@
 ```
-
-```{r, child = "supplementary-outcome-tables.Rmd"}
-```

### 2025-07-22 supplementary-outcome-tables.Rmd
Move results to an additional online material file

@@
 
-## Table S3. Outcomes in all patients during the entire study period, by treatment arm (95% CI)
+## Additional Online Table 1. Outcomes in all patients during the entire study period, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
+
 ```
@@
 
-## Table S4. Outcomes in all patients before training, by treatment arm (95% CI)
+## Additional Online Table 2. Outcomes in all patients before training, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients before training, by treatment arm"]]
+
 ```
@@
 
-## Table S5. Outcomes in all patients after training, by treatment arm (95% CI)
+## Additional Online Table 3. Outcomes in all patients after training, by treatment arm (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Outcomes in all patients after training, by treatment arm"]]
+
 ```
@@
 
-## Table S6. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arm (95% CI)
+## Additional Online Table 4. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
 ```
@@
 
-## Table S7. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arm (95% CI)
+## Additional Online Table 5. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
 ```
@@
 
-## Table S8. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (95% CI)
+## Additional Online Table 6. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
+
 ```
@@
 
-## Table S9. Absolute and relative differences in outcomes after training, comparing standard care with PTC (95% CI)
+## Additional Online Table 7. Absolute and relative differences in outcomes after training, comparing standard care with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
+
 ```
@@
 
-## Table S10. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (95% CI)
+## Additional Online Table 8. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
+
 ```
@@
 
-## Table S11. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (95% CI)
+## Additional Online Table 9. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
+
 ```
@@
 
-## Table S12. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (95% CI)
+## Additional Online Table 10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
+
 ```
@@
 
-## Table S13. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (95% CI)
+## Additional Online Table 11. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (95% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
 
 ```{r, echo = FALSE}
+
 outcome.results.tables[["ci.level.0.95"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
+
 ```

### 2025-07-22 functions/save_tables_to_file.R

@@ -28,11 +28,11 @@ save_tables_to_file <- function(outcome.results.tables, supplementary.material =
                 paste0(
                     "\\pagebreak\n\n",
                     "# Results with ", ci.level, " confidence intervals (CI)\n\n",
-                    lapply(seq_along(ci.level.tables) + 1, function(ci.level.table.index) {
+                    lapply(seq_along(ci.level.tables), function(ci.level.table.index) {
                         ci.level.table.name <- names(ci.level.tables)[ci.level.table.index]
                         ci.level.table <- ci.level.tables[[ci.level.table.name]]
                         paste0(
-                            "## ", ifelse(supplementary.material, paste0("Table S", ci.level.table.index + 1, ". "), ""), ci.level.table.name, " (", ci.level, " CI)", "\n\n",
+                            "## ", ifelse(supplementary.material, paste0("Additional Online Table ", ci.level.table.index, ". "), ""), ci.level.table.name, " (", ci.level, " CI)", "\n\n",
                             "Empty cells indicate that there were no observations to calculate the statistic.", "\n\n",
                             "\`\`\`{r, echo = FALSE} \n\n",
                             "outcome.results.tables[[\"", ci.level.tables.name, "\"]]",

### 2025-07-22 manuscript.Rmd
Incorporate the additional online material in the manuscript

@@
 The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
@@
 There were a standard care arm and two intervention arms, ATLS^®^ and PTC training.
-We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
+We planned to collect data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
+The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters.
 We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
@@
 
-Data was collected over a four-month period.
+The planned data collection period was four months, but the actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters.
 A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care.
@@
 Qualitative variables are presented as absolute numbers and percentages.
-Additional analyses performed according to the original protocol are described in the Supplementary Material S2 and their results shown in Supplementary Material S3 and Supplementary Tables S1-12.
+Additional analyses performed according to the original protocol are available as additional online material [@warnberg_additional_2025].
 
@@
 We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`.
-The period of data collection and the number of patients included per month per cluster is shown in Table \@ref(tab:patients-per-month).
+The period of data collection and the number of patients included per month per cluster is shown in Figure \@ref(fig:patients-per-month-figure).
+Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively.
 The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients.
@@
 
-```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month."}
+```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month.
+Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively."}
 knitr::include_graphics("patients-per-month.png")
@@
 The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
+Extended patient sample characteristics are shown in Supplementary Table S2.
 Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
-The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.
+The variables with the maximum amount of missing data were in the cost of treatment, also reported in Supplementary Tables S2.
 
@@
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Some variables, especially the cost of treatment (reported in Supplementary Table S2) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
@@
 Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial.
-Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.
+Fifth, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.
+Finally, because of a data uploading error, there was only limited data available from two clusters.
+At the time of data collection, network issues and other technical issues were a challenge in some of the clusters, and this can be mitigated in the full scale trial by using a more robust data collection system with local offline backups.
 

### 2025-07-22 main.R

@@ -57,8 +57,7 @@ overall.sample.characteristics.table <- create_descriptive_table(table.data, sho
 
 ## Create table of sample characteristics stratified by arm
 stratified.table.data <- data %>%
-    select(all_of(table.variables)) %>%
-    select(-outcomes__discharge_alive, -outcomes__alive_after_30_days)
+    select(all_of(table.variables))
 sample.characteristics.table <- create_descriptive_table(stratified.table.data,
     strata = "arm", show.all.levels = FALSE, include.overall = TRUE
 ) %>%
@@ -139,10 +138,7 @@ if (use.saved) {
 outcome.results.tables <- create_outcome_results_tables(outcome.results)
 
 ## Save outcome results for supplementary material
-save_tables_to_file(outcome.results.tables, supplementary.material = TRUE, only.95.ci = TRUE)
-
-## Save all outcome results tables to file
-save_tables_to_file(outcome.results.tables, supplementary.material = FALSE, only.95.ci = FALSE)
+save_tables_to_file(outcome.results.tables, supplementary.material = TRUE, only.95.ci = FALSE)
 
 # Extract results from tables
 females <- inline_text(overall.sample.characteristics.table,

### 2025-07-23 point-by-point-response.qmd
Compile manuscript

@@
 Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial.
-We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and avoid any conclusions about a potential effect of the training.
+We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and to avoid any conclusions about a potential effect of the training.
 
@@
 We agree that the manuscript, and the supplementary material, includes more results than necessary to fulfil the aim of the study.
-As you have noticed, we struggled to find a balance between staying true to the protocol and reporting the results that were most relevant in a transparent manner.
+As you have noticed, we struggled to find a balance between adhering to the protocol and reporting the results that were most relevant in a transparent manner.
 We have now attempted to address this by clearly stating that we deviated from the original protocol in reporting the pilot, to focus on the feasibility of the full-scale trial.
@@
 We agree that the supplementary materials are very extensive.
-We included these to transparently report the results of the analyses specified in the original protocol, and we have therefore decided to keep them in the manuscript.
+We included these to transparently report the results of the analyses specified in the original protocol, but in response to your comment we have now published the majority of the supplementary materials as online additional files, and cite these in the manuscript.
 We have attempted to include parts of the supplementary materials in the main text where appropriate, in response to your earlier comments.
@@
 
-The data collection period in each hospital was four months, but patients were followed up for one month. We have clarified this in the methods section.
+The data collection period in each hospital was intended to be four months, and then patients were followed up for one month.
+However, the actual data collected periods varied slightly between clusters, and we have clarified this in the methods section.
 
@@
 
-We have added a table to the results section to report the number of patients included per month.
+We have added a figure to the result's section, reporting the number of patients included per month per cluster.
+Notably, the process of preparing this figure allowed us to identify a data collection error in two clusters, and we have added this to the methods section.
+We have also included this in the discussion as a limitation of this pilot study and an important lesson for the full scale trial.
 
@@
 
-We now only report the yes in the outcomes tables.
+We have removed the majority of the outcome tables from the supplementary materials and published them as online additional files, referenced in the manuscript.
 
@@
 
-We have removed the absolute difference versus relative difference from the main results section, but kept it in the supplementary materials as it was part of the original protocol.
+We have removed the absolute difference versus relative difference from the main results section, but report them in the additional online files as this analysis was part of the original protocol.
 
@@
 Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial.
+We now include this as a suggestion for the full scale trial in the discussion.
 
@@
 
-Yes, that is correct. Empty cells indicate that there were no observations. We have added a note to clarify this.
+Yes, that is correct.
+Empty cells indicate that there were no observations.
+We have now moved the majority of the tables from the supplementary materials and instead published them as online additional files, referenced in the manuscript.
+These files include a  note to clarify that empty cells indicate that there were no observations.

### 2025-07-23 manuscript.Rmd
Compile manuscript

@@
 ## Trial registration
-This pilot study was registered with ClinicalTrials.gov (reg.
-no NCT05417243).
-We intended to also register this pilot trial with Clinical Trials Registry - India, but this was not done because of a communication error in the management group.
+This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).
 
@@
 Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
-Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016].
+Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025].
 In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
@@
 This difference  may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial.
+
 A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.
+In addition, educating physicians in trauma life support through programmes like ATLS^®^ and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propogating outdated practices [@wiles_atls_2015].
+Several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], and in recognition of their wide use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best option.
 
-Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.

### 2025-07-23 supplementary-materials.Rmd
Compile manuscript

@@
 
-| Cluster                                                       | City       | Beds | Type    | Study period | 
-+---------------------------------------------------------------+------------+------+---------+--------------+
-| Dr R N Cooper Municipal General Hospital                      | Mumbai     | 600  | Public  |              |
-| Seth GS Medical College and KEM Hospital                      | Mumbai     | 1800 | Public  |              |
-| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     | 1850 | Public  |              |
-| MEDICA Superspeciality Hospital                               | Kolkata    | 300  | Private |              |
-| Medical College                                               | Kolkata    | ??   | Public  |              |
-| Nilratan Sircar Medical College & Hospital                    | Kolkata    | 1910 | Public  |              |
-| Postgraduate Institute of Medical Education and Research      | Chandigarh | 1400 | Public  |              |
-+---------------------------------------------------------------+------------+------+---------+--------------+
+| Cluster                                                       | City       | Beds | Type    |
++---------------------------------------------------------------+------------+------+---------+
+| Dr R N Cooper Municipal General Hospital                      | Mumbai     | 600  | Public  |
+| Seth GS Medical College and KEM Hospital                      | Mumbai     | 1800 | Public  |
+| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     | 1850 | Public  |
+| MEDICA Superspeciality Hospital                               | Kolkata    | 300  | Private |
+| Medical College                                               | Kolkata    | ??   | Public  |
+| Nilratan Sircar Medical College & Hospital                    | Kolkata    | 1910 | Public  |
+| Postgraduate Institute of Medical Education and Research      | Chandigarh | 1400 | Public  |
++---------------------------------------------------------------+------------+------+---------+
 

### 2025-07-23 supplementary-outcome-tables.Rmd
Compile manuscript

@@
+\pagebreak
+
+# Results with 75% confidence intervals (CI)
+
+## Additional Online Table 1. Outcomes in all patients during the entire study period, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 2. Outcomes in all patients before training, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients before training, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 3. Outcomes in all patients after training, by treatment arm (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients after training, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 4. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 5. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 6. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 7. Absolute and relative differences in outcomes after training, comparing standard care with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 8. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 9. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 11. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (75% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
+
+```
+
+\pagebreak
+
+\pagebreak
+
+# Results with 85% confidence intervals (CI)
+
+## Additional Online Table 1. Outcomes in all patients during the entire study period, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients during the entire study period, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 2. Outcomes in all patients before training, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients before training, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 3. Outcomes in all patients after training, by treatment arm (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients after training, by treatment arm"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 4. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 5. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 6. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 7. Absolute and relative differences in outcomes after training, comparing standard care with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 8. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 9. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]
+
+```
+
+\pagebreak
+
+## Additional Online Table 11. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (85% CI)
+
+Empty cells indicate that there were no observations to calculate the statistic.
+
+```{r, echo = FALSE}
+
+outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
+
+```
+
+\pagebreak
+
 \pagebreak

### 2025-07-24 manuscript.Rmd
Fixes and compile supplementary material
Update manuscript

@@
 martin.gerdin@ki.se  
-+46 708 53 95 98 (mobile)figure
++46 708 53 95 98 (mobile)
 
@@
 
-**Main Outcomes and Measures** The outcomes were consent rate, lost to follow up rate, pass rate, missing data rates, differences in distribution between observed and data extracted from medical records as well as all cause and in-hospital mortality at 30 days from the time of arrival to the emergency department.
+**Main Outcomes and Measures** The outcomes were consent rate, lost to follow up rate, missing data rates, differences in distribution between observed and data extracted from medical records, and resident pass rate.
 
@@
 Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
-`r number_to_text(m30d.standard.care.n, first = TRUE)` `r m30d.standard.care.p` patients died within 30 days in the standard care arm, `r number_to_text(m30d.atls.n)` `r m30d.atls.p` patient in the ATLS^®^ arm, and `r number_to_text(m30d.ptc.n)` `r m30d.ptc.p` patients in the PTC arm.
 
@@
 
-### Hospitals
-
-We included tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
-
 ### Clusters
 
-We defined a cluster as one or more units of physicians providing trauma care in the emergency department of Indian tertiary care hospitals.
-To be eligible, units could have no more than 25% of their physicians trained in either ATLS^®^, PTC, or similar training programs before the start of the pilot study.
+We defined a cluster as a tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
+In each cluster, we trained one or more units of physicians providing trauma care in the emergency department.
+To be eligible, units could have no more than 25% of their physicians trained in either ATLS^®^, PTC, or similar training programmes before the start of the pilot study.
 Those residents who had received training in the last five years were considered as trained.
@@
 These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
-In the emergency department, trauma patients are initially assessed by either surgical or emergency medicine residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+In the emergency department, trauma patients are initially assessed by residents in these units, and they also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
 Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
@@
 The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
-The ATLS^®^ programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK Charity PTC Foundation and is free of charge.
+The ATLS^®^ programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK based charity the PTC Foundation and is free of charge.
 
@@
 The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
-The variables with the maximum amount of missing data were in the cost of treatment, also reported in Supplementary Tables S2.
+The variables with the maximum amount of missing data were  the cost of treatment, complications and cause of death, also reported in Supplementary Tables S2.
 
@@
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables, especially the cost of treatment (reported in Supplementary Table S2) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
+Some variables had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
 The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
@@
 
-During the course of this pilot study we deviated from the protocol in several ways. We have reported these deviations where appropriate in the methods section, and provide a list as Supplementary
-material S1.
+During the course of this pilot study we deviated from the protocol in several ways.
 The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes.

### 2025-08-04 manuscript.Rmd
Remove trademark symbols according to BMJ Open author guidelines

@@
 
-## Non-author contributors
+## Nonauthor contributors
 
@@
 
-**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.
 
@@
 
-**Interventions** ATLS^®^ or PTC training for residents in the intervention arms.
+**Interventions** ATLS or PTC training for residents in the intervention arms.
 
@@
 
-**Results** Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care.
+**Results** Two hospitals were randomised to ATLS, two to PTC, and three to standard care.
 We included `r n.patients` patients and `r n.residents` residents.
@@
 
-**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
+**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
 
@@
 
-- Cluster randomised controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
+- Cluster randomised controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
 - Prospective data collection with direct observations by dedicated research officers.
@@
 
-The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+The proprietary Advanced Trauma Life Support (ATLS) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
 Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
-Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved knowledge and skills among providers [@Mohammad2013].
+Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013].
 Observational evidence suggests that PTC also leads to improved provider skills [@Kadhum2020].
@@
 
-We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.
+We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS and PTC with standard care on outcomes in adult trauma patients.
 
@@
 We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022].
-There were a standard care arm and two intervention arms, ATLS^®^ and PTC training.
+There were a standard care arm and two intervention arms, ATLS and PTC training.
 We planned to collect data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
@@
 
-We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught.
+We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS, PTC, nor any other established trauma life support training program is routinely taught.
 Details about each cluster is included as Supplementary Table S1.
@@
 In each cluster, we trained one or more units of physicians providing trauma care in the emergency department.
-To be eligible, units could have no more than 25% of their physicians trained in either ATLS^®^, PTC, or similar training programmes before the start of the pilot study.
+To be eligible, units could have no more than 25% of their physicians trained in either ATLS, PTC, or similar training programmes before the start of the pilot study.
 Those residents who had received training in the last five years were considered as trained.
@@
 We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
-Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.
+Consent was sought from the residents in each of the intervention groups before they underwent the ATLS or PTC training.
 In the published protocol, we stated that only surgical residents would be trained.
@@
 
-In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC at the beginning of the three month intervention phase.
+In each intervention arm the residents in one or two units were trained in either ATLS or PTC at the beginning of the three month intervention phase.
 For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
 We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
-The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi.
+The ATLS training was conducted in an ATLS certified training centre in Mumbai and the PTC training was conducted in New Delhi.
 Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.
@@
 The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
-The ATLS^®^ programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK based charity the PTC Foundation and is free of charge.
+The ATLS programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK based charity the PTC Foundation and is free of charge.
 
@@
 Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively.
-The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients.
-We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.
+The standard care arm included `r n.control` patients, the ATLS arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients.
+We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS, and  `r number_to_text(n.ptc.residents)` in PTC.
 
@@
 
-After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS^®^ and PTC arms respectively.
+After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
+The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively.
 
@@
 
-We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS^®^ with PTC and standard care.
+We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS with PTC and standard care.
 Missing data were low for key variables, including the primary outcome and many secondary outcomes.
@@
 
-Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
+Previous studies on the effect of ATLS or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
 Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025].
@@
 A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.
-In addition, educating physicians in trauma life support through programmes like ATLS^®^ and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propogating outdated practices [@wiles_atls_2015].
+In addition, educating physicians in trauma life support through programmes like ATLS and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propogating outdated practices [@wiles_atls_2015].
 Several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], and in recognition of their wide use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best option.
@@
 
-Several authors are ATLS^®^ and/or PTC instructors.
+Several authors are ATLS and/or PTC instructors.
 

### 2025-08-04 manuscript.Rmd
Edit manuscript according to AJE proofreading

@@
 
-**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.
+**Objective** The aim was to assess the feasibility of conducting a cluster randomised controlled trial comparing the effects of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.
 
-**Design** A pilot pragmatic three-armed parallel, cluster randomised, controlled trial between `r start.date` and `r end.date`. Patient follow up was 30 days.
+**Design** This was a pilot pragmatic three-armed parallel, cluster randomised, controlled trial conducted between `r start.date` and `r end.date`. Patient were followed up for 30 days.
 
-**Setting** Tertiary care hospitals across metropolitan areas in India.
+**Setting** The study setting was tertiary care hospitals across metropolitan areas in India.
 
-**Participants** Adult trauma patients and residents managing these patients.
+**Participants** Adult trauma patients and residents managing these patients were included.
 
-**Interventions** ATLS or PTC training for residents in the intervention arms.
+**Interventions** ATLS or PTC training was provided for residents in the intervention arms.
 
-**Main Outcomes and Measures** The outcomes were consent rate, lost to follow up rate, missing data rates, differences in distribution between observed and data extracted from medical records, and resident pass rate.
+**Main Outcomes and Measures** The outcomes were the consent rate, lessons to follow up rate, missing data rates, differences in the distribution between observed data and data extracted from medical records, and the resident pass rate.
 
-**Results** Two hospitals were randomised to ATLS, two to PTC, and three to standard care.
+**Results** Two hospitals were randomised to the ATLS arm, two to the PTC arm, and three to the standard care arm.
 We included `r n.patients` patients and `r n.residents` residents.
-The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%.
-The lost to follow up rate was `r rate.lost.to.follow.up`%.
+The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to receive training was `r recruitment.rate.residents`%.
+The loss to follow up rate was `r rate.lost.to.follow.up`%.
 The pass rate was `r pass.rate`%.
-The missing data was overall low for key variables.
-Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.
+Overall, the amount of missing data for key variables was low.
+The data collected through observations were similar to data extracted from medical records, but there were more missing extracted data.
 
-**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.
+**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes appears feasible, especially if such a trial would use data and outcomes available in medical records.
 
@@
 
-- Prospective data collection with direct observations by dedicated research officers.
-- Lack of a priori defined success criteria and thresholds for the feasibility outcomes.
-- Use of sealed envelopes potentially compromised allocation concealment.
-- Participating centers' heterogeneity may affect the study estimates and bias the results.
+- Prospective data collection with direct observations made by dedicated research officers.
+- A lack of a priori defined success criteria and thresholds for feasibility outcomes.
+- The use of sealed envelopes potentially compromised allocation concealment.
+- Heterogeneity of the participating centers may affect the study estimates and introduce bias.
 
@@
 
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year [@injuries2020].
-Several trauma life support training programs have been developed to improve the early management of patients in hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@injuries2020].
+Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-The proprietary Advanced Trauma Life Support (ATLS) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
+The proprietary Advanced Trauma Life Support (ATLS) and low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
 Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
-Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013].
-Observational evidence suggests that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is then whether these improved knowledge and skills translate into measurably improved patient outcomes.
+Several studies, including at least two randomised studies [@Ali1996; @Ali1999], have shown that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013].
+Observational evidence also suggests that PTC also leads to improved provider skills [@Kadhum2020].
+The missing link is whether this improved knowledge and skill set translates into measurably improved patient outcomes.
 
-Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible.
-Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.
+Systematic reviews have called for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], as conducting such effectiveness trials in settings where they are part of the standard of care is not feasible.
+Many of these settings are in low- and middle income countries, where trial logistics can be more challenging.
 
-We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS and PTC with standard care on outcomes in adult trauma patients.
+We therefore conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.
 
@@
 We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022].
-There were a standard care arm and two intervention arms, ATLS and PTC training.
-We planned to collect data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm).
-The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters.
-We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
-In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].
+The trial included a standard care arm and two intervention arms, ATLS and PTC training.
+We planned to collect data for four months in all three arms, first during a one month observation phase, followed by a three-month intervention phase (or continued observation in the standard care arm).
+The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection in the intervention clusters post-training.
+We included a one month observation phase to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in changes from baseline.
+In the published protocol, we also aimed to estimate probable effect sizes and other parameters needed for sample size calculations for a full-scale trial [@GerdinWärnberg2022].
+However, we revised this aim in light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].
 
@@
 We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS, PTC, nor any other established trauma life support training program is routinely taught.
-Details about each cluster is included as Supplementary Table S1.
-Originally, we intented to include six hospitals as clusters, but decided to include a seventh when they expressed interest and as we had the budget to accommodate this.
-These seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.
+Details about each cluster are provided in Supplementary Table S1.
+We initially intented to include six hospitals as clusters, but added a seventh hospital when management expressed interest, as we had the budget to accommodate the request.
+These seven hospitals represented a convenience sample that fulfilled the inclusion criteria, and had existing connections to the research team.
 
@@
 
-We defined a cluster as a tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
+We defined a cluster as a tertiary care hospital in metropolitan areas in India that admit more than 400 adult patients with trauma annually, and has operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock.
 In each cluster, we trained one or more units of physicians providing trauma care in the emergency department.
-To be eligible, units could have no more than 25% of their physicians trained in either ATLS, PTC, or similar training programmes before the start of the pilot study.
-Those residents who had received training in the last five years were considered as trained.
-The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results.
+To be eligible, units could have no more than 25% of their physicians with previous training in either ATLS, PTC, or similar training programmes.
+Residents who had received training in the last five years were considered trained.
+The 25% threshold was determined through consensus within the research team, to balance feasibility and the risk of contamination.
 The principal investigator at each hospital selected the units for training.
-We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.
+We randomised at the hospital level to avoid contamination between the intervention and standard care arms.
 
@@
 
-We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training.
-Consent was sought from the residents in each of the intervention groups before they underwent the ATLS or PTC training.
+We trained resident doctors undertaking speciality training in surgery or emergency medicine who managed trauma patients in the emergency department, and were expected to remain in the participating hospitals for at least one year after training.
+Consent was obtained from the residents in each intervention arm before ATLS or PTC training.
 In the published protocol, we stated that only surgical residents would be trained.
-However, in some of the participating hospitals, emergency medicine residents performed the initial resuscitation and management of trauma patients, and we therefore decided to train them instead.
+However, in some of the participating hospitals, emergency medicine residents led the initial resuscitation and management of trauma patients, and we therefore included them in the training.
 
@@
 
-We included persons who were 15 years or older and presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
-History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter 20 of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.
+We included persons aged 15 years or older who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty.
+A history of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter 20 of the International Classification of Disease version 10 (ICD-10) codebook as the reason for presentation.
 
@@
 
-Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units.
-These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department.
-In the emergency department, trauma patients are initially assessed by residents in these units, and they also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
-Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.
+Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians into units.
+These units include both faculty members and residents, who are assigned a specific day of the week to work in the emergency department.
+Trauma patients are initially assessed by residents in these units, who also resuscitate patients, perform interventions and refer patients for imaging or other investigations.
+Compared with settings that adopt a trauma team approach, nurses and other healthcare professionals are involved to a limited extent during initial management.
 We did not collect data on how standard care varied between the participating hospitals.
@@
 
-In each intervention arm the residents in one or two units were trained in either ATLS or PTC at the beginning of the three month intervention phase.
-For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit.
-We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients.
-The ATLS training was conducted in an ATLS certified training centre in Mumbai and the PTC training was conducted in New Delhi.
+In each intervention arm, residents from one or two units were trained in either ATLS or PTC at the beginning of the three month intervention phase.
+For the purpose of this pilot study, our target was to train at least 75% of residents in each unit.
+Faculty members were not trained, because they are typically not directly involved in the initial management of trauma patients.
+ATLS training was conducted at an ATLS certified training centre in Mumbai and PTC training was conducted in New Delhi.
 Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.
 
-The provider courses of both programmes are two days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations.
-The teaching is based on case discussions and skill stations.
-There are some important differences between the two programmes.
-The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018].
-The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].
-The ATLS programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK based charity the PTC Foundation and is free of charge.
+The provider courses of both programmes take place across two days, and focus on the assessment, resuscitation and stabilisation of trauma patients, with adaptions for different patient populations.
+Teaching is based on case discussions and skill stations.
+There are several important differences between the two programmes.
+The ATLS course focuses more on inter-hospital patient transfers and includes a greater emphasis on the trauma team [@acsAtls2018].
+In contrast, the PTC course focuses on trauma care in the low resource setting [@ptcfPtc2018].
+The ATLS programme is run by the American College of Surgeons and requires a participant fee, whereas the PTC programme is run by the UK-based PTC Foundation and is provided free of charge.
 
-## Feasibility outcomes
+## Feasibility Outcomes
 
-Our feasibility outcomes were:
+Our feasibility outcomes were as follows:
 
-- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
-- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
-- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
+- Consent rates of patients and residents. This was defined as the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
+- Loss to follow-up rate. This applied only to patients and was defined as the percentage of patients among all included patients who did not complete the 30 day follow up.
+- Missing data rate. This applied to each outcome and variable and was defined as the percentage of missing values.
 - Differences in distributions between directly observed data and data extracted from medical records.
 Distribution refers to summary statistics and directly observed data refers to data collected by project officers while observing the delivery of care.
-This outcome applied to all variables that could be reasonably expected to be present in the medical records, but we decided to only extract this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
-- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
+This outcome applied to all variables that could be reasonably expected to be present in the medical records.
+To reduce workload, these data were extracted from a convenience sample of patients only.
+- Pass rate. This applied only to residents in the intervention arms and was defined as the percentage of residents who passed the training programme, among all residents who received training.
 
-We did not prespecify criteria to judge whether to continue with the full scale trial.
+We did not prespecify criteria to determine whether to proceed with a full-scale trial.
 
@@
 
-## Sample size
+## Sample Size
 
-We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training.
-We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods.
-We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.
+We aimed to include at least two clusters per arm to avoid basing conclusions on single centres.
+We also aimed to train at least two units per intervention cluster to evaluate the logistics of sending residents for training.
+We did not conduct a formal power calculation for this pilot, as the primary purpose of the study was to assess the feasibility of the trial logistics and research methods.
+We anticipated variation in the number of patients included per cluster depending on hospital patient volume.
 
@@
 
-Arriving patients were screened for eligibility and consented, if conscious.
-Unconscious patients were consented by the patient’s representative.
-This proxy consent was reaffirmed by the patient, on regaining consciousness.
-Patients who did not regain consciousness were included based on their representative's consent.
-We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department.
-The follow up period for each patient was therefore one month.
+Incoming patients were screened for eligibility and consented, if they were conscious and able to provide consent.
+For unconscious patients, consent was provided by a patient representative.
+These patients reaffirmed this proxy consent, upon regaining consciousness.
+Patients who did not regain consciousness remained included based on their representative's consent.
+We followed up patients at 24 hours and at 30 days after arrival at the emergency department.
+The follow-up period for each patient was therefore one month.
 
@@
 
-Participating units were screened for eligibility once hospitals confirmed their participation.
-All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms.
-The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training.
-The training was conducted approximately one month after the study started in that hospital.
-Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator.
-The number of residents trained in each intervention cluster varied because the size of the units varied.
+Participating units were screened for eligibility once the hospitals confirmed participation.
+All residents in these units were approached for consent to training if their hospital was randomised to one of the intervention arms.
+The protocol stated that residents would be approached for consent before randomisation, but this proved not to be feasible.
+Instead, we asked residents for consent after the hospitals were randomised but before training.
+Training took place approximately one month after study initiation in that hospital.
+We initially planned to use simple random sampling to select the units to be trained, but for pragmatic reasons this decision was left to the site principal investigator.
+The number of residents trained in each intervention cluster varied based on the unit size.
 
-## Allocation and blinding
+## Allocation and Blinding
 
-We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms.
+We used simple randomisation implemented using sealed envelopes to allocate sites to the trial arms.
 It was not possible to blind investigators, residents or patients to the intervention.
-We did not blind the data analysts.
+Data analysts were not blinded.
 
@@
 
-The planned data collection period was four months, but the actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters.
-A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care.
-The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records.
-We followed up admitted patients for their complications and other in-hospital outcome measures.
-Patients who were not admitted or who were discharged before the end of the study were followed up telephonically for mortality outcomes and quality of life outcomes.
+The planned data collection period was four months.
+However, the actual period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters.
+Research officers collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care.
+The research officers observed care, interviewed residents and patients, and extracted data from hospital records.
+Admitted patients were followed up for complications and other in-hospital outcome measures.
+Patients who were not admitted or who were discharged before the end of the study were followed up by telephone for mortality and quality-of-life.
 
@@
 
-The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained.
+Research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained.
 All injuries were coded according to the International Classification of Diseases version 10 (ICD-10).
-Based on these ICD-10 codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022].
-The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cutoff of 16 often being used to define major trauma, and 75 representing unsurvivable trauma.
-We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L).
-We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022].
+Based on these codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022].
+The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cut-off score of 16 often used to define major trauma and 75 representing unsurvivable trauma.
+We also collected data on potential outcomes for the full-scale trial, including 30-day and in-hospital mortality, complications and health related quality of life (assessed using the EQ-5D-3L).
+We did not calculate an EQ-5D-3L index score, because no Indian value set is currently available [@jyani_development_2022].
 We also attempted to collect data on cause of death.
@@
 
-## Patient and public involvement
+## Patient and Public Involvement
 
-We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial.
-The results of these consultations are published separately [@david_patient-reported_2024].
-We initially planned to distribute periodic surveys to residents and follow them up after 30 days after training, but we changed this to interviews after the end of the study period as we thought this would provide richer data (not published).
+We conducted community consultations with patients, their caregivers, patient groups, and resident doctors to inform the selection of outcome measures and implementation of the full-scale trial.
+Results of these consultations are published separately [@david_patient-reported_2024].
+We initially planned to distribute periodic surveys to residents and follow them up 30 days after training, but this was later changed to end-of-study interviews to allow for richer data (not published).
 
@@
 
-## Data monitoring
+## Data Monitoring
 
-We conducted weekly online meetings to monitor the study and data collection.
-We conducted one interim analysis approximately halfway through the study, and decided to complete the study as residents and patients were consenting to be included in the study and key variables including mortality outcomes could be collected.
-We did not use a data monitoring committee.
+Weekly online meetings were held to monitor study progress and data collection.
+One interim analysis was conducted approximately halfway through the study, and it was decided to complete the study, as residents and patients were consenting to be included in the study and key variables including mortality, could be collected.
+A formal data monitoring committee was not used.
 
@@
 
-We used the `r version$version.string` Statistical Software for all analyses [@R_2024].
-We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014].
-Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results.
-Quantitative variables are summarised as median and interquartile range.
+All analyses were conducted using `r version$version.string` statistical software [@R_2024].
+Feasibility outcomes and other data were analysed using descriptive statistics and no formal hypothesis testing was performed.
+Initially, we planned to analyse feasibility outcomes at both the overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results.
+Quantitative variables are summarised as medians and interquartile ranges.
 Qualitative variables are presented as absolute numbers and percentages.
@@
 We were granted research ethics approval from the institutional ethics committees at each participating hospital.
-For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for the Postgraduate Institute of Medical Education and Research, Chandigarh.
+For each hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for the Postgraduate Institute of Medical Education and Research, Chandigarh.
 
@@
 We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`.
-The period of data collection and the number of patients included per month per cluster is shown in Figure \@ref(fig:patients-per-month-figure).
-Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively.
+The  data collection period and the number of patients included per month per cluster are shown in Figure \@ref(fig:patients-per-month-figure).
+Owing to an error in the data uploading process, data were available only for one and three months in two clusters respectively.
 The standard care arm included `r n.control` patients, the ATLS arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients.
-We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS, and  `r number_to_text(n.ptc.residents)` in PTC.
+A total of  `r number_to_text(n.residents)` residents were trained, `r number_to_text(n.atls.residents)` in ATLS and  `r number_to_text(n.ptc.residents)` in PTC.
 
 ```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month.
-Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively."}
+Owing to an error in the data uploading process, data were available for only one and three months in two clusters respectively."}
 knitr::include_graphics("patients-per-month.png")
@@
 
-The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
+The study flow diagram is shown in Figure \@ref(fig:flow-diagram), and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics).
 Extended patient sample characteristics are shown in Supplementary Table S2.
-Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
+Overall, `r females` patients were females, the median (interquartile range, IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`.
 These prognostic factors differed between the trial arms.
-A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.
+A total of `r m30d` patients died within 30 days of arrival at the emergency department, and `r inhosp` patients died in the hospital.
 
-After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
-The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively.
+After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared with `r m30d.atls` patient in the ATLS arm and `r m30d.ptc` patients in the PTC arm.
+The corresponding rates for in-hospital mortality were `r inhosp.standard.care`, `r inhosp.atls`, and `r inhosp.ptc` for the standard care, ATLS and PTC arms, respectively.
 
@@
 
-The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%.
-The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
-The variables with the maximum amount of missing data were  the cost of treatment, complications and cause of death, also reported in Supplementary Tables S2.
+The percentage of patients who consented to follow-up was `r p.consent`% and the percentage lost to follow-up was `r rate.lost.to.follow.up`%.
+The missing data rate ranged from `r min.missing.data`% to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2.
+The variables with the largest amount of missing data were  the cost of treatment, complications and cause of death, also reported in Supplementary Table S2.
 
-The differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted).
-Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.
+Differences in distributions between directly observed data and data extracted from medical records for selected variables collected through observation or interviews are shown in Table \@ref(tab:observed-vs-extracted).
+Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records than in the directly observed data.
 
-The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was `r pass.rate`%.
+The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was also `r pass.rate`%.
 
-```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview in a convenience sample of patients."}
+```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables collected through observation or interview in a convenience sample of patients."}
 observed.vs.extracted.comparison.table
@@
 
-We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS with PTC and standard care.
-Missing data were low for key variables, including the primary outcome and many secondary outcomes.
-Some variables had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods.
-The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.
-To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.
+We demonstrated that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS with PTC and standard care.
+The missing data rate was low for key variables.
+However, some variables had very high missing data rates and may not be feasible to include in a full-scale trial, or may require different data collection methods.
+The missing data rate was substantially higher when data were extracted from medical records rather than directly observed, although the distributions were similar.
+Thus suggests that data collected from medical records are reliable even if they are less complete.
+To increase the completeness of data extracted from the medical records, a full-scale trial should limit the number of variables extracted from medical records and emphasise the importance of having these variables recorded to the participating hospitals.
 
@@
 
-All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients.
-This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016].
-Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge.
-Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials.
-Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome.
+All-cause 30-day mortality data were missing for `r rate.lost.to.follow.up`% of patients.
+This rate may be high, especially compared to, for example, the CRASH-2 and REACT-2 trials, which reported missing primary outcomes for fewer than 0.01% of patients [@Roberts2013; @Sierink2016].
+Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow patients after discharge.
+Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to those in previous trials.
+Following patients after discharge is notoriously challenging in this setting, and the full-scale trial may need to focus on in-hospital mortality as the primary outcome.
 
@@
 
-During the course of this pilot study we deviated from the protocol in several ways.
-The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes.
-However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [@eldridge_how_2015; @teresi_guidelines_2022].
-We therefore decided to only report patient outcomes descriptively.
-Another important deviation from the protocol was the training of emergency medicine residents.
-We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines.
-The full scale study will need to accommodate this variation as part of the protocol.
+During this pilot study we deviated from the protocol in several ways.
+The most significant deviation was a revision of the study aim, as we initially intended to estimate potential effect sizes and other parameters to help sample size calculations for a full-scale trial, in addition to assessing the feasibility outcomes.
+However, current guidance advises against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [@eldridge_how_2015; @teresi_guidelines_2022].
+We therefore chose to report patient outcomes descriptively.
+Another significant deviation was the training of emergency medicine residents.
+We originally planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we adapted to local routines.
+A full-scale study will need to accommodate this variation as part of the protocol.
 
-There are some important limitations of this pilot study and therefore additional lessons to be learned and factored into the design of the full-scale trial.
-First, the volumes of patients included at some of the participating hospitals were lower than expected.
-A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial.
-Second, data on complications and cause of death were almost universally missing.
-To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available.
-Third, we did not collect in-depth data on the standard care at each hospital.
-This data should be collected as part of the screening process for the full-scale trial.
+There are several significant limitations of this pilot study and, therefore, additional lessons to be learned and factored into the design of a full-scale trial.
+First, the patient volumes at some of the participating hospitals were lower than expected.
+A careful assessment of patient volumes as part of the screening process should be included for a full-scale trial.
+Second, data on complications and causes of death were almost universally missing.
+Collecting data on these variables will require alternative methods, as these were not explicitly described in the medical records and autopsy reports were not readily available.
+Third, we did not collect detailed data on the standard care at each hospital.
+These data should be collected as part of the screening process for a full-scale trial.
 Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors.
-The full scale trial should use a computer-generated randomisation system.
-Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial.
-Fifth, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.
-Finally, because of a data uploading error, there was only limited data available from two clusters.
-At the time of data collection, network issues and other technical issues were a challenge in some of the clusters, and this can be mitigated in the full scale trial by using a more robust data collection system with local offline backups.
+A full-scale trial should use a computer-generated randomisation system.
+Fifth, we did not blind the data analysts, but recommend doing so in a full-scale trial.
+Sixth, we assessed a large number of potential outcomes, and a full-scale trial should focus on the most relevant outcomes.
+Finally, owing to a data uploading error, limited data were available from two clusters.
+At the time of data collection, network and other technical issues were present in some of the clusters, and should be mitigated in a full-scale trial by using a more robust data collection system with local offline backups.
 
@@
 
-Previous studies on the effect of ATLS or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
-Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025].
-In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
-We found fewer deaths in the intervention arms compared to the standard arm.
-This difference may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial.
+Previous studies on the effects of ATLS or PTC training on patient outcomes have been observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
+Most suggest that these programmes are associated with improved outcomes, although not all report significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025].
+In contrast, some studies have shown potential associations with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
+We observed fewer deaths in the intervention arms than in the standard care arm.
+This difference may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing a full-scale trial.
 
-A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.
-In addition, educating physicians in trauma life support through programmes like ATLS and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propogating outdated practices [@wiles_atls_2015].
-Several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], and in recognition of their wide use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best option.
+A full-scale trial remains ethically justifiable after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.
+In addition, educating physicians in trauma life support through programmes such as ATLS and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propagating outdated practices [@wiles_atls_2015].
+Several systematic reviews have called for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+In recognition of their widespread use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best trial design.
 
-Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.
+Our study represents the first published attempt to pilot a controlled trial evaluating the effect of trauma life support training on patient outcomes.
+We conclude that a full-scale cluster randomised trial is feasible after incorporating the lessons of this pilot study.
 
@@
 
-MGW conceived of the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
+MGW conceived the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
 study. MGV, DKV, KDS, and MK drafted the first version of the
@@
 DB, JB, SC, LFT, GG, MK, TK, CJ, NR, RS, KDS, LS and VP interpreted the results and revised the manuscript.
-AR, AC, C, DK, GG, MK, MT, VK and VP are representatives of participating hospitals.
+AR, AC, C, DK, GG, MK, MT, VK and VP are representatives of the participating hospitals.
 
@@
 
-The code for analysis is released publicly. The final anonymized dataset is available from the corresponding author on request.
+The code for the analysis is released publicly. The final anonymised dataset is available from the corresponding author upon request.
 

### 2025-08-04 supplementary-materials.Rmd
Update additional documents

@@
       toc: true
+      toc_depth: 1
 ---
@@
 | Cluster | City | Beds | Type |
-+---------------------------------------------------------------+------------+------+---------+
+|---------|------|------|------|
 | Dr R N Cooper Municipal General Hospital | Mumbai | 600 | Public |
@@
 | MEDICA Superspeciality Hospital | Kolkata | 300 | Private |
-| Medical College                                               | Kolkata    | ??   | Public  |
+| Medical College | Kolkata | 2140 | Public |
 | Nilratan Sircar Medical College & Hospital | Kolkata | 1910 | Public |
 | Postgraduate Institute of Medical Education and Research | Chandigarh | 1400 | Public |
-+---------------------------------------------------------------+------------+------+---------+
 

### 2025-08-06 point-by-point-response.Rmd
Compile point by point response
Update additional documents
Update response letter

--- 2025-07-23 point-by-point-response.qmd
@@
+---
+output: word_document
+---
+
 Dear Editor and Reviewers,
@@
 
-## Response to Editor's Comments:
+### Response to Editor's Comments:
 
-> 1. Please revise the title of your manuscript to include the setting. This is the preferred format of the journal.
+QE.1 Please revise the title of your manuscript to include the setting. This is the preferred format of the journal.
 
-We have revised the title to include the setting. The new title is "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India".
+> RE.1 We have revised the title to include the setting. The new title is "Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India".
 
-> 2.
-Please revise the ‘Strengths and limitations of this study’ section of your manuscript (after the abstract).
+QE.2 Please revise the ‘Strengths and limitations of this study’ section of your manuscript (after the abstract).
 This section should contain up to five short bullet points, no longer than one sentence each, that relate specifically to the methods.
@@
 
-The strengths and limitations section now focuses on the methods, and have been revised in the light of the comments from the reviewers.
+> RE.2 The strengths and limitations section now focuses on the methods, and have been revised in the light of the comments from the reviewers.
 
->3. Please ensure that you have fully discussed the methodological limitations of the study in the Discussion section of the main text.
+QE.3 Please ensure that you have fully discussed the methodological limitations of the study in the Discussion section of the main text.
 
-We have revised the discussion to include the methodological limitations highlighted by the reviewers.
+> RE.3 We have revised the discussion to include the methodological limitations highlighted by the reviewers.
 
-## Response to Reviewer 1, Dr.  Feroze  Sidhwa, San Joaquin General Hospital:
+### Response to Reviewer 1, Dr.  Feroze  Sidhwa, San Joaquin General Hospital:
 
-> “We found that the ATLS ® and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
+Q1.1 "We found that the ATLS ® and PTC arms had lower 30-day mortality compared to the PTC and standard care arms.
 This finding could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes.
 The arms differed considerably in sample size, with the ATLS ® arm having the smallest sample size.
-This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.”
+This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial."
 
-> If this is the case then these outcomes simply should not be reported.
+If this is the case then these outcomes simply should not be reported.
 Since this is a feasibility trial the major concern should be assessing missing data, which it seems to be, but this statement will be controversial and may distract from the otherwise important and better-supported conclusions.
@@
 
-> The above is the only concern I had when I chose "no" for "Are the discussion and conclusions justified by the results?"
+The above is the only concern I had when I chose "no" for "Are the discussion and conclusions justified by the results?"
 
-Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial.
+> R1.1 Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial.
 We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and to avoid any conclusions about a potential effect of the training.
 
-## Response to Reviewer 2, Dr. Wietske H.W. Ham, University of Applied Sciences:
+### Response to Reviewer 2, Dr. Wietske H.W. Ham, University of Applied Sciences:
 
-> Thank you for submitting this relevant paper and the impressive work behind it.
+Q2.1 Thank you for submitting this relevant paper and the impressive work behind it.
 It was an honor to review the paper.
@@
 
-Thank you for your comments, they were very helpful. We have revised the manuscript in the light of your comments.
+> R2.1 Thank you for your comments, they were very helpful. We have revised the manuscript in the light of your comments.
 
-> The aim of the study is to gain insight into the feasibility of a larger study. At the same time, more results are presented than necessary to answer this question. This is not clear to the reader.
+Q2.2 The aim of the study is to gain insight into the feasibility of a larger study.
+At the same time, more results are presented than necessary to answer this question.
+This is not clear to the reader.
 
-We agree that the manuscript, and the supplementary material, includes more results than necessary to fulfil the aim of the study.
+> R2.2 We agree that the manuscript, and the supplementary material, includes more results than necessary to fulfil the aim of the study.
 As you have noticed, we struggled to find a balance between adhering to the protocol and reporting the results that were most relevant in a transparent manner.
@@
 
-> It is unclear to me whether patients who ultimately did not regain consciousness were also included in this study. This should be described more clearly.
+Q2.3 It is unclear to me whether patients who ultimately did not regain consciousness were also included in this study. This should be described more clearly.
 
-We have clarified that patients who did not regain consciousness were included based on the consent by their representative.
+> R2.3 We have clarified that patients who did not regain consciousness were included based on the consent by their representative.
 
-> It also seems that, given the research period, the study sites were included sequentially; if so, the order and period should be described.
+Q2.4 It also seems that, given the research period, the study sites were included sequentially; if so, the order and period should be described.
 
-That is correct, data collection started in each cluster as they received the necessary approvals. The information about the study period in each cluster has been added as supplementary material.
+> R2.4 That is correct, data collection started in each cluster as they received the necessary approvals.
+The information about the study period in each cluster has been added as supplementary material.
 
-> It is not sufficiently clear to me how the statistical methods used contribute to answering the research question, namely the feasibility of an RCT and determining an effect size (and other measures, which are too vague).
+Q2.5 It is not sufficiently clear to me how the statistical methods used contribute to answering the research question, namely the feasibility of an RCT and determining an effect size (and other measures, which are too vague).
 
-We have clarified how we used descriptive statistics to summarize patient characteristics and feasibility outcomes.
+> R2.5 We have clarified how we used descriptive statistics to summarize patient characteristics and feasibility outcomes.
 We now describe the methods used for the additional analyses according to the original protocol in the supplementary materials.
 
-> It seems that in the description of this manuscript, the pilot outcomes are mixed with the outcomes of the trial itself, which is confusing.
+Q2.6 It seems that in the description of this manuscript, the pilot outcomes are mixed with the outcomes of the trial itself, which is confusing.
 This is likely due to the way it is written (it needs to be better explained and specified): the authors describe results, but also the outcomes.
@@
 
-We agree that this was confusing and have therefore attempted to streamline the paper further by presenting mortality descriptively along with the other patient characteristics, and only report the feasibility outcomes in the outcomes section.
+> R2.6 We agree that this was confusing and have therefore attempted to streamline the paper further by presenting mortality descriptively along with the other patient characteristics, and only report the feasibility outcomes in the outcomes section.
 
-> The flow diagram does not clearly show how many participants did not participate or were missed. The design of the study itself seems to have been conducted properly.
+Q2.7 The flow diagram does not clearly show how many participants did not participate or were missed. The design of the study itself seems to have been conducted properly.
 
-Thank you for pointing this out. We have added the number of patients who were lost to follow up in each arm to the flow diagram.
+> R2.7 Thank you for pointing this out. We have added the number of patients who were lost to follow up in each arm to the flow diagram.
 
-> In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.
+Q2.8 In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.
 
-We have revised the manuscript to describe the deviations where they are most relevant, and retained a list in the supplementary materials for completeness.
+> R2.8 We have revised the manuscript to describe the deviations where they are most relevant, and retained a list in the supplementary materials for completeness.
 
-> I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question.
+Q2.9 I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question.
 Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.
 
-We agree that the supplementary materials are very extensive.
+> R2.9 We agree that the supplementary materials are very extensive.
 We included these to transparently report the results of the analyses specified in the original protocol, but in response to your comment we have now published the majority of the supplementary materials as online additional files, and cite these in the manuscript.
@@
 
-## Response to Reviewer 3, Dr. Sheila Sprague, McMaster University:
+### Response to Reviewer 3, Dr. Sheila Sprague, McMaster University:
 
-> Thank you for the opportunity to review the manuscript titled “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study”.
+Q3.1 Thank you for the opportunity to review the manuscript titled “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study”.
 This manuscript describes an important trial with the potential to save many lives at hospitals across India.
@@
 
-Thank you for your thorough review. We have made major revisions in response to your comments, and we hope that the manuscript is now clearer.
+> R3.1 Thank you for your thorough review. We have made major revisions in response to your comments, and we hope that the manuscript is now clearer.
 
-> Introduction
-> - Provide clear objectives for the both the pilot phase and the definitive trial phase
+Introduction
 
-We have revised the objective of the pilot study to also include the aim of the full scale trial.
-The revised objective is "to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.".
+Q3.2 Provide clear objectives for the both the pilot phase and the definitive trial phase
 
-> Methods
-> - Describe what information is being collected on standard of care at each hospital as it varies.
+> R3.2 We have revised the objective of the pilot study to also include the aim of the full scale trial.
+The revised objective is "to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients."
 
-We did not collect in-depth data on the standard care at each hospital, and we have added a note in the methods section to clarify this.
+Methods
+
+Q3.3 Describe what information is being collected on standard of care at each hospital as it varies.
+
+> R3.3 We did not collect in-depth data on the standard care at each hospital, and we have added a note in the methods section to clarify this.
 We have also added this as a limitation of this pilot study, and suggested that the full scale trial should include a more detailed assessment of the standard care at each hospital.
 
-> - Provide additional details on ATLS and PTC training interventions.
+Q3.4 Provide additional details on ATLS and PTC training interventions.
 
-We have added more details on the ATLS and PTC training interventions to the methods section.
+> R3.4 We have added more details on the ATLS and PTC training interventions to the methods section.
 
-> - Include details on the cluster selection.
+Q3.5 Include details on the cluster selection.
 
-We have revised the methods section to clarify that the seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.
+> R3.5 We have revised the methods section to clarify that the seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.
 
-> - The authors should comment on the sample size for a pilot study.
+Q3.6 The authors should comment on the sample size for a pilot study.
 For example, why did they select seven clusters?
@@
 
-We have added the rationale for the number of clusters and the number of units per cluster to the methods section.
+> R3.6 We have added the rationale for the number of clusters and the number of units per cluster to the methods section.
 We did not conduct a formal power calculation, but deemed that the anticipated sample sizes would be sufficient to evaluate the feasibility outcomes.
 
-> - The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.
+Q3.7 The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.
 
-We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
+> R3.7 We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.*
 
-> - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment.
+Q3.8 Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment.
 Authors should have used the computer-generated randomization system at time of randomization for each site.
 
-We recognise that the use of sealed envelopes for randomisation increases the risk of bias and errors, and we have added this as a limitation of this pilot study.
+> R3.8 We recognise that the use of sealed envelopes for randomisation increases the risk of bias and errors, and we have added this as a limitation of this pilot study.
 
-> - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.
+Q3.9 For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.
 
-We have added that we did not blind the data analysts in this pilot study, but recommend doing so in the full scale trial.
+> R3.9 We have added that we did not blind the data analysts in this pilot study, but recommend doing so in the full scale trial.
 
-> - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.
+Q3.10 The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.
 
-The data collection period in each hospital was intended to be four months, and then patients were followed up for one month.
+> R3.10 The data collection period in each hospital was intended to be four months, and then patients were followed up for one month.
 However, the actual data collected periods varied slightly between clusters, and we have clarified this in the methods section.
+The 3-day mortality outcome was a typo, and we have corrected this.
 
-> - Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.
+Q3.11 Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.
 
-We did not specify success criteria for the feasibility outcomes, and we have added this to the methods section.
+> R3.11 We did not specify success criteria for the feasibility outcomes, and we have added this to the methods section.
 
-> - The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.
+Q3.12 The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.
 
-We now specify that we collected data on potential outcomes for the full scale trial, including mortality within 30 days, in-hospital mortality, complications and health related quality of life.
+> R3.12 We now specify that we collected data on potential outcomes for the full scale trial, including mortality within 30 days, in-hospital mortality, complications and health related quality of life.
 
-> - The statistical analysis section is unclear.
+Q3.13 The statistical analysis section is unclear.
 It should list the descriptive analyses plans and then the analyses used to assess the feasibility objectives.
@@
 
-Thank you for pointing this out.
+> R3.13 Thank you for pointing this out.
 We acknowledge that the estimation of potential effect sizes is not appropriate for this pilot study, and we have revised the aim of the study to reflect this and report this change as a deviation from the original protocol.
@@
 
-> - The statistical analyses section also describes inconsistent times for outcome comparisons.
+Q3.14 The statistical analyses section also describes inconsistent times for outcome comparisons.
 The intervention arms measure change from baseline over one month pre-training vs.
@@
 
-We agree.
+> R3.14 We agree.
 
-> Results
-> - Add a description of each cluster.
+Results
 
-We have added a description of each cluster as Supplementary Table S1.
+Q3.15 Add a description of each cluster.
 
-> - Comment on why the number of residents and patients varied across the treatment groups.
+> R3.15 We have added a description of each cluster as Supplementary Table S1.
 
-We have clarified that the number of residents trained in each intervention cluster depended on the size of the units, and that the number of patients included per cluster varied between clusters depending on the volume of patients at each hospital.
+Q3.16 Comment on why the number of residents and patients varied across the treatment groups.
 
-> - A few key prognostic factors differ across the treatment groups (age, ISS).
+> R3.16 We have clarified that the number of residents trained in each intervention cluster depended on the size of the units, and that the number of patients included per cluster varied between clusters depending on the volume of patients at each hospital.
 
-Thank you for pointing this out and we now highlight this in the results section.
+Q3.17 A few key prognostic factors differ across the treatment groups (age, ISS).
 
-> - Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.
+> R3.17 Thank you for pointing this out and we now highlight this in the results section.
 
-We have added a figure to the result's section, reporting the number of patients included per month per cluster.
+Q3.18 Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.
+
+> R3.18 We have added a figure to the result's section, reporting the number of patients included per month per cluster.
 Notably, the process of preparing this figure allowed us to identify a data collection error in two clusters, and we have added this to the methods section.
@@
 
-> - The results of feasibility outcomes should be reported clearly in a table, with a column indicating whether the feasibility outcome was successfully achieved.
+Q3.19 The results of feasibility outcomes should be reported clearly in a table, with a column indicating whether the feasibility outcome was successfully achieved.
 
-We have decided not to report the feasibility outcomes in a table, because some of them, for example the comparison of directly observed data and data extracted from medical records, were assessed qualitatively.
+> R3.19 We have decided not to report the feasibility outcomes in a table, because some of them, for example the comparison of directly observed data and data extracted from medical records, were assessed qualitatively.
 We have however removed the description of the clinical outcomes from the outcomes section to avoid confusion.
 
-> - Mortality should only be reported once in the results section.
+Q3.20 Mortality should only be reported once in the results section.
 
-Mortality is now reported together with the other patient characteristics in the results section. We have removed the table comparing clinical outcomes from the results section.
+> R3.20 Mortality is now reported together with the other patient characteristics in the results section. We have removed the table comparing clinical outcomes from the results section.
 
-> - Report the resident outcomes and the participant outcomes separately.
+Q3.21 Report the resident outcomes and the participant outcomes separately.
 
-We now report the resident outcomes and the participant outcomes separately.
+> R3.21 We now report the resident outcomes and the participant outcomes separately.
 
-> - List the protocol deviations in the results section, not in the discussion.
+Q3.22 List the protocol deviations in the results section, not in the discussion.
 
-We now report the protocol deviations where relevant in the methods section, and include a list as supplementary material for completeness and easy reference.
+> R3.22 We now report the protocol deviations where relevant in the methods section, and include a list as supplementary material for completeness and easy reference.
 
-> - It is unclear the importance and relevance of the before training after training data (Table S1).
+Q3.23 It is unclear the importance and relevance of the before training after training data (Table S1).
 
-We included the one month before training and three month after training data to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
+> R3.23 We included the one month before training and three month after training data to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline.
 We have attempted to clarify this in the methods section.
 
-> - The EQ-5D should be scored as opposed to showing the results for each question.
+Q3.24 The EQ-5D should be scored as opposed to showing the results for each question.
 
-Currently, there is no Indian value set available for the EQ-5D-3L, and we have therefore not scored the data. We now explain this in the methods section.
+> R3.24 Currently, there is no Indian value set available for the EQ-5D-3L, and we have therefore not scored the data. We now explain this in the methods section.
 
-> - Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.
+Q3.25 Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.
 
-We have removed the majority of the outcome tables from the supplementary materials and published them as online additional files, referenced in the manuscript.
+> R3.25 We have removed the majority of the outcome tables from the supplementary materials and published them as online additional files, referenced in the manuscript.
 
-> - The absolute difference versus relative difference does not add to the analyses and could be removed.
+Q3.26 The absolute difference versus relative difference does not add to the analyses and could be removed.
 
-We have removed the absolute difference versus relative difference from the main results section, but report them in the additional online files as this analysis was part of the original protocol.
+> R3.26 We have removed the absolute difference versus relative difference from the main results section, but report them in the additional online files as this analysis was part of the original protocol.
 
-> - As this a pilot study, outcome data should be reported descriptively in a clearer manner.
+Q3.27 As this a pilot study, outcome data should be reported descriptively in a clearer manner.
 
-We have attempted to clarify the reporting of the outcome data in the results section.
+> R3.27 We have attempted to clarify the reporting of the outcome data in the results section.
 
-> - There are too many outcomes.
+Q3.28 There are too many outcomes.
 
-We agree and suggest that the full scale trial should focus on the most relevant outcomes.
+> R3.28 We agree and suggest that the full scale trial should focus on the most relevant outcomes.
 
-> Discussion
-> - Remove all results from the discussion and place them in the results section, then discuss the results in the discussion section.
+Discussion
 
-We now only discuss results already reported in the results section in the discussion.
+Q3.29 Remove all results from the discussion and place them in the results section, then discuss the results in the discussion section.
 
-> - Discuss the implication of adding the seventh site in more detail.
+> R3.29 We now only discuss results already reported in the results section in the discussion.
 
-We have clarified that the seventh site was included for pragmatic reasons.
+Q3.30 Discuss the implication of adding the seventh site in more detail.
 
-> - Discuss if there are methods to improve participant follow-up for the definitive trial.
+> R3.30 We have clarified that the seventh site was included for pragmatic reasons.
 
-We now highlight the challenges of following up patients after discharge in the discussion and suggest that the full scale trial should focus on in-hospital mortality as the primary outcome.
+Q3.31 Discuss if there are methods to improve participant follow-up for the definitive trial.
 
-> - Discuss if there are methods to reduce missing data for the definitive phase.
+> R3.31 We now highlight the challenges of following up patients after discharge in the discussion and suggest that the full scale trial should focus on in-hospital mortality as the primary outcome.
 
-We have added a suggestion to the discussion to improve the completeness of the data extracted from the medical records.
+Q3.32 Discuss if there are methods to reduce missing data for the definitive phase.
 
-> - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?
+> R3.32 We have added a suggestion to the discussion to improve the completeness of the data extracted from the medical records.
 
-We have added a sentence on the ethics of the full trial.
+Q3.33 Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?
 
-> - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.
+> R3.33 We have added a sentence on the ethics of the full trial.
 
-Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial.
+Q3.34 Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.
+
+> R3.34 Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial.
 We now include this as a suggestion for the full scale trial in the discussion.
 
-> - Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).
+Q3.35 Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).
 
-We have attempted to align our reporting with these guidelines.
+> R3.35 We have attempted to align our reporting with these guidelines.
 
-## Response to Reviewer 4, Dr. Kyle Dack, University of Bristol:
+### Response to Reviewer 4, Dr. Kyle Dack, University of Bristol:
 
@@
 >
-> I have only minor comments about certain study design aspects which I found unclear - particularly the observed/extracted data comparison.
+Q4.1 I have only minor comments about certain study design aspects which I found unclear - particularly the observed/extracted data comparison.
 
-Thank you for reviewing the manuscript!
+> R4.1 Thank you for reviewing the manuscript!
 
-> Methods:
-> 1.
-Intervention – how long was the training programme, and who was it delivered by?
+Methods:
+
+Q4.2 Intervention – how long was the training programme, and who was it delivered by?
 Was the training performed in stages, e.g.
@@
 
-We have clarified that the training was conducted at the beginning of the three month intervention phase.
+> R4.2 We have clarified that the training was conducted at the beginning of the three month intervention phase.
 
-> 2.
-Variables – more information on the injury severity score would be useful to readers, because I am unsure how this variable was planned to be used (it is not an outcome), and how precisely it was calculated.
+Q4.3 Variables – more information on the injury severity score would be useful to readers, because I am unsure how this variable was planned to be used (it is not an outcome), and how precisely it was calculated.
 A reference is provided, but a high-level overview of the approach (including the range of potential scores) would help readers understand this measurement better.
 
-We have added a description of the ISS to the methods section.
+> R4.3 We have added a description of the ISS to the methods section.
 
-> 3.
-“Differences in distributions of observed and extracted data” – I am unsure how observed data differs from extracted data – these terms were not previously explained or defined.
+Q4.4 "Differences in distributions of observed and extracted data" – I am unsure how observed data differs from extracted data – these terms were not previously explained or defined.
 Does it mean the observation period vs the 3 month intervention period?
-“distributions” is also somewhat unclear, histogram distributions, or a specific tendency statistic?
+"distributions" is also somewhat unclear, histogram distributions, or a specific tendency statistic?
 
-We have attempted to clarify what we mean by distributions (summary statistics) and directly observed data, in the section where we describe the feasibility outcomes.
+> R4.4 We have attempted to clarify what we mean by distributions (summary statistics) and directly observed data, in the section where we describe the feasibility outcomes.
 
-> 4. Statistical analysis: “We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient” which outcomes does this apply to?
+Q4.5 Statistical analysis: "We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient" which outcomes does this apply to?
 
-We have moved the description of this analysis to the supplementary materials, and clarified that it applies to all cause 30-day and in-hospital mortality.
+> R4.5 We have moved the description of this analysis to the supplementary materials, and clarified that it applies to all cause 30-day and in-hospital mortality.
 
-> 5.
-The analysis involved comparing the 1st month and 2nd-4th months, but it is not described how outcomes will be summed – e.g.
+Q4.6 The analysis involved comparing the 1st month and 2nd-4th months, but it is not described how outcomes will be summed – e.g.
 consent rate, I imagine this the total consent rate in month 1 vs months 2-4?
@@
 
-We have attempted to clarify that feasibility outcomes were not assessed as differences between the observation and intervention phases, but summarised across the entire period.
+> R4.6 We have attempted to clarify that feasibility outcomes were not assessed as differences between the observation and intervention phases, but summarised across the entire period.
 
-> Results
-> 6. Table 1. Missing counts are reported for some but not all measurements. Does the absence of a count indicate complete data? If so, a footnote stating this would help readers.
+Results
 
-We have added a footnote to Table 1 to clarify that the absence of a count indicates complete data.
+Q4.7 Table 1. Missing counts are reported for some but not all measurements. Does the absence of a count indicate complete data? If so, a footnote stating this would help readers.
 
-> 7.
-Table 2.
+> R4.7 We have added a footnote to Table 1 to clarify that the absence of a count indicates complete data.
+
+Q4.8 Table 2.
 As mentioned in a prior comment it is unclear to me how these two groups are defined.
@@
 
-We only extracted data from a convenience sample of patients to reduce the workload of the project officers collecting data.
+> R4.8 We only extracted data from a convenience sample of patients to reduce the workload of the project officers collecting data.
 We have attempted to clarify this in the methods section and in the table caption.
 
-> 8. Table S1. Suggest clarifying that this is patient characteristics in the title.
+Q4.9 Table S1. Suggest clarifying that this is patient characteristics in the title.
 
-We have added "patient" to the title.
+> R4.9 We have added "patient" to the title.
 
-> 9. Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)
+Q4.10 Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)
 
-Yes, that is correct.
+> R4.10 Yes, that is correct.
 Empty cells indicate that there were no observations.

### 2025-08-06 functions/create_patients_per_month_figure.R
Update manuscript and functions

@@ -38,5 +38,11 @@ create_patients_per_month_figure <- function(data) {
     scale_y_continuous(breaks = c(0, 15, 30))
 
   # Save the figure
-  ggsave(patients.per.month.figure, filename = "patients-per-month.png", width = 85, height = 130, units = "mm")
+  ggsave(patients.per.month.figure,
+    filename = "patients-per-month.png",
+    width = 85, height = 130,
+    units = "mm",
+    bg = "white",
+    dpi = 300
+  )
 }

### 2025-08-28 point-by-point-response-revision-2.Rmd
Update manuscript based on the second round of comments

---
output: word_document
---

Dear Editor and Reviewers,

Thank you for your second review of our manuscript and for allowing us to submit a minor revision. We have addressed all of the comments and provided a detailed response below.

On behalf of all the authors,

Martin Gerdin Wärnberg, MD, PhD
Department of Global Public Health
Karolinska Institutet
Stockholm, Sweden

### Response to Reviewer 1, Dr.  Feroze  Sidhwa, San Joaquin General Hospital:

Q1 Improvements in patient outcomes could take much longer than a few months to manifest, and may also be dependent on a certain level of training being achieved in the unit. These limitations should be addressed.

> R1 We agree, especially some outcomes may take substantially longer than the follow up of one month. We have added a note to the discussion section to address this.

Q2 “The principal investigator at each hospital selected the units for training.” How? Was this based on who conducts the trauma resuscitation?

> R2 Only units involved in the trauma resuscitation were eligible for training. The principal investigator selected the units based on their own mandate to involve the units in the study.

Q3 How do you know the accuracy of ICD-10 coding at these hospitals?

> R3 The project officers conducting the ICD-10 coding have been trained using the WHO module on ICD-10 coding and their coding reviewed by the research team in previous projects.

Q4 Faculty not being included in the study because they are not typically directly involved in the initial management of trauma patients will be a major factor limiting generalizability of the findings. This cannot be helped and not every trial can be generalizable to the entire universe, but this should be noted in future conclusions and limitations.

> R4 Good point. We have a highlighted this in the appropriate section in the discussion.

Q5 “We initially planned to use simple random sampling to select the units to be trained, but for pragmatic reasons this decision was left to the site principal investigator.” Does this mean that the site principal investigator decided what kind of sampling to use, or that they decided on what units to be trained instead of using any kind of randomization?

> R5 The principal investigator at each site decided what units to be trained instead of randomising these units. We have clarified this in the manuscript.

Q6 An “error in the data uploading process” that led to 1/10th of the data being collected in two out of seven clusters is very alarming. Isn’t it possible this data uploading process error wasn’t a source of random error? For example maybe it drastically reduced the data available from the two most incompetent clusters, regardless of training status? Unless this issue is addressed it would be a major source of potential error in a larger trial. This was mentioned briefly in the conclusion but I would encourage the authors to make clear that they understand that this is an issue of grave concern for the accuracy of the study.

> R6 We agree that this is a major concern that cannot be allowed to happen in a full-scale trial. We have strengthened the language in the discussions section accordingly.

Q7 Overall this is a very useful pilot study and I hope the larger study will shed light on some important questions in the trauma world.

> R7 Thank you very much and we agree.

### 2025-08-28 manuscript.Rmd
Update manuscript and functions
Update manuscript based on the second round of comments

@@
 Training took place approximately one month after study initiation in that hospital.
-We initially planned to use simple random sampling to select the units to be trained, but for pragmatic reasons this decision was left to the site principal investigator.
+We initially planned to use simple random sampling to select the units to be trained, but for pragmatic reasons the decision on which units to train was left to the site principal investigator.
 The number of residents trained in each intervention cluster varied based on the unit size.
@@
 Sixth, we assessed a large number of potential outcomes, and a full-scale trial should focus on the most relevant outcomes.
+Seventh, the follow-up period was only one month, and changes in some of the outcomes may take longer than that to manifest.
+The effect on the outcomes is also likely to depend on the adherence to the training, which is not assessed in this pilot study.
 Finally, owing to a data uploading error, limited data were available from two clusters.
-At the time of data collection, network and other technical issues were present in some of the clusters, and should be mitigated in a full-scale trial by using a more robust data collection system with local offline backups.
+At the time of data collection, network and other technical issues were present in some of the clusters.
+This is a major concern that must be mitigated in a full-scale trial by using a more robust data collection system with local offline backups and careful centralised monitoring.
 
@@
 In recognition of their widespread use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best trial design.
+With regards to generalisability, the study was conducted in India, and the results are likely to be generalisable to other settings with similar trauma care systems.
+The findings may be less generalisable to settings where senior faculty are more directly involved in the initial management of trauma patients, as these were not trained in this pilot study.
 
@@
 
-MGW conceived the study, performed the analysis and drafted and revised the manuscript. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the
-study. MGV, DKV, KDS, and MK drafted the first version of the
-protocol. AG, HS, and SD drafted the first version of the patient and
-public involvement activities. JB and PP drafted the first versions of
-the data management sections and wrote the data management plan. PB
-and PP drafted the first versions of the statistical analysis
-section. AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB,
-PP, RS, SC, SD, and VK contributed to the refinement of the
-protocol.
+MGW conceived the study, performed the analysis and drafted and revised the manuscript.
+AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MK, NR, PB, PP, RS, SD, and VK contributed to the design of the study.
+MGV, DKV, KDS, and MK drafted the first version of the protocol.
+AG, HS, and SD drafted the first version of the patient and public involvement activities.
+JB and PP drafted the first versions of the data management sections and wrote the data management plan.
+PB and PP drafted the first versions of the statistical analysis section.
+AG, AM, CJ, DKV, HS, JB, KDS, LFT, LS, MH, MGW, MK, NR, PB, PP, RS, SC, SD, and VK contributed to the refinement of the protocol.
 DB, JB, SC, LFT, GG, MK, TK, CJ, NR, RS, KDS, LS and VP interpreted the results and revised the manuscript.
 AR, AC, C, DK, GG, MK, MT, VK and VP are representatives of the participating hospitals.
+MGW is the guarantor.
 

### 2025-08-28 manuscript.Rmd
Update and render manuscript after reviewers comments

@@
 The proprietary Advanced Trauma Life Support (ATLS) and low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+Observational studies and a recent small cluster randomised trial indicate that such programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025; @lule_impact_2025], but there is no large controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 

### 2025-08-29 manuscript.Rmd
Work on revision

--- 2025-08-28 manuscript.Rmd
@@
 
-Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@injuries2020].
+Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@gbd_2021_collaborators_injuries_2024].
 Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-The proprietary Advanced Trauma Life Support (ATLS) and low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].
-Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) are the most established trauma life support training programmes [@acsAtls2018; @ptcfPtc2018].
+Both programmes improve provider knowledge and skills [@Mohammad2013; @Kadhum2020], and while most observational studies associate them with reduced mortality [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025], some report associations with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
 
-Several studies, including at least two randomised studies [@Ali1996; @Ali1999], have shown that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013].
-Observational evidence also suggests that PTC also leads to improved provider skills [@Kadhum2020].
-The missing link is whether this improved knowledge and skill set translates into measurably improved patient outcomes.
+There are no controlled trials of the effect of ATLS or PTC on patient outcomes, and no trials evaluating any trauma life support training programme in a general trauma population [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024], although a recent cluster randomised trial found that the rural trauma team development course reduces mortality in patients with motorcycle injuries [@lule_impact_2025].
+Systematic reviews call for controlled trials of ATLS and PTC in settings where these programmes are not standard care [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
 
-Systematic reviews have called for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], as conducting such effectiveness trials in settings where they are part of the standard of care is not feasible.
-Many of these settings are in low- and middle income countries, where trial logistics can be more challenging.
-
-We therefore conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.
+Large scale cluster randomised trials can be complex and challenging, so before embarking on a full scale trial, we conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.
 
@@
 Finally, owing to a data uploading error, limited data were available from two clusters.
-At the time of data collection, network and other technical issues were present in some of the clusters.
-This is a major concern that must be mitigated in a full-scale trial by using a more robust data collection system with local offline backups and careful centralised monitoring.
+At the time of data collection, network and other technical issues were present in some of the clusters, which could explain this error.
+Regardless, this is a major concern that must be mitigated in a full-scale trial by using a more robust data collection system with local offline backups and careful centralised monitoring.
 
@@
 
-Previous studies on the effects of ATLS or PTC training on patient outcomes have been observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016].
+Previous studies on the effects of ATLS or PTC training on patient outcomes have been observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025].
 Most suggest that these programmes are associated with improved outcomes, although not all report significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025].
@@
 
-Our study represents the first published attempt to pilot a controlled trial evaluating the effect of trauma life support training on patient outcomes.
+Our study represents the first published attempt to pilot a controlled trial evaluating the effect of ATLS and PTC on patient outcomes.
 We conclude that a full-scale cluster randomised trial is feasible after incorporating the lessons of this pilot study.
@@
 
-The code for the analysis is released publicly. The final anonymised dataset is available from the corresponding author upon request.
+The code for the analysis is released publicly on GitHub (https://github.com/martingerdin/tern-pilot). The final anonymised dataset is available from the corresponding author upon request.
 

### 2025-09-01 manuscript.Rmd
Finalise revising the manuscript

@@
 Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@gbd_2021_collaborators_injuries_2024].
-Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
 
@@
 
-There are no controlled trials of the effect of ATLS or PTC on patient outcomes, and no trials evaluating any trauma life support training programme in a general trauma population [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024], although a recent cluster randomised trial found that the rural trauma team development course reduces mortality in patients with motorcycle injuries [@lule_impact_2025].
-Systematic reviews call for controlled trials of ATLS and PTC in settings where these programmes are not standard care [@Mohammad2013; @Jayaraman2014; @Kadhum2020].
+There are no controlled trials of the effect of ATLS or PTC on patient outcomes, and no trials evaluating any trauma life support training programme in a general trauma population [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].
+However, a recent cluster randomised trial found that the rural trauma team development course reduced mortality in patients with motorcycle injuries [@lule_impact_2025].
 
-Large scale cluster randomised trials can be complex and challenging, so before embarking on a full scale trial, we conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.
+Systematic reviews call for controlled trials of ATLS and PTC [@Mohammad2013; @Jayaraman2014; @Kadhum2020], but large scale cluster randomised trials can be complex.
+We therefore conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.
 

### 2025-09-01 manuscript.Rmd
Minor fixes in manuscript

@@
 
-##### Debojit Basak, MS `r insert_orcid_link("0000-0002-8378-9689")` <!-- Accepted, debojit88@gmail.com -->
+##### Debojit Basak, MParm (DB) `r insert_orcid_link("0000-0002-8378-9689")` <!-- Accepted, debojit88@gmail.com -->
 Project Research Scientist, Institute of Post Graduate Medical Education and Research, Kolkata, India.
